---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "A"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "B"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "C"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "D"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "E"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "F"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "G"
    orcid: 
    affiliations:
      - ref: stanford
    corresponding: true
    email: x@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["Cognitive Development", "Mental Rotation", "Spatial Reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    extra_dependencies: ["booktabs", "longtable"]
    # fig-pos: H        
    # tbl-pos: H  
    

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

From fitting puzzle pieces together at age three to navigating complex diagrams in school, spatial reasoning underpins a broad array of cognitive achievements — and children differ substantially in how these skills develop. Such variability is detectable early, across a wide range of spatial tasks in primary school-aged children [@hodgkiss2021developmental; @lauer2019development], and its consequences are far-reaching: spatial ability predicts outcomes spanning creative domains such as music [@padulo2020effects; @hetland2000learning] and the visual arts [@kell2013creativity; @winner2013relationship], mathematical achievement [@gunderson2012relation], science learning [@gilligan2020first], and long-term educational and occupational outcomes in STEM fields [@wai2009spatial]. Understanding what drives this variability is therefore a question of both theoretical and practical importance. The present study takes a step toward that goal by examining children's performance on a mental rotation task across three culturally diverse countries, asking both what is consistent across children and what varies.

Spatial reasoning encompasses abilities for representing, transforming, and reasoning about visuospatial information [@lohman1996spatial]. Rather than a single unified capacity, it is better understood as a family of related but distinguishable skills, including mental rotation, spatial visualization, perspective-taking, and navigation, that differ in the type of spatial information involved and the cognitive operations required [@uttal2013malleability; @newcombe2018three]. Skills are organized along two dimensions: whether the spatial relations are intrinsic to an object or extrinsic between objects, and whether the task requires dynamic mental transformation or only static representation [@uttal2013malleability]. Mental rotation — the focus of the present study — falls in the intrinsic-dynamic quadrant of this framework, involving the mental transformation of an object's orientation in space.

Mental rotation is the ability to mentally transform an object by rotating it in space to predict how it would appear from a different orientation or viewpoint [@shepard1971mental]. It is one of the most studied spatial skills in cognitive and developmental psychology, in part because it yields one of the most robust and replicable findings in the field: response times increase approximately linearly with the angular disparity between two objects, suggesting that mental rotation operates as a spatially continuous process rather than a discrete rule-based one [@shepard1971mental; @cooper1973chronometric]. We return to this finding in detail below.

Mental rotation ability develops across a protracted period beginning in infancy. Precursors of successful rotation are detectable in the first year of life [@frick2014development; @moore2008mental], and performance on explicit rotation tasks improves substantially between ages 3-12  [@frick2013development; @marmor1975development], with continued refinement into adolescence. This extended developmental trajectory makes the primary school years, ages 5-12, a particularly informative window for studying how mental rotation skills are acquired and how they vary across children.

The standard mental rotation task, as developed by @shepard1971mental, presents participants with pairs of objects at varying angular disparities and asks them to judge whether the objects are identical or mirror-reversed. Response time and accuracy are the key dependent variables. While this paradigm has generated decades of consistent findings, an important interpretive caveat applies: performance on the task may not reflect a single, unitary process of mental rotation. Research with adults suggests that solvers draw on a range of strategies, spanning a holistic-analytic continuum [@gluck2003spatial]. At the holistic end, the stimulus is transformed as an integrated mental representation — the so-called "pure rotation" approach. At the analytic end, solvers decompose the object into sub-components and verify features sequentially, a more piecemeal strategy that places greater demands on working memory [@just1985cognitive; @khooshabeh2013individual]. Strategy selection is not fixed: individuals differ in their preferred approach, and the same individual may shift strategies depending on task demands such as the complexity of the stimulus or the degree of angular disparity [@nazareth2019strategy]. Importantly, far less is known about which strategies children employ when solving mental rotation problems, and whether strategy use varies with age, ability, or cultural context. This question motivates the exploratory component of the present study, which we describe below.

The vast majority of developmental research on mental rotation has been conducted with children from Western, educated populations [@davis2019spatial; @peters2006mental], raising questions about whether findings about its developmental trajectory, psychometric properties, and underlying cognitive processes generalize beyond the populations in which they were established.

The limited cross-cultural evidence that does exist offers a mixed picture. A large-scale study of 3,367 students across Canada, Germany, and Japan found that a male performance advantage on mental rotation was consistent across all three nations, suggesting some cross-cultural robustness in group-level patterns [@peters2006mental]. However, cross-cultural comparisons of children specifically are rare. One notable exception compared children in the Ivory Coast and Switzerland and found that while both groups showed evidence of mental rotation, Swiss children demonstrated substantially higher speed and accuracy — and crucially, age predicted performance in the Swiss group but not in the Ivorean group [@nunez1998mental]. More recent work in non-WEIRD (i.e., Western, Educated, Industrialized, Rich, and Democratic) populations has shown that environmental and cultural factors — including access to formal schooling, freedom to explore the physical environment, and culturally-specific spatial practices — can substantially shape both the level and the pattern of spatial performance in children [@davis2019spatial; @davis2021cultural].

Taken together, this evidence highlights two distinct but related gaps. First, we have limited data on whether mental rotation tasks developed and validated in Western samples are measuring the same underlying construct when administered to children in different cultural contexts — a question of measurement invariance rather than simply mean-level differences. Second, even where cross-cultural comparisons have been made, they have rarely examined whether the core cognitive signatures of mental rotation replicate across diverse populations of children. The linear relationship between reaction time and angular disparity — one of the most robust findings in cognitive psychology — has been demonstrated repeatedly in Western adult samples, but its replication in children across culturally diverse sites remains largely untested. Beyond these two gaps, relatively little is known about the cognitive strategies children employ when solving mental rotation problems, and whether these vary across cultural contexts — a question we explore tentatively here. The present study attempts to address these gaps.

Returning to the initial question: why do children vary in spatial reasoning, and what does that variation reflect? Cross-cultural data offer a potentially powerful way to answer this. But how confidently we can interpret performance on a mental rotation task across cultural contexts depends on two prior questions about the task itself — and addressing these questions across diverse sites is the central aim of the present study.

The first concerns measurement invariance. If a task is not measuring the same underlying construct across different cultural contexts, then observed performance becomes difficult to interpret — it may reflect genuine spatial reasoning, or it may reflect something more specific to the cultural and educational context in which the task was developed. Finding evidence of measurement invariance across diverse sites therefore substantially strengthens the case that the task is capturing a construct that is genuinely spatial in nature, and that conclusions drawn from it are likely to generalize beyond any single cultural setting.

The second concerns process-level validity. Even where measurement invariance holds, our confidence that the task is tracking spatial reasoning specifically is increased by showing that performance is sensitive to the spatial properties of the stimuli in the way theory predicts. The key benchmark here is the angular disparity effect, first demonstrated by @shepard1971mental: as the angular difference between two objects increases, so too does the time required to judge whether they match. If children across three culturally diverse sites show this same systematic pattern, this constitutes good evidence that the task is engaging the spatial demands it was designed to capture, and that this signature is not specific to the Western samples in which it was originally established. Alongside these two primary aims, we tentatively explore a secondary question: whether there is any indication that children are drawing on different cognitive strategies when solving mental rotation problems, and whether this varies across contexts. We treat this component with appropriate caution, given the limitations of the data for addressing it directly.

## The Present Study

The present study uses data from the Learning Variability Network Exchange (LEVANTE), an international research collaboration designed to capture how children's learning and development vary across ages and cultural contexts. LEVANTE's central aim is to build a large, openly shared dataset of children's cognitive development across diverse settings, with a particular focus on individual variability — what differs across children, and why. Data were collected freom three geographically and culturally diverse pilot sites: Canada, Germany, and Colombia. Here we use data from a single task in the LEVANTE battery: a mental rotation task developed using item response theory, which allows directly comparable ability estimates to be derived across children who completed different subsets of items [@Kachergis2025LEVANTEcoretasks].

Against this backdrop, the present study pursues three aims. First, we examine whether the mental rotation task achieves measurement invariance across the three LEVANTE pilot sites, establishing that it captures a consistent underlying construct across cultural contexts. Second, we test whether the canonical Shepard-Metzler angular disparity effect replicates across all three sites in children aged 5 to 12 — extending this foundational finding to a culturally diverse developmental sample. Third, in an exploratory analysis, we ask what cognitive strategies children appear to employ when solving mental rotation problems, and whether strategy use varies with age or angular disparity. Taken together, these analyses move from establishing what is consistent across children and contexts, toward a first look at what varies — and how that variation might be characterized.

# Methods

## Participants

Data for the present study were drawn from three pilot sites participating in LEVANTE: uniandes-co, mpieva-de, and western-ca. Collectively, these sites supported task localization and technical implementation, iterative piloting, and deployment in varied testing contexts. The resulting pilot datasets enabled preliminary evaluation of construct validity, measurement invariance, test–retest reliability, and calibration for computer-adaptive administration, as well as descriptive assessment of age-related differences across the targeted constructs (refer to @Kachergis2025LEVANTEcoretasks for a full description).

### uniandes-co

The Universidad de Los Andes team carried out school-based data collection in Colombia, working with four schools in Bogotá and two public schools in the rural departments of Caquetá and Boyacá. Schools were selected pragmatically from institutions that had previously collaborated with the university. In Bogotá, three schools were affiliated with two family compensation funds and one was operated by a charter-school organisation. Across locations, participating schools largely served children from low-income and lower-middle-income backgrounds. Testing was conducted on tablets provided by the research team using a group-administration format, with group sizes adjusted by age (larger groups for older children; smaller groups for younger children).

### mpieva-de

At the mpieva-de site, the Max Planck Institute for Evolutionary Anthropology implemented a remote, family-based assessment protocol. Families were recruited via an existing registry of Leipzig residents and were contacted by telephone to invite participation. Families who enrolled were given login details and completed the assigned LEVANTE tasks at home using their own computer or tablet.

### western-ca

The Western University (Ontario) team recruited children from the surrounding community, primarily via online outreach (e.g., Facebook advertisements and posts in community groups), complemented by physical flyers and announcements distributed through local school newsletters. Families self-enrolled by registering their interest in a participant database. Children were assessed individually by a researcher in a dedicated on-campus testing space.

## Measure

The LEVANTE mental rotation task is adapted from @frick2013development. On each trial, children view a target image (e.g., a rabbit) and then choose between two alternatives (e.g., two rabbit silhouettes): one alternative is the same object shown at a different orientation, whereas the other is a non-matching foil. Children indicate which option corresponds to the target, and we record both accuracy and response time. Angular disparity is manipulated across trials (e.g., 40°, 80°, 120°, 160°), producing systematic effects on performance. Stimuli include 2D animal figures (ducks and rabbits) and more challenging 3D geometric shapes drawn from the classic Shepard and Metzler (1971) set.

```{r}
#| label: fig-mrot-example-items
#| fig-cap: "Example items for 2D and 3D rotation"
#| out-width: "100%"
#| fig-align: "center"

knitr::include_graphics("mrot_item_examples.pdf")
```

All LEVANTE tasks including mental rotation were originally developed in English and later adapted for Spanish and German through a structured internationalization workflow carried out with collaborators at the uniandes-co and mpieva-de sites, respectively. We first produced draft translations of all verbal content—including instructions, item text, and recurring terminology (e.g., encouragement prompts)—using DeepL. These drafts were then refined by a professional translator to ensure grammatical accuracy and natural phrasing. Next, a site researcher who was a native speaker of the relevant language and dialect reviewed the materials to confirm clarity and suitability for the specific task context. To check for semantic and conceptual alignment with the English originals, we used AI-assisted back-translation to identify wording that required further revision and native-speaker review.

The mental rotation task has already been implemented in a computer-adaptive task (CAT) format. In the adaptive version, participant ability is summarized by a latent proficiency estimate (theta) that is re-estimated after each response. The updated theta then guides selection of the next item so that it is appropriately challenging for each individual child. This trial-by-trial targeting typically makes the task more engaging and increases measurement efficiency, enabling reliable ability estimation with fewer administered items. Refer to @Kachergis2025LEVANTEcoretasks for more information on the CAT versions of all LEVANTE tasks.

# Results

## Cross-Cultural Measurement Invariance of the Mental Rotation Task

We first established whether the mental rotation task measures the same underlying construct across all three sites. To  evaluate measurement invariance, we used multi-group IRT [@putnick2016measurement]. Multi-group models supported scalar invariance under a 2PL model, indicating that item difficulty is aligned across sites and that children in each country are being placed on a comparable latent scale. Practically, this means the groups share the same relative ordering of item difficulty — items with larger rotation angles are consistently more difficult across all three countries (see @tbl-mrot-2pl-scalar). The core structure of mental rotation performance is thus largely shared across the cultural settings represented here: the same item features drive difficulty in a similar way regardless of site. With this foundation established, we turn to the second precondition: whether the task shows the process-level sensitivity to angular disparity that theory predicts. 

It should be noted that while we use psychometric modelling to place children’s performance on a common scale, the LEVANTE project is not designed to estimate absolute differences between sites. In practice, LEVANTE does not support site-level performance comparisons: sites vary substantially in sampling, recruitment, and test administration, so any apparent cross-site differences in raw performance are heavily confounded. Accordingly, although we report descriptive patterns that may look like absolute site differences, we emphasize that these should not be interpreted as meaningful.

```{r}
#| include: false

# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(umap)
```

```{r}
#| label: setup
#| include: false
#| cache: false

source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)

# coefficients
all_coefs <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs/all_coefs.rds"))

```

```{r}
#| label: setup-palettes
#| include: false

# site_pal is keyed by pilot_* names
# site_labels is keyed by readable names, values are pilot_* names
# index site_pal by the pilot_* values, then assign readable names
site_pal_labelled <- setNames(
  site_pal[unname(site_labels)],
  names(site_labels)
  )
```

```{r}
#| label: irt-anova
#| include: false

# IRT model output - import object
github_dir <- file.path(path.expand("~"), "Documents", "GitHub")
task_anova_path <- file.path(github_dir, "levante-pilots", "02_scoring_outputs", "task_anova.rds")

stopifnot(file.exists(task_anova_path))
task_anova <- readRDS(task_anova_path)
#colnames(task_anova)
#table(task_anova$itemtype)

# extract
mrot_2pl_scalar <- task_anova %>%
  filter(task == "mrot", itemtype == "2pl", invariance == "scalar")
mrot_2pl_scalar
```

```{r}
#| label: tbl-mrot-2pl-scalar
#| tbl-cap: "Model fit indices for 2PL scalar invariance models. Note: 2PL = 2-parameter logistic model; scalar invariance indicates that item intercepts and loadings are constrained across groups."
#| echo: false

mrot_2pl_scalar %>%
  transmute(
    Task       = task,
    Factors    = nfact,
    Model      = itemtype,
    Invariance = invariance,
    logLik     = logLik,
    AIC        = AIC,
    BIC        = BIC,
    `# params` = df
  ) %>%
  gt() %>%
  # Ensures the gt title area is clear and doesn't overlap the Quarto caption
  tab_header(title = md("")) %>% 
  fmt_number(columns = c(logLik, AIC, BIC), decimals = 2) %>%
  fmt_number(columns = `# params`, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Task, Factors, Model, Invariance, `# params`)) %>%
  cols_align(align = "right",  columns = c(logLik, AIC, BIC)) %>%
  tab_options(
    #table.font.size = px(12), # Increased slightly for better legibility
    table.width = pct(100),   # Ensures it fills the page width nicely
    data_row.padding = px(3)   # Tighter row spacing often looks better for stats
  )
```

```{r}
#| label: fig-item-diff
#| fig-cap: "**Item Difficulty Parameters Across Sites (Configural Model).** Points represent individual items; close alignment along the diagonal indicates that item difficulty parameters are consistent across cultural contexts."
#| include: false
#| fig-width: 12
#| fig-height: 8

# Load configural parameters
mrot_configural <- all_coefs |>
  dplyr::filter(task == "mrot",
                itemtype == "2pl",
                invariance == "configural") |>
  dplyr::mutate(group = dplyr::recode(group,
                  "pilot_western_ca"  = "Canada",
                  "pilot_mpieva_de"   = "Germany",
                  "pilot_uniandes_co" = "Colombia"))

# pivot wide
mrot_wide_d <- mrot_configural |>
  dplyr::select(item, group, d) |>
  tidyr::pivot_wider(names_from = group, values_from = d) |>
  dplyr::mutate(item_label = item |>
                  stringr::str_remove("mrot_2d_") |>
                  stringr::str_remove("mrot_3d_") |>
                  stringr::str_replace("_", " ") |>
                  stringr::str_replace("(\\d+)$", "\\1°"))

plot_scatter <- function(data, x_var, y_var, x_lab, y_lab) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]],
                   label = item_label)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = pal$grey) +
    geom_point(size = 2.5, na.rm = TRUE) +
    ggrepel::geom_text_repel(size = 2.8,
                    max.overlaps = Inf,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    na.rm = TRUE) +
    coord_equal(xlim = c(-2.5, 4), ylim = c(-2.5, 4)) +
    labs(x = x_lab, y = y_lab)
}

p1 <- plot_scatter(mrot_wide_d, "Canada", "Germany",   "Canada", "Germany")
p2 <- plot_scatter(mrot_wide_d, "Canada", "Colombia",  "Canada", "Colombia")
p3 <- plot_scatter(mrot_wide_d, "Germany", "Colombia", "Germany", "Colombia")

(p1 | p2 | p3) +
  plot_annotation(
    title = "Item Difficulty Parameters Across Sites (Configural Model)",
    theme = theme(plot.title = element_text(size = 12))
  )

# checks
all_coefs |>
  dplyr::filter(task == "mrot", itemtype == "2pl", invariance == "configural") |>
  dplyr::count(subset, group)

# only 12 items appearing: multigroup IRT model was fitted only on the overlap items by design. The non-overlap items were either fitted in separate by-site models, or weren't included in the multigroup analysis at all.

# note to check this with Doria. 

```

@fig-dev-trends plots participants' mental rotation ability estimates (IRT theta) against age, with curves by site modeled using generalized additive models (GAMs). GAMs allow the developmental association with age to be non-linear, capturing periods of more rapid change and relative flattening without imposing a fixed form. Across all three sites, theta increases with age, indicating that the task captures developmental change consistently across cultural contexts. The dispersion of points within age bands underscores substantial individual variability in ability among same-aged children. Note that vertical offsets between sites reflect differences in sample composition and recruitment rather than true cross-cultural performance differences, and are not interpreted further here.

```{r}
#| label: fig-dev-trends
#| fig-cap: "**Developmental trajectories of mental rotation ability across sites.** Points represent individual children's IRT-estimated ability (θ); curves are GAM-smoothed age trends with 95% confidence intervals. Vertical offsets between sites should not be interpreted as performance differences; LEVANTE sites differ in sampling and recruitment in ways that heavily confound absolute comparisons."
#| fig-width: 8
#| fig-height: 6

mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!setNames(names(site_labels), 
                                                    unname(site_labels)),
                                  .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  scale_colour_manual(values = site_pal_labelled) +
  labs(
    x      = "Age (years)",
    y      = "Ability (theta)",
    colour = "Site"
  )
```

## Angular disparity and the RT–angle relationship

Our aim with this analysis was to test whether the canonical reaction time–rotation angle relationship first reported by @shepard1971mental replicates in our data. @fig-rt-angle-per-run-no-cat plots mean reaction times (±95% confidence intervals) as a function of rotation angle, stratified by age group (5–7, 7–9, and 9–12 years) and site, with separate series by stimulus and item type (2D vs 3D). CAT trials are excluded from this analysis, as adaptive item selection introduces a systematic confound: higher-ability children are disproportionately routed to harder items at larger angles, which would artificially compress the RT–angle slope. To show the underlying dispersion and sample density alongside the summary trends, we provide a second version of @fig-rt-angle-per-run-no-cat in the Appendix that overlays the raw per-run data (using the same age bins and 95% CIs; see @apx-irt-ability-age). 

The pattern with respect to angular disparity depends substantially on stimulus type. For 3D shape stimuli, RTs are slower overall and show a consistent positive relationship with rotation angle across all three sites and age groups, closely matching the classic RT–angle effect [@shepard1971mental]. For 2D stimuli (duck and rabbit), RT–angle functions are notably flat across all sites and age groups, suggesting these items do not impose sufficient rotational demands to elicit the angular disparity effect reliably. This stimulus-dependence is an interesting finding in its own right: the angular disparity effect is sensitive to the complexity and dimensionality of the stimuli, rather than being an invariant feature of mental rotation performance in children. Patterns for the 5–7 age group are more irregular, particularly for 3D stimuli in Colombia and Germany, though the wide confidence intervals in these cells suggest this reflects limited sample size rather than a substantively different pattern.

```{r}
#| label: trial-level-rts
#| include: false

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%
  tidyr::unnest(data) %>%
  dplyr::mutate(
    correct = dplyr::case_when(
      is.logical(correct) ~ as.integer(correct),
      TRUE ~ as.integer(as.numeric(correct))
    ),
    rt_s   = as.numeric(rt_numeric) / 1000,
    log_rt = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))
```

```{r}
#| label: remove-retest-data
#| include: false

baseline_runs <- mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%  
  group_by(site, user_id) %>%
  arrange(age, run_id) %>%           
  slice(1) %>%
  ungroup() %>%
  select(site, user_id, run_id)

# keep only trials from those baseline runs
mrot_rt_trial_no_retest <- mrot_rt_trial %>%
  semi_join(baseline_runs, by = c("site", "user_id", "run_id"))

# check: each child should now have exactly 1 run (within site)
mrot_rt_trial_no_retest %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs)

# how many re-testers?
mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs) # 133
```

```{r}
#| label: fig-rt-angle-per-run
#| include: false
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 7, 9, 12)

mrot_items_rt_runs <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = dplyr::n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  ) + scale_x_continuous(breaks = c(40, 80, 120, 160)) +
  .scale_colour_stimulus()
```

```{r}
#| label: fig-rt-angle-per-run-no-cat
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 7
#| fig-height: 6.5

age_breaks <- c(5, 7, 9, 12)

# load CAT metadata if not already in environment
cat_data <- read_rds(here("..", "levante-pilots", "02_scoring_outputs", "scores", "scores_combined.rds")) %>%
  group_by(user_id, item_task) %>%
  arrange(time_started) %>%
  mutate(run_number = row_number(),
         age_gap = age - age[1]) %>%
  ungroup() %>%
  filter(item_task == "mrot") %>%
  filter(run_number == 1 | (run_number == 2 & age_gap > 1/6)) %>%
  distinct(user_id, run_id, adaptive)

mrot_items_rt_runs <- mrot_rt_trial_no_retest %>%
  left_join(cat_data, by = c("user_id", "run_id")) %>%
  filter(adaptive == FALSE) %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = dplyr::n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  ) +
scale_x_continuous(breaks = c(40, 80, 120, 160)) +
  .scale_colour_stimulus()

p_mrot_rt_angle_age_1
```
\clearpage

## Within-Person Dynamics

Having established that 3D stimuli elicit a clear angular disparity effect while 2D stimuli do to a lesser extent, we next examined how RTs evolved within individuals over the course of the task. This analysis asks a complementary question: not whether children are sensitive to rotation angle, but whether they show sensitivity to task demands as they unfold in real time. To isolate time-on-task effects from item-driven RT changes, we residualized log-RTs on angle, stimulus, and item type prior to analysis.

@fig-within-person-trials reveals two consistent patterns across all three sites. First, there is a pronounced jump in residual RT at trial 48, corresponding to the transition from 2D to 3D stimuli. This is consistent with the between-stimulus RT difference established in the previous analysis, and extends it: the increase is immediate rather than gradual, suggesting children recalibrate their processing demands at the block transition rather than adapting incrementally. Second, within each block, RTs decrease across trials, indicating a practice effect that operates independently of angular disparity. This is notable given the flat RT–angle functions observed for 2D stimuli earlier: children are getting faster with practice on these items even in the absence of a detectable rotation effect, suggesting the two measures are capturing distinct aspects of performance. Both patterns — the block transition jump and the within-block acceleration — are qualitatively consistent across all three sites, further extending the cross-cultural consistency observed in the angular disparity analysis. Germany shows a somewhat flatter trajectory overall, with less acceleration in the 2D block and a less pronounced jump at the transition; this may reflect differences in sample composition or task familiarity,

```{r}
#| label: fig-within-person-trials
#| fig-cap: "**Trial-level Reaction Time Dynamics (angle-residualised).** Mean residual log-RT across trials for each non-adaptive site ±SE. The dashed line marks the 2D→3D transition. Re-test and CAT data removed."
#| fig-width: 6
#| fig-height: 4

# prep: remove CAT data
trial_resid_data <- mrot_rt_trial_no_retest %>%
  left_join(cat_data, by = c("user_id", "run_id")) %>%
  filter(adaptive == FALSE) %>%
  filter(!is.na(log_rt), correct == 1) %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle    = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(!is.na(angle), !is.na(stimulus), !is.na(item_group))

# residualise log_rt on angle + stimulus + item_group
resid_model <- lm(log_rt ~ angle + stimulus + item_group, 
                  data = trial_resid_data)

trial_resid_data <- trial_resid_data %>%
  mutate(log_rt_resid = residuals(resid_model))

# add block variable
transition_trial <- 48 

trial_resid_data <- trial_resid_data %>%
  mutate(
site_label = dplyr::recode(site, !!!setNames(names(site_labels), unname(site_labels))),    
block = if_else(trial_number < transition_trial, "2D", "3D")
  )

# plot
ggplot(trial_resid_data, 
       aes(x = trial_number, y = log_rt_resid, 
           color = site_label, fill = site_label)) +
  stat_summary(fun = mean, geom = "point", size = 1.2, alpha = 0.5) +
  geom_smooth(aes(group = interaction(site_label, block)),
              method = "loess", span = 0.75, se = TRUE, alpha = 0.2) +
  geom_vline(xintercept = transition_trial, linetype = "dashed",
             color = pal$grey, linewidth = 0.7) +
  annotate("text", x = transition_trial + 1, y = Inf,
           label = "Start of 3D block", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_colour_manual(values = site_pal_labelled) +
  scale_fill_manual(values = site_pal_labelled) +
  labs(
    x     = "Trial Number",
    y     = "Residual reaction time",
    color = "Site",
    fill  = "Site"
  ) +
  theme(legend.position = "bottom")
```
\clearpage

```{r}
#| label: strategy-analysis-prep
#| results: "hide"

# ability
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# Prepare trial-level data for the rich model
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    angle    = as.integer(str_extract(item, "\\d+")),
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(angle)),
    trial_s  = as.numeric(scale(trial_number))
  )

# set sum-to-zero contrasts - by default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the grand mean slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# extract subject-level random effects
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# compute instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(site, user_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

# join
rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(instability_data, by = c("site", "user_id")) %>% 
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  # Filter to ensure all needed features
  filter(if_all(c(ability, Rel_Slope, Abs_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))
```

```{r}
#| label: clustering-execution
#| include: false

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1], # red
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], # blue
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  # green
)

analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, 
                  Instability, Learning_Rate), is.finite)) 

global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )

# FIX: as.character() on all pull() calls
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% 
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

effective_clusters <- cluster_stats %>%
  filter(as.character(Global_Cluster) != low_eff_id) %>%
  arrange(mean_slope) 

analytical_id <- effective_clusters %>%
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

holistic_id <- effective_clusters %>%
  slice(2) %>% 
  pull(Global_Cluster) %>%
  as.character()

analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    as.character(Global_Cluster) == low_eff_id    ~ "Ineffective Strategies",
    as.character(Global_Cluster) == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    as.character(Global_Cluster) == holistic_id   ~ "Mental Rotation"
  ),
  Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  labs(title = "Cluster Result: Ability vs. Relative Slope")
```

## Strategy Identification via Clustering

The preceding analyses established that the mental rotation task measures a consistent underlying construct across sites, and that the canonical angular disparity effect partially replicates — robustly for 3D stimuli, and less consistently for 2D stimuli. Having established what is consistent across children, we now turn to an exploratory question: do children vary in the cognitive strategies they employ, and can meaningful strategy groups be identified from trial-level behavioral signatures? We emphasize throughout that this analysis is exploratory. The clusters we identify are inferred from behavioral proxies rather than directly observed strategies, the solution is sensitive to analytic choices, and replication with larger samples is required before strong conclusions can be drawn.

To characterize individual differences in how children approach the task, we fitted a mixed-effects model predicting log-RT from rotation angle, stimulus type, and trial number, with by-participant random slopes for angle and trial number. This allowed us to extract three person-level behavioral signatures for each child: their baseline response speed, the slope of their RT–angle function (both absolute and relative to their baseline speed), and their rate of within-task acceleration (learning rate). We additionally computed response instability as the standard deviation of model residuals, capturing trial-to-trial variability in RT not explained by item properties. These features were submitted to k-means clustering (k = 3).

Cluster labels were assigned based on mean ability and relative slope: a low-ability cluster with a near-zero RT–angle slope was labeled "Ineffective Strategies" (n = r ineff$N); a higher-ability cluster with a moderate positive slope was labeled "Mental Rotation" (n = r mrot$N); and the largest cluster was labeled "Other Effective Strategies (Flipping/Matching)" (n = r flip$N). Summary statistics are presented in @tbl-cluster-summary.

```{r}
#| label: tbl-cluster-summary
#| fig-width: 6
#| fig-height: 5
#| tbl-cap: "Cluster summary statistics. Mental Rotation shows the steepest RT-angle slope despite lower ability than the Flipping/Matching group, whose elevated RT reflects uniformly slower but angle-insensitive processing."

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    n              = n(),
    mean_ability   = mean(ability, na.rm = TRUE),
    sd_ability     = sd(ability, na.rm = TRUE),
    mean_rel_slope = mean(Rel_Slope, na.rm = TRUE),
    sd_rel_slope   = sd(Rel_Slope, na.rm = TRUE),
    mean_abs_slope = mean(Abs_Slope, na.rm = TRUE),
    sd_abs_slope   = sd(Abs_Slope, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_rel_slope)) %>%
  transmute(
    Strategy       = Strategy_Label,
    N              = n,
    `Mean Ability` = mean_ability,
    `SD Ability`   = sd_ability,
    `Mean Rel. Slope` = mean_rel_slope,
    `SD Rel. Slope`   = sd_rel_slope,
    `Mean Abs. Slope` = mean_abs_slope,
    `SD Abs. Slope`   = sd_abs_slope
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>%
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, 
                `Mean Rel. Slope`, `SD Rel. Slope`,
                `Mean Abs. Slope`, `SD Abs. Slope`),
    decimals = 2
  ) %>%
  fmt_number(columns = N, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = where(is.numeric)) %>%
  cols_width(
    Strategy ~ px(160),
    N        ~ px(40),
    everything() ~ px(65)
  ) %>%
  tab_spanner(label = "Ability",         columns = c(`Mean Ability`, `SD Ability`)) %>%
  tab_spanner(label = "Relative Slope",  columns = c(`Mean Rel. Slope`, `SD Rel. Slope`)) %>%
  tab_spanner(label = "Absolute Slope",  columns = c(`Mean Abs. Slope`, `SD Abs. Slope`)) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

```{r}
#| label: strategy-reporting-objects
#| include: false

ineff <- analysis_data %>%
  filter(Strategy_Label == "Ineffective Strategies") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )

flip <- analysis_data %>%
  filter(Strategy_Label == "Other Effective Strategies (Flipping/Matching)") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )

mrot <- analysis_data %>%
  filter(Strategy_Label == "Mental Rotation") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )
```
The three clusters differed significantly on both dimensions (@fig-validation-boxplots). Ability differed across all pairwise comparisons (all *p*s < .0001), with the Ineffective Strategies group showing markedly lower ability (M = `r round(ineff$Mean_Ability, 2)`, SD = `r round(ineff$SD_Ability, 2)`) than both the Flipping/Matching (M = `r round(flip$Mean_Ability, 2)`, SD = `r round(flip$SD_Ability, 2)`) and Mental Rotation (M = `r round(mrot$Mean_Ability, 2)`, SD = `r round(mrot$SD_Ability, 2)`) groups. Relative RT slope also differed significantly across all pairwise comparisons (all *p*s < .001), with the Mental Rotation cluster showing the steepest slope (M = `r round(mrot$Mean_Slope, 2)`), consistent with analog rotation processes that scale with angular disparity.

Inspection of the RT–angle functions (@fig-rt-by-disparity) reveals qualitatively distinct profiles across the three groups. The Ineffective Strategies group (n = `r ineff$N`) is characterized by very low ability (M = `r round(ineff$Mean_Ability, 2)`, SD = `r round(ineff$SD_Ability, 2)`), a flat or slightly negative RT–angle slope (M = `r round(ineff$Mean_Slope, 2)`), and high response instability (M = `r round(ineff$Mean_Instab, 2)`). This group's RT decreases with rotation angle — they respond more quickly to harder items, not more slowly. Combined with steeply declining accuracy (@fig-accuracy-by-angle), this pattern suggests these children are not engaging with the spatial structure of the items, and may be disengaging at higher angular disparities. This is a qualitatively different failure mode from simply being slow.

The Other Effective Strategies (Flipping/Matching) group (n = `r flip$N`) is the largest cluster and shows the highest mean ability (M = `r round(flip$Mean_Ability, 2)`, SD = `r round(flip$SD_Ability, 2)`). Their RT–angle function rises to a peak around 120° before declining at 160°, consistent with exploiting reflective symmetry at high angular disparities rather than performing full analog rotation — a pattern sometimes described as "flipping" in the mental rotation literature — as items approach 180°, mirror images become easier to distinguish perceptually without requiring full rotation. Their sustained accuracy across angles supports this interpretation: this is a genuinely effective strategy, not a speed-accuracy trade-off.

The Mental Rotation group (n = `r mrot$N`) shows the classic monotonically increasing RT–angle function consistent with the Shepard-Metzler signature, with the steepest relative slope of the three groups (M = `r round(mrot$Mean_Slope, 2)`, SD = `r round(mrot$SD_Slope, 2)`). Interestingly, this group has lower mean ability than the Flipping/Matching group (M = `r round(mrot$Mean_Ability, 2)`, SD = `r round(mrot$SD_Ability, 2)`), despite appearing to use the strategy the task was designed to elicit. This may reflect the greater cognitive cost of analog rotation relative to feature-matching strategies, particularly for the 2D stimuli where rotational demands are lower.

Taken together, these profiles offer a tentative explanation for the pattern observed in the angular disparity analysis: the flat RT–angle functions for 2D stimuli likely reflect the predominance of feature-matching strategies in the sample, with only a minority of children showing clear evidence of analog rotation.

```{r}
#| label: fig-rt-by-disparity
#| fig-width: 7
#| fig-height: 5
#| include: true
#| fig-cap: "**Participant-weighted mean log RT by rotation angle and strategy cluster.** Shaded bands show 95% CIs. The Flipping/Matching group shows a characteristic inverted-U function, with RT declining after 120°, consistent with exploitation of reflective symmetry at high angular disparities."

rt_user_disparity <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors))) %>%
  group_by(Strategy_Label, user_id, angle) %>%      
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_disparity_summary <- rt_user_disparity %>%
  group_by(Strategy_Label, angle) %>%                 
  summarise(
    n_users  = n_distinct(user_id),
    mean_rt  = mean(user_mean_log_rt, na.rm = TRUE),
    se       = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    lwr      = mean_rt - 1.96 * se,
    upr      = mean_rt + 1.96 * se,
    .groups  = "drop"
  )

ggplot(rt_disparity_summary, aes(x = angle, y = mean_rt,
                                  colour = Strategy_Label, fill = Strategy_Label)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(20, 40, 80, 120, 160)) +
  theme(legend.position = "bottom") +
  labs(
    x      = "Rotation Angle (°)",
    y      = "Mean log RT (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: fig-accuracy-by-angle
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Participant-weighted Accuracy-angle functions by strategy cluster."

acc_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_user_angle <- acc_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(
    user_mean_acc = mean(correct, na.rm = TRUE),
    .groups = "drop"
  )

acc_summary <- acc_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users   = n_distinct(user_id),
    mean_acc  = mean(user_mean_acc, na.rm = TRUE),
    se_acc    = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

ggplot(acc_summary, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc, ymax = mean_acc + 1.96 * se_acc, fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(20, 40, 80, 120, 160)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean Accuracy (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: tbl-descriptive-validation
#| tbl-cap: "Descriptive statistics by Strategy Cluster."
#| include: false

validation_table <- analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE),
    .groups = "drop"
  )

validation_table %>%
  transmute(
    Strategy      = Strategy_Label,
    N             = N,
    `Mean Ability`= Mean_Ability,
    `SD Ability`  = SD_Ability,
    `Mean Slope`  = Mean_Slope,
    `SD Slope`    = SD_Slope,
    `Instability` = Mean_Instab
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>% 
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability), 
    decimals = 2
  ) %>%
  fmt_number(columns = c(N), decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability)) %>%
  cols_width(
    Strategy     ~ px(180),
    N            ~ px(40),
    everything() ~ px(75)
  ) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

```{r}
#| label: fig-validation-boxplots
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Cluster validation. Left: ability (theta) by strategy group. Right: relative RT slope by strategy group. Significance bars show pairwise t-tests; non-significant comparisons hidden."

my_comparisons <- list(
  c("Ineffective Strategies", "Other Effective Strategies (Flipping/Matching)"),
  c("Ineffective Strategies", "Mental Rotation"),
  c("Other Effective Strategies (Flipping/Matching)", "Mental Rotation")
)

p_ability <- ggplot(analysis_data, aes(x = Strategy_Label, y = ability, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "1. Defined by Ability", y = "Ability (Theta)", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 32, hjust = 1))

p_slope <- ggplot(analysis_data, aes(x = Strategy_Label, y = Rel_Slope, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = pal$grey) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "2. Defined by RT Slope", y = "Relative Slope", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 32, hjust = 1))

p_ability + p_slope

```
\clearpage

### Cross-Cultural Consistency of Strategy Clusters
```{r}
#| label: local-clustering-agreement
#| include: false

vars <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")

stopifnot(all(vars %in% names(analysis_data)))
stopifnot(all(c("Site", "user_id", "Strategy_Label") %in% names(analysis_data)))
stopifnot(nrow(analysis_data) == nrow(dplyr::distinct(analysis_data, Site, user_id)))

# ARI function
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

get_local_strategy_labels <- function(df_site,
                                      k = 3,
                                      seed = 123,
                                      nstart = 25,
                                      min_n = 30) {

  if (nrow(df_site) < min_n) return(NULL)
  set.seed(seed)

  X <- df_site %>%
    dplyr::select(dplyr::all_of(vars)) %>%
    scale() %>%
    as.matrix()

  km <- kmeans(X, centers = k, nstart = nstart)

  tmp <- df_site %>%
    mutate(Local_Cluster = factor(km$cluster))

  local_stats <- tmp %>%
    group_by(Local_Cluster) %>%
    summarise(
      mean_abil  = mean(ability,   na.rm = TRUE),
      mean_slope = mean(Rel_Slope, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(mean_abil)

  ineff_id     <- local_stats %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  other_two    <- local_stats %>% filter(as.character(Local_Cluster) != ineff_id) %>% arrange(mean_slope)
  other_eff_id <- other_two   %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  mrot_id      <- other_two   %>% slice(2) %>% pull(Local_Cluster) %>% as.character()

  tmp %>%
    mutate(
      Strategy_Local = dplyr::case_when(
        as.character(Local_Cluster) == ineff_id     ~ "Ineffective Strategies",
        as.character(Local_Cluster) == other_eff_id ~ "Other Effective Strategies (Flipping/Matching)",
        as.character(Local_Cluster) == mrot_id      ~ "Mental Rotation"
      ),
      Strategy_Local = factor(Strategy_Local, levels = levels(analysis_data$Strategy_Label))
    ) %>%
    select(Site, user_id, Strategy_Local)
}

# fit local models independently within each site
local_labels_all <- analysis_data %>%
  group_by(Site) %>%
  group_split() %>%
  purrr::map_dfr(get_local_strategy_labels)

analysis_local <- analysis_data %>%
  select(-dplyr::any_of("Strategy_Local")) %>%
  left_join(local_labels_all, by = c("Site", "user_id"))

# agreement (3-class)
agreement_3class <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_by(Site) %>%
  summarise(
    n         = n(),
    pct_agree = mean(Strategy_Local == Strategy_Label),
    ARI       = ARI(Strategy_Label, Strategy_Local),
    .groups   = "drop"
  )

agreement_3class

# confusion matrices per site — save split first to avoid pipe . scoping issue
site_splits <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_split(Site)

confusion_by_site <- site_splits %>%
  purrr::set_names(purrr::map_chr(site_splits, ~ unique(.x$Site))) %>%
  purrr::map(~ table(.x$Strategy_Local, .x$Strategy_Label))

confusion_by_site

# recall + purity
recall_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Label, Strategy_Local) %>%
  group_by(Site, Strategy_Label) %>%
  mutate(recall = n / sum(n)) %>%
  ungroup()

purity_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Local, Strategy_Label) %>%
  group_by(Site, Strategy_Local) %>%
  mutate(purity = n / sum(n)) %>%
  ungroup()

diag_summary <- recall_tbl %>%
  filter(Strategy_Label == Strategy_Local) %>%
  transmute(Site, Strategy = Strategy_Label, recall) %>%
  left_join(
    purity_tbl %>%
      filter(Strategy_Local == Strategy_Label) %>%
      transmute(Site, Strategy = Strategy_Local, purity),
    by = c("Site", "Strategy")
  )

diag_summary
```

```{r}
#| label: local-vs-pooled-reporting
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_pct <- function(x, d = 1) paste0(formatC(100*x, format = "f", digits = d), "%")

get_row <- function(df, site) df[df$Site == site, , drop = FALSE]

# sites
s_mpi <- "mpieva-de"
s_uni <- "uniandes-co"
s_wes <- "western-ca"

# agreement rows
a_mpi <- get_row(agreement_3class, s_mpi)
a_uni <- get_row(agreement_3class, s_uni)
a_wes <- get_row(agreement_3class, s_wes)

# label strings
lab_ineff <- "Ineffective Strategies"
lab_other <- "Other Effective Strategies (Flipping/Matching)"
lab_mrot  <- "Mental Rotation"

get_diag <- function(site, label) {
  out <- diag_summary[diag_summary$Site == site & diag_summary$Strategy == label, , drop = FALSE]
  if (nrow(out) == 0) {
    return(data.frame(Site = site, Strategy = label, recall = NA_real_, purity = NA_real_))
  }
  stopifnot(nrow(out) == 1)
  out
}

d_mpi_ineff <- get_diag(s_mpi, lab_ineff)
d_mpi_other <- get_diag(s_mpi, lab_other)
d_mpi_mrot  <- get_diag(s_mpi, lab_mrot)

d_uni_ineff <- get_diag(s_uni, lab_ineff)
d_uni_other <- get_diag(s_uni, lab_other)
d_uni_mrot  <- get_diag(s_uni, lab_mrot)

d_wes_ineff <- get_diag(s_wes, lab_ineff)
d_wes_other <- get_diag(s_wes, lab_other)
d_wes_mrot  <- get_diag(s_wes, lab_mrot)

# off-diagonal moves
get_move <- function(tab, local_lab, global_lab) {
  stopifnot(local_lab %in% rownames(tab), global_lab %in% colnames(tab))
  n     <- tab[local_lab, global_lab]
  denom <- sum(tab[, global_lab])
  list(n = as.numeric(n), denom = as.numeric(denom), prop = as.numeric(n) / as.numeric(denom))
}

tab_mpi <- confusion_by_site[[s_mpi]]
tab_uni <- confusion_by_site[[s_uni]]
tab_wes <- confusion_by_site[[s_wes]]

# Mental Rotation globally assigned to Other Effective locally
mpi_mrot2other <- get_move(tab_mpi, lab_other, lab_mrot)
wes_mrot2other <- get_move(tab_wes, lab_other, lab_mrot)

# Other Effective globally assigned to Mental Rotation locally
uni_other2mrot <- get_move(tab_uni, lab_mrot, lab_other)
agreement_3class
confusion_by_site

```

To assess whether the identified clusters are recoverable within each site independently — a basic test of their robustness — we re-ran the k-means procedure separately within each site and computed agreement between the local and global solutions using percent agreement and the Adjusted Rand Index (ARI; see Appendix). Moderate agreement was observed across all three sites (Canada: `r fmt_pct(a_wes$pct_agree)`, ARI = `r fmt_num(a_wes$ARI)`; Colombia: `r fmt_pct(a_uni$pct_agree)`, ARI = `r fmt_num(a_uni$ARI)`; Germany: `r fmt_pct(a_mpi$pct_agree)`, ARI = `r fmt_num(a_mpi$ARI)`). The Ineffective Strategies and Flipping/Matching clusters were recovered consistently across sites, but the Mental Rotation cluster was not independently recoverable in Germany or Canada, where locally-clustered children predominantly mapped onto the Flipping/Matching group instead. This suggests the Mental Rotation cluster may partly emerge from pooling across sites rather than reflecting a robustly distinct behavioral profile within any single site. Colombia showed the most consistent local recovery across all three clusters.

These findings reinforce the exploratory status of the strategy analysis. The broad distinction between ineffective and effective strategies appears stable, but the finer-grained distinction between analog rotation and feature-matching strategies should be treated with caution pending replication in larger, site-specific samples.

### Robustness checks
To assess the robustness of the three-cluster solution, we conducted three additional checks reported in full in the appendices. First, evaluation of cluster quality across k = 1–8 showed that while silhouette analysis favored k = 2, the k = 2 solution collapsed two qualitatively distinct RT–angle profiles into a single cluster, justifying the adoption of k = 3 on theoretical grounds (see @apx-cluster). Second, dimensionality reduction using PCA, t-SNE, and UMAP consistently recovered the separation of the Ineffective Strategies group from the two effective strategy groups, though the boundary between the Mental Rotation and Flipping/Matching groups was less clear across all three methods, suggesting these may represent ends of a continuum rather than fully discrete clusters (see @apx-dimreduc). Third, re-plotting RT and accuracy functions using the full 0–360° rotation angles provided direct support for the assumption that participants rotate stimuli via the shortest available path: both effective strategy groups showed near-symmetric inverted-U RT functions peaking around 200–240° before declining as angles approached 360°, consistent with shortest-path rotation throughout. The flipping signature of the Flipping/Matching group was again evident, and the Ineffective Strategies group continued to show a broadly decreasing RT function with accuracy near chance at peak angles, replicating the main analysis findings under this alternative parameterization (see @apx-360-angle).

# Discussion

The central aim of this study was to establish whether the LEVANTE mental rotation task measures the same underlying construct across three culturally diverse sites. The answer, across three distinct levels of analysis, is largely yes. The task achieves scalar measurement invariance indicating that its psychometric structure is consistent across cultural contexts. The canonical angular disparity effect replicates across all three sites for 3D stimuli, providing process-level evidence that performance is driven by the spatial demands of the task in the way theory predicts [@shepard1971mental]. And at the level of individual behavioral profiles, the broad distinction between children who engage with the spatial structure of the task and those who do not is recoverable within each site independently. That these three forms of convergent evidence emerge across geographically and culturally diverse samples extends the cross-cultural generalizability of mental rotation research beyond the predominantly Western samples in which most developmental work has been conducted [@davis2021cultural; @henrich2010weirdest], and supports the interpretation that the task is capturing a genuinely spatial construct rather than one that is specific to the cultural or educational contexts in which it was developed.

The demonstration of scalar invariance under a 2PL model establishes that the mental rotation task functions comparably across pilot sites — that items are ordered by difficulty in the same way for children in Bogotá as for children in Leipzig or Ontario, and that ability estimates can be placed on a common latent scale [@putnick2016measurement]. The bar here is a meaningful one. Tasks developed in Western contexts and subsequently exported to non-Western settings do not automatically achieve psychometric equivalence; apparent performance differences can reflect differences in task familiarity, testing format, or culturally-specific knowledge rather than the target construct itself [@citation needed on this point]. The fact that scalar invariance holds here suggests the LEVANTE mental rotation task is not heavily dependent on such factors — the same item features drive difficulty in a similar way regardless of cultural context. This provides a necessary foundation for the analyses that follow, but it also carries a broader implication: with appropriate localization and piloting, spatial cognitive tasks developed in Western samples can achieve genuine psychometric equivalence across diverse populations of children. It should be noted that scalar invariance establishes metric comparability rather than cultural equivalence in any deeper sense. It does not indicate that mental rotation ability is equally developed across sites, and given that LEVANTE sites differ substantially in sampling, recruitment, and testing format, site-level performance differences are not interpreted here.

Beyond establishing that the task measures the same construct across sites, a second and independent question is whether performance is being driven by the spatial properties of the stimuli in the way theory predicts. The key benchmark here is the angular disparity effect first reported by @shepard1971mental: as the rotational distance between a 3D target and its match increases, so too does the time required to judge whether they correspond. This relationship — among the most replicated findings in cognitive psychology [@cooper1973chronometric; @peters2008applications] — has been demonstrated extensively in Western adult samples, but its replication across culturally diverse samples of children had not previously been established. Its consistent emergence here, across all three sites, suggests that the underlying cognitive process it indexes — the spatially continuous transformation of object orientation — is not specific to the cultural or educational contexts in which it was originally documented. This matters because psychometric equivalence and process-level validity are distinct forms of evidence. It is one thing to show that a task achieves comparable difficulty ordering across sites; it is another to show that performance is sensitive to the spatial structure of the stimuli in precisely the way that mental rotation theory predicts. The angular disparity effect provides the latter, and its cross-cultural replication substantially strengthens the case that the task is engaging spatial cognition specifically, rather than some culturally-variable combination of task-general skills.

The picture is less straightforward for 2D stimuli, where RT–angle functions were notably flat across all three sites and age groups. This is a straightforward failure of replication — the original Shepard-Metzler paradigm was developed exclusively using 3D perspective drawings. However, research with adults has shown that the angular disparity effect does extend to 2D stimuli, including alphanumeric characters, letter-like asymmetrical drawings, and pictures of common objects [@borst2011mental; @cooper1973chronometric; @tarr1989mental], indicating that dimensionality alone does not determine whether the effect emerges. The null finding for 2D stimuli in the present study therefore warrants some consideration. Two complementary accounts from the adult literature suggest why simple, familiar animal silhouettes may be particularly unlikely to elicit the effect. Folk and Luce [-@folk1987effects] showed that when stimuli are easy to tell apart, people can make the same/different judgment by simply comparing surface features without rotating at all — bypassing the rotation process entirely and producing flat RT-angle functions. Bethell-Fox and Shepard [-@bethellfox1988mental] showed that even when rotation does occur, familiar stimuli are rotated as integrated wholes rather than piece by piece, producing faster and less angle-sensitive responses than unfamiliar stimuli. These two accounts operate at different levels: the first concerns whether rotation happens at all, and the second concerns how it is carried out when it does. Both suggest flatter RT-angle functions but through different routes. Duck and rabbit silhouettes are both easy to discriminate and highly familiar to young children, making them susceptible to both effects simultaneously. The 3D geometric shapes, by contrast, are genuinely unfamiliar and harder to discriminate, which is likely what drives children toward the full, spatially continuous rotation process that produces the classic RT-angle signature.

A complementary window onto cross-cultural consistency comes from the within-task dynamics analysis, which asked not whether children are sensitive to rotation angle, but whether they show sensitivity to task demands as they unfold in real time. Two patterns emerged consistently across all three sites. First, there was a pronounced jump in residual reaction times at the transition from 2D to 3D stimuli, and this increase was immediate rather than gradual — suggesting that children recalibrate their processing demands at the block boundary rather than adapting incrementally. This is consistent with the between-stimulus difference established in the angular disparity analysis and extends it: children appear sensitive to the qualitative shift in task demands imposed by 3D stimuli, and this sensitivity is cross-culturally robust. Second, within each block, reaction times decreased across trials, indicating a practice effect that operates independently of angular disparity. This is notable given the flat RT-angle functions observed for 2D stimuli: children are getting faster with practice on these items even in the absence of a detectable rotation effect, suggesting that practice-related acceleration and angular disparity sensitivity are capturing distinct aspects of performance. The cross-cultural consistency of both patterns — the block transition jump and the within-block acceleration — extends the picture of convergence established by the measurement invariance and angular disparity analyses, suggesting that not only the psychometric structure and the core cognitive signature of mental rotation, but also the real-time dynamics of task engagement, are broadly stable across cultural contexts.

Having established consistency at the levels of psychometric structure, process-level signatures, and real-time task dynamics, we turn to the exploratory question of whether children vary in the cognitive strategies they employ. Research with adults has established that solvers draw on a range of strategies when performing mental rotation tasks, spanning a holistic-analytic continuum [@gluck2003spatial], and that strategy selection varies across individuals and shifts with task demands [@nazareth2019strategy]. Importantly, such strategies can be identified from behavioral signatures in RT data even in the absence of direct observation — an approach with established precedent in the mental rotation literature [@geiser2006separating; ]. We emphasize, however, that the present analysis is exploratory: the clusters are inferred from behavioral proxies rather than directly observed strategies, the three-cluster solution receives only partial support from silhouette analysis, and replication with larger samples is required before strong conclusions can be drawn. With those caveats in place, the clustering analysis identified three groups with qualitatively distinct behavioral profiles. The Ineffective Strategies group showed very low ability, a flat or slightly negative RT-angle slope, and high response instability, suggesting these children are not engaging with the spatial structure of the items. The Flipping/Matching group — the largest cluster — showed the highest mean ability and an inverted-U RT-angle function consistent with exploiting reflective symmetry (i.e. flipping) at high angular disparities rather than performing full rotation. The Mental Rotation group showed the classic monotonically increasing RT-angle function consistent with spatially continuous rotation, but had lower mean ability than the Flipping/Matching group. This may reflect the greater cognitive cost of full rotation relative to feature-matching strategies [@just1985cognitive; @khooshabeh2013individual], particularly for the 2D stimuli where rotational demands are lower: children who have discovered the more efficient feature-matching approach may outperform those performing the more demanding rotation process. At the cross-cultural level, the broad effective/ineffective distinction was recoverable within each site independently, but the finer boundary between the Mental Rotation and Flipping/Matching clusters was not, suggesting these two groups may represent ends of a continuum rather than discrete categories — a finding consistent with the substantial overlap between them across all three dimensionality reduction methods.

Several limitations of the present study warrant acknowledgment. First, the strategy clustering analysis is exploratory and should be treated with appropriate caution. The clusters are inferred from behavioral proxies rather than directly observed strategies, the three-cluster solution receives only partial support from silhouette analysis, and the Mental Rotation cluster was not independently recoverable in two of the three sites. Replication in larger, site-specific samples is needed before strong conclusions about strategy distributions — and their cultural variation — can be drawn. Second, the present study is limited to a single task from the LEVANTE battery, and it remains to be seen whether the patterns of cross-cultural consistency observed here extend to other spatial tasks, or whether they are specific to mental rotation. Finally, the sample spans ages 5–12 but the youngest age group showed more irregular patterns, particularly for 3D stimuli, likely reflecting limited sample sizes at the lower age range rather than a substantively different developmental trajectory. Future work with larger samples of younger children would help clarify the developmental onset of the cross-cultural signatures observed here.

The present study set out to ask whether a LEVANTE mental rotation task captures the same underlying construct when administered to children across culturally diverse settings. The answer, at multiple levels of analysis, is largely affirmative. The task achieves scalar measurement invariance across three continents, the canonical angular disparity effect replicates robustly for 3D stimuli across all three sites, and the trial-by-trial dynamics of task performance — practice effects and sensitivity to the shift from 2D to 3D stimuli — are consistent across all three sites. These findings do not mean that mental rotation development is uniform across children or contexts — the substantial individual variability within sites, and the tentative evidence for cross-cultural differences in strategy distributions, suggest there is much still to understand about what drives variation in spatial ability across children. But they do mean that the task is measuring something real and consistent — that when a child in Bogotá and a child in Leipzig and a child in Ontario sit down to judge whether two objects match, they are, by and large, engaging the same cognitive process. That is a necessary foundation for the kind of cross-cultural developmental science that the LEVANTE project aims to build.

\clearpage

# References

::: {#refs}
:::

# IRT Ability by Age {.appendix #apx-irt-ability-age}
```{r}
#| label: fig-rt-angle_raw
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI. CAT trials excluded."
#| fig-width: 8
#| fig-height: 8
  
  # Add raw data underneath
  p_mrot_rt_angle_age_2 <- ggplot() +
    # raw per-run medians
    geom_point(
      data = mrot_items_rt_runs,
      aes(x = angle, y = rt_med,
          colour = stimulus),
      size = 0.4,
      alpha = 0.25,
      position = position_jitter(width = 3, height = 0)
    ) +
    # summary CI + lines
    geom_errorbar(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y, ymin = lwr, ymax = upr,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      width = 2,
      linewidth = 0.35,
      alpha = 0.6
    ) +
    geom_point(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      size = 1
    ) +
    geom_line(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      alpha = 0.85
    ) +
    facet_grid(site ~ age_group) +
    labs(
      #title = "Mental Rotation: RT by angle within age groups ±95% CI",
      x = "Rotation angle (°)",
      y = "RT (s)",
      colour = "Stimulus",
      #linetype = "Item Type"
    )+ scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) 
  p_mrot_rt_angle_age_2
```

# Cluster Solution Validation {.appendix #apx-cluster}
To assess the appropriateness of the three-cluster solution, we evaluated cluster quality across k = 1–8 using two complementary metrics: total within-cluster sum of squares (the elbow method) and average silhouette width (@fig-k-sweep-validation). The silhouette analysis identified k = 2 as the statistically optimal solution, with average silhouette width peaking at 0.35 before declining steeply at k = 3 and remaining low thereafter. The elbow plot showed a moderate inflection at k = 3, after which additional clusters yielded diminishing reductions in within-cluster variance. To interrogate whether k = 2 was substantively adequate, we examined the RT–angle functions of the resulting two-cluster solution (@fig-k2-cluster-check). The higher-ability cluster displayed a characteristic inverted-U RT function, rising to a peak around 120° before declining at 160° — a pattern that, in the three-cluster solution, distinguishes the Flipping/Matching and Mental Rotation groups. This indicates that k = 2 collapses two qualitatively distinct effective strategies into a single cluster, obscuring a theoretically meaningful behavioral difference. We therefore adopted k = 3 on both empirical grounds (the elbow inflection) and theoretical grounds (the interpretability of the resulting RT profiles).

```{r}
#| label: fig-k-sweep-validation
#| eval: true
#| include: true
#| fig-width: 6
#| fig-height: 3
#| fig-cap: "**Cluster validation metrics: Elbow Method (left) and Silhouette Analysis (right).** The elbow plot shows a bend at k=3, and silhouette width peaks at k=2 before declining, providing partial support for the 3-cluster solution."

p_elbow <- fviz_nbclust(global_input, kmeans, method = "wss", k.max = 8) +
  geom_vline(xintercept = 3, linetype = 2, color = pal$red) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

p_sil <- fviz_nbclust(global_input, kmeans, method = "silhouette", k.max = 8) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

grid.arrange(p_elbow, p_sil, nrow = 1)
```

```{r}
#| label: fig-k2-cluster-check
#| eval: true
#| include: true
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "**RT–angle functions for the k=2 cluster solution.** The higher-ability cluster displays an inverted-U RT curve consistent with a mix of Mental Rotation and Flipping/Matching strategies, suggesting that k=2 collapses two qualitatively distinct effective strategies into a single group."

# --- 1. Set up Input (Same as before, removing Learning_Rate for clarity) ---
analysis_data_k2 <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, Instability, Learning_Rate), is.finite))

global_input_k2 <- analysis_data_k2 %>%
  select(Baseline_Speed, Rel_Slope, Instability, Abs_Slope, Learning_Rate) %>%
  scale()

# --- 2. Run K-Means with K=2 ---
set.seed(123)
k_global_2 <- kmeans(global_input_k2, centers = 2, nstart = 25)
analysis_data_k2$Global_Cluster <- as.factor(k_global_2$cluster)

# --- 3. Label the 2 Clusters ---
# We define them simply by Ability (Low vs High) since we can't split by slope yet
cluster_stats_k2 <- analysis_data_k2 %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability), 
    .groups = "drop"
  )

# Identify Lower Ability Cluster
low_eff_id <- cluster_stats_k2 %>% arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
# The other one is High Efficiency
high_eff_id <- cluster_stats_k2 %>% arrange(desc(mean_ability)) %>% slice(1) %>% pull(Global_Cluster)

analysis_data_k2 <- analysis_data_k2 %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id ~ "Low Efficiency (Cluster 1)",
    Global_Cluster == high_eff_id ~ "High Efficiency (Cluster 2)"
  ))

k2_colors <- c(
  "Low Efficiency (Cluster 1)"  = RColorBrewer::brewer.pal(3, "Set1")[1],
  "High Efficiency (Cluster 2)" = RColorBrewer::brewer.pal(3, "Set1")[2]
)

# --- 4. Plot the Curves (Visual Check) ---
# Join labels to raw data
rt_by_strategy_k2 <- mrot_rich_data %>%
  left_join(
    analysis_data_k2 %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle))

# Calculate User Means
rt_user_angle_k2 <- rt_by_strategy_k2 %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

# Calculate Group Means
rt_summary_k2 <- rt_user_angle_k2 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# plot
ggplot(rt_summary_k2, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_log_rt - 1.96*se_log_rt, 
                  ymax = mean_log_rt + 1.96*se_log_rt, 
                  fill = Strategy_Label), alpha = 0.2, colour = NA) +
  scale_colour_manual(values = k2_colors) +
  scale_fill_manual(values = k2_colors) +
  labs(
    title    = NULL,
    subtitle = NULL,
    y = "Mean log(RT) (participant-weighted)", 
    x = "Rotation angle (degrees)",
    colour = "Strategy", fill = "Strategy"
  ) +
  theme(legend.position = "bottom")
```
\clearpage
  
# Dimensionality Reduction Checks {.appendix #apx-dimreduc}
To further validate the cluster structure, we examined whether the three-cluster solution was recoverable using alternative dimensionality reduction methods: PCA (@fig-pca-check), t-SNE (@fig-tsne-check), and UMAP (@fig-umap-check). PCA loadings were also inspected to characterize the underlying feature structure (@fig-pca-loadings).

Results were consistent across all three methods in one key respect: the Ineffective Strategies group was clearly and reliably separated from the two effective strategy groups, appearing as a distinct region in all three projections — the top-left in the t-SNE, the negative end of PC1 in the PCA, and the bottom-left in the UMAP. This convergent separation provides strong evidence that this group occupies a genuinely distinct region of the feature space. The boundary between the Mental Rotation and Flipping/Matching groups was, however, substantially less clear. Green and blue points overlapped considerably across all three visualizations, suggesting these two groups may represent ends of a continuum rather than fully discrete clusters — a finding consistent with the silhouette analysis favouring k = 2.

Inspection of the PCA loadings offered interpretive support for the feature set. PC1 loaded positively on Baseline Speed, Learning Rate, and Absolute Slope, and negatively on Instability, reflecting a general processing efficiency dimension: children who respond faster, improve more across trials, and show less trial-to-trial noise score higher on this component. PC2 loaded most strongly on Relative Slope and Absolute Slope — both positive — and negatively on Learning Rate, reflecting a rotation cost dimension that is largely independent of general speed. The fact that these two components map onto theoretically meaningful and distinct dimensions provides confidence that the clustering features are not redundant.
```{r}
#| label: fig-tsne-check
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**t-SNE visualization of strategy clusters.** The Ineffective Strategies group forms a distinct region in the top-left, while the Mental Rotation and Flipping/Matching groups show considerable overlap in the central and right regions, suggesting softer boundaries between the two effective strategy groups."

library(Rtsne)

tsne_input <- analysis_data %>%
    select(user_id, Learning_Rate, Baseline_Speed, Rel_Slope, Abs_Slope, Instability) %>%
    distinct() %>% 
    na.omit()

set.seed(42) 
tsne_result <- Rtsne(
    as.matrix(tsne_input %>% select(-user_id)), 
    dims = 2, 
    perplexity = 30, 
    verbose = FALSE, 
    max_iter = 500
)

tsne_plot_data <- tsne_input %>%
    mutate(
        tSNE_1 = tsne_result$Y[,1],
        tSNE_2 = tsne_result$Y[,2]
    ) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(tsne_plot_data, aes(x = tSNE_1, y = tSNE_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "Dimension 1",
        y = "Dimension 2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-check
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**PCA projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group separates clearly along PC1, while the Mental Rotation and Flipping/Matching groups overlap along both components."

pca_result <- prcomp(tsne_input %>% select(-user_id), center = TRUE, scale. = TRUE)

pca_plot_data <- tsne_input %>%
    mutate(PC1 = pca_result$x[,1], PC2 = pca_result$x[,2]) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "PC1",
        y = "PC2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-loadings
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**PCA loadings for PC1 and PC2.** PC1 reflects a general processing efficiency dimension, loading positively on Baseline Speed, Learning Rate, and Absolute Slope, and negatively on Instability. PC2 reflects a rotation cost dimension, driven primarily by Relative and Absolute Slope."

library(tidytext)

loadings_df <- as.data.frame(pca_result$rotation) %>%
  rownames_to_column(var = "Angle_Condition")

loadings_long <- loadings_df %>%
  select(Angle_Condition, PC1, PC2) %>%
  pivot_longer(
    cols = c(PC1, PC2), 
    names_to = "Principal_Component", 
    values_to = "Loading"
  )

ggplot(loadings_long, aes(
    x = Loading, 
    y = reorder_within(Angle_Condition, Loading, Principal_Component), 
    fill = Loading > 0
  )) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Principal_Component, scales = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("TRUE" = pal$blue, "FALSE" = pal$red)) +
  labs(
    title    = NULL,
    subtitle = NULL,
    x = "Loading Weight",
    y = "Feature"
  ) +
  theme(
    #strip.text = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11),
    #panel.grid.minor = element_blank()
  )
```

::: {.landscape}
```{r}
#| label: fig-umap-check
#| fig-width: 8.5
#| fig-height: 5
#| fig-cap: "**UMAP projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group forms a distinct region in the bottom-left, while the Mental Rotation and Flipping/Matching groups show substantial overlap in the central and right regions."

set.seed(42) 
umap_result <- umap(as.matrix(tsne_input %>% select(-user_id)))

umap_plot_data <- tsne_input %>%
    mutate(UMAP_1 = umap_result$layout[,1], UMAP_2 = umap_result$layout[,2]) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(umap_plot_data, aes(x = UMAP_1, y = UMAP_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) + 
    labs(x = "UMAP Dimension 1", y = "UMAP Dimension 2") +
    theme(legend.position = "bottom")
```
:::

# Full 0–360° Angle Functions check {.appendix #apx-360-angle}
The primary analyses modeled angular disparity — the shortest rotational distance to upright, capped at 180° — on the assumption that participants rotate stimuli via the most efficient path. To test this assumption, we re-ran the clustering using full 0–360° rotation angles and plotted RT and accuracy functions across the complete angular range (@fig-rt-angle-by-strategy-360, @fig-accuracy-angle-by-strategy-360). We verified that the angular disparity variable was correctly computed, confirming that full rotation angles were accurately recoded to their shortest-path equivalents (e.g., a 320° rotation has a disparity of 40°).

The results were broadly consistent with the disparity-based analyses and provided direct support for the shortest-path assumption. Both the Mental Rotation and Flipping/Matching groups showed RT functions that increased from 20° to a peak around 200–240°, then declined as angles approached 320°, forming the expected near-symmetric inverted-U around 180°. This pattern is precisely what would be predicted if participants always rotate via the shorter available path, and would not be expected under a fixed-direction rotation strategy.

The flipping signature observed in the Flipping/Matching group was again evident in the full-angle data: their RT function rose more steeply than the Mental Rotation group and remained elevated through 240° before declining, consistent with a strategy that exploits reflective symmetry at high disparities. Accuracy for both effective strategy groups showed a corresponding U-shaped function, declining to a minimum around 160–200° before recovering at higher angles, with the Flipping/Matching group maintaining the highest accuracy throughout. The Ineffective Strategies group again showed a broadly decreasing RT function across the full angular range, with accuracy remaining near chance at peak angles and showing only a modest recovery above 240°, further supporting the interpretation that these children are not engaging with the spatial structure of the items.
```{r}
#| label: robustness-check-all-angles
#| results: "hide"

# Ability estimate - keep site + user
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    
    # Extract full 0-360° angle directly from item_original
    angle = as.integer(str_extract(item_original, "\\d{3}$")),
    
    # Calculate shortest distance to center (disparity) for the model
    ang_disparity = pmin(angle, 360 - angle),
    
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(ang_disparity)), 
    trial_s  = as.numeric(scale(trial_number))
  )

# --- Doria's FIX: Set Sum-to-Zero Contrasts ---
# By default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the GRAND MEAN slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
# Fit the rich mixed model
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# Extract subject-level random effects (deviations)
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# Fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# Compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope      = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope      = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# 4. Compute Instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(site, user_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

# 5. Join and Prepare for Clustering
rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(instability_data, by = c("site", "user_id")) %>% 
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  # Filter to ensure all needed features are finite
  # UPDATED: Added Learning_Rate to the list
  filter(if_all(c(ability, Abs_Slope, Rel_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))

# Check: ang_disparity should never exceed 180
summary(mrot_rich_data$ang_disparity)

# Spot check a few items
mrot_rich_data %>%
  select(item, item_original, angle, ang_disparity) %>%
  distinct() %>%
  arrange(item_original) %>%
  head(20)
```

```{r}
#| label: clustering-execution-360
#| include: false

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1],
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], 
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  
)

analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Rel_Slope, Abs_Slope, 
                  Instability, Learning_Rate), is.finite))
global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()
set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)
cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
effective_clusters <- cluster_stats %>%
  filter(Global_Cluster != low_eff_id) %>%
  arrange(mean_slope)
analytical_id <- effective_clusters %>% slice(1) %>% pull(Global_Cluster)
holistic_id   <- effective_clusters %>% slice(2) %>% pull(Global_Cluster)
analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id    ~ "Ineffective Strategies",
    Global_Cluster == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    Global_Cluster == holistic_id   ~ "Mental Rotation"
  ))
ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Cluster Result: Ability vs. Relative Slope (full 0-360° angles)",
    x = "Ability", y = "Relative Slope"
  )
```

```{r}
#| label: fig-rt-angle-by-strategy-360
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Participant-weighted RT–angle functions by strategy cluster (numbers indicate unique participants contributing at each angle; shaded bands show 95% CIs) (Note: Re-test data removed)."

rt_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

rt_user_angle <- rt_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_angle_person_summary <- rt_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

missing_cols <- setdiff(levels(rt_angle_person_summary$Strategy_Label), names(my_colors))
stopifnot(length(missing_cols) == 0)

ggplot(rt_angle_person_summary, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_log_rt - 1.96 * se_log_rt,
        ymax = mean_log_rt + 1.96 * se_log_rt,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean log(RT) (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: fig-accuracy-angle-by-strategy-360
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Participant-weighted Accuracy-angle functions by strategy cluster."

acc_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_user_angle <- acc_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_acc = mean(correct, na.rm = TRUE), .groups = "drop")

acc_summary <- acc_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users  = n_distinct(user_id),
    mean_acc = mean(user_mean_acc, na.rm = TRUE),
    se_acc   = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_users),
    .groups  = "drop"
  )

ggplot(acc_summary, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc,
        ymax = mean_acc + 1.96 * se_acc,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean Accuracy (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

<!--
####################### ROBUSTNESS CHECK - BOOTSTRAPPING ##########################
```{r}
#| label: cluster-stability-bootstrap
#| include: false

vars   <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")
k      <- 3
B      <- 500
nstart <- 25
set.seed(123)

df_full <- analysis_data %>%
  select(user_id, Strategy_Label, all_of(vars)) %>%
  distinct(user_id, .keep_all = TRUE) %>%
  filter(if_all(all_of(vars), is.finite)) %>%
  mutate(
    Strategy_Label = factor(
      Strategy_Label,
      levels = names(my_colors)   
    )
  )

stopifnot(nlevels(df_full$Strategy_Label) == 3)
stopifnot(n_distinct(df_full$user_id) == nrow(df_full))

ref_levels <- levels(df_full$Strategy_Label)
y_ref      <- df_full$Strategy_Label
ids        <- df_full$user_id
n          <- nrow(df_full)

# scaling: lock to full-sample mean/sd so bootstrap runs are comparable
mu  <- df_full %>% summarise(across(all_of(vars), \(x) mean(x, na.rm = TRUE))) %>% as.list()
sdv <- df_full %>% summarise(across(all_of(vars), \(x) sd(x,   na.rm = TRUE))) %>% as.list()

X_full <- df_full %>%
  mutate(across(all_of(vars), ~ (. - mu[[cur_column()]]) / sdv[[cur_column()]])) %>%
  select(all_of(vars)) %>%
  as.matrix()

# helpers
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

jaccard <- function(set_a, set_b) {
  inter <- length(intersect(set_a, set_b))
  uni   <- length(union(set_a, set_b))
  if (uni == 0) NA_real_ else inter / uni
}

all_perms_3 <- list(
  c(1,2,3), c(1,3,2),
  c(2,1,3), c(2,3,1),
  c(3,1,2), c(3,2,1)
)

# Match bootstrap cluster IDs to reference labels by maximum membership overlap
best_perm_by_overlap <- function(cl_hat, y_ref, ref_levels, all_perms_3) {
  tab    <- table(cl_hat, factor(y_ref, levels = ref_levels))
  scores <- sapply(all_perms_3, function(p) sum(tab[cbind(1:3, p)]))
  all_perms_3[[which.max(scores)]]
}

# reference k-means (reproduces clustering-execution exactly)
# UPDATED: label assignment now mirrors the two-step logic in clustering-execution:
#   step 1 — lowest mean ability  → "Ineffective Strategies"
#   step 2 — of the remainder, flatter Rel_Slope → "Other Effective...", 
#             steeper → "Mental Rotation"
# We use this to build ref_centers aligned to ref_levels.

# need ability for label assignment (not a clustering var, so grab separately)
ability_lookup <- df_full %>%
  left_join(analysis_data %>% distinct(user_id, ability), by = "user_id")

set.seed(123)
km_ref <- kmeans(X_full, centers = k, nstart = nstart)
cl_ref <- km_ref$cluster

# replicate the production label-assignment rule on the reference run
cluster_stats_ref <- tibble(
  cluster      = 1:k,
  mean_ability = tapply(ability_lookup$ability, cl_ref, mean, na.rm = TRUE),
  mean_rel_slope = tapply(df_full$Rel_Slope,   cl_ref, mean, na.rm = TRUE)
)

low_eff_id_ref  <- cluster_stats_ref %>% arrange(mean_ability) %>% slice(1) %>% pull(cluster)
effective_ref   <- cluster_stats_ref %>% filter(cluster != low_eff_id_ref) %>% arrange(mean_rel_slope)
analytical_id_ref <- effective_ref %>% slice(1) %>% pull(cluster)
holistic_id_ref   <- effective_ref %>% slice(2) %>% pull(cluster)

# map cluster integer to ref_levels index (used to order ref_centers rows)
label_order_ref <- integer(k)
label_order_ref[low_eff_id_ref]   <- which(ref_levels == "Ineffective Strategies")
label_order_ref[analytical_id_ref] <- which(ref_levels == "Other Effective Strategies (Flipping/Matching)")
label_order_ref[holistic_id_ref]   <- which(ref_levels == "Mental Rotation")

# ref_centers[i, ] corresponds to ref_levels[i]
ref_centers <- km_ref$centers[order(label_order_ref), , drop = FALSE]

# bootstrap loop
res <- map_dfr(1:B, function(b) {

  set.seed(1000 + b)

  idx        <- sample.int(n, size = n, replace = TRUE)
  idx_unique <- unique(idx)

  tab_ref_unique <- table(factor(y_ref[idx_unique], levels = ref_levels))
  min_ref_unique <- min(tab_ref_unique)
  min_ref_prop   <- min_ref_unique / length(idx_unique)
  n_unique       <- length(idx_unique)

  X_b <- X_full[idx, , drop = FALSE]
  km  <- kmeans(X_b, centers = k, nstart = nstart)

  # assign all full-sample participants to nearest bootstrap center
  centers_b <- km$centers
  dists <- sapply(1:k, function(j) {
    rowSums((X_full - matrix(centers_b[j, ],
                             nrow = nrow(X_full), ncol = ncol(X_full),
                             byrow = TRUE))^2)
  })
  cl_hat <- max.col(-dists)

  # resolve label switching via maximum overlap with reference labels
  perm  <- best_perm_by_overlap(cl_hat, y_ref, ref_levels, all_perms_3)
  y_hat <- factor(ref_levels[perm[cl_hat]], levels = ref_levels)

  ari <- ARI(y_ref, y_hat)

  # per-cluster Jaccard
  jac <- map_dbl(ref_levels, function(lab) {
    jaccard(ids[y_ref == lab], ids[y_hat == lab])
  })
  names(jac) <- paste0("J_", make.names(ref_levels))

  # mean centroid displacement vs reference
  centers_b_aligned <- centers_b[perm, , drop = FALSE]
  center_shift      <- mean(sqrt(rowSums((centers_b_aligned - ref_centers)^2)))

  tibble(
    replicate      = b,
    ARI            = ari,
    n_unique       = n_unique,
    min_ref_unique = min_ref_unique,
    min_ref_prop   = min_ref_prop,
    center_shift   = center_shift,
    !!!as.list(jac)
  )
})

# summaries
stability_summary <- res %>%
  summarise(
    B                   = n(),
    ARI_median          = median(ARI, na.rm = TRUE),
    ARI_IQR             = IQR(ARI, na.rm = TRUE),
    center_shift_median = median(center_shift, na.rm = TRUE),
    center_shift_IQR    = IQR(center_shift, na.rm = TRUE),
    across(starts_with("J_"),
           list(median = ~ median(.x, na.rm = TRUE),
                IQR    = ~ IQR(.x, na.rm = TRUE)))
  )

stability_summary

ari_tail <- res %>%
  summarise(
    pct_ARI_below_0.70 = mean(ARI < 0.70, na.rm = TRUE) * 100,
    pct_ARI_below_0.60 = mean(ARI < 0.60, na.rm = TRUE) * 100,
    min_ARI            = min(ARI, na.rm = TRUE)
  )

ari_tail
```

```{r}
#| label: cluster-stability-reporting
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)

ari_med    <- stability_summary$ARI_median
ari_iqr    <- stability_summary$ARI_IQR
pct_ari_70 <- ari_tail$pct_ARI_below_0.70
pct_ari_60 <- ari_tail$pct_ARI_below_0.60
min_ari    <- ari_tail$min_ARI

# UPDATED: pull Jaccard stats using new label names
j_med <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_median")]]
})
j_iqr <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_IQR")]]
})
names(j_med) <- ref_levels
names(j_iqr) <- ref_levels

j_ineff_med  <- j_med["Ineffective Strategies"]
j_ineff_iqr  <- j_iqr["Ineffective Strategies"]
j_other_med  <- j_med["Other Effective Strategies (Flipping/Matching)"]
j_other_iqr  <- j_iqr["Other Effective Strategies (Flipping/Matching)"]
j_mrot_med   <- j_med["Mental Rotation"]
j_mrot_iqr   <- j_iqr["Mental Rotation"]

rho_shift <- suppressWarnings(
  cor(res$ARI, res$center_shift, method = "spearman", use = "complete.obs")
)

# ARI distribution across bootstrap resamples
ggplot(res, aes(x = ARI)) +
  geom_histogram(bins = 40, fill = "#3498db", alpha = 0.8) +
  geom_vline(xintercept = median(res$ARI, na.rm = TRUE), 
             linetype = "dashed", colour = "black") +
  labs(x = "ARI (vs reference clustering)", y = "Count",
       title = paste0("Bootstrap stability (B = ", nrow(res), ")"),
       subtitle = paste0("Median ARI = ", round(median(res$ARI, na.rm = TRUE), 3)))

# Per-cluster Jaccard
res %>%
  select(replicate, starts_with("J_")) %>%
  pivot_longer(-replicate, names_to = "Cluster", values_to = "Jaccard") %>%
  mutate(Cluster = str_remove(Cluster, "^J_")) %>%
  ggplot(aes(x = Jaccard, fill = Cluster)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  facet_wrap(~ Cluster, ncol = 1) +
  labs(x = "Jaccard index", y = "Count") +
  theme(legend.position = "none")

```
# checks
```{r}
analysis_data %>%
  count(Strategy_Label)

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(across(all_of(c("Baseline_Speed", "Rel_Slope", "Abs_Slope", 
                             "Instability", "Learning_Rate")), 
                   ~ round(mean(.x, na.rm = TRUE), 3)))

analysis_data %>%
  select(Strategy_Label, Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  pivot_longer(-Strategy_Label, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Strategy_Label, y = value, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~ variable, scales = "free_y") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

####################### ROBUSTNESS CHECK - WITHIN PERSON STRATEGY SWITCHING (DORIA'S ANALYSIS) ##########################

### Strategy Stability Analysis (Split-Half Reliability)

To investigate whether children consistently use a single cognitive strategy throughout the task or switch strategies over time (e.g., discovering a shortcut), we performed a **Split-Half Reliability Analysis**.

**Methodology:**
We partitioned each participant's data into two sequential halves: the first 50% of trials (Early Phase) and the last 50% of trials (Late Phase). For each half, we estimated a "Raw Slope" using a simple linear regression of reaction time on angle (`log_rt ~ angle`). This slope serves as a proxy for the participant's rotation strategy during that specific period. 

We then correlated the *First Half Slope* with the *Second Half Slope* to quantify stability. A high correlation ($r \approx 1$) would indicate a fixed strategy, while a low correlation ($r \approx 0$) suggests dynamic behavior or strategy switching.

```{r}
#| label: fig-strategy-switching
#| fig-cap: "Strategy Stability: Comparison of Rotation Slope in First Half vs. Second Half of trials. Points below the diagonal suggest a switch to a flatter strategy (e.g., discovering the shortcut)."


# 1. Split Data into Halves
# We determine the midpoint trial for each user
split_data <- mrot_rich_data %>%
  group_by(user_id) %>%
  mutate(
    Midpoint = median(trial_number, na.rm = TRUE),
    Half = ifelse(trial_number <= Midpoint, "First_Half", "Second_Half")
  ) %>%
  ungroup() %>%
  # Filter out bad data points just to be safe
  filter(is.finite(log_rt), is.finite(angle))

# 2. Calculate Simple Slope for Each Half (Using simple lm per subject)
# This is a fast proxy for the complex mixed-model slope
slope_stability <- split_data %>%
  group_by(user_id, Half) %>%
  summarise(
    # We only care about the slope (coefficient for angle)
    # We use a safe lm() call wrapped in tryCatch to handle kids with too few trials
    Raw_Slope = tryCatch(
      #coef(lm(log_rt ~ angle, data = pick(everything())))["angle"], 
      # Use distance from 0 so the slope represents 'cost of rotation'
  coef(lm(log_rt ~ pmin(angle, 360 - angle), data = pick(everything())))[2],
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  # Reshape to Wide Format (Columns: Slope_First, Slope_Second)
  tidyr::pivot_wider(names_from = Half, values_from = Raw_Slope, names_prefix = "Slope_") %>%
  # Join with our Final Strategy Labels to color-code
  left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id") %>%
  filter(!is.na(Strategy_Label), !is.na(Slope_First_Half), !is.na(Slope_Second_Half))

# 3. Calculate Correlation (Quantify Stability)
correlation <- cor(slope_stability$Slope_First_Half, slope_stability$Slope_Second_Half, use = "complete.obs")

# 4. Visualization
ggplot(slope_stability, aes(x = Slope_First_Half, y = Slope_Second_Half, color = Strategy_Label)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") + # The "Stability Line"
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Strategy Switching Check (Split-Half Reliability)",
    subtitle = paste0("Correlation between halves: r = ", round(correlation, 2), 
                      ". Points below diagonal = Strategy became flatter (faster)."),
    x = "Slope (First Half of Trials)",
    y = "Slope (Second Half of Trials)"
  ) +
  coord_fixed() # Keeps the aspect ratio square so the diagonal is 45 degrees
```

```{r}
# Doria's original code for examining up to 360 
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    
    # 1. Extract the base angle (e.g., 40)
    base_angle = as.integer(str_extract(item, "\\d+")), 
    
    # 2. Detect if the original item was a "Negative/Counter-Clockwise" rotation
    # Looking for 'rn' prefix in item_original (e.g., rn040)
    is_ccw = str_detect(item_original, "^rn"),
    
    # 3. RECONSTRUCT the full 0-360 angle
    angle = ifelse(is_ccw, 360 - base_angle, base_angle),
    
    # 4. Calculate shortest distance to center (disparity) for the model
    # This ensures the model still measures 'Rotation Cost' correctly
    ang_disparity = pmin(angle, 360 - angle),
    
    stimulus = as.factor(stimulus),
    # SCALE THE DISPARITY, NOT THE RAW ANGLE
    angle_s  = as.numeric(scale(ang_disparity)), 
    trial_s  = as.numeric(scale(trial_number))
  )

#  Issues: the above code is only only catches rabbit counter-clockwise items (rn040, rn080 etc.), and misses duck and shape items. It is also not necessary to deconstruct item_original as the info. we need is already there. item_original already contains the full angle directly (e.g., an11240 = 240).

# Fionnuala checking
# Check if column exists
"item_original" %in% names(mrot_rt_trial_no_retest)

# If it does, see what's in it
mrot_rt_trial_no_retest$item_original |> unique() |> sort()

# problem
# str_detect(item_original, "^rn") is wrong — it only catches rabbit counter-clockwise, missing dn, an, bn etc.

#diagnose
# Step 1: Check is_ccw detection by stimulus
mrot_rich_data %>%
  distinct(stimulus, item_original, is_ccw, base_angle, angle) %>%
  arrange(stimulus, item_original) %>%
  print(n = 50)

mrot_rich_data %>%
  count(stimulus, angle) %>%
  pivot_wider(names_from = stimulus, values_from = n) %>%
  arrange(angle) %>%
  print(n = 20)
```
-->