---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "Fionnuala O'Reilly"
    orcid: 0000-0002-4355-9088
    affiliations:
      - ref: stanford
  - name: "Xingyao Xiao"
    orcid: 0000-0001-8430-0438
    affiliations:
      - ref: stanford
  - name: "Rebecca Zhu"
    orcid: 0000-0003-0957-5962
    affiliations:
      - ref: stanford
  - name: "Mika Braginsky"
    orcid: 0000-0001-9039-3220
    affiliations:
      - ref: stanford
  - name: "Nilàm Ram"
    orcid: 0000-0003-1671-5257
    affiliations:
      - ref: stanford
  - name: "Benjamin W. Domingue"
    orcid: 0000-0002-3894-9049
    affiliations:
      - ref: stanford
  - name: "Michael C. Frank"
    orcid: 0000-0002-7551-4378
    affiliations:
      - ref: stanford
    corresponding: true
    email: mcfrank@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["cognitive development", "mental rotation", "spatial reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    # fig-pos: H        
    # tbl-pos: H  

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

Spatial reasoning, broadly defined as the ability to generate, retain, and manipulate visuospatial information is a fundamental cognitive skill [@lohman2013spatial]. It allows us to form and transform mental images of 2D shapes and 3D objects (e.g., by rotating, bending, or folding them), and to represent how spatial relations between locations, shapes, and objects change over time [@frick2014development; @newcombe2014thinking].

Spatial ability is important for a wide range of educational outcomes [@wai2018spatial], including creative domains such as music [@padulo2020effects; @hetland2000learning] and visual arts [@kell2013creativity; @winner2013relationship] as well as STEM subjects [@wai2009spatial] and, most notably, math [@atit2022examining; @xie2020spatial; @mix2012relation; @gunderson2012relation]. Given it’s importance for educational outcomes, a number of attempts have been made to train spatial reasoning skills. For example, @hawes2015effects reported that, compared with an active control group, 6–8-year-old children who received mental rotation training showed significant gains on two separate 2D mental rotation measures, as well as more general improvements in spatial reasoning on an untrained spatial task. @uttal2013malleability conducted a meta-analysis of 217 studies with child and adult samples to examine the size and generalizability of spatial skills training effects. Across studies, the mean effect size for training compared with control conditions was Hedges’ g = 0.47 (SE = 0.04). @gilligan2020first showed that training spatial reasoning skills had	near,	intermediate	and	far	transfer	effects in 8-year old children [@gilligan2020first]. Similarly,in a sample of elementary school students, @lowrie2017visuospatial showed that those who participated in a 10-week visuospatial reasoning (VSR) program achieved substantially higher spatial reasoning scores than students in the control group, and also demonstrated improved mathematics performance.  	

to be continued

## Reaction time

## Cross-country comparisons

## Strategies

```{r}
# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
```

```{r}
#| echo: false
source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)
```

```{r}
#| label: dev-trends
#| fig-cap: "Developmental trends in Mental Rotation ability by site."
#| fig-width: 8
#| fig-height: 6

# colnames(all_scores)
# table(all_scores$model)
mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!site_labels, .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

mrot_irt <- ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  labs(
    #title = "Developmental trends in Mental Rotation ability by site",
    subtitle = "Ability estimates smoothed with GAM",
    x = "Age (years)",
    y = "Ability (theta)",
    colour = "Site"
  )
mrot_irt
```

```{r}
#| label: ability-rt
#| fig-cap: "IRT Ability (Theta) vs log RT."
#| fig-width: 8
#| fig-height: 6

# get shape and rotation data
mrot_rt_sum <- all_sum_scores %>%
  filter(item_task == "mrot") %>%
  select(site, language, data) %>%
  unnest(data) %>%                         
  mutate(
    correct = as.numeric(correct),
    rt_s    = as.numeric(rt_numeric) / 1000
  ) %>%
  filter(is.finite(rt_s), rt_s > 0) %>%     
  mutate(log_rt = log(rt_s))

# collapse to run level RT summaries
mrot_rt_runs <- mrot_rt_sum %>%
  left_join(run_data %>% 
              select(site, user_id, run_id, age), 
            by = c("site", "user_id", "run_id")) %>%
  group_by(site, user_id, run_id) %>%
  summarise(
    # trial-level RT summaries for correct items
    logrt_median = median(log_rt[correct == 1], na.rm = TRUE),
    logrt_mean   = mean(log_rt[correct == 1],  na.rm = TRUE),
    age = first(age),
    n_items = n_distinct(item_uid),
    .groups = "drop"
  )

ability_runs <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>%
  select(site, user_id, run_id, ability = metric_value, age) %>%
  group_by(site, user_id, run_id) %>%               
  summarise(
    ability = mean(ability, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# join RT and ability
mrot_speed_ability <- ability_runs %>%
  inner_join(
    mrot_rt_runs %>% select(site, user_id, run_id, logrt_mean),
    by = c("site", "user_id", "run_id")
  )

p_mrot_speed_ability <- mrot_speed_ability %>%
  mutate(age_binned = as_factor(floor(age))) %>%
  ggplot(aes(x = logrt_mean, y = ability, colour = age_binned)) +
  geom_point(alpha = 0.55, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ site) +
  labs(
    #title = "IRT Ability (Theta) vs log RT",
    subtitle = "Each point = participant run; site-wise linear fits",
    x = "log(RT, seconds)",
    y = "IRT ability (theta)",
    colour = "Age (years)"
  )
p_mrot_speed_ability
```
\clearpage

```{r}
#| label: trial-level-rts
#| echo: false
#| message: false
#| cache: true

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%        
  tidyr::unnest(data) %>%              
  dplyr::mutate(
    correct = as.numeric(correct),
    rt_s    = as.numeric(rt_numeric) / 1000,
    log_rt  = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))

```

```{r}
#| label: fig-rt-angle
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 8, 11, 14)

mrot_items_rt_runs <- mrot_rt_trial %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle    = as.integer(stringr::str_extract(item, "\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med, na.rm = TRUE),
    sd  = sd(rt_med, na.rm = TRUE),
    n   = dplyr::n(),
    se  = sd / sqrt(pmax(n, 1)),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  # raw data layer added below (step 2) ...
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  )
p_mrot_rt_angle_age
```
\clearpage

```{r}
#| label: fig-rt-angle-per-run
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI."
#| fig-width: 8
#| fig-height: 8

# Add raw data underneath
p_mrot_rt_angle_age <- ggplot() +
  # raw per-run medians
  geom_point(
    data = mrot_items_rt_runs,
    aes(x = angle, y = rt_med,
        colour = stimulus),
    size = 0.4,
    alpha = 0.25,
    position = position_jitter(width = 3, height = 0)
  ) +
  # summary CI + lines
  geom_errorbar(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y, ymin = lwr, ymax = upr,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    width = 2,
    linewidth = 0.35,
    alpha = 0.6
  ) +
  geom_point(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    size = 1
  ) +
  geom_line(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    alpha = 0.85
  ) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",
    colour = "Stimulus",
    #linetype = "Item Type"
  )
p_mrot_rt_angle_age
```
\clearpage

# RT Residualization
To isolate pure processing speed from item difficulty (e.g., the added difficulty of 3D objects), we fitted a Linear Mixed Model (LMM) residualizing log reaction times against rotation angle, trial number, and stimulus type. 

```{r}
#| label: residualization
#| echo: false
#| message: false
#| cache: true

# 1. Prepare Data for Modeling
# Ensure mrot_rt_trial has the necessary columns derived from 'item'
model_rt_data <- mrot_rt_trial %>%
  mutate(
    # Extract "rabbit" from "rabbit_040" if not already present
    stimulus = stringr::str_extract(item, "^[a-z]+"), 
    angle = as.integer(stringr::str_extract(item, "\\d+")) 
  ) %>%
  filter(!is.na(angle), !is.na(log_rt))

# 2. FIT FULL MODEL (Residualization)
# Controlling for angle, trial, stimulus, and item grouping
m_rt_full <- lmer(
  log_rt ~ angle + trial_number + stimulus + item_group + (1|item), 
  data = model_rt_data
)

# 3. Extract Residuals (Person Speed)
# These represent speed after accounting for item difficulty
model_rt_data <- model_rt_data %>%
  mutate(rt_resid_item = residuals(m_rt_full))

rt_residuals_person <- model_rt_data %>%
  group_by(user_id, run_id) %>%
  summarise(
    rt_residual_mean = mean(rt_resid_item, na.rm=TRUE),
    rt_residual_sd   = sd(rt_resid_item, na.rm=TRUE),
    n_items = n(),
    .groups = "drop"
  )

# Save for later
#write_csv(rt_residuals_person, here::here("..", "levante-pilots", "02_scoring_outputs", "scores", "rt_residuals_corrected.csv"))
```

```{r}
#| label: within-person-rt
#| echo: false
#| message: false
#| cache: true

# 1. Decompose RT for Within-Person Analysis
within_person_data <- model_rt_data %>%
  filter(!is.na(log_rt), !is.na(correct)) %>%
  group_by(user_id) %>%
  mutate(
    rt_person_mean = mean(log_rt, na.rm = TRUE),
    rt_within = log_rt - rt_person_mean # >0 means slower than my usual
  ) %>%
  ungroup()

# 2. Fit GLMM (Logistic Regression)
# Does slowing down predict correctness?
m_within <- glmer(
  correct ~ rt_within + rt_person_mean + angle + item_group + trial_number + (1|user_id) + (1|item),
  data = within_person_data,
  family = binomial,
  nAGQ = 0 # Optimization for speed
)
```



\clearpage

# Cross-Country Comparisons (Bivariate Model)

## Cross-Population Differences in Ability and Speed

To simultaneously evaluate developmental trends and cross-cultural differences, we fitted a multivariate Bayesian regression model. This model jointly predicts Spatial Ability ($\theta$) and Residualized Processing Speed, accounting for the correlation between the two traits.

```{r}
#| label: bivariate-prep
#| echo: false
#| message: false

# Join existing scores with new residuals
model_data_corrected <- ability_runs %>%
  inner_join(rt_residuals_person %>% 
               select(user_id, run_id, rt_resid = rt_residual_mean),
             by = c("user_id", "run_id")) %>%
  filter(!is.na(ability), !is.na(rt_resid), !is.na(age)) %>%
  mutate(
    # Recode sites for cleaner output
    site_model = dplyr::recode(site, !!!site_labels)
  )
```

```{r}
#| label: fig-development
#| fig-cap: "Developmental Trajectories. Panel A shows the increase in Spatial Ability (Theta) with age. Panel B shows the decrease in Residualized Reaction Time (log seconds) with age, indicating faster processing speed."
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(patchwork)

# 1. Define APA Theme
theme_apa <- theme_classic(base_size = 12) +
  theme(
    text = element_text(family = "serif", color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(face = "bold", size = 12),
    plot.tag = element_text(face = "bold", size = 16), 
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

# 2. Panel A: Spatial Ability by Age
plot_a <- ggplot(model_data_corrected, aes(x = age, y = ability)) +
  geom_jitter(alpha = 0.25, size = 1.2, width = 0.2, color = "grey40") +
  geom_smooth(method = "lm", color = "black", linewidth = 1, se = TRUE, fill = "grey80") +
  labs(
    tag = "A",
    x = "Age (Years)",
    y = expression(paste("Spatial Ability (", theta, ")")) 
  ) +
  theme_apa

# 3. Panel B: Processing Speed by Age
plot_b <- ggplot(model_data_corrected, aes(x = age, y = rt_resid)) +
  geom_jitter(alpha = 0.25, size = 1.2, width = 0.2, color = "grey40") +
  geom_smooth(method = "lm", color = "black", linewidth = 1, se = TRUE, fill = "grey80") +
  labs(
    tag = "B",
    x = "Age (Years)",
    y = "Residualized RT (log seconds)\n(Negative = Faster)"
  ) 
# 4. Combine and Print
plot_a + plot_b
```


\clearpage

```{r}
#| label: run-brms
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| results: hide

# NOTE: This chunk uses cache=TRUE because brms takes time to run.

m_joint_corrected <- brm(
  mvbind(ability, rt_resid) ~ site + age,
  data = model_data_corrected,
  chains = 4, 
  cores = 4, 
  iter = 2000, 
  seed = 123
)
```


```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(dplyr)
library(tibble)

# 1. Extract results
fixef_results <- fixef(m_joint_corrected) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter")

# 2. Define Helper Functions
get_est <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Estimate)
  if(length(val) == 0) return(NA) else return(val)
}
get_low <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Q2.5)
  if(length(val) == 0) return(NA) else return(val)
}
get_up <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Q97.5)
  if(length(val) == 0) return(NA) else return(val)
}

# 3. Build Dataframe
site_results <- tibble(
  Outcome = c("Spatial Ability", "Spatial Ability", 
              "Response Speed", "Response Speed"),
  Comparison = c("Colombia vs. Germany", "Canada vs. Germany", 
                 "Colombia vs. Germany", "Canada vs. Germany"),
  Param_Name = c("ability_sitepilot_uniandes_co", "ability_sitepilot_western_ca",
                 "rtresid_sitepilot_uniandes_co", "rtresid_sitepilot_western_ca")
) %>%
  rowwise() %>%
  mutate(
    Beta = get_est(Param_Name),
    CI_Lower = get_low(Param_Name),
    CI_Upper = get_up(Param_Name),
    Result = case_when(
      is.na(Beta) ~ "Error",
      TRUE ~ "Valid"
    )
  ) %>%
  ungroup() %>%
  filter(Result == "Valid") %>%
  mutate(
    Estimate = sprintf("%.2f", Beta),
    `95% CI` = paste0("[", sprintf("%.2f", CI_Lower), ", ", sprintf("%.2f", CI_Upper), "]")
  ) %>%
  select(Outcome, Comparison, Estimate, `95% CI`)

# 4. Simple Table (No Caption = No Error)
kable(site_results, booktabs = TRUE)
```

<!--

```{r}
#| label: fig-site-forest
#| fig-cap: "Site-Specific Processing Styles. Forest plot showing posterior estimates for Spatial Ability (Panel A) and Processing Speed (Panel B) relative to the German baseline (dashed line at 0). Error bars represent 95% confidence intervals."
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(dplyr)
library(tibble)

# 1. Re-construct the data safely (Self-contained)
# This ensures we have the raw numbers (Beta, CI) needed for plotting
fixef_results <- fixef(m_joint_corrected) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter")

get_est <- function(p) { val <- fixef_results %>% filter(Parameter == p) %>% pull(Estimate); if(length(val)==0) NA else val }
get_low <- function(p) { val <- fixef_results %>% filter(Parameter == p) %>% pull(Q2.5); if(length(val)==0) NA else val }
get_up  <- function(p) { val <- fixef_results %>% filter(Parameter == p) %>% pull(Q97.5); if(length(val)==0) NA else val }

plot_data <- tibble(
  Outcome = c("Spatial Ability", "Spatial Ability", 
              "Response Speed", "Response Speed"),
  Comparison = c("Colombia vs. Germany", "Canada vs. Germany", 
                 "Colombia vs. Germany", "Canada vs. Germany"),
  Param_Name = c("ability_sitepilot_uniandes_co", "ability_sitepilot_western_ca",
                 "rtresid_sitepilot_uniandes_co", "rtresid_sitepilot_western_ca")
) %>%
  rowwise() %>%
  mutate(
    Beta = get_est(Param_Name),
    CI_Lower = get_low(Param_Name),
    CI_Upper = get_up(Param_Name),
    Site = ifelse(grepl("Colombia", Comparison), "Colombia", "Canada")
  ) %>%
  ungroup() %>%
  filter(!is.na(Beta))

# 2. Define Theme
theme_apa_forest <- theme_classic(base_size = 11) +
  theme(
    text = element_text(family = "serif"),
    axis.text.y = element_text(face = "bold", size = 10),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )

# 3. Panel A: Ability Differences
p1 <- ggplot(subset(plot_data, Outcome == "Spatial Ability"), 
             aes(x = Beta, y = Site)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "#27ae60") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2, color = "#27ae60") +
  labs(
    title = "A. Ability Differences",
    subtitle = "(Reference: Germany = 0)",
    x = "Difference in SD units"
  ) +
  theme_apa_forest

# 4. Panel B: Speed Differences
p2 <- ggplot(subset(plot_data, Outcome == "Response Speed"), 
             aes(x = Beta, y = Site)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "#2980b9") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2, color = "#2980b9") +
  labs(
    title = "B. Speed Differences",
    subtitle = "(Negative = Faster)",
    x = "Difference in log(seconds)"
  ) +
  theme_apa_forest

# 5. Combine and Print
p1 + p2
```

\clearpage

## Differential Effectiveness of Processing Speed

We further tested whether the accuracy benefits of slowing down varied across cultures by modeling the interaction between site and processing speed. In contrast to the site-specific hypothesis, results indicated a universal benefit to deliberative processing.

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| results: hide
# 1. Fit the Interaction Model
# NOTE: Using cache=TRUE to avoid re-running this slow model every render
m_strategy <- brm(
  ability ~ age + site_model * rt_resid, 
  data = model_data_corrected,
  chains = 4, 
  cores = 4, 
  iter = 2000, 
  seed = 123
)
```

```{r}
#| label: fig-strategy-interaction
#| fig-cap: "Universal Speed-Accuracy Trade-off. The relationship between residualized reaction time and spatial ability is consistent across sites. In all three populations, slower processing (higher residualized RT) predicts higher spatial ability scores, as indicated by the parallel positive slopes."
#| echo: false
#| message: false
#| warning: false


# 2. Extract trend lines for plotting
effects_data <- conditional_effects(m_strategy, effects = "rt_resid:site_model")[[1]]

# 3. Define APA Theme
theme_apa <- theme_classic(base_size = 12) +
  theme(
    text = element_text(family = "serif", color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Prepare data for plotting
# Note: Ensure site_labels vector exists from previous chunks
effects_data2 <- effects_data %>%
  mutate(site_label = dplyr::recode(as.character(site_model), 
                                    "pilot_uniandes_co" = "Colombia",
                                    "pilot_mpieva_de" = "Germany",
                                    "pilot_western_ca" = "Canada"))

# Position for line-end labels (rightmost x for each site)
label_positions <- effects_data2 %>%
  group_by(site_label) %>%
  slice_max(rt_resid, with_ties = FALSE) %>%
  ungroup()

# 4. Create the Plot
ggplot(effects_data2, aes(x = rt_resid, y = estimate__, linetype = site_label)) +
  # Ribbons (Confidence Intervals)
  geom_ribbon(aes(ymin = lower__, ymax = upper__, fill = site_label), 
              alpha = 0.15, color = NA) +
  # Trend Lines
  geom_line(linewidth = 1.2, color = "black") +
  # Direct labels near lines
  geom_text(
    data = label_positions,
    aes(x = rt_resid + 0.05, y = estimate__, label = site_label),
    inherit.aes = FALSE, hjust = 0, size = 4.5, fontface = "bold"
  ) +
  # Manual Scales
  scale_linetype_manual(values = c("Germany" = "solid", "Colombia" = "longdash", "Canada" = "dotted")) +
  scale_fill_manual(values = c("Germany" = "black", "Colombia" = "black", "Canada" = "black")) +
  # Labels and Layout
  labs(
    x = "Residualized RT (log seconds)\n(Negative = Fast | Positive = Slow)",
    y = "Predicted Spatial Ability (Theta)"
  ) +
  coord_cartesian(clip = "off") + 
  theme_apa +
  theme(legend.position = "none", plot.margin = margin(r = 80))

```
\clearpage

-->

