---
title: "Mental rotation in children aged 5–12: a cross-site comparison across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "Fionnuala O'Reilly"
    orcid: 0000-0002-4355-9088
    affiliations:
      - ref: stanford
  - name: "Xingyao Xiao"
    orcid: 0000-0001-8430-0438
    affiliations:
      - ref: stanford
  - name: "Rebecca Zhu"
    orcid: 0000-0003-0957-5962
    affiliations:
      - ref: stanford
  - name: "Mika Braginsky"
    orcid: 0000-0001-9039-3220
    affiliations:
      - ref: stanford
  - name: "George Kachergis"
    orcid: 0000-0003-4153-4167
    affiliations:
      - ref: stanford
  - name: "Daniel Ansari"
    orcid: 0000-0002-7625-618X
    affiliations:
      - ref: western
  - name: "Manuel Bohn"
    orcid: 0000-0001-6006-1348
    affiliations:
      - ref: mpieva 
      - ref: leuphana
  - name: "Daniel Haun"
    orcid: 0000-0002-3262-645X
    affiliations:
      - ref: mpieva
  - name: "Maria Camila Perfetti"
    affiliations:
      - ref: uniandes
  - name: "Julián P. Mariño"
    affiliations:
      - ref: uniandes
  - name: "Nilàm Ram"
    orcid: 0000-0003-1671-5257
    affiliations:
      - ref: stanford
  - name: "Benjamin W. Domingue"
    orcid: 0000-0002-3894-9049
    affiliations:
      - ref: stanford
  - name: "Michael C. Frank"
    orcid: 0000-0002-7551-4378
    affiliations:
      - ref: stanford
    corresponding: true
    email: mcfrank@stanford.edu

   # roles:
   #    - conceptualization
   #    - data curation
   #    - formal analysis
   #    - funding acquisition
   #    - investigation
   #    - methodology
   #    - project administration
   #    - resources
   #    - software
   #    - supervision
   #    - validation
   #    - visualization
   #    - writing – original draft
   #    - writing – review & editing

affiliations:
  - id: stanford
    name: "Stanford University"
  - id: western
    name: "Western University"
  - id: mpieva
    name: "Max Planck Institute for Evolutionary Anthropology"
  - id: leuphana
    name: "Leuphana University"
  - id: uniandes
    name: "Universidad de los Andes"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    conflict-of-interest: "The author has no conflict of interest to declare."

abstract: "Mental rotation — the ability to mentally transform an object's orientation in space — is one of the most studied spatial skills in cognitive and developmental psychology, yet the cross-cultural generalizability of mental rotation tasks in children remains largely untested. The present study examines data from 1,141 children aged 5–12 across threeinternational study sites (Canada, Germany, and Colombia) participating in the Learning Variability Network Exchange (LEVANTE), building a layered case for the task's cross-cultural validity. First, multi-group item response theory supported scalar measurement invariance across sites, indicating that item difficulty is ordered consistently and on a comparable scale regardless of cultural context. Second, the canonical angular disparity effect first reported by Shepard and Metzler (1971) replicated across all three sites for 3D stimuli, with a significant but attenuated effect for 2D stimuli. Third, an exploratory clustering analysis identified three profiles of task engagement — ranging from disengagement to effective feature-matching and analog rotation — providing preliminary evidence that the task elicits theoretically meaningful variation in how children engage with its spatial demands, though the boundary between the latter two profiles was not robustly recoverable within sites. Taken together, these findings establish that the LEVANTE mental rotation task captures a consistent and spatially-driven construct across culturally diverse samples of children; a necessary precondition for using cross-cultural data to understand variation in spatial reasoning."

keywords: ["Cognitive Development", "Mental Rotation", "Spatial Reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    extra_dependencies: ["booktabs", "longtable"]
    # fig-pos: H        
    # tbl-pos: H  
    
knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: false
  cache: true
---

# Introduction
From fitting puzzle pieces together at age three to navigating complex diagrams in school, spatial reasoning underpins a broad array of cognitive achievements — and children differ substantially in how these skills develop. Such variability is detectable early, across a wide range of spatial tasks in primary school-aged children [@hodgkiss2021developmental; @lauer2019development], and its consequences are far-reaching: spatial ability predicts outcomes spanning creative domains such as music [@padulo2020effects; @hetland2000learning] and the visual arts [@kell2013creativity; @winner2013relationship], mathematical achievement [@gunderson2012relation], science learning [@gilligan2020first], and long-term educational and occupational outcomes in STEM fields [@wai2009spatial]. Understanding what drives this variability is therefore a question of both theoretical and practical importance. The present study takes a step toward that goal by examining children's performance on a mental rotation task across three international study sites, asking both what is consistent across cultures and what varies.

Spatial reasoning encompasses abilities for representing, transforming, and reasoning about visuospatial information [@lohman1996spatial]. Rather than a single unified capacity, it is better understood as a family of related but distinguishable skills, including mental rotation, spatial visualization, perspective-taking, and navigation, that differ in the type of spatial information involved and the cognitive operations required [@uttal2013malleability; @newcombe2018three]. These skills can be organized along two dimensions: whether the spatial relations are intrinsic to an object or extrinsic between objects, and whether the task requires dynamic mental transformation or only static representation [@uttal2013malleability]. Mental rotation — the focus of the present study — falls in the intrinsic-dynamic quadrant of this framework, involving the mental transformation of a single object's orientation in space.

Classic mental rotation tasks, first developed by @shepard1971mental, present participants with pairs of objects at varying angular disparities and ask them to judge whether the objects are identical or mirror-reversed, with response time and accuracy as the key dependent variables. It is one of the most studied spatial skills in cognitive and developmental psychology. In 'pure' or analog rotation, the solver constructs a holistic mental image of the object and continuously rotates it until it aligns with the target — because the mental image must traverse each intermediate angle, larger rotations take proportionally longer [@shepard1971mental; @cooper1973chronometric]. The result is one of the most robust and replicable findings in the field: response times increase approximately linearly with angular disparity (i.e., how far an object has been rotated from its original orientation).

However, this aggregate pattern may obscure meaningful individual variation in the processes underlying performance. Performance on the task does not necessarily reflect mental rotation per se — participants may instead rely on alternative spatial strategies that produce similar accuracy and response time patterns. Research with adults suggests that solvers draw on a range of strategies, spanning a holistic-analytic continuum [@gluck2003spatial]. At the holistic end, solvers use 'pure' or analog rotation as described above. Feature-based strategies, by contrast, operate by identifying discrete local features — a particular arm, corner, or protrusion — and rotating or comparing these components piecemeal rather than rotating the whole object [@just1985cognitive; @khooshabeh2013individual]. A related approach, 'flipping', involves mentally reflecting one object along an axis to check for mirror symmetry, again engaging a different type of spatial transformation. Because these approaches decompose the problem into parts, they can produce correct responses even when the angular disparity effect is attenuated or absent — making them distinguishable from analog rotation at the level of RT patterns. Importantly, this distinction concerns process, not proficiency: feature-based strategies draw on genuine spatial ability, they simply engage it differently [@hegarty2018ability]. More recent work suggests that solvers do not consistently adopt a single strategy but instead shift flexibly between holistic and analytical/piecemeal approaches depending on the demands of a particular item [@nazareth2019strategy].

Importantly, far less is known about which strategies children employ when solving mental rotation problems, and whether strategy use varies with age, ability, or cultural context. Recent work by @mckee2025malleability finds that children rely on non-rotation strategies — flipping and feature matching — more often than analog rotation, with strategy choice shaped by how the task is framed. This question motivates the exploratory component of the present study, which we describe below. These questions take on added significance given how narrow the populations studied to date have been.

Mental rotation ability develops across a protracted period beginning in infancy. Precursors of successful rotation are detectable in the first year of life [@frick2014development; @moore2008mental], and performance on explicit rotation tasks improves substantially between ages 3-12 years [@frick2013development; @marmor1975development; @kail1980developmental], with continued refinement into adolescence. This extended developmental trajectory makes the primary school years, ages 5--12 years, a particularly informative window for studying how mental rotation skills are acquired and how they vary across children.

One prominent source of variation is child sex.^[In the current paper, we refer to sex differences (in the sense of the biological variable), but some reported differences could be related to gender (e.g., driven by differences in socialization and experience). We are agnostic about this question here.] Sex differences in mental rotation are well documented in adults [@voyer1995magnitude; @zell2015evaluating], but in children the picture is considerably less clear. @lauer2019development synthesized data from over 30,000 children aged 3--17, finding no reliable sex difference in the preschool years, a small male advantage emerging around age 6 (g ≈ 0.20), and a moderate effect only by age 14, with task and stimulus characteristics explaining substantial variability independently of age. This sensitivity to task format has been documented most systematically in German primary school children. @neuburger2011gender found the gap absent at age 8 but small and significant at age 10, concentrated in cube figure conditions where girls showed no improvement across grades; @jansen2013mental found cube figures too demanding for either grade but reliable male advantages in speed with more accessible stimuli. Going further, @ruthsatz2017gender and @ruthsatz2019girls showed that the direction of the difference depended on stimulus-gender congruence, reversing entirely under female-congruent conditions. Studies from other countries present a similarly inconsistent picture. A male advantage has been reported in Italian primary school children from as early as age 8 [@moe2018mental], and in Dutch children aged 7–9 using a simplified child-friendly task, though this difference disappeared in older children in the same sample [@van2019sex]. In French children, no sex difference was found at age 7--8 but one emerged by age 11--12 [@hoyek2011use]. In contrast, @hodgkiss2021developmental found no significant gender difference across the full primary school age range in a UK sample. Whether these discrepancies reflect genuine cultural variation, differences in task format, or both remains unclear.

The vast majority of developmental research on mental rotation has been conducted with children from WEIRD (Western, Educated, Industrialized, Rich, Democratic) populations [@henrich2010weirdest], raising questions about whether findings about its developmental trajectory, psychometric properties, and underlying cognitive processes generalize beyond the populations in which they were established.

The limited cross-cultural evidence that does exist spans different populations, developmental periods, and spatial tasks, making it difficult to draw firm conclusions about how cultural context shapes performance. Studies with non-WEIRD children — including Tsimané forager-horticulturalists in Bolivia [@davis2019spatial] and seminomadic pastoralist-forager children in northern Namibia [@davis2021cultural] — have focused primarily on broader spatial skills rather than mental rotation specifically. In one of the few studies to examine mental rotation cross-culturally in children, researchers compared children in the Ivory Coast and Switzerland and found that the overall shape of the RT–angle curve was similar across both groups, suggesting the basic mental rotation signature is present in both cultural contexts [@nunez1998mental]. However, age predicted performance in the Swiss group but not the Ivorian group, pointing to the possibility that the developmental trajectory of mental rotation differs across cultures even when its basic signature holds. Yet the study has important limitations — it was small (n = 151) and relied on a computerized task likely more familiar to Swiss children. Notably, the authors themselves attributed much of the observed RT difference to cultural differences in familiarity with technology and the value placed on speed, rather than to differences in spatial cognition per se. Taken together, the cross-cultural validity of the mental rotation paradigm with children remains largely unestablished.

Taken together, this evidence highlights three distinct but related gaps. First, we have limited data on whether mental rotation tasks developed and validated in resource-rich contexts measure the same underlying construct when administered to children from different cultural and socioeconomic backgrounds. Second, even where cross-cultural comparisons have been made, they have rarely examined whether core cognitive signatures replicate across diverse sites. The linear relationship between reaction time and angular disparity has been demonstrated repeatedly in WEIRD adult and child samples [@kaltner2016developmental; @jansen2015role], but its replication in children across varied international sites remains largely untested. Third, whether and when a gender difference in mental rotation emerges in primary school children remains unclear, with findings varying substantially across studies and tasks — making it a natural question to examine in a cross-cultural dataset. Beyond these three gaps, relatively little is known about the cognitive strategies children employ when solving mental rotation problems, and whether these vary across cultural contexts — a question we explore tentatively here. We address these gaps with four sets of analyses, corresponding to four research questions, each building toward a broader understanding of how and why children's spatial reasoning varies across cultures.

## The Present Study
The present study uses the pilot dataset from the Learning Variability Network Exchange (LEVANTE), an international research collaboration designed to capture how children's learning and development vary across ages and cultural contexts. LEVANTE's central aim is to build a large, openly shared dataset of children's cognitive development across diverse settings, with a particular focus on individual variability — what differs across children, and why. Data were collected from three geographically and culturally diverse pilot sites: the Universidad de los Andes in Bogotá, Colombia (uniandes-co); the Max Planck Institute for Evolutionary Anthropology in Leipzig, Germany (mpieva-de); and the University of Western Ontario in London, Canada (western-ca). Here we use data from a single task in the LEVANTE battery: a mental rotation task developed using item response theory, which allows directly comparable ability estimates to be derived across children who completed different subsets of items [@Kachergis2025LEVANTEcoretasks]. LEVANTE is not designed to support site-level performance comparisons — sites vary substantially in sampling, recruitment, and test administration in ways that heavily confound absolute comparisons. Because of this confounding, differences in overall scores across sites should not be interpreted as reflecting population differences. This interpretive caveat is an important limitation of LEVANTE data.

Our first research question asks whether the LEVANTE mental rotation task achieves measurement invariance across three international study sites — that is, whether it captures the same underlying construct regardless of where it is administered. Our second asks whether performance is sensitive to the spatial properties of the stimuli in the way theory predicts [@shepard1971mental], using the angular disparity effect as the key benchmark: as the angular difference between two objects increases, so too should the time required to judge whether they match. If children across three culturally diverse sites show this same systematic pattern, this constitutes good evidence that the task is engaging the spatial demands it was designed to capture. Our third research question asks whether a gender difference in mental rotation is detectable in this sample, and whether it interacts with age. Alongside these three primary research questions, we tentatively explore a fourth: whether children draw on different cognitive strategies when solving mental rotation problems, and whether their strategies vary across contexts. These four questions move from psychometric structure (*Cross-Site Invariance*), to process-level validity (*Angular Sensitivity*), to individual variation (*Gender and Age*), to profiles of engagement (*Strategy Variation*) — building a layered case for the task's cross-cultural validity.

```{r}
#| include: false

# packages
library(tidyverse)
library(glue)
library(here)
library(tidytext)

library(ggeffects)
library(patchwork)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(umap)
library(gt)
library(kableExtra)

library(brms)
library(lme4)
library(lmerTest) # p-values via satterthwaite
library(Rtsne)
```

```{r}
#| label: setup
#| cache: false

source(here::here("helper", "plot_settings.R"))
source(here::here("helper", "explore_helper.R"))
```

```{r}
#| label: site-info

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# site_pal is keyed by pilot_* names
# site_labels is keyed by readable names, values are pilot_* names
# index site_pal by the pilot_* values, then assign readable names
site_pal_labelled <- setNames(
  site_pal[unname(site_labels)],
  names(site_labels)
)
```

```{r}
#| label: load-data

# Run data
run_data <- readr::read_rds(here::here("data/runs_mrot.rds"))

# RT data
task_data <- readr::read_rds(here::here("data/task_data_nested.rds"))

# IRT data
mrot_scores <- readr::read_rds(here::here("models/mrot_scores.rds")) |>
  filter(model_set == "multigroup_site") # also contains "multigroup_dataset"

# coefficients
mrot_coefs <- readr::read_rds(here::here("models/mrot_coef.rds"))

# survey data for gender analysis
survey_gender <- readr::read_rds(
  here::here("survey", "survey_gender_conditions.rds")
)
```

```{r}
#| label: filter-runs

# apply filters to runs for use in all subsequent analyses
min_trials <- 10
min_gap_months <- 3
run_data_filtered <- run_data |>
  filter(num_attempted >= min_trials) |>
  filter(!is.na(age), age >= 5, age <= 14) |>
  arrange(user_id, age) |>
  group_by(user_id) |>
  mutate(run_number = row_number(),
         age_gap = age - lag(age)) |>
  ungroup() |>
  filter_out(age_gap < min_gap_months / 12) |>
  select(site, user_id, run_id, age, adaptive)

# create age groups based on roughly equal sample sizes
n_groups <- 3
ages <- run_data_filtered |> pull(age)
age_breaks <- quantile(ages, probs = seq(0, 1, length.out = n_groups + 1))

# add age groups and site labels
runs_filtered <- run_data_filtered |>
  mutate(age_group = cut(age, breaks = age_breaks, include.lowest = TRUE),
         site_label = site |>
           recode_values(from = site_labels, to = names(site_labels)))
```

# Methods

## Participants

Data for the present study were drawn from three pilot sites participating in LEVANTE: the Universidad de los Andes in Bogotá, Colombia (uniandes-co); the Max Planck Institute for Evolutionary Anthropology in Leipzig, Germany (mpieva-de); and the University of Western Ontario in London, Canada (western-ca). Site codes rather than country names are used throughout to reflect that each site represents a single institution/region within a large, diverse country, and should not be taken as representative of national populations. Collectively, these sites supported task localization and technical implementation, iterative piloting, and deployment in varied testing contexts. The resulting pilot datasets enabled preliminary evaluation of construct validity, measurement invariance, test–retest reliability, and calibration for computer-adaptive administration, as well as descriptive assessment of age-related differences across the targeted constructs.

### uniandes-co

The Universidad de los Andes site (uniandes-co) recruited 1,222 [note to udate] children aged 5–12 years across six schools in Colombia: four in Bogotá — three affiliated with family compensation funds and one managed by a charter school organisation — and two public schools in the rural departments of Caquetá and Boyacá. Schools were selected based on existing university partnerships and primarily serve children from low- and lower-middle-income families. Data were collected in groups using tablets provided by the research team, with group size altered for the age of the children (smaller groups for younger children). Due to time constraints inherent to school-based administration, children typically completed a subset of tasks.

### mpieva-de

The Max Planck Institute for Evolutionary Anthropology site (mpieva-de) collected data from 292 [note to udate] children aged 5–12 years in Leipzig, Germany, a mid-sized urban centre of approximately 620,000 residents. Leipzig is characterized by relatively high educational attainment and cultural diversity, with approximately 17% of residents having a migration background. Families were recruited from an existing database of Leipzig residents via telephone and completed tasks remotely at home using their own devices. 

### western-ca

The Western University site (western-ca) recruited 171 [note to udate] children aged 5–12 years primarily from London, Ontario, with a smaller number from the Greater Toronto Area. London is a predominantly urban region; census data indicate a median household income of approximately $80,000 CAD and mean education levels in the post secondary range. Recruitment relied on online outreach through social media and community groups, supplemented by local posters and school newsletters. Children were assessed individually by a researcher in a dedicated campus testing space. 

```{r}
#| label: sample-counts

mrot_df <- mrot_scores |>
  inner_join(runs_filtered)

n_excluded <- mrot_scores |>
  anti_join(mrot_df) |>
  summarise(n_users = n_distinct(user_id),
            n_runs = n_distinct(run_id))

n_final <- mrot_df |>
  summarise(n_users = n_distinct(user_id),
            n_runs = n_distinct(run_id))

n_repeat_users <- runs_filtered %>%
  group_by(user_id) %>%
  summarise(n_assessments = n_distinct(run_id)) %>%
  filter(n_assessments > 1) %>%
  nrow()
```

Site-level sampling characteristics are described in @tbl-participants. Due to de-identification protocols, individual-level demographic data — including socioeconomic indicators — are not available for the current sample. `r n_excluded$n_runs` assessments (from `r n_excluded$n_users` participants) were excluded due to missing age, being outside the target age range (5–12 years), having too few trials (<`r min_trials`), or having too short a gap between administrations (<`r min_gap_months` months), leaving a final sample of n = `r n_final$n_runs` assessments (from `r n_final$n_users` participants). A total of `r n_repeat_users` children contributed two assessments each; each assessment was treated as a separate observation, with participant included as a random effect in all mixed-effects models to account for the non-independence of repeated assessments. Data and analysis code for this paper are available at https://github.com/levante-framework/mental-rotation. For more information about LEVANTE public data releases, see https://researcher.levante-network.org/data.

```{r}
#| label: trial-level-rts

mrot_rt_trial <- task_data |>
  select(site, data) |>
  unnest(data) |>
  inner_join(runs_filtered) |>
  select(site, site_label, user_id, run_id, adaptive, age, age_group, trial_id,
         trial_number, item_uid, item_group, item, item_original, correct, rt_numeric) |>
  mutate(
    correct    = as.integer(correct),
    rt_s       = as.numeric(rt_numeric) / 1000,
    log_rt     = log(rt_s),
    angle      = as.integer(stringr::str_extract(item, "(?<=_)\\d+")),
    angle_full = as.integer(stringr::str_extract(item_original, "\\d{3}$")),
    stimulus   = str_extract(item, "^[a-z]+")
  ) |>
  filter(is.finite(rt_s), rt_s > 0)
```

```{r}
#| label: cat-exclusion-counts

cat_counts <- mrot_rt_trial |>
  group_by(adaptive) |>
  summarise(trials = n(), runs = n_distinct(run_id)) |>
  mutate(adaptive = if_else(adaptive, "n_cat", "n_non_cat")) |>
  pivot_longer(cols = c(trials, runs), names_to = "count") |>
  pivot_wider(names_from = adaptive, values_from = value) |>
  mutate(n_total = n_cat + n_non_cat,
         pct_cat = round(n_cat / n_total * 100, 1)) |>
  nest(counts = -count) |>
  deframe()
```

```{r}
#| label: tbl-participants
#| tbl-cap: "**Participant/assessment characteristics by site.** Age in years."
#| echo: false

trials_per_run <- mrot_rt_trial %>%
  # left_join(mrot_df %>% select(run_id, site_label), by = "run_id") %>%
  group_by(site_label, user_id, run_id) %>%
  summarise(n_trials = n(), .groups = "drop") %>%
  group_by(site_label) %>%
  summarise(
    Trials_mean = mean(n_trials, na.rm = TRUE),
    Trials_min  = min(n_trials, na.rm = TRUE),
    Trials_max  = max(n_trials, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(`Trials: M [range]` = sprintf("%.0f [%d–%d]",
                                        Trials_mean, Trials_min, Trials_max))

mrot_df |>
  # distinct(user_id, .keep_all = TRUE) |>
  group_by(site_label) |>
  summarise(
    N_run    = n(),
    N_user   = n_distinct(user_id),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd   = sd(age, na.rm = TRUE),
    Age_min  = min(age, na.rm = TRUE),
    Age_max  = max(age, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  mutate(
    `Age: M (SD) [range]` = sprintf("%.1f (%.1f) [%.1f–%.1f]",
                                     Age_mean, Age_sd, Age_min, Age_max),
    `Testing format` = case_when(
      site_label == "uniandes-co" ~ "In school; group; tablet (researcher-provided)",
      site_label == "mpieva-de"   ~ "Remote (home-based); individual, family device",
      site_label == "western-ca"  ~ "On-campus; individual, table (researcher-provided)"
    )
  ) |>
  left_join(trials_per_run %>% 
              select(site_label, `Trials: M [range]`), 
            by = "site_label") |>
  select(site_label, N_run, N_user, `Age: M (SD) [range]`, `Trials: M [range]`, `Testing format`) |>
  rename(Site = site_label,
         `N assessments` = N_run,
         `N participants` = N_user) |>
  knitr::kable(align = c("l", "c", "c", "c", "c", "l"), booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                            full_width = TRUE) |>
  kableExtra::column_spec(1, width = "2cm") |>
  kableExtra::column_spec(2, width = "1.8cm") |>
  kableExtra::column_spec(3, width = "1.8cm") |>
  kableExtra::column_spec(4, width = "3cm") |>
  kableExtra::column_spec(5, width = "2cm") |>
  kableExtra::column_spec(6, width = "3.5cm")

```

## Measure

The LEVANTE mental rotation task [@frick2013development] is administered digitally on a tablet or computer as part of a battery of nine core tasks. It employs two stimulus categories of increasing difficulty: 2D animal figures (ducks and rabbits), and 3D block configurations from the @shepard1971mental set. On each trial, a target image is displayed alongside two alternatives: one shows the same object rotated to a different angle, and the other is a mirror-reversed version that cannot be matched to the target by rotation alone. Children select whichever alternative they judge to correspond to the target by tapping or clicking on their device. The angular disparity between target and correct alternative varies systematically across trials (40°, 80°, 120°, 160°), enabling analysis of how increasing rotation demands affect both accuracy and response latency.

```{r}
#| label: fig-mrot-example-items
#| fig-cap: "Example items for 2D and 3D rotation"
#| out-width: "100%"
#| fig-align: "center"

knitr::include_graphics("mrot_item_examples.pdf")

# library(pdftools)
# pdf_convert("mrot_item_examples.pdf", format = "png", dpi = 300,
#             filenames = "mrot_item_examples.png")
```
\clearpage

The nine core tasks in LEVANTE, including mental rotation, were initially designed in English and then subsequently adapted for Spanish and German following a set of internationalization procedures in collaboration with researchers at the uniandes-co and mpieva-de sites, respectively - see @Kachergis2025LEVANTEcoretasks for more detail.

The LEVANTE team has implemented the mental rotation task in a computer-adaptive testing (CAT) format. In the adaptive version, participant ability is summarized by a latent proficiency estimate (θ) that is re-estimated after each response. The updated θ then guides selection of the next item so that it is appropriately challenging for each individual child. This trial-by-trial targeting typically makes the task more engaging and increases measurement efficiency, enabling reliable ability estimation with fewer administered items. In the present sample, `r cat_counts$runs$pct_cat`% of administrations (`r cat_counts$runs$n_cat` of `r cat_counts$runs$n_total`) used the CAT format. In the fixed format, testing stopped after either six consecutive incorrect responses or a maximum of 10 minutes, whichever came first. Refer to @Kachergis2025LEVANTEcoretasks for more information on the CAT versions of all LEVANTE tasks.

## Analysis

### Item Screening

Prior to fitting psychometric models, we examined item-level performance within each site to identify poorly functioning items. Items showing average accuracy near or below chance, low item–total correlations, high missingness, or other anomalous patterns were excluded from further analysis. For full details of the item screening procedure, see @Kachergis2025LEVANTEcoretasks. 

### Item Response Theory Models

Item response theory (IRT) is a psychometric framework that models the relationship between a latent trait and performance on individual task items. Unlike classical test theory, which relies on total scores and treats all items as equivalent contributors to measurement, IRT estimates item-level parameters independently of the specific sample tested, providing more precise measurement across the ability range. All models assumed a single underlying factor, following standard practice in psychometric analyses of mental rotation tasks and consistent with the task's design as a measure of a single spatial ability.

We compared two different IRT models. The Rasch model (1-parameter logistic; 1PL) estimates only item difficulty, holding discrimination constant across items — that is, it assumes all items are equally good at differentiating children of higher versus lower ability. The two-parameter logistic model (2PL) relaxes this constraint, additionally estimating each item's discrimination parameter. A better-fitting 2PL therefore indicates that items vary meaningfully in how well they distinguish between ability levels.

### Cross-Site Invariance

To assess cross-site measurement invariance, we fitted both models with varying constraints across the three sites: configural invariance (all parameters free to vary), metric invariance (item discriminations constrained equal across sites), and scalar invariance (both discriminations and difficulties constrained equal). We compared models using BIC as the selection criterion [@putnick2016measurement]. Overlapping items across sites served as anchors to establish a common scale, with non-overlapping items freely estimated within the same multigroup model. We estimated all models in R using the mirt package [@chalmers2012mirt] via marginal maximum likelihood with 500 quadrature points. We specified Gaussian priors on item difficulty (M = 0, SD = 3) and discrimination (M = 1, SD = 0.3) in all models. We extracted ability scores (θ) using expected a posteriori (EAP) scoring.

### Angular Sensitivity

To investigate whether performance was sensitive to rotation angle in the theoretically predicted direction, we fitted linear mixed-effects models predicting log response time (RT) from rotation angle, with participant level intercepts and slopes for angle. Practice trials (a single 20° item) were excluded, leaving four rotation angles (40°, 80°, 120°, and 160°). Rotation angle was standardized (M = 0, SD = 1) prior to analysis. We fitted separate models for 2D and 3D stimuli. For each, we additionally tested site as a moderator of the angle slope by including a site × angle interaction. A combined model including both stimulus types was used to test whether the angle slope differed between 2D and 3D items. Only correct trials were included in RT analyses. CAT trials were excluded from all RT models because adaptive item selection produces non-random angle exposure: higher-ability children disproportionately receive high-angle items and lower-ability children receive low-angle items. This correlation between ability and angle makes it impossible to obtain unbiased estimates of the RT–angle relationship, regardless of whether angular sensitivity itself varies with ability. Of `r cat_counts$trials$n_total` trials from all assessments, `r cat_counts$trials$n_cat` (`r cat_counts$trials$pct_cat`%) were administered in CAT format and excluded, leaving `r cat_counts$trials$n_non_cat` trials for analysis. This proportion is lower than the `r cat_counts$runs$pct_cat`% of administrations using CAT, reflecting the fact that adaptive runs contain fewer trials by design.

### Strategy Variation

To characterize individual differences in how children engaged with the task's spatial demands, we first fitted a mixed-effects model predicting log RT from rotation angle, stimulus type, their interaction, and trial number, with by-participant random intercepts and slopes for angle and trial number, and a random intercept by item. Unlike the angular sensitivity analysis, CAT trials were retained here: the strategy analysis extracts relative person-level features rather than estimating the angle effect at the population level, and including all available trials yields more reliable individual-level estimates. From this model, we extracted five person-level features for each child. Baseline response speed reflects general processing efficiency independent of rotation demands. The absolute RT–angle slope captures the raw magnitude of angular sensitivity, while the RT–angle slope relative to baseline speed captures the degree to which rotation demands slow a child down proportional to their typical pace, enabling comparison across children with different baseline speeds. The trial-by-trial RT trend indexes within-session learning or fatigue. Finally, response instability — the standard deviation of model residuals — captures trial-to-trial variability unexplained by angle or trial position; children who disengage from the task's spatial demands or shift inconsistently between approaches are expected to show greater residual variability than those applying a consistent strategy. These features were identified through iterative exploration and retained on the basis that each captures a distinct and interpretable aspect of how children engage with the task's spatial demands. These five features were then submitted to k-means clustering. 

# Results

## Cross-Site Invariance
```{r}
#| label: irt-anova

task_anova <- read_rds(here("models/mrot_anova.rds"))
mrot_models <- task_anova |> filter(model_set == "multigroup_site")
```

```{r}
#| label: mrot-irt-values

mrot_2pl_scalar <- mrot_models %>%
  filter(task == "mrot", itemtype == "2pl", invariance == "scalar")

mrot_rasch_scalar <- mrot_models %>%
  filter(task == "mrot", itemtype == "rasch", invariance == "scalar")

delta_bic_rasch <- round(mrot_rasch_scalar$BIC - mrot_2pl_scalar$BIC, 1)
bic_2pl_scalar  <- round(mrot_2pl_scalar$BIC, 2)
aic_2pl_scalar  <- round(mrot_2pl_scalar$AIC, 2)
npar_2pl_scalar <- mrot_2pl_scalar$df
```

Our first research question asked whether the mental rotation task measures the same underlying construct across all three sites — that is, whether it achieves measurement invariance. The 2PL scalar invariance model provided the best fit (BIC = `r bic_2pl_scalar`, AIC = `r aic_2pl_scalar`, *k* = `r npar_2pl_scalar`), outperforming the best-fitting Rasch model by `r delta_bic_rasch` BIC points (see @tbl-mrot-irt). The superiority of the 2PL model indicates that mental rotation items vary meaningfully in their discrimination. Scalar invariance — constraining both item discriminations and difficulties across sites — was supported over more flexible models, indicating not only that items are similarly ordered in difficulty across sites, but that the absolute spacing between items on the latent scale is also equivalent. 

This means that children across sites are being measured on a comparable scale, and that differences in θ are not attributable to site-specific measurement artefacts. Practically, the relative difficulty of items — for example, the greater challenge posed by larger rotation angles and 3D figures — is consistent in magnitude across all three sites. Scalar invariance is a psychometric property of the instrument: it establishes that scores can be meaningfully compared across sites, but does not by itself speak to what cognitive process is driving performance. Evidence that children are engaging the spatial demands the task was designed to capture — specifically, the predicted increase in response time with angular disparity — is addressed in the following section.

```{r}
#| label: tbl-mrot-irt
#| tbl-cap: "**Model fit indices for IRT models fit to overlapping items.** Note: 2PL = 2-parameter logistic model; Rasch model constrains all item discriminations to be equal. Configural invariance allows all parameters to vary across groups; metric constrains loadings; scalar constrains loadings and intercepts. The selected model (bold) had the lowest BIC."

best_row <- which.min(mrot_models$BIC)

invar_table <- mrot_models %>%
  transmute(
    Model       = itemtype,
    Invariance  = invariance,
    logLik      = logLik,
    AIC         = AIC,
    BIC         = BIC,
    `Parameters` = df,
    model_order = model_order
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>%
  cols_hide(columns = model_order) %>%
  fmt_number(columns = c(logLik, AIC, BIC), decimals = 2) %>%
  fmt_number(columns = Parameters, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Model, Invariance, Parameters)) %>%
  cols_align(align = "right",  columns = c(logLik, AIC, BIC)) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold"),
      cell_fill(color = "#f0f7ff")
    ),
    locations = cells_body(rows = best_row)
  ) %>%
  tab_footnote(
    footnote = "Degrees of freedom not computed for the baseline configural model.",
    locations = cells_body(columns = Parameters, 
                           rows = Model == "rasch" & Invariance == "configural")
  ) %>%
  tab_footnote(
    footnote = "Number of parameters estimated relative to the previous model in the sequence.",
    locations = cells_column_labels(columns = Parameters)
  ) %>%
  tab_row_group(label = "2PL",   rows = model_order %in% 3:5) %>%
  tab_row_group(label = "Rasch", rows = model_order %in% 1:2) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3),
    row_group.font.weight = "bold"
  )
invar_table

# gtsave(invar_table, "table2_model_fit.png")

```

```{r}
#| label: fig-item-diff
#| fig-cap: "**Item Difficulty Parameters Across Sites (Configural Model).** Points represent individual items; close alignment along the diagonal indicates that item difficulty parameters are consistent across cultural contexts."
#| fig-width: 12
#| fig-height: 8
#| eval: false

# Load configural parameters
mrot_configural <- mrot_coefs |>
  filter(task == "mrot",
                itemtype == "2pl",
                invariance == "configural",
                model_set == "multigroup_site") |>
  mutate(group = recode(group,
                  "pilot_western_ca"  = "Canada",
                  "pilot_mpieva_de"   = "Germany",
                  "pilot_uniandes_co" = "Colombia"))

# pivot wide
mrot_wide_d <- mrot_configural |>
  select(item, group, d) |>
  pivot_wider(names_from = group, values_from = d) |>
  mutate(item_label = item |>
                  stringr::str_remove("mrot_2d_") |>
                  stringr::str_remove("mrot_3d_") |>
                  stringr::str_replace("_", " ") |>
                  stringr::str_replace("(\\d+)$", "\\1°"))

plot_scatter <- function(data, x_var, y_var, x_lab, y_lab) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]],
                   label = item_label)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = pal$grey) +
    geom_point(size = 2.5, na.rm = TRUE) +
    ggrepel::geom_text_repel(size = 2.8,
                    max.overlaps = Inf,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    na.rm = TRUE) +
    coord_equal(xlim = c(-2.5, 4), ylim = c(-2.5, 4)) +
    labs(x = x_lab, y = y_lab)
}

p1 <- plot_scatter(mrot_wide_d, "Canada", "Germany",   "Canada", "Germany")
p2 <- plot_scatter(mrot_wide_d, "Canada", "Colombia",  "Canada", "Colombia")
p3 <- plot_scatter(mrot_wide_d, "Germany", "Colombia", "Germany", "Colombia")

(p1 | p2 | p3) +
  plot_annotation(
    title = "Item Difficulty Parameters Across Sites (Configural Model)",
    theme = theme(plot.title = element_text(size = 12))
  )

# checks
coef_ns <- mrot_coefs |>
  filter(task == "mrot", itemtype == "2pl", invariance == "configural") |>
  count(model_set, subset, group)

# only 12 items appearing: multigroup IRT model was fitted only on the overlap items by design. The non-overlap items were either fitted in separate by-site models, or weren't included in the multigroup analysis at all.

```

@fig-dev-trends plots participants' mental rotation ability estimates (θ) against age. Curves show GAM-smoothed age trends within each site, used here as a visualization aid to highlight the general shape of the developmental trajectory without imposing a linear form; formal statistical modelling of these developmental trends is reported in @Kachergis2025LEVANTEcoretasks. Across all three sites, ability increases with age, indicating that the task captures developmental change across cultural contexts. The dispersion of points within age bands underscores substantial individual variability among same-aged children. 

```{r}
#| label: fig-dev-trends
#| fig-cap: "**Developmental trajectories of mental rotation ability across sites.** Points represent individual children's IRT-estimated ability (θ); curves are GAM-smoothed age trends with 95% confidence intervals."
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 4.5

ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1, shape = 16) +
  # geom_linerange(aes(ymin = metric_value - metric_se, ymax = metric_value + metric_se),
  #                alpha = 0.35, size = 1.1) +
  geom_line(aes(group = user_id), alpha = .3) + 
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  scale_colour_manual(values = site_pal_labelled) +
  labs(
    x      = "Age (years)",
    y      = "Ability (θ)",
    colour = "Site"
  ) +
  theme(legend.position = "bottom")

# ggsave("mrot_age.png", width = 7, height = 5, dpi = 300)
# ggsave("fig2_age_plot.png", plot = age_plot,
#        width = 7, height = 5, dpi = 300)
```

```{r}
#| label: rt-angle-lme

# filter to non-CAT, add age group, stimulus, angle
mrot_lme_data <- mrot_rt_trial %>%
  filter(!adaptive, correct == 1) %>%
  mutate(
    item_group = factor(item_group) |> fct_relevel("3d"), # 2D vs 3D
    site       = factor(site),
    angle_sc   = as.numeric(scale(angle)), 
  )
```

```{r}
#| label: rt-angle-models

# 3D stimuli: main test of angular disparity effect

mrot_3d <- mrot_lme_data %>%
  filter(item_group == "3d") %>%
  # filter(!is.na(site), !is.na(age_group), !is.na(angle_sc), !is.na(log_rt)) %>%
  droplevels()  # drops empty factor levels

# check size
# nrow(mrot_3d)

# fit all models on the same object
m_3d_main <- lmer(
  log_rt ~ angle_sc + (1 + angle_sc | user_id),
  data = mrot_3d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

m_3d_site <- lmer(
  log_rt ~ angle_sc * site + (1 + angle_sc | user_id),
  data = mrot_3d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

m_3d_age <- lmer(
  log_rt ~ angle_sc * age_group + (1 + angle_sc | user_id),
  data = mrot_3d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

m_3d_full <- lmer(
  log_rt ~ angle_sc * site * age_group + (1 | user_id),
  data = mrot_3d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

anova_3d <- anova(m_3d_main, m_3d_site, m_3d_age, m_3d_full)

mrot_2d <- mrot_lme_data %>%
  filter(item_group == "2d") %>%
  # filter(!is.na(site), !is.na(age_group), !is.na(angle_sc), !is.na(log_rt)) %>%
  droplevels()

# nrow(mrot_2d)

m_2d_main <- lmer(
  log_rt ~ angle_sc + (1 + angle_sc | user_id),
  data = mrot_2d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

m_2d_site <- lmer(
  log_rt ~ angle_sc * site + (1 + angle_sc | user_id),
  data = mrot_2d, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

```

```{r}
#| label: rt-angle-combined

# mrot_combined <- bind_rows(mrot_3d, mrot_2d) %>%
#   mutate(item_group = factor(item_group, levels = c("3d", "2d")))

m_combined_main <- lmer(
  log_rt ~ angle_sc + item_group + (1 + angle_sc | user_id),
  data = mrot_lme_data, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

m_combined_interaction <- lmer(
  log_rt ~ angle_sc * item_group + (1 + angle_sc | user_id),
  data = mrot_lme_data, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

# assign rather than just print
anova_combined <- anova(m_combined_main, m_combined_interaction)
```

```{r}
#| label: rt-angle-results

# site model numbers (for results text)
m_3d_site_coefs <- coef(summary(m_3d_site))
beta_site_co <- m_3d_site_coefs["angle_sc:sitepilot_uniandes_co", "Estimate"]
se_site_co <- m_3d_site_coefs["angle_sc:sitepilot_uniandes_co", "Std. Error"]
p_site_co <- m_3d_site_coefs["angle_sc:sitepilot_uniandes_co", "Pr(>|t|)"]
beta_site_ca <- m_3d_site_coefs["angle_sc:sitepilot_western_ca", "Estimate"]
se_site_ca <- m_3d_site_coefs["angle_sc:sitepilot_western_ca", "Std. Error"]
p_site_ca <- m_3d_site_coefs["angle_sc:sitepilot_western_ca", "Pr(>|t|)"]

anova_site_chisq <- anova(m_3d_main, m_3d_site)$Chisq[2]
anova_site_df <- anova(m_3d_main, m_3d_site)$Df[2]
anova_site_p <- anova(m_3d_main, m_3d_site)[["Pr(>Chisq)"]][2]

# age moderation - tested over and above site
anova_age_chisq <- anova(m_3d_site, m_3d_full)$Chisq[2]
anova_age_df    <- anova(m_3d_site, m_3d_full)$Df[2]
anova_age_p     <- anova(m_3d_site, m_3d_full)[["Pr(>Chisq)"]][2]

# age group contrasts from m_3d_age
m_3d_age_coefs <- coef(summary(m_3d_age))
beta_age_mid   <- m_3d_age_coefs["angle_sc:age_group(8.3,10.6]", "Estimate"]
se_age_mid     <- m_3d_age_coefs["angle_sc:age_group(8.3,10.6]", "Std. Error"]
p_age_mid      <- m_3d_age_coefs["angle_sc:age_group(8.3,10.6]", "Pr(>|t|)"]
beta_age_old   <- m_3d_age_coefs["angle_sc:age_group(10.6,13.7]", "Estimate"]
se_age_old     <- m_3d_age_coefs["angle_sc:age_group(10.6,13.7]", "Std. Error"]
p_age_old      <- m_3d_age_coefs["angle_sc:age_group(10.6,13.7]", "Pr(>|t|)"]

# 3D main model numbers
fe_3d <- fixef(m_3d_main)
beta_3d_main <- fe_3d["angle_sc"]
se_3d_main <- sqrt(diag(vcov(m_3d_main)))["angle_sc"]
t_3d_main <- coef(summary(m_3d_main))["angle_sc", "t value"]
df_3d_main <- coef(summary(m_3d_main))["angle_sc", "df"]
p_3d_main <- coef(summary(m_3d_main))["angle_sc", "Pr(>|t|)"]

# 2D main model numbers
fe_2d <- fixef(m_2d_main)
beta_2d_main <- fe_2d["angle_sc"]
se_2d_main <- sqrt(diag(vcov(m_2d_main)))["angle_sc"]
t_2d_main <- coef(summary(m_2d_main))["angle_sc", "t value"]
df_2d_main <- coef(summary(m_2d_main))["angle_sc", "df"]
p_2d_main <- coef(summary(m_2d_main))["angle_sc", "Pr(>|t|)"]

# 2D site model numbers
m_2d_site_coefs <- coef(summary(m_2d_site))
beta_2d_site_co <- m_2d_site_coefs["angle_sc:sitepilot_uniandes_co", "Estimate"]
se_2d_site_co   <- m_2d_site_coefs["angle_sc:sitepilot_uniandes_co", "Std. Error"]
p_2d_site_co    <- m_2d_site_coefs["angle_sc:sitepilot_uniandes_co", "Pr(>|t|)"]
beta_2d_site_ca <- m_2d_site_coefs["angle_sc:sitepilot_western_ca", "Estimate"]
se_2d_site_ca   <- m_2d_site_coefs["angle_sc:sitepilot_western_ca", "Std. Error"]
p_2d_site_ca    <- m_2d_site_coefs["angle_sc:sitepilot_western_ca", "Pr(>|t|)"]

anova_2d_site_chisq <- anova(m_2d_main, m_2d_site)$Chisq[2]
anova_2d_site_df    <- anova(m_2d_main, m_2d_site)$Df[2]
anova_2d_site_p     <- anova(m_2d_main, m_2d_site)[["Pr(>Chisq)"]][2]

# children who completed 3D trials
n_children_3d <- mrot_3d %>% distinct(user_id) %>% nrow()
age_3d_mean <- mrot_3d %>% distinct(user_id, .keep_all = TRUE) %>% pull(age) %>% mean(na.rm = TRUE) %>% round(1)
age_3d_sd   <- mrot_3d %>% distinct(user_id, .keep_all = TRUE) %>% pull(age) %>% sd(na.rm = TRUE) %>% round(1)
age_3d_min  <- mrot_3d %>% distinct(user_id, .keep_all = TRUE) %>% pull(age) %>% min(na.rm = TRUE) %>% round(1)
age_3d_max  <- mrot_3d %>% distinct(user_id, .keep_all = TRUE) %>% pull(age) %>% max(na.rm = TRUE) %>% round(1)

# combined model numbers
anova_combined_chisq <- anova_combined$Chisq[2]
anova_combined_df    <- anova_combined$Df[2]
anova_combined_p     <- anova_combined[["Pr(>Chisq)"]][2]

fe_comb <- fixef(m_combined_interaction)
se_comb <- sqrt(diag(vcov(m_combined_interaction)))

beta_3d_comb <- fe_comb["angle_sc"]
se_3d_comb   <- se_comb["angle_sc"]
beta_2d_comb <- fe_comb["angle_sc"] + fe_comb["angle_sc:item_group2d"]
beta_diff    <- fe_comb["angle_sc:item_group2d"]
se_diff      <- se_comb["angle_sc:item_group2d"]

comb_coefs <- coef(summary(m_combined_interaction))
t_diff <- comb_coefs["angle_sc:item_group2d", "t value"]
p_diff <- comb_coefs["angle_sc:item_group2d", "Pr(>|t|)"]

# p-value formatter
fmt_p <- function(p) {
  if (p < .001) "< .001" else paste0("= ", sub("^0", "", round(p, 3)))
}

# 90-degree RT interpretation
sd_angle       <- mrot_lme_data |> pull(angle) |> sd()
beta_per_90    <- beta_3d_main * (90 / sd_angle)
rt_multiplier_90 <- exp(beta_per_90)
pct_increase_90  <- round((rt_multiplier_90 - 1) * 100, 1) |> unname()

```

## Angular Sensitivity

Our second research question asked whether performance is sensitive to the spatial properties of the stimuli in the way theory predicts — specifically, whether the canonical angular disparity effect first reported by @shepard1971mental replicates across all three sites. A total of `r n_children_3d` children contributed 3D trials (M age = `r age_3d_mean`, SD = `r age_3d_sd`, range = `r age_3d_min`--`r age_3d_max` years). For 3D stimuli, there was a strong positive effect of rotation angle (β = `r round(beta_3d_main, 2)`, SE = `r round(se_3d_main, 2)`, t(`r round(df_3d_main, 1)`) = `r round(t_3d_main, 2)`, p `r fmt_p(p_3d_main)`), confirming that response times increased systematically with angular disparity across the full sample. To aid interpretation, a 90-degree increase in rotation angle was associated with approximately `r pct_increase_90`% longer response times for 3D stimuli, consistent with the predicted linear increase in mental rotation time. Adding site as a moderator significantly improved model fit (χ²(`r anova_site_df`) = `r round(anova_site_chisq, 2)`, p `r fmt_p(anova_site_p)`), indicating that the magnitude of the angle slope differed across sites. The angle slope `r ifelse(p_site_co < .05, "differed significantly", "did not differ significantly")` between Germany and Colombia (β = `r round(beta_site_co, 2)`, SE = `r round(se_site_co, 2)`, p `r fmt_p(p_site_co)`), `r ifelse(p_site_ca < .05, "and was significantly shallower in Canada than Germany", "but did not differ significantly between Canada and Germany")` (β = `r round(beta_site_ca, 2)`, SE = `r round(se_site_ca, 2)`, p `r fmt_p(p_site_ca)`). Age, operationalized as a three-level grouping, did not independently moderate the angle slope once site was accounted for (χ²(`r anova_age_df`) = `r round(anova_age_chisq, 2)`, p `r fmt_p(anova_age_p)`).

For 2D stimuli, we observed a significant but smaller positive angle effect (β = `r round(beta_2d_main, 2)`, SE = `r round(se_2d_main, 2)`, t(`r round(df_2d_main, 1)`) = `r round(t_2d_main, 2)`, p `r fmt_p(p_2d_main)`). Adding site as a moderator `r ifelse(anova_2d_site_p < .05, "significantly improved model fit", "did not significantly improve model fit")` (χ²(`r anova_2d_site_df`) = `r round(anova_2d_site_chisq, 2)`, p `r fmt_p(anova_2d_site_p)`), indicating that the angle slope `r ifelse(anova_2d_site_p < .05, "differed", "did not differ significantly")` across sites. The angle slope `r ifelse(p_2d_site_co < .05, "was significantly shallower", "did not differ significantly")` in Colombia than Germany (β = `r round(beta_2d_site_co, 2)`, SE = `r round(se_2d_site_co, 2)`, p `r fmt_p(p_2d_site_co)`), and `r ifelse(p_2d_site_ca < .05, "differed significantly", "did not differ significantly")` between Canada and Germany (β = `r round(beta_2d_site_ca, 2)`, SE = `r round(se_2d_site_ca, 2)`, p `r fmt_p(p_2d_site_ca)`). `r ifelse(p_site_co >= .05 & p_2d_site_co < .05, "This pattern is the reverse of that observed for 3D stimuli, where Colombia did not differ from Germany but Canada showed a shallower slope, suggesting that site differences in angular sensitivity may be stimulus-type dependent.", "The pattern of site differences differed between 2D and 3D stimuli, suggesting that angular sensitivity may be stimulus-type dependent.")` A combined model confirmed that the angle slope was significantly steeper for 3D than 2D stimuli (β difference = `r round(beta_diff, 2)`, SE = `r round(se_diff, 2)`, χ²(`r anova_combined_df`) = `r round(anova_combined_chisq, 2)`, p `r fmt_p(anova_combined_p)`), with the 3D slope (β = `r round(beta_3d_comb, 2)`) approximately `r round(beta_3d_comb / beta_2d_comb, 1)` times larger than the 2D slope (β = `r round(beta_2d_comb, 2)`).

@fig-rt-angle-per-run-no-cat plots mean reaction times (±95% confidence intervals) as a function of rotation angle, stratified by age group (5–7, 7–9, and 9–12 years) and site, separately for 2D and 3D stimuli. Cell-level sample sizes underlying this figure are reported in @apx-rt-angle-sample-sizes. Although the primary analyses modeled angle as a linear predictor over the 40–160° range, inspection of @fig-rt-angle-per-run-no-cat suggests some non-linearity in certain subgroups. The full 0–360° rotation functions — reported in @apx-360-angle in the context of the strategy variation analysis — show the expected inverted-U, with RT increasing to a peak around 200–240° before declining symmetrically, consistent with participants taking the shortest rotational path. The 40–160° window analysed here captures the approximately linear rising limb of this function, supporting the use of a linear model for the angular disparity effect.

Raw per-run data underlying these trends are reported in @apx-irt-ability-age. Within-person dynamics — including trial-level variability and individual trajectories across runs are shown in @apx-within-person-dynamics. 

```{r}
#| label: fig-rt-angle-per-run-no-cat
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-align: "center"
#| include: true
#| fig-width: 7
#| fig-height: 6.5

mrot_items_rt_runs <- mrot_rt_trial %>%
  group_by(site, site_label, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, site_label, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

min_n <- 10

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci %>% filter(n >= min_n),
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site_label ~ age_group) +
  labs(x = "Rotation angle (°)", y = "RT (s)",
       colour = "Stimulus", linetype = "Item Type") +
  scale_x_continuous(breaks = c(40, 80, 120, 160)) +
  # .scale_colour_stimulus() +
  theme(legend.position = "bottom")
p_mrot_rt_angle_age_1

# ggsave("fig3_rt_by_angle_age_site.png", plot = p_mrot_rt_angle_age_1,
#        width = 8, height = 7, dpi = 300)
```

\clearpage

```{r}
#| label: gender-analysis-prep
#| include: false

# note: survey data loaded in 'load data' chunk above.

survey_gender_wide <- survey_gender |>
  filter(variable %in% c("ChildSexBirth", "ChildGenderCurrent")) |>
  select(child_id, variable, string_response) |>
  group_by(child_id, variable) |>
  # keep non-NA if available, otherwise NA
  summarise(string_response = na.omit(string_response) |> unique() |> 
              (\(x) if(length(x) == 1) x else NA_character_)(),
            .groups = "drop") |>
  tidyr::pivot_wider(names_from = variable, values_from = string_response) |>
  rename(sex = ChildSexBirth, 
         gender = ChildGenderCurrent)

# # distributions
# table(survey_gender_wide$sex, useNA = "always")
# table(survey_gender_wide$gender, useNA = "always")
# 
# # do any child_ids from survey match user_ids in your main data?
# sum(survey_gender_wide$child_id %in% mrot_scores$user_id)
# sum(survey_gender_wide$child_id %in% mrot_rt_trial$user_id)
```

```{r}
#| label: german-data
#| include: false

de_sex <- readxl::read_excel(here::here("demographics_data","all_Leipzig_participants_nonames.xlsx")) |>
  rename(user_id = uid, sex = Gender) |>
  mutate(sex_demo = recode_values(sex, "m" ~ "Male", "f" ~ "Female"))

# number of DE users
n_distinct(mrot_df$user_id[mrot_df$site_label == "mpieva-de"])

# number overlapping with data
length(intersect(de_sex$user_id, mrot_df$user_id))

# number overlapping with surveys
length(intersect(de_sex$user_id, survey_gender_wide$child_id))

sex_data <- full_join(de_sex, 
          rename(survey_gender_wide, sex_survey = sex), 
          by = c("user_id" = "child_id"))

sex_data |>
  filter_out(is.na(sex_survey) | is.na(sex_demo)) |>
  filter(sex_demo != sex_survey)


sex_data_cleaned <- sex_data |>
  filter_out(sex_demo != sex_survey) |>
  transmute(user_id = user_id,
    sex = ifelse(is.na(sex_survey), sex_demo, sex_survey))
```

```{r}
#| label: sex-data-coverage
#| include: false

mrot_rt_trial |> 
  distinct(user_id) |> 
  nrow()

mrot_rt_trial |>
  left_join(sex_data_cleaned) |>
  distinct(user_id, site, sex) |>
  group_by(site) |>
  summarise(
    n_total = n(),
    n_with_sex = sum(!is.na(sex)),
    pct = round(100 * n_with_sex / n_total, 1)
  )

```

```{r}
#| label: gender-models
#| include: false

mrot_df_gender <- inner_join(mrot_df, survey_gender_wide, 
                             by = c("user_id" = "child_id")) |>
  filter_out(is.na(sex))

# Mike's model
# sex_mod <- lmer(metric_value ~ age * sex + (1 | user_id) + (age * sex | site),
#   data = mrot_df_gender)
# 
# summary(sex_mod)

# more simple model (Fionnuala) to avoid singular fits
sex_mod <- lmer(metric_value ~ age * sex + (1 | user_id) + (1 | site),
                 data = mrot_df_gender)
summary(sex_mod)
# no evidence of sex differences in either overall performance or developmental trajectories

# German model
# sex_mod_de <- lmer(metric_value ~ age * sex + (1 | user_id),
#   data = filter(mrot_df_gender, site == "pilot_mpieva_de"))
# 
# summary(sex_mod_de)

# Bayesian model -- left for completeness but not used
# updated model with priors (not reported)
# sex_mod_bayes <- brm(
#   metric_value ~ age * sex + (1 | user_id) + (1 | site),
#   data = mrot_df_gender,
#   prior = c(
#     prior(normal(0, 1), class = "b"),
#     prior(normal(0, 1), class = "Intercept"),
#     prior(exponential(1), class = "sd")
#   ),
#   control = list(adapt_delta = 0.99, max_treedepth = 15)
# )
# for compariosn
# Regression Coefficients:
#             Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept      -2.63      0.53    -3.64    -1.50 1.00     1534     1921
# age             0.24      0.03     0.18     0.30 1.00     1252     1990
# sexMale        -0.02      0.38    -0.78     0.70 1.00     1153     1946
# age:sexMale     0.01      0.04    -0.06     0.09 1.00     1152     1976
```

## Sex Differences

Our third research question examined whether sex differences in mental rotation ability are detectable in this sample, and whether any such differences interact with age. Ability increased with age (β = `r round(lme4::fixef(sex_mod)["age"], 2)`, SE = `r round(coef(summary(sex_mod))["age","Std. Error"], 2)`, p < .001), with no evidence for a main effect of sex (β = `r round(lme4::fixef(sex_mod)["sexMale"], 2)`, p = `r format.pval(coef(summary(sex_mod))["sexMale","Pr(>|t|)"], digits = 2, eps = .001)`) or an age-by-sex interaction (β = `r round(lme4::fixef(sex_mod)["age:sexMale"], 2)`, p = `r format.pval(coef(summary(sex_mod))["age:sexMale","Pr(>|t|)"], digits = 2, eps = .001)`).

```{r}
#| label: gender-plot
#| fig-cap: "Mental rotation ability (θ) as a function of age and sex across three sites."
#| include: true
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5

ggplot(mrot_df_gender, aes(x = age, y = metric_value, colour = sex)) +
  geom_point(alpha = 0.35, size = 1.1) +
  # geom_line(aes(group = user_id), alpha = .3) + 
  geom_smooth(method = "lm") +
  labs(
    x = "Age (years)",
    y = "Ability (θ)",
    colour = "Sex"
  ) +
  facet_wrap(~site_label) + 
  theme(legend.position = "bottom")
```

```{r}
#| label: strategy-analysis-prep
#| results: "hide"

# ability
ability_run <- mrot_df %>%
  group_by(site_label, user_id, run_id, age, age_group) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    .groups = "drop"
  )

# prep trial-level data for the rich model
mrot_rich_data <- mrot_rt_trial %>%
  # filter(!adaptive) %>% # note: added by Fionnuala, april 12
  mutate(
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(angle)),
    trial_s  = as.numeric(scale(trial_number))
  )

# set sum-to-zero contrasts - by default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the grand mean slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | run_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# extract subject-level random effects
ran_effs <- ranef(m_rich)$run_id %>%
  tibble::rownames_to_column("run_id") %>%
  as_tibble() %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# compute instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(run_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

# join
rich_features_final <- rich_features_raw %>%
  left_join(instability_data, by = c("run_id")) %>% 
  left_join(ability_run,  by = c("run_id")) %>%
  mutate(Site = site_label) %>%
  # filter to ensure all needed features
  filter(if_all(c(ability, Rel_Slope, Abs_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))
```

```{r}
#| label: clustering-execution

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1], # red
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], # blue
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  # green
)

analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, 
                  Instability, Learning_Rate), is.finite)) 

global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )

# as.character() on all pull() calls
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% 
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

effective_clusters <- cluster_stats %>%
  filter(as.character(Global_Cluster) != low_eff_id) %>%
  arrange(mean_slope) 

analytical_id <- effective_clusters %>%
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

holistic_id <- effective_clusters %>%
  slice(2) %>% 
  pull(Global_Cluster) %>%
  as.character()

analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    as.character(Global_Cluster) == low_eff_id ~ "Ineffective Strategies",
    as.character(Global_Cluster) == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    as.character(Global_Cluster) == holistic_id ~ "Mental Rotation"
  ),
  Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

# ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
#   geom_point(alpha = 0.6) +
#   scale_color_manual(values = my_colors) +
#   labs(title = "Cluster Result: Ability vs. Relative Slope")
```

```{r}
#| label: strategy-reporting-objects

ineff <- analysis_data %>%
  filter(Strategy_Label == "Ineffective Strategies") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )

flip <- analysis_data %>%
  filter(Strategy_Label == "Other Effective Strategies (Flipping/Matching)") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )

mrot_strat <- analysis_data %>%
  filter(Strategy_Label == "Mental Rotation") %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE)
  )
```

## Strategy Variation
Our fourth research question asked, exploratorily, whether children's trial-level behavioral signatures reveal meaningful variation in how they engage with the task's spatial demands — and whether any such variation differs across cultural contexts. We approach this as an attempt to characterize a continuum of engagement rather than identify discrete strategy types — from children who appear to disengage from the spatial structure of the task entirely, to those who engage effectively in varying ways. All analyses in this section are exploratory and descriptive; findings should be treated as hypothesis-generating rather than confirmatory.

K=3 clusters (see the [Robustness Checks](#sec-robustness-checks) section below for justification of K=3) showed clearly differentiated profiles on both defining dimensions (@fig-validation-boxplots). Ability differed across all pairwise comparisons (all *p*s < .0001), with the 'Ineffective Strategies' group showing markedly lower ability (M = `r round(ineff$Mean_Ability, 2)`, SD = `r round(ineff$SD_Ability, 2)`) than both the 'Other Effective Strategies (Flipping/Matching)' (M = `r round(flip$Mean_Ability, 2)`, SD = `r round(flip$SD_Ability, 2)`) and 'Mental Rotation' (M = `r round(mrot_strat$Mean_Ability, 2)`, SD = `r round(mrot_strat$SD_Ability, 2)`) groups. Relative RT slopes also differed significantly across all pairwise comparisons (all *p*s < .001), with the Mental Rotation cluster showing the steepest slope (M = `r round(mrot_strat$Mean_Slope, 2)`), consistent with analog rotation processes that scale with angular disparity. We note that flipping and feature matching are theoretically distinct processes [@mckee2025malleability; @hegarty2018ability], but are treated as a single cluster here on pragmatic grounds that both tend to produce a weaker or less systematic angular disparity effect than analog rotation. 

Inspection of the RT–angle functions (@fig-rt-by-disparity) reveals qualitatively distinct profiles across the three groups. The Ineffective Strategies group (n = `r ineff$N`) is characterized bylow ability (M = `r round(ineff$Mean_Ability, 2)`, SD = `r round(ineff$SD_Ability, 2)`), a near-zero RT–angle slope (M = `r round(ineff$Mean_Slope, 2)`), and high response instability (M = `r round(ineff$Mean_Instab, 2)`). Unlike the other groups, this near-zero slope is not accompanied by accurate performance — combined with steeply declining accuracy (@fig-accuracy-by-angle), the absence of angular sensitivity represents a meaningfully different pattern from simply being slow. The label 'Ineffective Strategies' captures a behaviorally heterogeneous group united by the absence of angular sensitivity rather than any shared approach to the task: children assigned here may include those who disengage entirely at higher angular disparities, those who rely on non-spatial heuristics such as position-based responding, and those who attempt rotation but abandon it before completion — the RT signature alone cannot distinguish between these possibilities.

The 'Other Effective Strategies (Flipping/Matching)' group (n = `r flip$N`) is the second largest cluster and shows the highest mean ability (M = `r round(flip$Mean_Ability, 2)`, SD = `r round(flip$SD_Ability, 2)`). Note that although this group's absolute RT values are lower than the Mental Rotation group for most of the angle range (as visible in @fig-rt-by-disparity), this does not indicate faster or more efficient rotation — rather, it reflects the Mental Rotation group's steeper relative slope (M = `r round(mrot_strat$Mean_Slope, 2)`, SD = `r round(mrot_strat$SD_Slope, 2)`), which drives their RT substantially higher at larger angular disparities. Their RT–angle function rises to a peak around 120° before declining at 160°, consistent with exploiting reflective symmetry at high angular disparities (i.e. flipping) rather than performing full analog rotation. As items approach 180°, mirror images become easier to distinguish perceptually without requiring full rotation.

The Mental Rotation group (n = `r mrot_strat$N`) shows the classic monotonically increasing RT–angle function consistent with the Shepard-Metzler signature, with the steepest relative slope of the three groups (M = `r round(mrot_strat$Mean_Slope, 2)`, SD = `r round(mrot_strat$SD_Slope, 2)`). Interestingly, this group has lower mean ability than the Flipping/Matching group (M = `r round(mrot_strat$Mean_Ability, 2)`, SD = `r round(mrot_strat$SD_Ability, 2)`), despite showing the RT signature the task was designed to elicit. This may reflect the greater cognitive cost of analog rotation relative to feature-matching strategies, particularly for the 2D stimuli where rotational demands are lower.

```{r}
#| label: fig-validation-boxplots
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
#| fig-cap: "**Cluster profiles.** A) Ability (θ) by strategy group. B) Relative RT slope by strategy group. Significance bars show pairwise t-tests."

my_comparisons <- list(
  c("Ineffective Strategies", "Other Effective Strategies (Flipping/Matching)"),
  c("Ineffective Strategies", "Mental Rotation"),
  c("Other Effective Strategies (Flipping/Matching)", "Mental Rotation")
)
# wrap x-axis labels to two lines, including n and %
wrapped_labels <- analysis_data %>%
  count(Strategy_Label) %>%
  mutate(
    pct = round(n / sum(n) * 100),
    label = case_when(
      Strategy_Label == "Ineffective Strategies" ~ 
        glue::glue("Ineffective Strategies\n(n = {n}, {pct}%)"),
      Strategy_Label == "Other Effective Strategies (Flipping/Matching)" ~ 
        glue::glue("Other Effective\nStrategies (Flipping/\nMatching) (n = {n}, {pct}%)"),
      Strategy_Label == "Mental Rotation" ~ 
        glue::glue("Mental Rotation\n(n = {n}, {pct}%)")
    )
  ) %>%
  pull(label, name = Strategy_Label)

p_ability <- ggplot(analysis_data, aes(x = Strategy_Label, y = ability, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  scale_x_discrete(labels = setNames(wrapped_labels, 
                     c("Ineffective Strategies",
                       "Other Effective Strategies (Flipping/Matching)",
                       "Mental Rotation"))) +
  geom_signif(comparisons = my_comparisons,
              map_signif_level = TRUE,
              y_position = c(2.5, 3.0, 3.5),
              tip_length = 0.01,
              linewidth = 0.25) +
  labs(title = "A) Ability by cluster", y = "Ability (θ)", x = NULL) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))

p_slope <- ggplot(analysis_data, aes(x = Strategy_Label, y = Rel_Slope, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  scale_x_discrete(labels = setNames(wrapped_labels,
                     c("Ineffective Strategies",
                       "Other Effective Strategies (Flipping/Matching)",
                       "Mental Rotation"))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = pal$grey) +
  geom_signif(comparisons = my_comparisons,
              map_signif_level = TRUE,
              y_position = c(1.0, 1.5, 2.0),
              tip_length = 0.01,
              linewidth = 0.25) +
  labs(title = "B) RT slope by cluster", y = "Relative Slope", x = NULL) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  coord_cartesian(ylim = c(-2, 2.5))

p_combined <- p_ability + p_slope
p_combined

# ggsave("fig4_cluster_profiles.png", plot = p_combined,
#        width = 10, height = 6, dpi = 300)
```

```{r}
#| label: fig-rt-by-disparity
#| fig-width: 7
#| fig-height: 4
#| fig-align: "center"
#| include: true
#| fig-cap: "**Participant-weighted mean log RT by rotation angle and strategy cluster.** Shaded bands show 95% CIs. Note that raw means for the Ineffective Strategies group appear to decline slightly with angle; however, this reflects confounds with stimulus type, item difficulty, and trial position that are controlled in the mixed model. The model-adjusted relative slope for this group is near-zero positive (M = `r round(ineff$Mean_Slope, 2)`), indicating absent rather than negative angular sensitivity."

rt_run_disparity <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(Site, run_id, Strategy_Label),
    by = c("run_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors))) %>%
  group_by(Strategy_Label, run_id, angle) %>%      
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_disparity_summary <- rt_run_disparity %>%
  group_by(Strategy_Label, angle) %>%                 
  summarise(
    n_runs   = n_distinct(run_id),
    mean_rt  = mean(user_mean_log_rt, na.rm = TRUE),
    se       = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_runs),
    lwr      = mean_rt - 1.96 * se,
    upr      = mean_rt + 1.96 * se,
    .groups  = "drop"
  )

p_rt_disparity <- ggplot(rt_disparity_summary, aes(x = angle, y = mean_rt,
                                  colour = Strategy_Label, fill = Strategy_Label)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(20, 40, 80, 120, 160)) +
  theme(legend.position = "bottom",
      plot.margin = margin(t = 5, r = 5, b = 5, l = 15, unit = "pt")) +
  labs(
    x = "Rotation Angle (°)",
    y = "Mean log RT",
    colour = "Strategy", fill = "Strategy"
  )
p_rt_disparity

# ggsave("fig5_rt_by_angle_strategy.png", plot = p_rt_disparity,
#        width = 7, height = 5, dpi = 300)
```

```{r}
#| label: fig-accuracy-by-angle
#| fig-width: 7
#| fig-height: 4
#| fig-align: "center"
#| fig-cap: "**Participant-weighted accuracy-angle functions by strategy cluster.** Shaded bands show 95% CIs."

acc_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(Site, run_id, Strategy_Label),
    by = c("run_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_run_angle <- acc_by_strategy %>%
  group_by(Strategy_Label, run_id, angle) %>%
  summarise(
    user_mean_acc = mean(correct, na.rm = TRUE),
    .groups = "drop"
  )

acc_summary <- acc_run_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_runs    = n_distinct(run_id),
    mean_acc  = mean(user_mean_acc, na.rm = TRUE),
    se_acc    = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_runs),
    .groups = "drop"
  )

ggplot(acc_summary, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc, ymax = mean_acc + 1.96 * se_acc, fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  scale_colour_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, 
    drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(20, 40, 80, 120, 160)) +
  theme(legend.position = "bottom",
      plot.margin = margin(t = 5, r = 5, b = 5, l = 15, unit = "pt")) +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean Accuracy",
    colour = "Strategy", fill = "Strategy"
  )
```
\clearpage

```{r}
#| label: tbl-descriptive-validation
#| tbl-cap: "**Descriptive statistics by Strategy Cluster.**"
#| include: false

validation_table <- analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE),
    .groups = "drop"
  )

validation_table %>%
  transmute(
    Strategy      = Strategy_Label,
    N             = N,
    `Mean Ability`= Mean_Ability,
    `SD Ability`  = SD_Ability,
    `Mean Slope`  = Mean_Slope,
    `SD Slope`    = SD_Slope,
    `Instability` = Mean_Instab
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>% 
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability), 
    decimals = 2
  ) %>%
  fmt_number(columns = c(N), decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability)) %>%
  cols_width(
    Strategy     ~ px(180),
    N            ~ px(40),
    everything() ~ px(75)
  ) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

### Robustness checks {#sec-robustness-checks}

Three robustness checks are reported in full in the appendices. First, comparison of k=2 and k=3 solutions confirmed that k=2 collapsed two distinguishable RT–angle profiles into a single region, justifying the three-profile solution (see @apx-cluster-summary-stats and @apx-cluster). Cluster membership by age group is reported in @apx-strategy-age-group. Second, dimensionality reduction via PCA, t-SNE, and UMAP consistently recovered the separation between the Ineffective Strategies group and the two effective engagement profiles, while the boundary between the Mental Rotation and Other Effective Strategies (Flipping/Matching) groups remained less clear — supporting the continuum interpretation (see @apx-dimreduc). Third, re-analysis using full 0–360° rotation angles replicated the main findings and provided direct support for shortest-path rotation, with both the Mental Rotation and Other Effective Strategies (Flipping/Matching) groups showing near-symmetric inverted-U RT functions peaking around 200–240° (see @apx-360-angle).

```{r}
#| label: local-clustering-agreement
#| include: false

cluster_vars <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")

stopifnot(all(cluster_vars %in% names(analysis_data)))
stopifnot(all(c("Site", "user_id", "run_id", "Strategy_Label") %in% names(analysis_data)))
stopifnot(nrow(analysis_data) == nrow(distinct(analysis_data, Site, user_id, run_id)))

# ARI function
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

get_local_strategy_labels <- function(df_site,
                                      k = 3,
                                      seed = 123,
                                      nstart = 25,
                                      min_n = 30) {

  if (nrow(df_site) < min_n) return(NULL)
  set.seed(seed)

  X <- df_site %>%
    select(all_of(cluster_vars)) %>%
    scale() %>%
    as.matrix()

  km <- kmeans(X, centers = k, nstart = nstart)

  tmp <- df_site %>%
    mutate(Local_Cluster = factor(km$cluster))

  local_stats <- tmp %>%
    group_by(Local_Cluster) %>%
    summarise(
      mean_abil  = mean(ability,   na.rm = TRUE),
      mean_slope = mean(Rel_Slope, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(mean_abil)

  ineff_id     <- local_stats %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  other_two    <- local_stats %>% filter(as.character(Local_Cluster) != ineff_id) %>% arrange(mean_slope)
  other_eff_id <- other_two   %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  mrot_id      <- other_two   %>% slice(2) %>% pull(Local_Cluster) %>% as.character()

  tmp %>%
    mutate(
      Strategy_Local = case_when(
        as.character(Local_Cluster) == ineff_id ~ "Ineffective Strategies",
        as.character(Local_Cluster) == other_eff_id ~ "Other Effective Strategies (Flipping/Matching)",
        as.character(Local_Cluster) == mrot_id ~ "Mental Rotation"
      ),
      Strategy_Local = factor(Strategy_Local, levels = levels(analysis_data$Strategy_Label))
    ) %>%
    select(Site, user_id, Strategy_Local)
}

# fit local models independently within each site
local_labels_all <- analysis_data %>%
  group_by(Site) %>%
  group_split() %>%
  purrr::map_dfr(get_local_strategy_labels)

analysis_local <- analysis_data %>%
  select(-any_of("Strategy_Local")) %>%
  left_join(local_labels_all, by = c("Site", "user_id"))

# agreement (3-class)
agreement_3class <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_by(Site) %>%
  summarise(
    n         = n(),
    pct_agree = mean(Strategy_Local == Strategy_Label),
    ARI       = ARI(Strategy_Label, Strategy_Local),
    .groups   = "drop"
  )

agreement_3class

# confusion matrices per site — save split first to avoid pipe . scoping issue
site_splits <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_split(Site)

confusion_by_site <- site_splits %>%
  purrr::set_names(purrr::map_chr(site_splits, ~ unique(.x$Site))) %>%
  purrr::map(~ table(.x$Strategy_Local, .x$Strategy_Label))

confusion_by_site

# recall + purity
recall_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Label, Strategy_Local) %>%
  group_by(Site, Strategy_Label) %>%
  mutate(recall = n / sum(n)) %>%
  ungroup()

purity_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Local, Strategy_Label) %>%
  group_by(Site, Strategy_Local) %>%
  mutate(purity = n / sum(n)) %>%
  ungroup()

diag_summary <- recall_tbl %>%
  filter(Strategy_Label == Strategy_Local) %>%
  transmute(Site, Strategy = Strategy_Label, recall) %>%
  left_join(
    purity_tbl %>%
      filter(Strategy_Local == Strategy_Label) %>%
      transmute(Site, Strategy = Strategy_Local, purity),
    by = c("Site", "Strategy")
  )

diag_summary
```

```{r}
#| label: local-vs-pooled-reporting
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_pct <- function(x, d = 1) paste0(formatC(100*x, format = "f", digits = d), "%")

get_row <- function(df, site) df[df$Site == site, , drop = FALSE]

# sites
s_mpi <- "mpieva-de"
s_uni <- "uniandes-co"
s_wes <- "western-ca"

# agreement rows
a_mpi <- get_row(agreement_3class, s_mpi)
a_uni <- get_row(agreement_3class, s_uni)
a_wes <- get_row(agreement_3class, s_wes)

# label strings
lab_ineff <- "Ineffective Strategies"
lab_other <- "Other Effective Strategies (Flipping/Matching)"
lab_mrot  <- "Mental Rotation"

get_diag <- function(site, label) {
  out <- diag_summary[diag_summary$Site == site & diag_summary$Strategy == label, , drop = FALSE]
  if (nrow(out) == 0) {
    return(data.frame(Site = site, Strategy = label, recall = NA_real_, purity = NA_real_))
  }
  stopifnot(nrow(out) == 1)
  out
}

d_mpi_ineff <- get_diag(s_mpi, lab_ineff)
d_mpi_other <- get_diag(s_mpi, lab_other)
d_mpi_mrot  <- get_diag(s_mpi, lab_mrot)

d_uni_ineff <- get_diag(s_uni, lab_ineff)
d_uni_other <- get_diag(s_uni, lab_other)
d_uni_mrot  <- get_diag(s_uni, lab_mrot)

d_wes_ineff <- get_diag(s_wes, lab_ineff)
d_wes_other <- get_diag(s_wes, lab_other)
d_wes_mrot  <- get_diag(s_wes, lab_mrot)

# off-diagonal moves
get_move <- function(tab, local_lab, global_lab) {
  stopifnot(local_lab %in% rownames(tab), global_lab %in% colnames(tab))
  n     <- tab[local_lab, global_lab]
  denom <- sum(tab[, global_lab])
  list(n = as.numeric(n), denom = as.numeric(denom), prop = as.numeric(n) / as.numeric(denom))
}

tab_mpi <- confusion_by_site[[s_mpi]]
tab_uni <- confusion_by_site[[s_uni]]
tab_wes <- confusion_by_site[[s_wes]]

# Mental Rotation globally assigned to Other Effective locally
mpi_mrot2other <- get_move(tab_mpi, lab_other, lab_mrot)
wes_mrot2other <- get_move(tab_wes, lab_other, lab_mrot)

# Other Effective globally assigned to Mental Rotation locally
uni_other2mrot <- get_move(tab_uni, lab_mrot, lab_other)
agreement_3class
confusion_by_site
```

### Cross-Cultural Consistency of Strategy Clusters

We next examined whether the continuum structure is recoverable within each site independently by re-running the k-means procedure separately within each site and computing agreement between the local and global solutions using percent agreement and the Adjusted Rand Index (ARI). The most fundamental distinction — between children who disengage from the spatial structure of the task and those who engage effectively — was recovered consistently across all three sites. The Ineffective Strategies and Other Effective Strategies (Flipping/Matching) profiles showed moderate-to-good local recovery across sites (Canada: `r fmt_pct(a_wes$pct_agree)`, ARI = `r fmt_num(a_wes$ARI)`; Colombia: `r fmt_pct(a_uni$pct_agree)`, ARI = `r fmt_num(a_uni$ARI)`; Germany: `r fmt_pct(a_mpi$pct_agree)`, ARI = `r fmt_num(a_mpi$ARI)`); Germany's lower percent agreement despite its larger sample may reflect greater RT variability introduced by its remote home-based testing format.

The finer-grained boundary between the Mental Rotation and Other Effective Strategies (Flipping/Matching) profiles was less stable: the Mental Rotation profile was not independently recoverable in Germany or Canada, where children locally assigned to this region predominantly mapped onto the Other Effective Strategies (Flipping/Matching) profile instead. Colombia showed the most consistent local recovery across all three regions of the continuum. These findings reinforce the interpretation that what is robustly recoverable across cultural contexts is the broad distinction between disengagement and effective spatial engagement — not the finer subdivision within the effective end of the continuum.

# Discussion

The central aim of this study was to establish whether the LEVANTE mental rotation task measures the same underlying construct across three culturally diverse sites. The answer, across three distinct levels of analysis, is largely yes. The task achieves scalar measurement invariance indicating that its psychometric structure is consistent across cultural contexts. The canonical angular disparity effect replicates across all three sites for 3D stimuli, providing process-level evidence that performance is driven by the spatial demands of the task in the way theory predicts [@shepard1971mental]. And at the level of individual behavioral profiles, the broad distinction between children who engage with the spatial structure of the task and those who do not is recoverable within each site independently. That these three forms of convergent evidence emerge across geographically and culturally diverse samples extends the cross-cultural generalizability of mental rotation research beyond solely WEIRD samples [@davis2021cultural; @henrich2010weirdest], and supports the interpretation that the task is capturing a genuinely spatial construct rather than one that is specific to the cultural or educational contexts in which it was developed.

## Cross-Site Invariance

Turning to our first research question, the demonstration of scalar invariance under a 2PL model establishes that the mental rotation task functions comparably across pilot sites — that items are ordered by difficulty in the same way for children in Bogotá as for children in Leipzig or Ontario, and that ability estimates can be placed on a common latent scale [@putnick2016measurement]. This is a meaningful result: tasks developed in one context do not automatically achieve psychometric equivalence when administered elsewhere [@van2004bias], and the inclusion of a low-to-middle-income site from the Global South makes this a stronger test than studies conducted entirely within high-income North American and European samples. That said, scalar invariance establishes metric comparability rather than cultural equivalence in any deeper sense. With only one non-WEIRD site — though one spanning both urban and rural schools, representing a meaningfully different cultural and socioeconomic context from the German and Canadian samples — the present findings should be treated as an encouraging foundation rather than evidence of broad cross-cultural portability. Further validation across more culturally and socioeconomically diverse sites will be needed before stronger claims can be made. Finally, given that LEVANTE sites differ substantially in sampling, recruitment, and testing format, site-level performance differences are not interpreted here regardless of the invariance result.

## Angular Sensitivity

Our second research question asked whether performance is driven by the spatial properties of the stimuli in the way theory predicts. The key benchmark here is the angular disparity effect first reported by @shepard1971mental: as the rotational distance between a 3D target and its match increases, so too does the time required to judge whether they correspond. This relationship, among the most replicated findings in cognitive psychology [@cooper1973chronometric; @peters2008applications], has been demonstrated extensively in WEIRD child and adult samples [@kaltner2016developmental; @jansen2015role], but its replication across culturally diverse samples of children had not previously been established. Its consistent emergence here, across all three sites, suggests that the underlying cognitive process it indexes — the spatially continuous transformation of object orientation — is not specific to the cultural or educational contexts in which it was originally documented. This matters because psychometric equivalence and process-level validity are distinct forms of evidence. It is one thing to show that a task achieves comparable difficulty ordering across sites; it is another to show that performance is sensitive to the spatial structure of the stimuli in the way that mental rotation theory predicts [@borsboom2004concept]. The angular disparity effect provides the latter, and its cross-cultural replication substantially strengthens the case that the task is engaging spatial cognition specifically, rather than some culturally-variable combination of task-general skills. The shallower slope for 2D stimuli likely reflects the greater familiarity of duck and rabbit silhouettes relative to the 3D geometric shapes: when stimuli are easy to tell apart, children can make same/different judgments by comparing surface features rather than rotating, producing a weaker RT–angle relationship [@folk1987effects; @bethell1988mental].

Children at mpieva-de showed both higher baseline response times and steeper absolute RT–angle slopes than the other two sites — a pattern consistent with a testing format explanation rather than a genuine cultural difference in angular sensitivity. Remote home-based testing on family devices, as used at mpieva-de, may introduce environmental variability that is likely to inflate both baseline RT and trial-to-trial noise; because absolute slopes scale with baseline speed, this would mechanically produce steeper slopes even without greater rotation sensitivity, a pattern visible in the wider raw RT distributions at mpieva-de (@apx-irt-ability-age). 

## Sex Differences

Our third research question asked whether sex differences in mental rotation ability are detectable in this sample, and whether any such differences interact with age. They are not: the absence of a main effect of sex or an age-by-sex interaction replicates the pattern reported in several recent studies of primary school-aged children using child-friendly mental rotation tasks [@hodgkiss2021developmental; @lauer2019development], and is consistent with the meta-analytic picture: reliable sex differences in mental rotation are small and inconsistent in this age range, particularly for tasks using 2D stimuli and child-friendly formats of the kind used here [@lauer2019development]. The LEVANTE task's design -- 2D animal silhouettes alongside 3D block figures -- places it squarely in the category of tasks where the meta-analytic expectation for a gender difference is modest at best. Notably, the null finding held across all three international study sites, extending prior null results beyond the WEIRD samples in which they have predominantly been documented [@hodgkiss2021developmental]. Taken together, the sex difference results are consistent with the broader argument of this paper: that the task captures a spatially-driven construct that is consistent across cultural contexts.

## Strategy Variation

Our fourth research question asked whether meaningful variation in how children engage with the spatial demands of the task can be identified from trial-level behavioral signatures. Research with adults has established that solvers draw on a range of strategies when performing mental rotation tasks, spanning a holistic-analytic continuum [@gluck2003spatial], and that strategy selection varies across individuals and shifts with task demands [@nazareth2019strategy]. Crucially, such strategies can be identified from behavioral signatures in RT data even in the absence of direct observation [@geiser2006separating], providing the methodological basis for the present analysis — though we treat the resulting profiles as descriptive anchors along a continuum of engagement rather than as evidence for discrete strategy types.

The clustering exercise identified three differentiable regions of this continuum, partly because the data resist clean separation and partly because, as discussed below, the Mental Rotation profile proves difficult to recover within individual sites. At one end, a subset of children — labeled 'Ineffective Strategies' — show clear disengagement from the spatial structure of the task: near-zero model-adjusted RT–angle slopes, high response instability, and near-chance accuracy at high angular disparities. Notably, this group's near-zero relative slope superficially resembles that of the 'Other Effective Strategies (Flipping/Matching)' group, but for a different reason: disengagement rather than an alternative spatial solution. At the other end, children engage effectively with the task, though they vary in the form that engagement takes.

The two effective profiles — Other Effective Strategies (Flipping/Matching) and Mental Rotation — are distinguished primarily by the shape of their RT–angle functions, but notably not by accuracy, which was similar across both profiles. Importantly, the two profile-defining dimensions do different separating work: ability distinguishes the Ineffective Strategies group from both effective profiles, while relative RT slope distinguishes the two effective profiles from each other — and it is this complementary structure that justifies retaining three profiles rather than two. A striking finding is that the Mental Rotation profile is both the most common globally (n = `r mrot_stratN‘,‘rround(mrotstratN`, `r round(mrot_strat N‘,‘rround(mrots​tratN / nrow(analysis_data) * 100)`%) and yet associated with lower mean ability than the Other Effective Strategies (Flipping/Matching) group — the reverse of what the broader literature might lead one to expect [@khooshabeh2013individual; @heil2008sex]. This fits with the interpretation that RT signatures consistent with full analog rotation carry a greater cognitive cost than feature-based engagement [@just1985cognitive; @khooshabeh2013individual], particularly for the 2D stimuli where rotational demands are lower. The overlap between these two profiles in the behavioral feature space — consistent across all three dimensionality reduction methods (see @apx-dimreduc) — reinforces this picture: the distinction between them is gradient rather than categorical.

The Mental Rotation profile was not independently recoverable in Germany or Canada, raising the question of whether it reflects a genuine and replicable behavioral pattern or emerges from the particular composition of the pooled sample. We retain the distinction in the global solution because collapsing to k=2 obscures a theoretically meaningful difference in RT–angle profile shape, but we treat it as a convenient marker of one end of the effective-engagement spectrum rather than a discrete strategy type. Colombia showed the strongest local recovery across all three regions of the continuum. Whether the failure to recover the Mental Rotation profile within Germany and Canada reflects a genuine difference in the distribution of engagement profiles across sites, or simply insufficient power to separate it from the Flipping/Matching profile at the available site-specific sample sizes, warrants investigation with larger samples. Taken together, the exploratory clustering exercise reveals a meaningful continuum of engagement with the task's spatial demands — from children who appear to disengage from its spatial elements entirely, to those who solve it effectively through different spatial processes. This continuum is recoverable, at least in its broad outline, across three culturally diverse sites.

## Limitations
Several limitations of the present study should be acknowledged. Although the three pilot sites span different continents and linguistic contexts, they are all relatively urban and educated populations — far from representative of the full range of cultural environments in which children develop. The LEVANTE project has already expanded to six additional sites since this pilot data collection, with future efforts prioritizing data collection in low- and middle-income countries, which will substantially broaden the cultural scope of findings like those reported here. Sample sizes were also unbalanced across sites, with the Canadian sample notably smaller, which limits the precision of site-level estimates and the power to detect cross-site differences. The cross-sectional design is a further constraint: developmental trajectories here are inferred from age differences between children rather than observed directly. Longitudinal data are currently being collected and processed, however, and will allow this question to be addressed more directly in due course. Finally, the strategy clustering analysis is exploratory and should be treated with caution. The clusters are inferred from behavioral proxies rather than directly observed strategies, and replication in larger and more diverse samples is needed before strong conclusions can be drawn.

# Conclusion

Three research questions motivated this study, each asking in a different way whether the LEVANTE mental rotation task is doing consistent and interpretable work across culturally diverse settings. Across all three, the answer is largely affirmative. These findings do not mean that mental rotation development is uniform across children or contexts — the substantial individual variability within sites, and the tentative evidence for cross-cultural differences in strategy distributions, suggest there is much still to understand about what drives variation in spatial ability across children. But they do mean that the task is measuring something consistent —  that when a child in Bogotá, a child in Leipzig, and a child in London (Ontario) sit down to judge whether two rotated objects are the same, they are, by and large, engaging the same cognitive process. Such consistency is the foundation on which meaningful cross-cultural comparisons can be built.

\clearpage

# References

::: {#refs}
:::

# Sample Sizes per Cell for RT–Angle Figure {.appendix #apx-rt-angle-sample-sizes}

```{r}
#| label: tbl-rt-angle-sample-sizes

n_table <- mrot_items_rt_agebin_ci %>%
  filter(angle != 20) %>%
  select(site_label, age_group, stimulus, item_group, angle, n) %>%
  mutate(
    cell = paste0(item_group, ".", stimulus),
    age_group = recode(as.character(age_group),
      "[5,7]"  = "5–7",
      "(7,9]"  = "7–9",
      "(9,12]" = "9–12"
    )
  ) %>%
  select(-stimulus, -item_group) %>%
  pivot_wider(names_from = angle, values_from = n) %>%
  arrange(site_label, desc(age_group), cell)

knitr::kable(
  n_table,
  col.names = c("Site", "Age Group", "Cell", "40", "80", "120", "160"),
  caption = "Sample sizes (n) per cell."
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    position = "center",
    full_width = FALSE
  )
```

# IRT Ability by Age {.appendix #apx-irt-ability-age}

```{r}
#| label: fig-rt-angle_raw
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI. CAT trials excluded."
#| fig-width: 8
#| fig-height: 8
#| fig-align: "center"

# Add raw data underneath
p_mrot_rt_angle_age_2 <- ggplot() +
  # raw per-run medians
  geom_point(
    data = mrot_items_rt_runs,
    aes(x = angle, y = rt_med,
        colour = stimulus),
    size = 0.4,
    alpha = 0.25,
    position = position_jitter(width = 3, height = 0)
  ) +
  # summary CI + lines
  geom_errorbar(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y, ymin = lwr, ymax = upr,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    width = 2,
    linewidth = 0.35,
    alpha = 0.6
  ) +
  geom_point(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    size = 1
  ) +
  geom_line(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    alpha = 0.85
  ) +
  facet_grid(site_label ~ age_group) +
  labs(
    #title = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",
    colour = "Stimulus",
    #linetype = "Item Type"
  ) + scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom")

p_mrot_rt_angle_age_2
```

# Within-Person Dynamics {.appendix #apx-within-person-dynamics}
Having established that 3D stimuli elicit a clear angular disparity effect while 2D stimuli do to a lesser extent, we next asked a complementary question: whether children show sensitivity to task demands as they unfold in real time — specifically, whether reaction times shift systematically at the transition from 2D to 3D stimuli, and whether practice effects are detectable within blocks. To isolate time-on-task effects from item-driven RT changes, we residualized log-RTs on angle, stimulus, and item type prior to analysis.
```{r}
#| label: fig-within-person-trials
#| fig-cap: "**Trial-level Reaction Time Dynamics (angle-residualised).** Mean residual log-RT across trials for each non-adaptive site ±SE. The dashed line marks the 2D→3D transition. Re-test and CAT data removed."
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 4

# prep: remove CAT data
trial_resid_data <- mrot_rt_trial %>%
  filter(!adaptive, correct == 1)

# residualize log_rt on angle + stimulus + item_group
resid_model <- lm(log_rt ~ angle + stimulus + item_group, 
                  data = trial_resid_data)

trial_resid_data <- trial_resid_data %>%
  mutate(log_rt_resid = residuals(resid_model),
         block = item_group |> str_to_upper())

transition_trial <- trial_resid_data |>
  filter(item_group == "3d") |>
  pull(trial_number) |>
  min()

# plot
ggplot(trial_resid_data, 
       aes(x = trial_number, y = log_rt_resid, 
           color = site_label, fill = site_label)) +
  stat_summary(fun = mean, geom = "point", size = 1.2, alpha = 0.5) +
  geom_smooth(aes(group = interaction(site_label, block)),
              method = "loess", span = 0.75, se = TRUE, alpha = 0.2) +
  geom_vline(xintercept = transition_trial - 0.5, linetype = "dashed",
             color = pal$grey, linewidth = 0.7) +
  annotate("text", x = transition_trial, y = Inf,
           label = "Start of 3D block", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_colour_manual(values = site_pal_labelled) +
  scale_fill_manual(values = site_pal_labelled) +
  labs(
    x     = "Trial Number",
    y     = "Residual reaction time",
    color = "Site",
    fill  = "Site"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: within-person-models
#| include: false

# center trial number within each block so the intercept at transition is interpretable
within_person_data <- trial_resid_data %>%
  mutate(
    block         = factor(block, levels = c("2D", "3D")),
    # trial number centered within block
    trial_c       = trial_number - transition_trial,
    site          = factor(site)
  ) %>%
  filter(!is.na(trial_number), !is.na(log_rt_resid))

# Model 1: continuous trial slope only 
m_wp_continuous <- lmer(
  log_rt_resid ~ trial_c + (1 + trial_c | run_id),
  data    = within_person_data,
  REML    = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

# Model 2: add block as main effect — tests for discontinuous jump at transition
m_wp_jump <- lmer(
  log_rt_resid ~ trial_c + block + (1 + trial_c | run_id),
  data    = within_person_data,
  REML    = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

# Model 3: add trial × block interaction — tests whether slope changes after transition
m_wp_slope <- lmer(
  log_rt_resid ~ trial_c * block + (1 + trial_c | run_id),
  data    = within_person_data,
  REML    = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

# model comparisons
anova_wp_jump  <- anova(m_wp_continuous, m_wp_jump)   # is there a jump?
anova_wp_slope <- anova(m_wp_jump, m_wp_slope)         # does slope also change?

# extract key numbers
wp_coefs       <- coef(summary(m_wp_slope))

# block jump (intercept shift at transition)
beta_wp_block  <- wp_coefs["block3D", "Estimate"]
se_wp_block    <- wp_coefs["block3D", "Std. Error"]
p_wp_block     <- wp_coefs["block3D", "Pr(>|t|)"]

# within-block trial slope (2D block)
beta_wp_trial  <- wp_coefs["trial_c", "Estimate"]
se_wp_trial    <- wp_coefs["trial_c", "Std. Error"]
p_wp_trial     <- wp_coefs["trial_c", "Pr(>|t|)"]

# interaction: does slope differ in 3D block?
beta_wp_int    <- wp_coefs["trial_c:block3D", "Estimate"]
se_wp_int      <- wp_coefs["trial_c:block3D", "Std. Error"]
p_wp_int       <- wp_coefs["trial_c:block3D", "Pr(>|t|)"]

# anova numbers for inline reporting
wp_jump_chisq  <- anova_wp_jump$Chisq[2]
wp_jump_df     <- anova_wp_jump$Df[2]
wp_jump_p      <- anova_wp_jump[["Pr(>Chisq)"]][2]

wp_slope_chisq <- anova_wp_slope$Chisq[2]
wp_slope_df    <- anova_wp_slope$Df[2]
wp_slope_p     <- anova_wp_slope[["Pr(>Chisq)"]][2]
```

\clearpage

@fig-within-person-trials plots mean residual log-RT across trials for each site. To test for a discontinuous shift in RT at the 2D→3D block transition and for within-block practice effects, we fitted a series of mixed-effects models predicting residual log-RT from trial number, with by-participant random intercepts and slopes.

First, a model including a block indicator (2D vs 3D) significantly improved fit over a continuous trial-slope model (χ²(`r wp_jump_df`) = `r round(wp_jump_chisq, 2)`, p `r fmt_p(wp_jump_p)`), confirming that the increase in RT at the 2D→3D transition is discontinuous rather than a continuation of a gradual trend. The block coefficient indicates a sudden upward shift in residual RT at trial 48 (β = `r round(beta_wp_block, 2)`, SE = `r round(se_wp_block, 2)`, p `r fmt_p(p_wp_block)`), consistent with children recalibrating their processing demands at the block boundary rather than adapting incrementally.

Second, within each block, RTs decreased across trials, indicating a practice effect operating independently of angular disparity (β = `r round(beta_wp_trial, 4)`, SE = `r round(se_wp_trial, 4)`, p `r fmt_p(p_wp_trial)`). Adding a trial × block interaction `r ifelse(wp_slope_p < .05, "significantly improved model fit", "did not significantly improve model fit")` (χ²(`r wp_slope_df`) = `r round(wp_slope_chisq, 2)`, p `r fmt_p(wp_slope_p)`) `r ifelse(wp_slope_p < .05, paste0(", with the within-block acceleration differing between the 2D and 3D blocks (β = ", round(beta_wp_int, 3), ", SE = ", round(se_wp_int, 3), ", p ", fmt_p(p_wp_int), ")"), "")`. Both patterns — the discontinuous block transition and the within-block acceleration — were qualitatively consistent across all three sites, further extending the cross-cultural consistency observed in the angular disparity analysis.

# Cluster Summary Statistics {.appendix #apx-cluster-summary-stats}
```{r}
#| label: tbl-cluster-summary
#| fig-width: 6
#| fig-height: 5
#| tbl-cap: "Profile summary statistics. FLAG TO UPDATE THIS CAPTION"

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    n              = n(),
    mean_ability   = mean(ability, na.rm = TRUE),
    sd_ability     = sd(ability, na.rm = TRUE),
    mean_rel_slope = mean(Rel_Slope, na.rm = TRUE),
    sd_rel_slope   = sd(Rel_Slope, na.rm = TRUE),
    mean_abs_slope = mean(Abs_Slope, na.rm = TRUE),
    sd_abs_slope   = sd(Abs_Slope, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_rel_slope)) %>%
  transmute(
    Strategy       = Strategy_Label,
    N              = n,
    `Mean Ability` = mean_ability,
    `SD Ability`   = sd_ability,
    `Mean Rel. Slope` = mean_rel_slope,
    `SD Rel. Slope`   = sd_rel_slope,
    `Mean Abs. Slope` = mean_abs_slope,
    `SD Abs. Slope`   = sd_abs_slope
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>%
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, 
                `Mean Rel. Slope`, `SD Rel. Slope`,
                `Mean Abs. Slope`, `SD Abs. Slope`),
    decimals = 2
  ) %>%
  fmt_number(columns = N, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = where(is.numeric)) %>%
  cols_width(
    Strategy ~ px(160),
    N        ~ px(40),
    everything() ~ px(65)
  ) %>%
  tab_spanner(label = "Ability",         columns = c(`Mean Ability`, `SD Ability`)) %>%
  tab_spanner(label = "Relative Slope",  columns = c(`Mean Rel. Slope`, `SD Rel. Slope`)) %>%
  tab_spanner(label = "Absolute Slope",  columns = c(`Mean Abs. Slope`, `SD Abs. Slope`)) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

# Cluster Solution Validation {.appendix #apx-cluster}

To assess the appropriateness of the three-profile solution, we evaluated cluster quality across k = 1–8 using two complementary metrics: total within-cluster sum of squares (the elbow method) and average silhouette width (@fig-k-sweep-validation). The silhouette analysis identified k = 2 as the statistically optimal solution, with average silhouette width peaking at 0.35 before declining steeply at k = 3 and remaining low thereafter. The elbow plot showed a moderate inflection at k = 3, after which additional clusters yielded diminishing reductions in within-cluster variance. To examine whether k = 2 was substantively adequate, we examined the RT–angle functions of the resulting two-cluster solution (@fig-k2-cluster-check). The higher-ability cluster displayed a characteristic inverted-U RT function, rising to a peak around 120° before declining at 160° — a pattern that, in the three-cluster solution, distinguishes the Other Effective Strategies (Flipping/Matching) and Mental Rotation groups. This indicates that k = 2 collapses two distinguishable effective engagement profiles into a single region, obscuring a theoretically meaningful behavioral difference. We therefore adopted k = 3 on both empirical grounds (the elbow inflection) and theoretical grounds (the interpretability of the resulting RT profiles), noting that the three profiles are treated as convenient descriptive anchors along a continuum rather than evidence for discrete strategy types.

```{r}
#| label: fig-k-sweep-validation
#| eval: true
#| include: true
#| fig-width: 6
#| fig-height: 3
#| fig-cap: "**Cluster validation metrics: Elbow Method (left) and Silhouette Analysis (right).** The elbow plot shows a bend at k=3, and silhouette width peaks at k=2 before declining, providing partial support for the 3-cluster solution."

p_elbow <- fviz_nbclust(global_input, kmeans, method = "wss", k.max = 8) +
  geom_vline(xintercept = 3, linetype = 2, color = pal$red) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

p_sil <- fviz_nbclust(global_input, kmeans, method = "silhouette", k.max = 8) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

grid.arrange(p_elbow, p_sil, nrow = 1)
```

```{r}
#| label: fig-k2-cluster-check
#| eval: true
#| include: true
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "**RT–angle functions for the k=2 cluster solution.** The higher-ability cluster displays an inverted-U RT curve consistent with a mix of Mental Rotation and Flipping/Matching strategies, suggesting that k=2 collapses two qualitatively distinct effective strategies into a single group."

# --- 1. Set up Input (Same as before, removing Learning_Rate for clarity) ---
analysis_data_k2 <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, Instability, Learning_Rate), is.finite))

global_input_k2 <- analysis_data_k2 %>%
  select(Baseline_Speed, Rel_Slope, Instability, Abs_Slope, Learning_Rate) %>%
  scale()

# --- 2. Run K-Means with K=2 ---
set.seed(123)
k_global_2 <- kmeans(global_input_k2, centers = 2, nstart = 25)
analysis_data_k2$Global_Cluster <- as.factor(k_global_2$cluster)

# --- 3. Label the 2 Clusters ---
# We define them simply by Ability (Low vs High) since we can't split by slope yet
cluster_stats_k2 <- analysis_data_k2 %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability), 
    .groups = "drop"
  )

# Identify Lower Ability Cluster
low_eff_id <- cluster_stats_k2 %>% arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
# The other one is High Efficiency
high_eff_id <- cluster_stats_k2 %>% arrange(desc(mean_ability)) %>% slice(1) %>% pull(Global_Cluster)

analysis_data_k2 <- analysis_data_k2 %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id ~ "Low Efficiency (Cluster 1)",
    Global_Cluster == high_eff_id ~ "High Efficiency (Cluster 2)"
  ))

k2_colors <- c(
  "Low Efficiency (Cluster 1)"  = RColorBrewer::brewer.pal(3, "Set1")[1],
  "High Efficiency (Cluster 2)" = RColorBrewer::brewer.pal(3, "Set1")[2]
)

# --- 4. Plot the Curves (Visual Check) ---
# Join labels to raw data
rt_by_strategy_k2 <- mrot_rich_data %>%
  left_join(
    analysis_data_k2 %>% select(run_id, Strategy_Label),
    by = c("run_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle))

# Calculate User Means
rt_user_angle_k2 <- rt_by_strategy_k2 %>%
  group_by(Strategy_Label, run_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

# Calculate Group Means
rt_summary_k2 <- rt_user_angle_k2 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(run_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# plot
ggplot(rt_summary_k2, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_log_rt - 1.96*se_log_rt, 
                  ymax = mean_log_rt + 1.96*se_log_rt, 
                  fill = Strategy_Label), alpha = 0.2, colour = NA) +
  scale_colour_manual(values = k2_colors) +
  scale_fill_manual(values = k2_colors) +
  labs(
    title    = NULL,
    subtitle = NULL,
    y = "Mean log(RT) (participant-weighted)", 
    x = "Rotation angle (degrees)",
    colour = "Strategy", fill = "Strategy"
  ) +
  theme(legend.position = "bottom")
```

\clearpage

```{r}
#| label: strategy-age-data
#| include: false

strategy_by_age <- analysis_data |>
  count(age_group, Strategy_Label) |>
  group_by(age_group) |>
  mutate(prop = round(n / sum(n), 2)) |>
  ungroup()

get_prop <- function(age_grp, strategy) {
  strategy_by_age[
    strategy_by_age$age_group == age_grp & 
    strategy_by_age$Strategy_Label == strategy, 
    "prop"
  ][[1]]
}
```

# Strategy Use Across Age Groups {.appendix #apx-strategy-age-group}
We examined whether the distribution of strategy clusters varied across age groups (@fig-strategy-by-age). This analysis is purely descriptive — cell sizes are too small for formal inferential testing, and the cluster solution itself carries the caveats described above. With those caveats in mind, a descriptive pattern is nonetheless visible. The proportion of children assigned to the Ineffective Strategies cluster declined substantially with age, from `r get_prop("[5,8.3]", "Ineffective Strategies") * 100`% in the youngest group (ages 5–8.3) to `r get_prop("(8.3,10.6]", "Ineffective Strategies") * 100`% (ages 8.3–10.6) and `r get_prop("(10.6,13.7]", "Ineffective Strategies") * 100`% 
(ages 10.6–13.7). This was accompanied by a corresponding increase in the Other Effective Strategies (Flipping/Matching) cluster, from 
`r get_prop("[5,8.3]", "Other Effective Strategies (Flipping/Matching)") * 100`% to `r get_prop("(8.3,10.6]", "Other Effective Strategies (Flipping/Matching)") * 100`% and `r get_prop("(10.6,13.7]", "Other Effective Strategies (Flipping/Matching)") * 100`%. By contrast, the proportion assigned to the Mental Rotation cluster remained stable across age groups (`r get_prop("[5,8.3]", "Mental Rotation") * 100`%, 
`r get_prop("(8.3,10.6]", "Mental Rotation") * 100`%, `r get_prop("(10.6,13.7]", "Mental Rotation") * 100`%). The primary developmental shift thus appears to be a move away from task disengagement toward effective feature-matching engagement profiles, with the proportion using analog rotation remaining comparatively stable across this age range.

```{r}
#| label: fig-strategy-by-age
#| fig-align: "center"
#| include: true
#| fig-width: 7
#| fig-height: 3.5
#| fig-cap: "**Strategy cluster proportions by age group.**"

strategy_by_age %>%
  ggplot(aes(x = age_group, y = prop, fill = Strategy_Label)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = my_colors,
                    labels = function(x) str_wrap(x, width = 25)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age group (years)",
       y = "Proportion",
       fill = "Strategy") +
  theme(legend.position = "bottom")
```
  
# Dimensionality Reduction Checks {.appendix #apx-dimreduc}
To validate the cluster structure, we examined whether the three-cluster solution was recoverable using alternative dimensionality reduction methods: PCA (@fig-pca-check), t-SNE (@fig-tsne-check), and UMAP (@fig-umap-check). PCA loadings were also inspected to characterize the underlying feature structure (@fig-pca-loadings).

Results were consistent across all three methods in one key respect: the Ineffective Strategies group was clearly and reliably separated from the two effective strategy groups, appearing as a distinct region in all three projections — the top-left in the t-SNE, the negative end of PC1 in the PCA, and the bottom-left in the UMAP. This convergent separation provides strong evidence that this group occupies a genuinely distinct region of the feature space. The boundary between the Mental Rotation and Other Effective Strategies (Flipping/Matching) groups was, however, substantially less clear. Green and blue points overlapped considerably across all three visualizations, suggesting these two groups may represent ends of a continuum rather than fully discrete clusters — a finding consistent with the silhouette analysis favoring k = 2.

Inspection of the PCA loadings offered interpretive support for the feature set. PC1 loaded positively on Baseline Speed, Learning Rate, and Absolute Slope, and negatively on Instability, reflecting a general processing efficiency dimension: children who respond faster, improve more across trials, and show less trial-to-trial noise score higher on this component. PC2 loaded most strongly on Relative Slope and Absolute Slope — both positive — and negatively on Learning Rate, reflecting a rotation cost dimension that is largely independent of general speed. The fact that these two components map onto theoretically meaningful and distinct dimensions provides confidence that the clustering features are not redundant.
```{r}
#| label: fig-tsne-check
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**t-SNE visualization of strategy clusters.** The Ineffective Strategies group forms a distinct region in the top-left, while the Mental Rotation and Flipping/Matching groups show considerable overlap in the central and right regions, suggesting softer boundaries between the two effective strategy groups."

tsne_input <- analysis_data %>%
    select(run_id, Learning_Rate, Baseline_Speed, Rel_Slope, Abs_Slope, Instability) %>%
    distinct() %>% 
    na.omit()

set.seed(42) 
tsne_result <- Rtsne(
    as.matrix(tsne_input %>% select(-run_id)), 
    dims = 2, 
    perplexity = 30, 
    verbose = FALSE, 
    max_iter = 500
)

tsne_plot_data <- tsne_input %>%
    mutate(
        tSNE_1 = tsne_result$Y[,1],
        tSNE_2 = tsne_result$Y[,2]
    ) %>%
    left_join(analysis_data %>% select(run_id, Strategy_Label), by = "run_id")

ggplot(tsne_plot_data, aes(x = tSNE_1, y = tSNE_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "Dimension 1",
        y = "Dimension 2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-check
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**PCA projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group separates clearly along PC1, while the Mental Rotation and Flipping/Matching groups overlap along both components."

pca_result <- prcomp(tsne_input %>% select(-run_id), center = TRUE, scale. = TRUE)

pca_plot_data <- tsne_input %>%
    mutate(PC1 = pca_result$x[,1], PC2 = pca_result$x[,2]) %>%
    left_join(analysis_data %>% select(run_id, Strategy_Label), by = "run_id")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "PC1",
        y = "PC2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-loadings
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**PCA loadings for PC1 and PC2.** PC1 reflects a general processing efficiency dimension, loading positively on Baseline Speed, Learning Rate, and Absolute Slope, and negatively on Instability. PC2 reflects a rotation cost dimension, driven primarily by Relative and Absolute Slope."

loadings_df <- as.data.frame(pca_result$rotation) %>%
  rownames_to_column(var = "Angle_Condition")

loadings_long <- loadings_df %>%
  select(Angle_Condition, PC1, PC2) %>%
  pivot_longer(
    cols = c(PC1, PC2), 
    names_to = "Principal_Component", 
    values_to = "Loading"
  )

ggplot(loadings_long, aes(
    x = Loading, 
    y = reorder_within(Angle_Condition, Loading, Principal_Component), 
    fill = Loading > 0
  )) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Principal_Component, scales = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("TRUE" = pal$blue, "FALSE" = pal$red)) +
  labs(
    title    = NULL,
    subtitle = NULL,
    x = "Loading Weight",
    y = "Feature"
  ) +
  theme(
    #strip.text = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11),
    #panel.grid.minor = element_blank()
  )
```

```{r}
#| label: fig-umap-check
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "**UMAP projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group forms a distinct region in the bottom-left, while the Mental Rotation and Flipping/Matching groups show substantial overlap in the central and right regions."

set.seed(42)
umap_result <- umap(as.matrix(tsne_input %>% select(-run_id)))

umap_plot_data <- tsne_input %>%
    mutate(UMAP_1 = umap_result$layout[,1], UMAP_2 = umap_result$layout[,2]) %>%
    left_join(analysis_data %>% select(run_id, Strategy_Label), by = "run_id")

ggplot(umap_plot_data, aes(x = UMAP_1, y = UMAP_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) + 
    labs(x = "UMAP Dimension 1", y = "UMAP Dimension 2") +
    theme(legend.position = "bottom")
```

# Full 0–360° Angle Functions check {.appendix #apx-360-angle}
The primary analyses modeled angular disparity as the shortest path to upright, capped at 180°, on the assumption that participants take the most efficient route when rotating stimuli. To test this assumption, we re-ran the clustering using full 0–360° rotation angles and plotted RT and accuracy functions across the complete angular range (@fig-rt-angle-by-strategy-360, @fig-accuracy-angle-by-strategy-360). We verified that the angular disparity variable was correctly computed, confirming that full rotation angles were accurately recoded to their shortest-path equivalents (e.g., a 320° rotation has a disparity of 40°).

The results were broadly consistent with the disparity-based analyses and provided direct support for the shortest-path assumption. Both the Mental Rotation and Other Effective Strategies (Flipping/Matching) groups showed RT functions that increased from 20° to a peak around 200–240°, then declined as angles approached 320°, forming the expected near-symmetric inverted-U around 180°. This pattern is what would be predicted if participants always rotate via the shorter available path, and would not be expected under a fixed-direction rotation strategy.

The flipping signature observed in the Other Effective Strategies (Flipping/Matching) group was again evident in the full-angle data: their RT function rose more steeply than the Mental Rotation group and remained elevated through 240° before declining, consistent with a strategy that exploits reflective symmetry at high disparities. Accuracy for both effective strategy groups showed a corresponding U-shaped function, declining to a minimum around 160–200° before recovering at higher angles, with the Other Effective Strategies (Flipping/Matching) group maintaining the highest accuracy throughout. The Ineffective Strategies group again showed a broadly decreasing RT function across the full angular range, with accuracy remaining near chance at peak angles and showing only a modest recovery above 240°, further supporting the interpretation that these children are not engaging with the spatial structure of the items.

```{r}
#| label: robustness-check-all-angles
#| results: "hide"

# NOTE: all objects in this and subsequent 360° chunks use _360 suffixes to avoid overwriting main analysis objects. These chunks must remain the
# last substantive analysis chunks in the document.

mrot_rich_data_360 <- mrot_rt_trial %>%
  rename(ang_disparity = angle,
         angle = angle_full) %>%
  mutate(
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(ang_disparity)),
    trial_s  = as.numeric(scale(trial_number))
  )

contrasts(mrot_rich_data_360$stimulus) <- contr.sum(levels(mrot_rich_data_360$stimulus))

m_rich_360 <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | run_id) +
    (1 | item),
  data = mrot_rich_data_360,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

ran_effs_360 <- ranef(m_rich_360)$run_id %>%
  tibble::rownames_to_column("run_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

fix_int_360 <- fixef(m_rich_360)["(Intercept)"]
fix_ang_360 <- fixef(m_rich_360)["angle_s"]
fix_tri_360 <- fixef(m_rich_360)["trial_s"]

rich_features_raw_360 <- ran_effs_360 %>%
  mutate(
    Baseline_Speed = fix_int_360 + int_dev,
    Abs_Slope      = fix_ang_360 + slope_dev,
    Rel_Slope      = Abs_Slope / Baseline_Speed,
    Learning_Rate  = fix_tri_360 + trial_dev
  )

mrot_rich_data_360$resid <- residuals(m_rich_360)

instability_data_360 <- mrot_rich_data_360 %>%
  group_by(run_id) %>%
  summarise(
    Instability = sd(resid, na.rm = TRUE),
    .groups = "drop"
  )

rich_features_final_360 <- rich_features_raw_360 %>%
  left_join(instability_data_360, by = "run_id") %>%
  left_join(ability_run, by = "run_id") %>%
  mutate(Site = site_label) %>%
  filter(if_all(c(ability, Rel_Slope, Abs_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))

# Check: ang_disparity should never exceed 180
summary(mrot_rich_data_360$ang_disparity)

# Spot check a few items
mrot_rich_data_360 %>%
  select(item, item_original, angle, ang_disparity) %>%
  distinct() %>%
  arrange(item_original) %>%
  head(20)
  
```

```{r}
#| label: clustering-execution-360
#| include: false

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1],
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], 
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  
)

analysis_data_360 <- rich_features_final_360 %>%
  filter(if_all(c(ability, Baseline_Speed, Rel_Slope, Abs_Slope, 
                  Instability, Learning_Rate), is.finite))

global_input_360 <- analysis_data_360 %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()

set.seed(123)
k_global_360 <- kmeans(global_input_360, centers = 3, nstart = 25)
analysis_data_360$Global_Cluster <- as.factor(k_global_360$cluster)

cluster_stats_360 <- analysis_data_360 %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )

low_eff_id_360 <- cluster_stats_360 %>%
  arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)

effective_clusters_360 <- cluster_stats_360 %>%
  filter(Global_Cluster != low_eff_id_360) %>%
  arrange(mean_slope)

analytical_id_360 <- effective_clusters_360 %>% slice(1) %>% pull(Global_Cluster)
holistic_id_360   <- effective_clusters_360 %>% slice(2) %>% pull(Global_Cluster)

analysis_data_360 <- analysis_data_360 %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id_360    ~ "Ineffective Strategies",
    Global_Cluster == analytical_id_360 ~ "Other Effective Strategies (Flipping/Matching)",
    Global_Cluster == holistic_id_360   ~ "Mental Rotation"
  ))

ggplot(analysis_data_360, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Cluster Result: Ability vs. Relative Slope (full 0-360° angles)",
    x = "Ability", y = "Relative Slope"
  )
```

```{r}
#| label: fig-rt-angle-by-strategy-360
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Participant-weighted RT–angle functions by strategy profile (numbers indicate unique participants contributing at each angle; shaded bands show 95% CIs)."

rt_by_strategy_360 <- mrot_rich_data_360 %>%
  left_join(
    analysis_data_360 %>% select(run_id, Strategy_Label),
    by = "run_id"
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

rt_user_angle_360 <- rt_by_strategy_360 %>%
  group_by(Strategy_Label, run_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_angle_person_summary_360 <- rt_user_angle_360 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(run_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

missing_cols <- setdiff(levels(rt_angle_person_summary_360$Strategy_Label), names(my_colors))
stopifnot(length(missing_cols) == 0)

ggplot(rt_angle_person_summary_360, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_log_rt - 1.96 * se_log_rt,
        ymax = mean_log_rt + 1.96 * se_log_rt,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(
    values = my_colors, drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean log(RT) (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: fig-accuracy-angle-by-strategy-360
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Participant-weighted accuracy-angle functions by strategy profile."

acc_by_strategy_360 <- mrot_rich_data_360 %>%
  left_join(
    analysis_data_360 %>% select(run_id, Strategy_Label),
    by = "run_id"
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_user_angle_360 <- acc_by_strategy_360 %>%
  group_by(Strategy_Label, run_id, angle) %>%
  summarise(user_mean_acc = mean(correct, na.rm = TRUE), .groups = "drop")

acc_summary_360 <- acc_user_angle_360 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users  = n_distinct(run_id),
    mean_acc = mean(user_mean_acc, na.rm = TRUE),
    se_acc   = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_users),
    .groups  = "drop"
  )

ggplot(acc_summary_360, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc,
        ymax = mean_acc + 1.96 * se_acc,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(
    values = my_colors, drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_fill_manual(
    values = my_colors, drop = FALSE,
    labels = function(x) stringr::str_wrap(x, width = 25)
  ) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean Accuracy (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

<!--
```{r}
#| label: survey-gender
#| include: false
#| eval: false

survey_gender_wide <- survey_gender |>
  filter(variable %in% c("ChildSexBirth", "ChildGenderCurrent")) |>
  select(child_id, variable, string_response) |>
  tidyr::pivot_wider(names_from = variable, values_from = string_response)

survey_gender |>
  filter(variable %in% c("ChildSexBirth", "ChildGenderCurrent")) |>
  dplyr::summarise(n = dplyr::n(), .by = c(child_id, variable)) |>
  dplyr::filter(n > 1L) # several children with 2 responses

# are there children appearing more than once for the same variable
survey_gender |>
  filter(variable %in% c("ChildSexBirth", "ChildGenderCurrent")) |>
  select(child_id, variable, string_response, site, dataset) |>
  group_by(child_id, variable) |>
  filter(n() > 1) |>
  arrange(child_id, variable)

survey_gender_wide <- survey_gender |>
  filter(variable %in% c("ChildSexBirth", "ChildGenderCurrent")) |>
  select(child_id, variable, string_response) |>
  group_by(child_id, variable) |>
  # keep non-NA if available, otherwise NA
  summarise(string_response = na.omit(string_response) |> unique() |> 
              (\(x) if(length(x) == 1) x else NA_character_)(),
            .groups = "drop") |>
  tidyr::pivot_wider(names_from = variable, values_from = string_response)

# how many kids have each variable
survey_gender_wide |> 
  summarise(
    n_sex = sum(!is.na(ChildSexBirth)),
    n_gender = sum(!is.na(ChildGenderCurrent))
  )

# distributions
table(survey_gender_wide$ChildSexBirth, useNA = "always")
table(survey_gender_wide$ChildGenderCurrent, useNA = "always")

# do any child_ids from survey match user_ids in your main data?
sum(survey_gender_wide$child_id %in% mrot_rt_trial_no_retest$user_id)

mrot_rt_trial_no_retest |> 
  distinct(user_id) |> 
  nrow()

mrot_rt_trial_no_retest |>
  left_join(survey_gender_wide |> select(child_id, ChildSexBirth),
            by = c("user_id" = "child_id")) |>
  distinct(user_id, site, ChildSexBirth) |>
  group_by(site) |>
  summarise(
    n_total = n(),
    n_with_sex = sum(!is.na(ChildSexBirth)),
    pct = round(100 * n_with_sex / n_total, 1)
  )
```

```{r}
#| label: gender-ca
#| include: false
#| eval: false

demographics_ca <- readxl::read_excel(here::here("demographics_data", "LEVANTE Demographics Western University (By Child).xlsx"))

# check dimensions and first few column names
dim(demographics_ca)
colnames(demographics_ca)

# look at the first few rows
head(demographics_ca, 3)

# check unique values across all Q columns
demographics_ca |>
  select(starts_with("Q")) |>
  summarise(across(everything(), \(x) paste(unique(na.omit(x)), collapse = " | "))) |>
  tidyr::pivot_longer(everything(), names_to = "column", values_to = "unique_values") |>
  filter(str_detect(unique_values, regex("male|female|boy|girl|man|woman", ignore_case = TRUE)))

# extract q3 and join
demographics_ca_gender <- demographics_ca |>
  select(child_id = `Child ID`, sex_birth_ca = `Q3...20`) |>
  filter(!is.na(sex_birth_ca))

# check distribution
table(demographics_ca_gender$sex_birth_ca, useNA = "always")

# check ID overlap with your main data
sum(demographics_ca_gender$child_id %in% mrot_rt_trial_no_retest$user_id)

demographics_ca_raw <- readxl::read_excel(
  here::here("demographics_data", "LEVANTE Demographics Western University (By Child).xlsx"),
  n_max = 2 
)
# glimpse(demographics_ca_raw) # no child data (only caregiver)

```
####################### ROBUSTNESS CHECK - BOOTSTRAPPING ##########################
```{r}
#| label: cluster-stability-bootstrap
#| include: false
#| eval: false

vars   <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")
k      <- 3
B      <- 500
nstart <- 25
set.seed(123)

df_full <- analysis_data %>%
  select(user_id, Strategy_Label, all_of(vars)) %>%
  distinct(user_id, .keep_all = TRUE) %>%
  filter(if_all(all_of(vars), is.finite)) %>%
  mutate(
    Strategy_Label = factor(
      Strategy_Label,
      levels = names(my_colors)   
    )
  )

stopifnot(nlevels(df_full$Strategy_Label) == 3)
stopifnot(n_distinct(df_full$user_id) == nrow(df_full))

ref_levels <- levels(df_full$Strategy_Label)
y_ref      <- df_full$Strategy_Label
ids        <- df_full$user_id
n          <- nrow(df_full)

# scaling: lock to full-sample mean/sd so bootstrap runs are comparable
mu  <- df_full %>% summarise(across(all_of(vars), \(x) mean(x, na.rm = TRUE))) %>% as.list()
sdv <- df_full %>% summarise(across(all_of(vars), \(x) sd(x,   na.rm = TRUE))) %>% as.list()

X_full <- df_full %>%
  mutate(across(all_of(vars), ~ (. - mu[[cur_column()]]) / sdv[[cur_column()]])) %>%
  select(all_of(vars)) %>%
  as.matrix()

# helpers
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

jaccard <- function(set_a, set_b) {
  inter <- length(intersect(set_a, set_b))
  uni   <- length(union(set_a, set_b))
  if (uni == 0) NA_real_ else inter / uni
}

all_perms_3 <- list(
  c(1,2,3), c(1,3,2),
  c(2,1,3), c(2,3,1),
  c(3,1,2), c(3,2,1)
)

# Match bootstrap cluster IDs to reference labels by maximum membership overlap
best_perm_by_overlap <- function(cl_hat, y_ref, ref_levels, all_perms_3) {
  tab    <- table(cl_hat, factor(y_ref, levels = ref_levels))
  scores <- sapply(all_perms_3, function(p) sum(tab[cbind(1:3, p)]))
  all_perms_3[[which.max(scores)]]
}

# reference k-means (reproduces clustering-execution exactly)
# UPDATED: label assignment now mirrors the two-step logic in clustering-execution:
#   step 1 — lowest mean ability  → "Ineffective Strategies"
#   step 2 — of the remainder, flatter Rel_Slope → "Other Effective...", 
#             steeper → "Mental Rotation"
# We use this to build ref_centers aligned to ref_levels.

# need ability for label assignment (not a clustering var, so grab separately)
ability_lookup <- df_full %>%
  left_join(analysis_data %>% distinct(user_id, ability), by = "user_id")

set.seed(123)
km_ref <- kmeans(X_full, centers = k, nstart = nstart)
cl_ref <- km_ref$cluster

# replicate the production label-assignment rule on the reference run
cluster_stats_ref <- tibble(
  cluster      = 1:k,
  mean_ability = tapply(ability_lookup$ability, cl_ref, mean, na.rm = TRUE),
  mean_rel_slope = tapply(df_full$Rel_Slope,   cl_ref, mean, na.rm = TRUE)
)

low_eff_id_ref  <- cluster_stats_ref %>% arrange(mean_ability) %>% slice(1) %>% pull(cluster)
effective_ref   <- cluster_stats_ref %>% filter(cluster != low_eff_id_ref) %>% arrange(mean_rel_slope)
analytical_id_ref <- effective_ref %>% slice(1) %>% pull(cluster)
holistic_id_ref   <- effective_ref %>% slice(2) %>% pull(cluster)

# map cluster integer to ref_levels index (used to order ref_centers rows)
label_order_ref <- integer(k)
label_order_ref[low_eff_id_ref]   <- which(ref_levels == "Ineffective Strategies")
label_order_ref[analytical_id_ref] <- which(ref_levels == "Other Effective Strategies (Flipping/Matching)")
label_order_ref[holistic_id_ref]   <- which(ref_levels == "Mental Rotation")

# ref_centers[i, ] corresponds to ref_levels[i]
ref_centers <- km_ref$centers[order(label_order_ref), , drop = FALSE]

# bootstrap loop
res <- map_dfr(1:B, function(b) {

  set.seed(1000 + b)

  idx        <- sample.int(n, size = n, replace = TRUE)
  idx_unique <- unique(idx)

  tab_ref_unique <- table(factor(y_ref[idx_unique], levels = ref_levels))
  min_ref_unique <- min(tab_ref_unique)
  min_ref_prop   <- min_ref_unique / length(idx_unique)
  n_unique       <- length(idx_unique)

  X_b <- X_full[idx, , drop = FALSE]
  km  <- kmeans(X_b, centers = k, nstart = nstart)

  # assign all full-sample participants to nearest bootstrap center
  centers_b <- km$centers
  dists <- sapply(1:k, function(j) {
    rowSums((X_full - matrix(centers_b[j, ],
                             nrow = nrow(X_full), ncol = ncol(X_full),
                             byrow = TRUE))^2)
  })
  cl_hat <- max.col(-dists)

  # resolve label switching via maximum overlap with reference labels
  perm  <- best_perm_by_overlap(cl_hat, y_ref, ref_levels, all_perms_3)
  y_hat <- factor(ref_levels[perm[cl_hat]], levels = ref_levels)

  ari <- ARI(y_ref, y_hat)

  # per-cluster Jaccard
  jac <- map_dbl(ref_levels, function(lab) {
    jaccard(ids[y_ref == lab], ids[y_hat == lab])
  })
  names(jac) <- paste0("J_", make.names(ref_levels))

  # mean centroid displacement vs reference
  centers_b_aligned <- centers_b[perm, , drop = FALSE]
  center_shift      <- mean(sqrt(rowSums((centers_b_aligned - ref_centers)^2)))

  tibble(
    replicate      = b,
    ARI            = ari,
    n_unique       = n_unique,
    min_ref_unique = min_ref_unique,
    min_ref_prop   = min_ref_prop,
    center_shift   = center_shift,
    !!!as.list(jac)
  )
})

# summaries
stability_summary <- res %>%
  summarise(
    B                   = n(),
    ARI_median          = median(ARI, na.rm = TRUE),
    ARI_IQR             = IQR(ARI, na.rm = TRUE),
    center_shift_median = median(center_shift, na.rm = TRUE),
    center_shift_IQR    = IQR(center_shift, na.rm = TRUE),
    across(starts_with("J_"),
           list(median = ~ median(.x, na.rm = TRUE),
                IQR    = ~ IQR(.x, na.rm = TRUE)))
  )

stability_summary

ari_tail <- res %>%
  summarise(
    pct_ARI_below_0.70 = mean(ARI < 0.70, na.rm = TRUE) * 100,
    pct_ARI_below_0.60 = mean(ARI < 0.60, na.rm = TRUE) * 100,
    min_ARI            = min(ARI, na.rm = TRUE)
  )

ari_tail
```

```{r}
#| label: cluster-stability-reporting
#| include: false
#| eval: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)

ari_med    <- stability_summary$ARI_median
ari_iqr    <- stability_summary$ARI_IQR
pct_ari_70 <- ari_tail$pct_ARI_below_0.70
pct_ari_60 <- ari_tail$pct_ARI_below_0.60
min_ari    <- ari_tail$min_ARI

# UPDATED: pull Jaccard stats using new label names
j_med <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_median")]]
})
j_iqr <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_IQR")]]
})
names(j_med) <- ref_levels
names(j_iqr) <- ref_levels

j_ineff_med  <- j_med["Ineffective Strategies"]
j_ineff_iqr  <- j_iqr["Ineffective Strategies"]
j_other_med  <- j_med["Other Effective Strategies (Flipping/Matching)"]
j_other_iqr  <- j_iqr["Other Effective Strategies (Flipping/Matching)"]
j_mrot_med   <- j_med["Mental Rotation"]
j_mrot_iqr   <- j_iqr["Mental Rotation"]

rho_shift <- suppressWarnings(
  cor(res$ARI, res$center_shift, method = "spearman", use = "complete.obs")
)

# ARI distribution across bootstrap resamples
ggplot(res, aes(x = ARI)) +
  geom_histogram(bins = 40, fill = "#3498db", alpha = 0.8) +
  geom_vline(xintercept = median(res$ARI, na.rm = TRUE), 
             linetype = "dashed", colour = "black") +
  labs(x = "ARI (vs reference clustering)", y = "Count",
       title = paste0("Bootstrap stability (B = ", nrow(res), ")"),
       subtitle = paste0("Median ARI = ", round(median(res$ARI, na.rm = TRUE), 3)))

# Per-cluster Jaccard
res %>%
  select(replicate, starts_with("J_")) %>%
  pivot_longer(-replicate, names_to = "Cluster", values_to = "Jaccard") %>%
  mutate(Cluster = str_remove(Cluster, "^J_")) %>%
  ggplot(aes(x = Jaccard, fill = Cluster)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  facet_wrap(~ Cluster, ncol = 1) +
  labs(x = "Jaccard index", y = "Count") +
  theme(legend.position = "none")

```
# checks
```{r}
#| eval: false

analysis_data %>%
  count(Strategy_Label)

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(across(all_of(c("Baseline_Speed", "Rel_Slope", "Abs_Slope", 
                             "Instability", "Learning_Rate")), 
                   ~ round(mean(.x, na.rm = TRUE), 3)))

analysis_data %>%
  select(Strategy_Label, Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  pivot_longer(-Strategy_Label, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Strategy_Label, y = value, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~ variable, scales = "free_y") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

####################### ROBUSTNESS CHECK - WITHIN PERSON STRATEGY SWITCHING (DORIA'S ANALYSIS) ##########################

### Strategy Stability Analysis (Split-Half Reliability)

To investigate whether children consistently use a single cognitive strategy throughout the task or switch strategies over time (e.g., discovering a shortcut), we performed a **Split-Half Reliability Analysis**.

**Methodology:**
We partitioned each participant's data into two sequential halves: the first 50% of trials (Early Phase) and the last 50% of trials (Late Phase). For each half, we estimated a "Raw Slope" using a simple linear regression of reaction time on angle (`log_rt ~ angle`). This slope serves as a proxy for the participant's rotation strategy during that specific period. 

We then correlated the *First Half Slope* with the *Second Half Slope* to quantify stability. A high correlation ($r \approx 1$) would indicate a fixed strategy, while a low correlation ($r \approx 0$) suggests dynamic behavior or strategy switching.

```{r}
#| label: fig-strategy-switching
#| fig-cap: "Strategy Stability: Comparison of Rotation Slope in First Half vs. Second Half of trials. Points below the diagonal suggest a switch to a flatter strategy (e.g., discovering the shortcut)."
#| eval: false

# 1. Split Data into Halves
# We determine the midpoint trial for each user
split_data <- mrot_rich_data %>%
  group_by(user_id) %>%
  mutate(
    Midpoint = median(trial_number, na.rm = TRUE),
    Half = ifelse(trial_number <= Midpoint, "First_Half", "Second_Half")
  ) %>%
  ungroup() %>%
  # Filter out bad data points just to be safe
  filter(is.finite(log_rt), is.finite(angle))

# 2. Calculate Simple Slope for Each Half (Using simple lm per subject)
# This is a fast proxy for the complex mixed-model slope
slope_stability <- split_data %>%
  group_by(user_id, Half) %>%
  summarise(
    # We only care about the slope (coefficient for angle)
    # We use a safe lm() call wrapped in tryCatch to handle kids with too few trials
    Raw_Slope = tryCatch(
      #coef(lm(log_rt ~ angle, data = pick(everything())))["angle"], 
      # Use distance from 0 so the slope represents 'cost of rotation'
  coef(lm(log_rt ~ pmin(angle, 360 - angle), data = pick(everything())))[2],
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  # Reshape to Wide Format (Columns: Slope_First, Slope_Second)
  pivot_wider(names_from = Half, values_from = Raw_Slope, names_prefix = "Slope_") %>%
  # Join with our Final Strategy Labels to color-code
  left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id") %>%
  filter(!is.na(Strategy_Label), !is.na(Slope_First_Half), !is.na(Slope_Second_Half))

# 3. Calculate Correlation (Quantify Stability)
correlation <- cor(slope_stability$Slope_First_Half, slope_stability$Slope_Second_Half, use = "complete.obs")

# 4. Visualization
ggplot(slope_stability, aes(x = Slope_First_Half, y = Slope_Second_Half, color = Strategy_Label)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") + # The "Stability Line"
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Strategy Switching Check (Split-Half Reliability)",
    subtitle = paste0("Correlation between halves: r = ", round(correlation, 2), 
                      ". Points below diagonal = Strategy became flatter (faster)."),
    x = "Slope (First Half of Trials)",
    y = "Slope (Second Half of Trials)"
  ) +
  coord_fixed() # Keeps the aspect ratio square so the diagonal is 45 degrees
```

```{r}
#| eval: false

# Doria's original code for examining up to 360 
mrot_rich_data <- mrot_rt_trial %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    
    # 1. Extract the base angle (e.g., 40)
    base_angle = as.integer(str_extract(item, "\\d+")), 
    
    # 2. Detect if the original item was a "Negative/Counter-Clockwise" rotation
    # Looking for 'rn' prefix in item_original (e.g., rn040)
    is_ccw = str_detect(item_original, "^rn"),
    
    # 3. RECONSTRUCT the full 0-360 angle
    angle = ifelse(is_ccw, 360 - base_angle, base_angle),
    
    # 4. Calculate shortest distance to center (disparity) for the model
    # This ensures the model still measures 'Rotation Cost' correctly
    ang_disparity = pmin(angle, 360 - angle),
    
    stimulus = as.factor(stimulus),
    # SCALE THE DISPARITY, NOT THE RAW ANGLE
    angle_s  = as.numeric(scale(ang_disparity)), 
    trial_s  = as.numeric(scale(trial_number))
  )

#  Issues: the above code is only only catches rabbit counter-clockwise items (rn040, rn080 etc.), and misses duck and shape items. It is also not necessary to deconstruct item_original as the info. we need is already there. item_original already contains the full angle directly (e.g., an11240 = 240).

# Fionnuala checking
# Check if column exists
"item_original" %in% names(mrot_rt_trial)

# If it does, see what's in it
mrot_rt_trial$item_original |> unique() |> sort()

# problem
# str_detect(item_original, "^rn") is wrong — it only catches rabbit counter-clockwise, missing dn, an, bn etc.

#diagnose
# Step 1: Check is_ccw detection by stimulus
mrot_rich_data %>%
  distinct(stimulus, item_original, is_ccw, base_angle, angle) %>%
  arrange(stimulus, item_original) %>%
  print(n = 50)

mrot_rich_data %>%
  count(stimulus, angle) %>%
  pivot_wider(names_from = stimulus, values_from = n) %>%
  arrange(angle) %>%
  print(n = 20)
```
-->