---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "A"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "B"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "C"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "D"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "E"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "F"
    orcid: 
    affiliations:
      - ref: stanford
  - name: "G"
    orcid: 
    affiliations:
      - ref: stanford
    corresponding: true
    email: x@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["cognitive development", "mental rotation", "spatial reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    extra_dependencies: ["booktabs", "longtable"]
    # fig-pos: H        
    # tbl-pos: H  
    

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

From fitting puzzle pieces together at age three to navigating complex diagrams in school, spatial reasoning underpins a broad array of cognitive achievements — and children differ substantially in how these skills develop. Such variability is detectable early, across a wide range of spatial tasks in primary school-aged children [@hodgkiss2021developmental; @lauer2019development], and its consequences are far-reaching: spatial ability predicts outcomes spanning creative domains such as music [@padulo2020effects; @hetland2000learning] and the visual arts [@kell2013creativity; @winner2013relationship], mathematical achievement [@gunderson2012relation], science learning [@gilligan2020first], and long-term educational and occupational outcomes in STEM fields [@wai2009spatial]. Understanding what drives this variability is therefore a question of both theoretical and practical importance. The present study takes a step toward that goal by examining children's performance on a mental rotation task across three culturally diverse countries, asking both what is consistent across children and what varies.

Spatial reasoning encompasses abilities for representing, transforming, and reasoning about visuospatial information [@lohman1996spatial]. Rather than a single unified capacity, it is better understood as a family of related but distinguishable skills, including mental rotation, spatial visualization, perspective-taking, and navigation, that differ in the type of spatial information involved and the cognitive operations required [@uttal2013malleability; newcombe2018three]. Skills are organized along two dimensions: whether the spatial relations in question are intrinsic to an object or extrinsic between objects, and whether the task requires dynamic mental transformation or only static representation [@uttal2013malleability]. Mental rotation — the focus of the present study — falls in the intrinsic-dynamic quadrant of this framework, involving the mental transformation of an object's orientation in space.

Mental rotation is the ability to mentally transform an object by rotating it in space to predict how it would appear from a different orientation or viewpoint [@shepard1971mental]. It is one of the most studied spatial skills in cognitive and developmental psychology, in part because it yields one of the most robust and replicable findings in the field: response times increase approximately linearly with the angular disparity between two objects, suggesting that mental rotation operates as a spatially continuous process rather than a discrete rule-based one [@shepard1971mental; @cooper1973chronometric]. We return to this finding in detail below.

Mental rotation ability develops across a protracted period beginning in infancy. Precursors of successful rotation are detectable in the first year of life [@frick2014development; moore2008mental], and performance on explicit rotation tasks improves substantially between ages three and twelve [@frick2013development; @marmor1975development], with continued refinement into adolescence. This extended developmental trajectory makes the primary school years, ages five to twelve, a particularly informative window for studying how mental rotation skills are acquired and how they vary across children.

The standard mental rotation task, as developed by @shepard1971mental, presents participants with pairs of objects at varying angular disparities and asks them to judge whether the objects are identical or mirror-reversed. Response time and accuracy are the key dependent variables. While this paradigm has generated decades of remarkably consistent findings, an important interpretive caveat applies: performance on the task may not reflect a single, unitary process of mental rotation. Research with adults suggests that solvers draw on a range of strategies, spanning a holistic-analytic continuum [@gluck2003spatial]. At the holistic end, the stimulus is transformed as an integrated mental representation — the so-called "pure rotation" approach. At the analytic end, solvers decompose the object into sub-components and verify features sequentially, a more piecemeal strategy that places greater demands on working memory [@just1985cognitive; @khooshabeh2013individual]. Strategy selection is not fixed: individuals differ in their preferred approach, and the same individual may shift strategies depending on task demands such as the complexity of the stimulus or the degree of angular disparity [@nazareth2019strategy]. Importantly, far less is known about which strategies children employ when solving mental rotation problems, and whether strategy use varies with age, ability, or cultural context. This question motivates the exploratory component of the present study, which we describe below.

The vast majority of developmental research on mental rotation has been conducted with children from Western, educated populations [@davis2019spatial; @peters2006mental], raising questions about whether findings about its developmental trajectory, psychometric properties, and underlying cognitive processes generalise beyond the populations in which they were established.

The limited cross-cultural evidence that does exist offers a mixed picture. A large-scale study of 3,367 students across Canada, Germany, and Japan found that a male performance advantage on mental rotation was consistent across all three nations, suggesting some cross-cultural robustness in group-level patterns [@peters2006mental]. However, cross-cultural comparisons of children specifically are rare. One notable exception compared children in the Ivory Coast and Switzerland and found that while both groups showed evidence of mental rotation, Swiss children demonstrated substantially higher speed and accuracy — and crucially, age predicted performance in the Swiss group but not in the Ivorean group [@nunez1998mental]. More recent work in non-WEIRD populations has shown that environmental and cultural factors — including access to formal schooling, freedom to explore the physical environment, and culturally-specific spatial practices — can substantially shape both the level and the pattern of spatial performance in children [@davis2019spatial; @davis2021cultural].

Taken together, this evidence highlights two distinct but related gaps. First, we have limited data on whether mental rotation tasks developed and validated in Western samples are measuring the same underlying construct when administered to children in different cultural contexts — a question of measurement invariance rather than simply mean-level differences. Second, even where cross-cultural comparisons have been made, they have rarely examined whether the core cognitive signatures of mental rotation replicate across diverse populations of children. The linear relationship between reaction time and angular disparity — one of the most robust findings in cognitive psychology — has been demonstrated repeatedly in Western adult samples, but its replication in children across culturally diverse sites remains largely untested. Beyond these two gaps, relatively little is known about the cognitive strategies children employ when solving mental rotation problems, and whether these vary across cultural contexts — a question we explore tentatively here. The present study addresses these gaps.

Returning to the question that opened this paper — why do children vary in spatial reasoning, and what does that variation reflect? Cross-cultural data offer a potentially powerful way to answer this. But before individual differences observed across cultural contexts can be meaningfully interpreted, two preconditions must be met.

The first is measurement invariance. If a task is not measuring the same underlying construct across different cultural contexts, then findings about its developmental trajectory — even within a single site — cannot be confidently attributed to the development of spatial reasoning per se. They may instead reflect something more specific to the cultural and educational context in which the task was developed. Establishing that a task achieves measurement invariance across diverse sites is therefore crucial evidence that the construct being measured is genuinely spatial in nature, and that developmental conclusions drawn from it are likely to generalize beyond any single cultural setting.

The second precondition is process-level validity. Even if a task achieves measurement invariance across sites, we need evidence that it is sensitive to the spatial properties of the stimuli in the way theory predicts. The key benchmark here is the angular disparity effect (described above), first demonstrated by @shepard1971mental: as the angular difference between two objects increases, so too does the time required to judge whether they match. If children across three culturally diverse sites show this same systematic pattern, this constitutes strong evidence that performance is being driven by the spatial demands of the task, and that this signature is not specific to the Western samples in which it was originally established. This does not require that all children are using identical strategies — what matters is that the angular disparity effect emerges consistently, indicating that children are engaging with the spatial structure of the problem. Only once these two preconditions are established does it become meaningful to ask what the sources of individual variability actually are — including, tentatively, whether children are employing different cognitive strategies.

The present study uses data from the Learning Variability Network Exchange (LEVANTE), an international research collaboration designed to capture how children's learning and development vary across ages and cultural contexts. LEVANTE's central aim is to build a large, openly shared dataset of children's cognitive development across diverse settings, with a particular focus on individual variability — what differs across children, and why. Data were collected across three geographically and culturally diverse pilot sites: Canada, Germany, and Colombia. Here we use data from a single task in the LEVANTE battery: a chronometric mental rotation task developed using item response theory, which allows directly comparable ability estimates to be derived across children who completed different subsets of items [@Kachergis2025LEVANTEcoretasks].

Against this backdrop, the present study pursues three aims. First, we examine whether the mental rotation task achieves measurement invariance across the three LEVANTE sites, establishing that it captures a consistent underlying construct across cultural contexts. Second, we test whether the canonical Shepard-Metzler angular disparity effect replicates across all three sites in children aged 5 to 12 — extending this foundational finding to a culturally diverse developmental sample. Third, in an exploratory analysis, we ask what cognitive strategies children appear to employ when solving mental rotation problems, and whether strategy use varies with age or angular disparity. Taken together, these analyses move from establishing what is consistent across children and contexts, toward a first look at what varies — and how that variation might be characterized.

# Methods

## Participants

Data for the present study were drawn from three pilot sites participating in LEVANTE: uniandes-co, mpieva-de, and western-ca. Collectively, these sites supported task localisation and technical implementation, iterative piloting, and deployment in varied testing contexts. The resulting pilot datasets enabled preliminary evaluation of construct validity, measurement invariance, test–retest reliability, and calibration for computer-adaptive administration, as well as descriptive assessment of age-related differences across the targeted constructs (refer to @Kachergis2025LEVANTEcoretasks for a full description).

### uniandes-co

The Universidad de Los Andes team carried out school-based data collection in Colombia, working with four schools in Bogotá and two public schools in the rural departments of Caquetá and Boyacá. Schools were selected pragmatically from institutions that had previously collaborated with the university. In Bogotá, three schools were affiliated with two family compensation funds and one was operated by a charter-school organisation. Across locations, participating schools largely served children from low-income and lower-middle-income backgrounds. Testing was conducted on tablets provided by the research team using a group-administration format, with group sizes adjusted by age (larger groups for older children; smaller groups for younger children).

### mpieva-de

At the mpieva-de site, the Max Planck Institute for Evolutionary Anthropology implemented a remote, family-based assessment protocol. Families were recruited via an existing registry of Leipzig residents and were contacted by telephone to invite participation. Families who enrolled were given login details and completed the assigned LEVANTE tasks at home using their own computer or tablet.

### western-ca

The Western University (Ontario) team recruited children from the surrounding community, primarily via online outreach (e.g., Facebook advertisements and posts in community groups), complemented by physical flyers and announcements distributed through local school newsletters. Families self-enrolled by registering their interest in a participant database. Children were assessed individually by a researcher in a dedicated on-campus testing space.

## Measure

The LEVANTE mental rotation task is adapted from @frick2013development. On each trial, children view a target image (e.g., a rabbit) and then choose between two alternatives (e.g., two rabbit silhouettes): one alternative is the same object shown at a different orientation, whereas the other is a non-matching foil. Children indicate which option corresponds to the target, and we record both accuracy and response time. Angular disparity is manipulated across trials (e.g., 40°, 80°, 120°, 160°), producing systematic effects on performance. Stimuli include 2D animal figures (ducks and rabbits) and more challenging 3D geometric shapes drawn from the classic Shepard and Metzler (1971) set.

```{r}
#| label: fig-mrot-example-items
#| fig-cap: "Example items for 2D and 3D rotation"
#| out-width: "100%"
#| fig-align: "center"

knitr::include_graphics("mrot_item_examples.pdf")
```

All LEVANTE tasks including mental rotation were originally developed in English and later adapted for Spanish and German through a structured internationalisation workflow carried out with collaborators at the uniandes-co and mpieva-de sites, respectively. We first produced draft translations of all verbal content—including instructions, item text, and recurring terminology (e.g., encouragement prompts)—using DeepL. These drafts were then refined by a professional translator to ensure grammatical accuracy and natural phrasing. Next, a site researcher who was a native speaker of the relevant language and dialect reviewed the materials to confirm clarity and suitability for the specific task context. To check for semantic and conceptual alignment with the English originals, we used AI-assisted back-translation to identify wording that required further revision and native-speaker review.

The mental rotation task has already been implemented in a computer-adaptive task (CAT) format. In the adaptive version, participant ability is summarized by a latent proficiency estimate (theta) that is re-estimated after each response. The updated theta then guides selection of the next item so that it is appropriately challenging for the individual. This trial-by-trial targeting typically makes the task more engaging and increases measurement efficiency, enabling reliable ability estimation with fewer administered items. Refer to @Kachergis2025LEVANTEcoretasks for more information on the CAT versions of all LEVANTE tasks.

# Results

## Cross-Cultural Measurement Invariance of the Mental Rotation Task

We first establish whether the mental rotation task measures the same underlying construct across all three sites. We addressed this using multi-group IRT to evaluate measurement invariance [@putnick2016measurement]. Multi-group models supported scalar invariance under a 2PL model, indicating that item difficulty is aligned across sites and that children in each country are being placed on a comparable latent scale. Practically, this means the groups share the same relative ordering of item difficulty — items with larger rotation angles are consistently more difficult across all three countries (see @tbl-mrot-2pl-scalar). The core structure of mental rotation performance is thus largely shared across the cultural settings represented here: the same item features drive difficulty in a similar way regardless of site. With this foundation established, we turn to the second precondition: whether the task shows the process-level sensitivity to angular disparity that theory predicts. 

It should be noted that while we use psychometric modelling to place children’s performance on a common scale, the LEVANTE project is not designed to estimate absolute differences between sites. In practice, LEVANTE does not support site-level performance comparisons: sites vary substantially in sampling, recruitment, and test administration, so any apparent cross-site differences in raw performance are heavily confounded. Accordingly, although we report descriptive patterns that may look like absolute site differences, we emphasize that these should not be interpreted as meaningful.

```{r}
#| include: false

# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(umap)
```

```{r}
#| label: setup
#| include: false
#| cache: false

source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)

# coefficients
all_coefs <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs/all_coefs.rds"))

```

```{r}
#| label: setup-palettes
#| include: false

site_pal_labelled <- site_pal
names(site_pal_labelled) <- names(site_labels)  
# site_labels is named by readable label, values are raw names
# so names(site_labels) gives us "uniandes-co", "mpieva-de", "western-ca"
```

```{r}
#| label: irt-anova
#| include: false

# IRT model output - import object
github_dir <- file.path(path.expand("~"), "Documents", "GitHub")
task_anova_path <- file.path(github_dir, "levante-pilots", "02_scoring_outputs", "task_anova.rds")

stopifnot(file.exists(task_anova_path))
task_anova <- readRDS(task_anova_path)
#colnames(task_anova)
#table(task_anova$itemtype)

# extract
mrot_2pl_scalar <- task_anova %>%
  filter(task == "mrot", itemtype == "2pl", invariance == "scalar")
mrot_2pl_scalar
```

```{r}
#| label: tbl-mrot-2pl-scalar
#| tbl-cap: "Model fit indices for 2PL scalar invariance models. Note: 2PL = 2-parameter logistic model; scalar invariance indicates that item intercepts and loadings are constrained across groups."
#| echo: false

mrot_2pl_scalar %>%
  transmute(
    Task       = task,
    Factors    = nfact,
    Model      = itemtype,
    Invariance = invariance,
    logLik     = logLik,
    AIC        = AIC,
    BIC        = BIC,
    `# params` = df
  ) %>%
  gt() %>%
  # Ensures the gt title area is clear and doesn't overlap the Quarto caption
  tab_header(title = md("")) %>% 
  fmt_number(columns = c(logLik, AIC, BIC), decimals = 2) %>%
  fmt_number(columns = `# params`, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Task, Factors, Model, Invariance, `# params`)) %>%
  cols_align(align = "right",  columns = c(logLik, AIC, BIC)) %>%
  tab_options(
    #table.font.size = px(12), # Increased slightly for better legibility
    table.width = pct(100),   # Ensures it fills the page width nicely
    data_row.padding = px(3)   # Tighter row spacing often looks better for stats
  )
```

```{r}
#| label: fig-item-diff
#| fig-cap: "**Item Difficulty Parameters Across Sites (Configural Model).** Points represent individual items; close alignment along the diagonal indicates that item difficulty parameters are consistent across cultural contexts."
#| include: false
#| fig-width: 12
#| fig-height: 8

# Load configural parameters
mrot_configural <- all_coefs |>
  dplyr::filter(task == "mrot",
                itemtype == "2pl",
                invariance == "configural") |>
  dplyr::mutate(group = dplyr::recode(group,
                  "pilot_western_ca"  = "Canada",
                  "pilot_mpieva_de"   = "Germany",
                  "pilot_uniandes_co" = "Colombia"))

# pivot wide
mrot_wide_d <- mrot_configural |>
  dplyr::select(item, group, d) |>
  tidyr::pivot_wider(names_from = group, values_from = d) |>
  dplyr::mutate(item_label = item |>
                  stringr::str_remove("mrot_2d_") |>
                  stringr::str_remove("mrot_3d_") |>
                  stringr::str_replace("_", " ") |>
                  stringr::str_replace("(\\d+)$", "\\1°"))

plot_scatter <- function(data, x_var, y_var, x_lab, y_lab) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]],
                   label = item_label)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", colour = "grey50") +
    geom_point(size = 2.5, na.rm = TRUE) +
    ggrepel::geom_text_repel(size = 2.8,
                    max.overlaps = Inf,
                    box.padding = 0.5,
                    point.padding = 0.3,
                    na.rm = TRUE) +
    coord_equal(xlim = c(-2.5, 4), ylim = c(-2.5, 4)) +
    labs(x = x_lab, y = y_lab) +
    theme_bw()
    #theme(panel.grid.minor = element_blank())
}

p1 <- plot_scatter(mrot_wide_d, "Canada", "Germany",   "Canada", "Germany")
p2 <- plot_scatter(mrot_wide_d, "Canada", "Colombia",  "Canada", "Colombia")
p3 <- plot_scatter(mrot_wide_d, "Germany", "Colombia", "Germany", "Colombia")

(p1 | p2 | p3) +
  plot_annotation(
    title = "Item Difficulty Parameters Across Sites (Configural Model)",
    theme = theme(plot.title = element_text(size = 12))
  )

# checks
all_coefs |>
  dplyr::filter(task == "mrot", itemtype == "2pl", invariance == "configural") |>
  dplyr::count(subset, group)

# only 12 items appearing: multigroup IRT model was fitted only on the overlap items by design. The non-overlap items were either fitted in separate by-site models, or weren't included in the multigroup analysis at all.

# note to check this with Doria. 

```

@fig-dev-trends plots participants' mental rotation ability estimates (IRT theta) against age, with curves by site modelled using generalized additive models (GAMs). GAMs allow the developmental association with age to be non-linear, capturing periods of more rapid change and relative flattening without imposing a fixed form. Across all three sites, theta increases with age, indicating that the task captures developmental change consistently across cultural contexts. The dispersion of points within age bands underscores substantial individual variability in ability among same-aged children. Note that vertical offsets between sites reflect differences in sample composition and recruitment rather than true cross-cultural performance differences, and are not interpreted further here.

```{r}
#| label: fig-dev-trends
#| fig-cap: "**Developmental trajectories of mental rotation ability across sites.** Points represent individual children's IRT-estimated ability (θ); curves are GAM-smoothed age trends with 95% confidence intervals. Vertical offsets between sites should not be interpreted as performance differences; LEVANTE sites differ in sampling and recruitment in ways that heavily confound absolute comparisons."
#| fig-width: 8
#| fig-height: 6

mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!setNames(names(site_labels), 
                                                    unname(site_labels)),
                                  .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  scale_colour_manual(values = site_pal_labelled) +
  labs(
    x      = "Age (years)",
    y      = "Ability (theta)",
    colour = "Site"
  )
```

## Angular disparity and the RT–angle relationship

Our aim with this analysis was to test whether the canonical reaction time–rotation angle relationship first reported by @shepard1971mental replicates in our data. @fig-rt-angle-per-run plots mean reaction times (±95% confidence intervals) as a function of rotation angle, stratified by age group (5–7, 7–9, and 9–12 years) and site, with separate series by stimulus and item type (2D vs 3D). To show the underlying dispersion and sample density alongside the summary trends, we provide a second version of @fig-rt-angle-per-run in the Appendix that overlays the raw per-run data (using the same age bins and 95% CIs; see @fig-rt-angle_raw).

Across sites, mean RTs generally decrease with age, with the 9–12 group responding faster overall than the younger groups, particularly for the 2D animal stimuli (duck/rabbit). RTs also tend to increase with rotation angle (angular disparity), consistent with the classic RT–angle effect (@shepard1971mental). This angular-disparity effect is most pronounced for the 3D shape stimuli, which show slower overall RTs and steeper angle-related increases, whereas the 2D stimuli exhibit comparatively flatter RT–angle functions. Despite some differences in absolute RT levels between sites, the same qualitative pattern—faster responses with age and a stronger angular-disparity effect for 3D items—appears across all three sites, suggesting broadly comparable task behaviour across contexts.

```{r}
#| label: trial-level-rts
#| include: false

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%
  tidyr::unnest(data) %>%
  dplyr::mutate(
    correct = dplyr::case_when(
      is.logical(correct) ~ as.integer(correct),
      TRUE ~ as.integer(as.numeric(correct))
    ),
    rt_s   = as.numeric(rt_numeric) / 1000,
    log_rt = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))
```

```{r}
#| label: remove-retest-data
#| include: false

baseline_runs <- mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%  
  group_by(site, user_id) %>%
  arrange(age, run_id) %>%           
  slice(1) %>%
  ungroup() %>%
  select(site, user_id, run_id)

# keep only trials from those baseline runs
mrot_rt_trial_no_retest <- mrot_rt_trial %>%
  semi_join(baseline_runs, by = c("site", "user_id", "run_id"))

# check: each child should now have exactly 1 run (within site)
mrot_rt_trial_no_retest %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs)

# how many re-testers?
mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs) # 133
```

```{r}
#| label: fig-rt-angle-per-run
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 7, 9, 12)

mrot_items_rt_runs <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = dplyr::n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  ) + scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) 
p_mrot_rt_angle_age_1
```

## Within-Person Dynamics

Having established the expected angular-disparity effect across age groups and sites, we next examined how RTs evolved within individuals over the course of the task. Because item difficulty increased adaptively—such that higher-performing children were more likely to encounter the harder 3D items—we analysed trial-by-trial RT trajectories across the 80 trials to distinguish item-driven changes from more gradual time-on-task effects (e.g., fatigue). @fig-within-person-trials illustrates a clear 'jump' in RT about halfway through the task (trial 48), corresponding to the transition from 2D to 3D stimuli, which required significantly more processing time across all three site populations. We observed consistent response acceleration within both the 2D and 3D blocks, as children became faster with practice.

```{r}
#| label: fig-within-person-trials
#| fig-cap: "**Trial-level Reaction Time Dynamics (angle-residualised).** Mean residual log-RT across trials for each non-adaptive site ±SE. The dashed line marks the 2D→3D transition. Re-test and CAT data removed."
#| fig-width: 6
#| fig-height: 4

# get CAT data
cat_data <- read_rds(here("..", "levante-pilots", "02_scoring_outputs", "scores", "scores_combined.rds")) %>%
  group_by(user_id, item_task) %>%
  arrange(time_started) %>%
  mutate(run_number = row_number(),
         age_gap = age - age[1]) %>%
  ungroup() %>%
  filter(item_task == "mrot") %>%
  filter(run_number == 1 | (run_number == 2 & age_gap > 1/6)) %>%
  distinct(user_id, run_id, adaptive)

# prep: remove CAT data
trial_resid_data <- mrot_rt_trial_no_retest %>%
  left_join(cat_data, by = c("user_id", "run_id")) %>%
  filter(adaptive == FALSE) %>%
  filter(!is.na(log_rt), correct == 1) %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle    = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(!is.na(angle), !is.na(stimulus), !is.na(item_group))

# residualise log_rt on angle + stimulus + item_group
resid_model <- lm(log_rt ~ angle + stimulus + item_group, 
                  data = trial_resid_data)

trial_resid_data <- trial_resid_data %>%
  mutate(log_rt_resid = residuals(resid_model))

# add block variable
transition_trial <- 48 

trial_resid_data <- trial_resid_data %>%
  mutate(
    site_label = dplyr::recode(site, !!!site_labels),
    block      = if_else(trial_number < transition_trial, "2D", "3D")
  )

# recode site_pal names to match site_label values
names(site_pal_labelled) <- dplyr::recode(names(site_pal), !!!site_labels)

# plot
ggplot(trial_resid_data, 
       aes(x = trial_number, y = log_rt_resid, 
           color = site_label, fill = site_label)) +
  stat_summary(fun = mean, geom = "point", size = 1.2, alpha = 0.5) +
  geom_smooth(aes(group = interaction(site_label, block)),
              method = "loess", span = 0.75, se = TRUE, alpha = 0.2) +
  geom_vline(xintercept = transition_trial, linetype = "dashed",
             color = pal$grey, linewidth = 0.7) +
  annotate("text", x = transition_trial + 1, y = Inf,
           label = "Start of 3D block", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_colour_manual(values = site_pal_labelled) +
  scale_fill_manual(values = site_pal_labelled) +
  labs(
    x     = "Trial Number",
    y     = "Residual reaction time",
    color = "Site",
    fill  = "Site"
  ) +
  theme(legend.position = "bottom")
```
\clearpage
## Cognitive Strategies

To characterize individual differences in how children solved the mental-rotation task, we derived participant-level 'strategy features; from trial-level reaction times. Consistent with the generally log-normal distribution of reaction times, we modeled log-transformed RTs. While the classic @shepard1971mental finding describes an approximately linear relationship with raw RT, fitting the model to log(RT) provides more robust parameter estimates for defining individual strategy profiles in a developing population. We fit a mixed-effects model to log(RT) that captured stimulus and angle-dependent responding alongside within-run practice effects, extracted participant-specific deviations (random effects), and combined these model-derived quantities with RT variability and IRT-estimated ability before clustering participants using k-means (k = 3) on standardized features.

The extracted features were: (1) Baseline Speed, defined as the intercept from the mixed-effects model (interpretable as expected speed at 0° rotation); (2) Rotation Slope, defined as the “cost” of rotation (the slope of angle on log RT), estimated as the average slope across stimulus types; (3) Response Variability (RT SD), defined as the intra-individual standard deviation of log RT; and (4) IRT-estimated ability (theta).

```{r}
#| label: strategy-analysis-prep
#| results: "hide"

# ability
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# Prepare trial-level data for the rich model
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    angle    = as.integer(str_extract(item, "\\d+")),
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(angle)),
    trial_s  = as.numeric(scale(trial_number))
  )

# set sum-to-zero contrasts - by default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the grand mean slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# extract subject-level random effects
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# compute instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(site, user_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

# join
rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(instability_data, by = c("site", "user_id")) %>% 
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  # Filter to ensure all needed features
  filter(if_all(c(ability, Rel_Slope, Abs_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))
```

```{r}
#| label: clustering-execution
#| include: false

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1], # red
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], # blue
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  # green
)

analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, 
                  Instability, Learning_Rate), is.finite)) 

global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )

# FIX: as.character() on all pull() calls
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% 
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

effective_clusters <- cluster_stats %>%
  filter(as.character(Global_Cluster) != low_eff_id) %>%
  arrange(mean_slope) 

analytical_id <- effective_clusters %>%
  slice(1) %>% 
  pull(Global_Cluster) %>%
  as.character()

holistic_id <- effective_clusters %>%
  slice(2) %>% 
  pull(Global_Cluster) %>%
  as.character()

analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    as.character(Global_Cluster) == low_eff_id    ~ "Ineffective Strategies",
    as.character(Global_Cluster) == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    as.character(Global_Cluster) == holistic_id   ~ "Mental Rotation"
  ),
  Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  labs(title = "Cluster Result: Ability vs. Relative Slope")
```

```{r}
#| label: tbl-cluster-summary
#| fig-width: 6
#| fig-height: 5
#| tbl-cap: "Cluster summary statistics. Mental Rotation shows the steepest RT-angle slope despite lower ability than the Flipping/Matching group, whose elevated RT reflects uniformly slower but angle-insensitive processing."

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    n              = n(),
    mean_ability   = mean(ability, na.rm = TRUE),
    sd_ability     = sd(ability, na.rm = TRUE),
    mean_rel_slope = mean(Rel_Slope, na.rm = TRUE),
    sd_rel_slope   = sd(Rel_Slope, na.rm = TRUE),
    mean_abs_slope = mean(Abs_Slope, na.rm = TRUE),
    sd_abs_slope   = sd(Abs_Slope, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_rel_slope)) %>%
  transmute(
    Strategy       = Strategy_Label,
    N              = n,
    `Mean Ability` = mean_ability,
    `SD Ability`   = sd_ability,
    `Mean Rel. Slope` = mean_rel_slope,
    `SD Rel. Slope`   = sd_rel_slope,
    `Mean Abs. Slope` = mean_abs_slope,
    `SD Abs. Slope`   = sd_abs_slope
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>%
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, 
                `Mean Rel. Slope`, `SD Rel. Slope`,
                `Mean Abs. Slope`, `SD Abs. Slope`),
    decimals = 2
  ) %>%
  fmt_number(columns = N, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = where(is.numeric)) %>%
  cols_width(
    Strategy ~ px(160),
    N        ~ px(40),
    everything() ~ px(65)
  ) %>%
  tab_spanner(label = "Ability",         columns = c(`Mean Ability`, `SD Ability`)) %>%
  tab_spanner(label = "Relative Slope",  columns = c(`Mean Rel. Slope`, `SD Rel. Slope`)) %>%
  tab_spanner(label = "Absolute Slope",  columns = c(`Mean Abs. Slope`, `SD Abs. Slope`)) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

```{r}
#| label: fig-rt-by-disparity
#| fig-width: 6
#| fig-height: 5
#| include: true
#| fig-cap: "**Participant-weighted mean log RT by rotation angle and strategy cluster.** Shaded bands show 95% CIs. The Flipping/Matching group shows a characteristic inverted-U function, with RT declining after 120°, consistent with exploitation of reflective symmetry at high angular disparities."

rt_user_disparity <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors))) %>%
  group_by(Strategy_Label, user_id, angle) %>%      
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_disparity_summary <- rt_user_disparity %>%
  group_by(Strategy_Label, angle) %>%                 
  summarise(
    n_users  = n_distinct(user_id),
    mean_rt  = mean(user_mean_log_rt, na.rm = TRUE),
    se       = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    lwr      = mean_rt - 1.96 * se,
    upr      = mean_rt + 1.96 * se,
    .groups  = "drop"
  )

ggplot(rt_disparity_summary, aes(x = angle, y = mean_rt,
                                  colour = Strategy_Label, fill = Strategy_Label)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  scale_x_continuous(breaks = c(20, 40, 80, 120, 160)) +
  theme(legend.position = "bottom") +
  labs(
    #title  = "Mean log RT by Rotation Angle and Strategy",
    x      = "Rotation Angle (°)",
    y      = "Mean log RT (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: fig-accuracy-by-angle
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Participant-weighted Accuracy-angle functions by strategy cluster."

acc_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_user_angle <- acc_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(
    user_mean_acc = mean(correct, na.rm = TRUE),
    .groups = "drop"
  )

acc_summary <- acc_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users   = n_distinct(user_id),
    mean_acc  = mean(user_mean_acc, na.rm = TRUE),
    se_acc    = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

ggplot(acc_summary, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc, ymax = mean_acc + 1.96 * se_acc, fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean Accuracy (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  ) +scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) + 
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-descriptive-validation
#| tbl-cap: "Descriptive statistics by Strategy Cluster."

validation_table <- analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    N            = n(),
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    Mean_Instab  = mean(Instability, na.rm = TRUE),
    .groups = "drop"
  )

validation_table %>%
  transmute(
    Strategy      = Strategy_Label,
    N             = N,
    `Mean Ability`= Mean_Ability,
    `SD Ability`  = SD_Ability,
    `Mean Slope`  = Mean_Slope,
    `SD Slope`    = SD_Slope,
    `Instability` = Mean_Instab
  ) %>%
  gt() %>%
  tab_header(title = md("")) %>% 
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability), 
    decimals = 2
  ) %>%
  fmt_number(columns = c(N), decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Strategy, N)) %>%
  cols_align(align = "right", columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability)) %>%
  cols_width(
    Strategy     ~ px(180),
    N            ~ px(40),
    everything() ~ px(75)
  ) %>%
  tab_options(
    table.width = pct(100),
    data_row.padding = px(3)
  )
```

```{r}
#| label: fig-validation-boxplots
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Cluster validation. Left: ability (theta) by strategy group. Right: relative RT slope by strategy group. Significance bars show pairwise t-tests; non-significant comparisons hidden."

my_comparisons <- list(
  c("Ineffective Strategies", "Other Effective Strategies (Flipping/Matching)"),
  c("Ineffective Strategies", "Mental Rotation"),
  c("Other Effective Strategies (Flipping/Matching)", "Mental Rotation")
)

p_ability <- ggplot(analysis_data, aes(x = Strategy_Label, y = ability, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "1. Defined by Ability", y = "Ability (Theta)", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 15, hjust = 1))

p_slope <- ggplot(analysis_data, aes(x = Strategy_Label, y = Rel_Slope, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = pal$grey) +
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "2. Defined by RT Slope", y = "Relative Slope", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 12, hjust = 1))

p_ability + p_slope

```
\clearpage

```{r}
#| label: inline-values
#| include: false

ineff <- validation_table %>% filter(Strategy_Label == "Ineffective Strategies")
flip  <- validation_table %>% filter(Strategy_Label == "Other Effective Strategies (Flipping/Matching)")
mrot  <- validation_table %>% filter(Strategy_Label == "Mental Rotation")
```

### Strategy Identification via Clustering

To identify distinct cognitive strategies used during the mental rotation task, we extracted five subject-level features from a mixed-effects model of log-transformed reaction times: baseline speed, absolute RT-angle slope, relative RT-angle slope, response instability, and learning rate across trials. These features were entered into a k-means clustering algorithm (k = 3, nstart = 25), yielding three distinct strategy groups.

Cluster labels were assigned based on mean ability and relative slope: a low-ability cluster with a near-zero RT-angle slope was labeled "Ineffective Strategies" (n = `` `r ineff$N` ``); a higher-ability cluster with a moderate positive slope was labeled "Mental Rotation" (n = `` `r mrot$N` ``); and the largest cluster was labeled "Other Effective Strategies (Flipping/Matching)" (n = `` `r flip$N` ``). Summary statistics are presented in @tbl-cluster-summary.

The three clusters differed significantly on both dimensions (@fig-validation-boxplots). Ability differed across all pairwise comparisons (all *p*s < .0001), with the Ineffective Strategies group showing markedly lower ability (M = `` `r round(ineff$Mean_Ability, 2)` ``, SD = `` `r round(ineff$SD_Ability, 2)` ``) than both the Flipping/Matching (M = `` `r round(flip$Mean_Ability, 2)` ``, SD = `` `r round(flip$SD_Ability, 2)` ``) and Mental Rotation (M = `` `r round(mrot$Mean_Ability, 2)` ``, SD = `` `r round(mrot$SD_Ability, 2)` ``) groups. Relative RT slope also differed significantly across all pairwise comparisons (all *p*s < .001), with the Mental Rotation cluster showing the steepest slope (M = `` `r round(mrot$Mean_Slope, 2)` ``), consistent with analog rotation processes that scale with angular disparity.

Inspection of the RT-by-angle functions (@fig-rt-by-disparity) provided further validation of the cluster solution and offered mechanistic insight into the nature of each strategy. Rather than slowing with increasing angle as would be expected under systematic rotation, the Ineffective Strategies group showed decreasing RTs, suggesting they were relying on a more holistic or perceptual strategy rather than engaging in analytic rotation, consistent with their steeply declining accuracy (@fig-accuracy-by-angle). In contrast, both effective strategy groups showed increasing RTs with angular disparity, consistent with effortful but accurate processing. Critically, however, the two effective groups diverged at high angular disparities: whereas the Mental Rotation group continued to slow monotonically, the Flipping/Matching group showed a decrease in RT at 160°, crossing below the Mental Rotation line. This pattern could suggest a flipping strategy — rather than rotating a stimulus through its full angular distance, participants may exploit the reflective symmetry of the task by flipping the stimulus and rotating through the shorter complementary angle (e.g., 180° − 160° = 20°). Under this account, flipping becomes increasingly advantageous at high angular disparities, producing the inverted-U RT function observed. The Flipping/Matching group also maintained the highest accuracy across all angles despite elevated overall RTs, consistent with a slower but highly reliable alternative strategy.



# References

::: {#refs}
:::

# Cluster Solution Validation {.appendix}

To assess the appropriateness of the three-cluster solution, we evaluated cluster quality across k = 1–8 using the elbow method and average silhouette width (@fig-k-sweep-validation). The silhouette analysis suggested k = 2 as the statistically optimal solution, with silhouette width declining thereafter, and the elbow plot identified k = 3 as a reasonable inflection point at which additional clusters yielded diminishing reductions in within-cluster variance. To further interrogate the k = 2 solution, we examined its resulting RT-angle functions (@fig-k2-cluster-check). The higher-ability cluster displayed the characteristic inverted-U RT curve that, in the three-cluster solution, distinguishes the Flipping/Matching and Mental Rotation groups — indicating that k = 2 collapses two qualitatively distinct effective strategies into a single cluster. We therefore adopted k = 3 on both empirical and theoretical grounds.
```{r}
#| label: fig-k-sweep-validation
#| eval: true
#| include: true
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**Cluster validation metrics: Elbow Method (left) and Silhouette Analysis (right).** The elbow plot shows a bend at k=3, and silhouette width peaks at k=2 before declining, providing partial support for the 3-cluster solution."

p_elbow <- fviz_nbclust(global_input, kmeans, method = "wss", k.max = 8) +
  geom_vline(xintercept = 3, linetype = 2, color = pal$red) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

p_sil <- fviz_nbclust(global_input, kmeans, method = "silhouette", k.max = 8) +
  labs(title = NULL, subtitle = NULL, x = "Number of clusters k")

grid.arrange(p_elbow, p_sil, nrow = 1)
```

```{r}
#| label: fig-k2-cluster-check
#| eval: true
#| include: true
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**RT–angle functions for the k=2 cluster solution.** The higher-ability cluster displays an inverted-U RT curve consistent with a mix of Mental Rotation and Flipping/Matching strategies, suggesting that k=2 collapses two qualitatively distinct effective strategies into a single group."

# --- 1. Set up Input (Same as before, removing Learning_Rate for clarity) ---
analysis_data_k2 <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Abs_Slope, Rel_Slope, Instability, Learning_Rate), is.finite))

global_input_k2 <- analysis_data_k2 %>%
  select(Baseline_Speed, Rel_Slope, Instability, Abs_Slope, Learning_Rate) %>%
  scale()

# --- 2. Run K-Means with K=2 ---
set.seed(123)
k_global_2 <- kmeans(global_input_k2, centers = 2, nstart = 25)
analysis_data_k2$Global_Cluster <- as.factor(k_global_2$cluster)

# --- 3. Label the 2 Clusters ---
# We define them simply by Ability (Low vs High) since we can't split by slope yet
cluster_stats_k2 <- analysis_data_k2 %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability), 
    .groups = "drop"
  )

# Identify Lower Ability Cluster
low_eff_id <- cluster_stats_k2 %>% arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
# The other one is High Efficiency
high_eff_id <- cluster_stats_k2 %>% arrange(desc(mean_ability)) %>% slice(1) %>% pull(Global_Cluster)

analysis_data_k2 <- analysis_data_k2 %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id ~ "Low Efficiency (Cluster 1)",
    Global_Cluster == high_eff_id ~ "High Efficiency (Cluster 2)"
  ))

k2_colors <- c(
  "Low Efficiency (Cluster 1)"  = RColorBrewer::brewer.pal(3, "Set1")[1],
  "High Efficiency (Cluster 2)" = RColorBrewer::brewer.pal(3, "Set1")[2]
)

# --- 4. Plot the Curves (Visual Check) ---
# Join labels to raw data
rt_by_strategy_k2 <- mrot_rich_data %>%
  left_join(
    analysis_data_k2 %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle))

# Calculate User Means
rt_user_angle_k2 <- rt_by_strategy_k2 %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

# Calculate Group Means
rt_summary_k2 <- rt_user_angle_k2 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# plot
ggplot(rt_summary_k2, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_log_rt - 1.96*se_log_rt, 
                  ymax = mean_log_rt + 1.96*se_log_rt, 
                  fill = Strategy_Label), alpha = 0.2, colour = NA) +
  scale_colour_manual(values = k2_colors) +
  scale_fill_manual(values = k2_colors) +
  labs(
    title    = NULL,
    subtitle = NULL,
    y = "Mean log(RT) (participant-weighted)", 
    x = "Rotation angle (degrees)",
    colour = "Strategy", fill = "Strategy"
  ) +
  theme(legend.position = "bottom")
```
\clearpage
  
# Tsne check {.appendix}

To further validate the cluster structure, we examined whether the three-cluster solution was recoverable using alternative dimensionality reduction methods: PCA (@fig-pca-check), t-SNE (@fig-tsne-check), and UMAP (@fig-umap-check). PCA loadings were also inspected to characterize the underlying feature structure (@fig-pca-loadings).

Results were consistent across all three methods in one key respect: the Ineffective Strategies group was clearly and reliably separated from the two effective strategy groups, appearing as a distinct region in all three visualizations. This provides strong convergent evidence that this group occupies a genuinely different region of the feature space. The boundary between the Mental Rotation and Flipping/Matching groups was, however, less clear-cut. In all three visualizations, green and blue points showed substantial overlap, suggesting these two groups may represent ends of a continuum rather than fully discrete clusters. This is consistent with the silhouette analysis above, which favoured k = 2, and should be noted as a limitation of the three-cluster solution.

Inspection of the PCA loadings offered interpretive support for the feature set. PC1 loaded positively on Baseline_Speed, Learning_Rate, and Abs_Slope, and negatively on Instability, reflecting a general processing efficiency dimension. PC2 loaded primarily on Rel_Slope and Abs_Slope, reflecting a rotation cost dimension. The fact that these two components map cleanly onto theoretically meaningful dimensions provides confidence that the clustering features are capturing distinct aspects of task performance.
```{r}
#| label: fig-tsne-check
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**t-SNE visualization of strategy clusters.** The Ineffective Strategies group forms a distinct region in the top-left, while the Mental Rotation and Flipping/Matching groups show considerable overlap in the central and right regions, suggesting softer boundaries between the two effective strategy groups."

library(Rtsne)

tsne_input <- analysis_data %>%
    select(user_id, Learning_Rate, Baseline_Speed, Rel_Slope, Abs_Slope, Instability) %>%
    distinct() %>% 
    na.omit()

set.seed(42) 
tsne_result <- Rtsne(
    as.matrix(tsne_input %>% select(-user_id)), 
    dims = 2, 
    perplexity = 30, 
    verbose = FALSE, 
    max_iter = 500
)

tsne_plot_data <- tsne_input %>%
    mutate(
        tSNE_1 = tsne_result$Y[,1],
        tSNE_2 = tsne_result$Y[,2]
    ) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(tsne_plot_data, aes(x = tSNE_1, y = tSNE_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "Dimension 1",
        y = "Dimension 2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-check
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**PCA projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group separates clearly along PC1, while the Mental Rotation and Flipping/Matching groups overlap along both components."

pca_result <- prcomp(tsne_input %>% select(-user_id), center = TRUE, scale. = TRUE)

pca_plot_data <- tsne_input %>%
    mutate(PC1 = pca_result$x[,1], PC2 = pca_result$x[,2]) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) +
    labs(
        title = NULL,
        x = "PC1",
        y = "PC2",
        color = "Strategy"
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: fig-pca-loadings
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**PCA loadings for PC1 and PC2.** PC1 reflects a general processing efficiency dimension, loading positively on Baseline Speed, Learning Rate, and Absolute Slope, and negatively on Instability. PC2 reflects a rotation cost dimension, driven primarily by Relative and Absolute Slope."

library(tidytext)

loadings_df <- as.data.frame(pca_result$rotation) %>%
  rownames_to_column(var = "Angle_Condition")

loadings_long <- loadings_df %>%
  select(Angle_Condition, PC1, PC2) %>%
  pivot_longer(
    cols = c(PC1, PC2), 
    names_to = "Principal_Component", 
    values_to = "Loading"
  )

ggplot(loadings_long, aes(
    x = Loading, 
    y = reorder_within(Angle_Condition, Loading, Principal_Component), 
    fill = Loading > 0
  )) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Principal_Component, scales = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("TRUE" = pal$blue, "FALSE" = pal$red)) +
  labs(
    title    = NULL,
    subtitle = NULL,
    x = "Loading Weight",
    y = "Feature"
  ) +
  theme(
    #strip.text = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11),
    #panel.grid.minor = element_blank()
  )
```

::: {.landscape}
```{r}
#| label: fig-umap-check
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "**UMAP projection of the five clustering features coloured by strategy group.** The Ineffective Strategies group forms a distinct region in the bottom-left, while the Mental Rotation and Flipping/Matching groups show substantial overlap in the central and right regions."

set.seed(42) 
umap_result <- umap(as.matrix(tsne_input %>% select(-user_id)))

umap_plot_data <- tsne_input %>%
    mutate(UMAP_1 = umap_result$layout[,1], UMAP_2 = umap_result$layout[,2]) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(umap_plot_data, aes(x = UMAP_1, y = UMAP_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) + 
    labs(x = "UMAP Dimension 1", y = "UMAP Dimension 2") +
    theme(legend.position = "bottom")
```
:::

# Local Clustering Agreement {.appendix}
```{r}
#| label: local-clustering-agreement
#| include: false

vars <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")

stopifnot(all(vars %in% names(analysis_data)))
stopifnot(all(c("Site", "user_id", "Strategy_Label") %in% names(analysis_data)))
stopifnot(nrow(analysis_data) == nrow(dplyr::distinct(analysis_data, Site, user_id)))

# ARI function
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

get_local_strategy_labels <- function(df_site,
                                      k = 3,
                                      seed = 123,
                                      nstart = 25,
                                      min_n = 30) {

  if (nrow(df_site) < min_n) return(NULL)
  set.seed(seed)

  X <- df_site %>%
    dplyr::select(dplyr::all_of(vars)) %>%
    scale() %>%
    as.matrix()

  km <- kmeans(X, centers = k, nstart = nstart)

  tmp <- df_site %>%
    mutate(Local_Cluster = factor(km$cluster))

  local_stats <- tmp %>%
    group_by(Local_Cluster) %>%
    summarise(
      mean_abil  = mean(ability,   na.rm = TRUE),
      mean_slope = mean(Rel_Slope, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(mean_abil)

  ineff_id     <- local_stats %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  other_two    <- local_stats %>% filter(as.character(Local_Cluster) != ineff_id) %>% arrange(mean_slope)
  other_eff_id <- other_two   %>% slice(1) %>% pull(Local_Cluster) %>% as.character()
  mrot_id      <- other_two   %>% slice(2) %>% pull(Local_Cluster) %>% as.character()

  tmp %>%
    mutate(
      Strategy_Local = dplyr::case_when(
        as.character(Local_Cluster) == ineff_id     ~ "Ineffective Strategies",
        as.character(Local_Cluster) == other_eff_id ~ "Other Effective Strategies (Flipping/Matching)",
        as.character(Local_Cluster) == mrot_id      ~ "Mental Rotation"
      ),
      Strategy_Local = factor(Strategy_Local, levels = levels(analysis_data$Strategy_Label))
    ) %>%
    select(Site, user_id, Strategy_Local)
}

# fit local models independently within each site
local_labels_all <- analysis_data %>%
  group_by(Site) %>%
  group_split() %>%
  purrr::map_dfr(get_local_strategy_labels)

analysis_local <- analysis_data %>%
  select(-dplyr::any_of("Strategy_Local")) %>%
  left_join(local_labels_all, by = c("Site", "user_id"))

# agreement (3-class)
agreement_3class <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_by(Site) %>%
  summarise(
    n         = n(),
    pct_agree = mean(Strategy_Local == Strategy_Label),
    ARI       = ARI(Strategy_Label, Strategy_Local),
    .groups   = "drop"
  )

agreement_3class

# confusion matrices per site — save split first to avoid pipe . scoping issue
site_splits <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_split(Site)

confusion_by_site <- site_splits %>%
  purrr::set_names(purrr::map_chr(site_splits, ~ unique(.x$Site))) %>%
  purrr::map(~ table(.x$Strategy_Local, .x$Strategy_Label))

confusion_by_site

# recall + purity
recall_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Label, Strategy_Local) %>%
  group_by(Site, Strategy_Label) %>%
  mutate(recall = n / sum(n)) %>%
  ungroup()

purity_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Local, Strategy_Label) %>%
  group_by(Site, Strategy_Local) %>%
  mutate(purity = n / sum(n)) %>%
  ungroup()

diag_summary <- recall_tbl %>%
  filter(Strategy_Label == Strategy_Local) %>%
  transmute(Site, Strategy = Strategy_Label, recall) %>%
  left_join(
    purity_tbl %>%
      filter(Strategy_Local == Strategy_Label) %>%
      transmute(Site, Strategy = Strategy_Local, purity),
    by = c("Site", "Strategy")
  )

diag_summary
```

```{r}
#| label: local-vs-pooled-reporting
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_pct <- function(x, d = 1) paste0(formatC(100*x, format = "f", digits = d), "%")

get_row <- function(df, site) df[df$Site == site, , drop = FALSE]

# sites
s_mpi <- "mpieva-de"
s_uni <- "uniandes-co"
s_wes <- "western-ca"

# agreement rows
a_mpi <- get_row(agreement_3class, s_mpi)
a_uni <- get_row(agreement_3class, s_uni)
a_wes <- get_row(agreement_3class, s_wes)

# label strings
lab_ineff <- "Ineffective Strategies"
lab_other <- "Other Effective Strategies (Flipping/Matching)"
lab_mrot  <- "Mental Rotation"

get_diag <- function(site, label) {
  out <- diag_summary[diag_summary$Site == site & diag_summary$Strategy == label, , drop = FALSE]
  if (nrow(out) == 0) {
    return(data.frame(Site = site, Strategy = label, recall = NA_real_, purity = NA_real_))
  }
  stopifnot(nrow(out) == 1)
  out
}

d_mpi_ineff <- get_diag(s_mpi, lab_ineff)
d_mpi_other <- get_diag(s_mpi, lab_other)
d_mpi_mrot  <- get_diag(s_mpi, lab_mrot)

d_uni_ineff <- get_diag(s_uni, lab_ineff)
d_uni_other <- get_diag(s_uni, lab_other)
d_uni_mrot  <- get_diag(s_uni, lab_mrot)

d_wes_ineff <- get_diag(s_wes, lab_ineff)
d_wes_other <- get_diag(s_wes, lab_other)
d_wes_mrot  <- get_diag(s_wes, lab_mrot)

# off-diagonal moves
get_move <- function(tab, local_lab, global_lab) {
  stopifnot(local_lab %in% rownames(tab), global_lab %in% colnames(tab))
  n     <- tab[local_lab, global_lab]
  denom <- sum(tab[, global_lab])
  list(n = as.numeric(n), denom = as.numeric(denom), prop = as.numeric(n) / as.numeric(denom))
}

tab_mpi <- confusion_by_site[[s_mpi]]
tab_uni <- confusion_by_site[[s_uni]]
tab_wes <- confusion_by_site[[s_wes]]

# Mental Rotation globally assigned to Other Effective locally
mpi_mrot2other <- get_move(tab_mpi, lab_other, lab_mrot)
wes_mrot2other <- get_move(tab_wes, lab_other, lab_mrot)

# Other Effective globally assigned to Mental Rotation locally
uni_other2mrot <- get_move(tab_uni, lab_mrot, lab_other)
agreement_3class
confusion_by_site

```

### Full 0–360° Angle Functions check
The primary analyses modeled angular disparity (i.e., the shortest rotational distance to upright, capped at 180°) on the assumption that participants rotate stimuli via the most efficient path. To test this assumption, we re-ran the clustering and plotted RT and accuracy functions using the full 0–360° rotation angles (@fig-rt-angle-by-strategy-360, @fig-accuracy-angle-by-strategy-360).

The results were broadly consistent with the disparity-based analyses. Critically, the RT functions for both effective strategy groups showed the expected symmetry around 180°: RTs increased from 0° to approximately 180–200°, then decreased again as angles approached 360°, forming an inverted-U consistent with participants always rotating via the shortest available path. This pattern provides direct support for the angular disparity assumption underlying the primary analysis.

The flipping signature observed in the Flipping/Matching group was again evident: their RT function rose more steeply and peaked later than the Mental Rotation group, before declining sharply after 240°, consistent with a strategy that exploits reflective symmetry at high disparities. The Ineffective Strategies group continued to show a broadly decreasing RT function across angles, and accuracy for this group deteriorated most steeply, remaining near chance at peak angles. The Flipping/Matching group again maintained the highest accuracy across the full angular range, reinforcing the interpretation that this strategy, while slower, is highly reliable.
```{r}
#| label: robustness-check-all-angles
#| results: "hide"

# Ability estimate - keep site + user
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    
    # Extract full 0-360° angle directly from item_original
    angle = as.integer(str_extract(item_original, "\\d{3}$")),
    
    # Calculate shortest distance to center (disparity) for the model
    ang_disparity = pmin(angle, 360 - angle),
    
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(ang_disparity)), 
    trial_s  = as.numeric(scale(trial_number))
  )

# --- Doria's FIX: Set Sum-to-Zero Contrasts ---
# By default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the GRAND MEAN slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
# Fit the rich mixed model
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# Extract subject-level random effects (deviations)
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# Fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# Compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope      = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope      = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# 4. Compute Instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(site, user_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

# 5. Join and Prepare for Clustering
rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(instability_data, by = c("site", "user_id")) %>% 
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  # Filter to ensure all needed features are finite
  # UPDATED: Added Learning_Rate to the list
  filter(if_all(c(ability, Abs_Slope, Rel_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))

# Check: ang_disparity should never exceed 180
summary(mrot_rich_data$ang_disparity)

# Spot check a few items
mrot_rich_data %>%
  select(item, item_original, angle, ang_disparity) %>%
  distinct() %>%
  arrange(item_original) %>%
  head(20)
```

```{r}
#| label: clustering-execution-360
#| include: false

my_colors <- c(
  "Ineffective Strategies" = RColorBrewer::brewer.pal(3, "Set1")[1],
  "Other Effective Strategies (Flipping/Matching)" = RColorBrewer::brewer.pal(3, "Set1")[2], 
  "Mental Rotation" = RColorBrewer::brewer.pal(3, "Set1")[3]  
)

analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Rel_Slope, Abs_Slope, 
                  Instability, Learning_Rate), is.finite))
global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()
set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)
cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
effective_clusters <- cluster_stats %>%
  filter(Global_Cluster != low_eff_id) %>%
  arrange(mean_slope)
analytical_id <- effective_clusters %>% slice(1) %>% pull(Global_Cluster)
holistic_id   <- effective_clusters %>% slice(2) %>% pull(Global_Cluster)
analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id    ~ "Ineffective Strategies",
    Global_Cluster == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    Global_Cluster == holistic_id   ~ "Mental Rotation"
  ))
ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Cluster Result: Ability vs. Relative Slope (full 0-360° angles)",
    x = "Ability", y = "Relative Slope"
  )
```

```{r}
#| label: fig-rt-angle-by-strategy-360
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Participant-weighted RT–angle functions by strategy cluster (numbers indicate unique participants contributing at each angle; shaded bands show 95% CIs) (Note: Re-test data removed)."

rt_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

rt_user_angle <- rt_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

rt_angle_person_summary <- rt_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

missing_cols <- setdiff(levels(rt_angle_person_summary$Strategy_Label), names(my_colors))
stopifnot(length(missing_cols) == 0)

ggplot(rt_angle_person_summary, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_log_rt - 1.96 * se_log_rt,
        ymax = mean_log_rt + 1.96 * se_log_rt,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean log(RT) (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```

```{r}
#| label: fig-accuracy-angle-by-strategy-360
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Participant-weighted Accuracy-angle functions by strategy cluster."

acc_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(angle)) %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = names(my_colors)))

acc_user_angle <- acc_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_acc = mean(correct, na.rm = TRUE), .groups = "drop")

acc_summary <- acc_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users  = n_distinct(user_id),
    mean_acc = mean(user_mean_acc, na.rm = TRUE),
    se_acc   = sd(user_mean_acc, na.rm = TRUE) / sqrt(n_users),
    .groups  = "drop"
  )

ggplot(acc_summary, aes(x = angle, y = mean_acc, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(ymin = mean_acc - 1.96 * se_acc,
        ymax = mean_acc + 1.96 * se_acc,
        fill = Strategy_Label),
    alpha = 0.15, colour = NA
  ) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) +
  theme(legend.position = "bottom") +
  labs(
    x      = "Rotation angle (degrees)",
    y      = "Mean Accuracy (participant-weighted)",
    colour = "Strategy", fill = "Strategy"
  )
```
# IRT Ability by Age {.appendix} - leaving out for now.

```{r}
#| label: fig-rt-angle_raw
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI."
#| fig-width: 8
#| fig-height: 8
  
  # Add raw data underneath
  p_mrot_rt_angle_age_2 <- ggplot() +
    # raw per-run medians
    geom_point(
      data = mrot_items_rt_runs,
      aes(x = angle, y = rt_med,
          colour = stimulus),
      size = 0.4,
      alpha = 0.25,
      position = position_jitter(width = 3, height = 0)
    ) +
    # summary CI + lines
    geom_errorbar(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y, ymin = lwr, ymax = upr,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      width = 2,
      linewidth = 0.35,
      alpha = 0.6
    ) +
    geom_point(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      size = 1
    ) +
    geom_line(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      alpha = 0.85
    ) +
    facet_grid(site ~ age_group) +
    labs(
      #title = "Mental Rotation: RT by angle within age groups ±95% CI",
      x = "Rotation angle (°)",
      y = "RT (s)",
      colour = "Stimulus",
      #linetype = "Item Type"
    )+ scale_x_continuous(breaks = c(0, 40, 80, 120, 160, 200, 240, 280, 320, 360)) 
  p_mrot_rt_angle_age_2
```

<!--

####################### ROBUSTNESS CHECK - BOOTSTRAPPING ##########################
```{r}
#| label: cluster-stability-bootstrap
#| include: false

vars   <- c("Baseline_Speed", "Rel_Slope", "Abs_Slope", "Instability", "Learning_Rate")
k      <- 3
B      <- 500
nstart <- 25
set.seed(123)

df_full <- analysis_data %>%
  select(user_id, Strategy_Label, all_of(vars)) %>%
  distinct(user_id, .keep_all = TRUE) %>%
  filter(if_all(all_of(vars), is.finite)) %>%
  mutate(
    Strategy_Label = factor(
      Strategy_Label,
      levels = names(my_colors)   
    )
  )

stopifnot(nlevels(df_full$Strategy_Label) == 3)
stopifnot(n_distinct(df_full$user_id) == nrow(df_full))

ref_levels <- levels(df_full$Strategy_Label)
y_ref      <- df_full$Strategy_Label
ids        <- df_full$user_id
n          <- nrow(df_full)

# scaling: lock to full-sample mean/sd so bootstrap runs are comparable
mu  <- df_full %>% summarise(across(all_of(vars), \(x) mean(x, na.rm = TRUE))) %>% as.list()
sdv <- df_full %>% summarise(across(all_of(vars), \(x) sd(x,   na.rm = TRUE))) %>% as.list()

X_full <- df_full %>%
  mutate(across(all_of(vars), ~ (. - mu[[cur_column()]]) / sdv[[cur_column()]])) %>%
  select(all_of(vars)) %>%
  as.matrix()

# helpers
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)
  comb2    <- function(x) x * (x - 1) / 2
  sum_ij   <- sum(comb2(tab))
  sum_i    <- sum(comb2(rowSums(tab)))
  sum_j    <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom    <- max_idx - expected
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

jaccard <- function(set_a, set_b) {
  inter <- length(intersect(set_a, set_b))
  uni   <- length(union(set_a, set_b))
  if (uni == 0) NA_real_ else inter / uni
}

all_perms_3 <- list(
  c(1,2,3), c(1,3,2),
  c(2,1,3), c(2,3,1),
  c(3,1,2), c(3,2,1)
)

# Match bootstrap cluster IDs to reference labels by maximum membership overlap
best_perm_by_overlap <- function(cl_hat, y_ref, ref_levels, all_perms_3) {
  tab    <- table(cl_hat, factor(y_ref, levels = ref_levels))
  scores <- sapply(all_perms_3, function(p) sum(tab[cbind(1:3, p)]))
  all_perms_3[[which.max(scores)]]
}

# reference k-means (reproduces clustering-execution exactly)
# UPDATED: label assignment now mirrors the two-step logic in clustering-execution:
#   step 1 — lowest mean ability  → "Ineffective Strategies"
#   step 2 — of the remainder, flatter Rel_Slope → "Other Effective...", 
#             steeper → "Mental Rotation"
# We use this to build ref_centers aligned to ref_levels.

# need ability for label assignment (not a clustering var, so grab separately)
ability_lookup <- df_full %>%
  left_join(analysis_data %>% distinct(user_id, ability), by = "user_id")

set.seed(123)
km_ref <- kmeans(X_full, centers = k, nstart = nstart)
cl_ref <- km_ref$cluster

# replicate the production label-assignment rule on the reference run
cluster_stats_ref <- tibble(
  cluster      = 1:k,
  mean_ability = tapply(ability_lookup$ability, cl_ref, mean, na.rm = TRUE),
  mean_rel_slope = tapply(df_full$Rel_Slope,   cl_ref, mean, na.rm = TRUE)
)

low_eff_id_ref  <- cluster_stats_ref %>% arrange(mean_ability) %>% slice(1) %>% pull(cluster)
effective_ref   <- cluster_stats_ref %>% filter(cluster != low_eff_id_ref) %>% arrange(mean_rel_slope)
analytical_id_ref <- effective_ref %>% slice(1) %>% pull(cluster)
holistic_id_ref   <- effective_ref %>% slice(2) %>% pull(cluster)

# map cluster integer to ref_levels index (used to order ref_centers rows)
label_order_ref <- integer(k)
label_order_ref[low_eff_id_ref]   <- which(ref_levels == "Ineffective Strategies")
label_order_ref[analytical_id_ref] <- which(ref_levels == "Other Effective Strategies (Flipping/Matching)")
label_order_ref[holistic_id_ref]   <- which(ref_levels == "Mental Rotation")

# ref_centers[i, ] corresponds to ref_levels[i]
ref_centers <- km_ref$centers[order(label_order_ref), , drop = FALSE]

# bootstrap loop
res <- map_dfr(1:B, function(b) {

  set.seed(1000 + b)

  idx        <- sample.int(n, size = n, replace = TRUE)
  idx_unique <- unique(idx)

  tab_ref_unique <- table(factor(y_ref[idx_unique], levels = ref_levels))
  min_ref_unique <- min(tab_ref_unique)
  min_ref_prop   <- min_ref_unique / length(idx_unique)
  n_unique       <- length(idx_unique)

  X_b <- X_full[idx, , drop = FALSE]
  km  <- kmeans(X_b, centers = k, nstart = nstart)

  # assign all full-sample participants to nearest bootstrap center
  centers_b <- km$centers
  dists <- sapply(1:k, function(j) {
    rowSums((X_full - matrix(centers_b[j, ],
                             nrow = nrow(X_full), ncol = ncol(X_full),
                             byrow = TRUE))^2)
  })
  cl_hat <- max.col(-dists)

  # resolve label switching via maximum overlap with reference labels
  perm  <- best_perm_by_overlap(cl_hat, y_ref, ref_levels, all_perms_3)
  y_hat <- factor(ref_levels[perm[cl_hat]], levels = ref_levels)

  ari <- ARI(y_ref, y_hat)

  # per-cluster Jaccard
  jac <- map_dbl(ref_levels, function(lab) {
    jaccard(ids[y_ref == lab], ids[y_hat == lab])
  })
  names(jac) <- paste0("J_", make.names(ref_levels))

  # mean centroid displacement vs reference
  centers_b_aligned <- centers_b[perm, , drop = FALSE]
  center_shift      <- mean(sqrt(rowSums((centers_b_aligned - ref_centers)^2)))

  tibble(
    replicate      = b,
    ARI            = ari,
    n_unique       = n_unique,
    min_ref_unique = min_ref_unique,
    min_ref_prop   = min_ref_prop,
    center_shift   = center_shift,
    !!!as.list(jac)
  )
})

# summaries
stability_summary <- res %>%
  summarise(
    B                   = n(),
    ARI_median          = median(ARI, na.rm = TRUE),
    ARI_IQR             = IQR(ARI, na.rm = TRUE),
    center_shift_median = median(center_shift, na.rm = TRUE),
    center_shift_IQR    = IQR(center_shift, na.rm = TRUE),
    across(starts_with("J_"),
           list(median = ~ median(.x, na.rm = TRUE),
                IQR    = ~ IQR(.x, na.rm = TRUE)))
  )

stability_summary

ari_tail <- res %>%
  summarise(
    pct_ARI_below_0.70 = mean(ARI < 0.70, na.rm = TRUE) * 100,
    pct_ARI_below_0.60 = mean(ARI < 0.60, na.rm = TRUE) * 100,
    min_ARI            = min(ARI, na.rm = TRUE)
  )

ari_tail
```

```{r}
#| label: cluster-stability-reporting
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)

ari_med    <- stability_summary$ARI_median
ari_iqr    <- stability_summary$ARI_IQR
pct_ari_70 <- ari_tail$pct_ARI_below_0.70
pct_ari_60 <- ari_tail$pct_ARI_below_0.60
min_ari    <- ari_tail$min_ARI

# UPDATED: pull Jaccard stats using new label names
j_med <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_median")]]
})
j_iqr <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_IQR")]]
})
names(j_med) <- ref_levels
names(j_iqr) <- ref_levels

j_ineff_med  <- j_med["Ineffective Strategies"]
j_ineff_iqr  <- j_iqr["Ineffective Strategies"]
j_other_med  <- j_med["Other Effective Strategies (Flipping/Matching)"]
j_other_iqr  <- j_iqr["Other Effective Strategies (Flipping/Matching)"]
j_mrot_med   <- j_med["Mental Rotation"]
j_mrot_iqr   <- j_iqr["Mental Rotation"]

rho_shift <- suppressWarnings(
  cor(res$ARI, res$center_shift, method = "spearman", use = "complete.obs")
)

# ARI distribution across bootstrap resamples
ggplot(res, aes(x = ARI)) +
  geom_histogram(bins = 40, fill = "#3498db", alpha = 0.8) +
  geom_vline(xintercept = median(res$ARI, na.rm = TRUE), 
             linetype = "dashed", colour = "black") +
  labs(x = "ARI (vs reference clustering)", y = "Count",
       title = paste0("Bootstrap stability (B = ", nrow(res), ")"),
       subtitle = paste0("Median ARI = ", round(median(res$ARI, na.rm = TRUE), 3)))

# Per-cluster Jaccard
res %>%
  select(replicate, starts_with("J_")) %>%
  pivot_longer(-replicate, names_to = "Cluster", values_to = "Jaccard") %>%
  mutate(Cluster = str_remove(Cluster, "^J_")) %>%
  ggplot(aes(x = Jaccard, fill = Cluster)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  facet_wrap(~ Cluster, ncol = 1) +
  labs(x = "Jaccard index", y = "Count") +
  theme(legend.position = "none")

```
# checks
```{r}
analysis_data %>%
  count(Strategy_Label)

analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(across(all_of(c("Baseline_Speed", "Rel_Slope", "Abs_Slope", 
                             "Instability", "Learning_Rate")), 
                   ~ round(mean(.x, na.rm = TRUE), 3)))

analysis_data %>%
  select(Strategy_Label, Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  pivot_longer(-Strategy_Label, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Strategy_Label, y = value, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~ variable, scales = "free_y") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

####################### ROBUSTNESS CHECK - WITHIN PERSON STRATEGY SWITCHING (DORIA'S ANALYSIS) ##########################

### Strategy Stability Analysis (Split-Half Reliability)

To investigate whether children consistently use a single cognitive strategy throughout the task or switch strategies over time (e.g., discovering a shortcut), we performed a **Split-Half Reliability Analysis**.

**Methodology:**
We partitioned each participant's data into two sequential halves: the first 50% of trials (Early Phase) and the last 50% of trials (Late Phase). For each half, we estimated a "Raw Slope" using a simple linear regression of reaction time on angle (`log_rt ~ angle`). This slope serves as a proxy for the participant's rotation strategy during that specific period. 

We then correlated the *First Half Slope* with the *Second Half Slope* to quantify stability. A high correlation ($r \approx 1$) would indicate a fixed strategy, while a low correlation ($r \approx 0$) suggests dynamic behavior or strategy switching.

```{r}
#| label: fig-strategy-switching
#| fig-cap: "Strategy Stability: Comparison of Rotation Slope in First Half vs. Second Half of trials. Points below the diagonal suggest a switch to a flatter strategy (e.g., discovering the shortcut)."


# 1. Split Data into Halves
# We determine the midpoint trial for each user
split_data <- mrot_rich_data %>%
  group_by(user_id) %>%
  mutate(
    Midpoint = median(trial_number, na.rm = TRUE),
    Half = ifelse(trial_number <= Midpoint, "First_Half", "Second_Half")
  ) %>%
  ungroup() %>%
  # Filter out bad data points just to be safe
  filter(is.finite(log_rt), is.finite(angle))

# 2. Calculate Simple Slope for Each Half (Using simple lm per subject)
# This is a fast proxy for the complex mixed-model slope
slope_stability <- split_data %>%
  group_by(user_id, Half) %>%
  summarise(
    # We only care about the slope (coefficient for angle)
    # We use a safe lm() call wrapped in tryCatch to handle kids with too few trials
    Raw_Slope = tryCatch(
      #coef(lm(log_rt ~ angle, data = pick(everything())))["angle"], 
      # Use distance from 0 so the slope represents 'cost of rotation'
  coef(lm(log_rt ~ pmin(angle, 360 - angle), data = pick(everything())))[2],
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  # Reshape to Wide Format (Columns: Slope_First, Slope_Second)
  tidyr::pivot_wider(names_from = Half, values_from = Raw_Slope, names_prefix = "Slope_") %>%
  # Join with our Final Strategy Labels to color-code
  left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id") %>%
  filter(!is.na(Strategy_Label), !is.na(Slope_First_Half), !is.na(Slope_Second_Half))

# 3. Calculate Correlation (Quantify Stability)
correlation <- cor(slope_stability$Slope_First_Half, slope_stability$Slope_Second_Half, use = "complete.obs")

# 4. Visualization
ggplot(slope_stability, aes(x = Slope_First_Half, y = Slope_Second_Half, color = Strategy_Label)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") + # The "Stability Line"
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Strategy Switching Check (Split-Half Reliability)",
    subtitle = paste0("Correlation between halves: r = ", round(correlation, 2), 
                      ". Points below diagonal = Strategy became flatter (faster)."),
    x = "Slope (First Half of Trials)",
    y = "Slope (Second Half of Trials)"
  ) +
  coord_fixed() # Keeps the aspect ratio square so the diagonal is 45 degrees
```

```{r}
# Doria's original code for examining up to 360 
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    
    # 1. Extract the base angle (e.g., 40)
    base_angle = as.integer(str_extract(item, "\\d+")), 
    
    # 2. Detect if the original item was a "Negative/Counter-Clockwise" rotation
    # Looking for 'rn' prefix in item_original (e.g., rn040)
    is_ccw = str_detect(item_original, "^rn"),
    
    # 3. RECONSTRUCT the full 0-360 angle
    angle = ifelse(is_ccw, 360 - base_angle, base_angle),
    
    # 4. Calculate shortest distance to center (disparity) for the model
    # This ensures the model still measures 'Rotation Cost' correctly
    ang_disparity = pmin(angle, 360 - angle),
    
    stimulus = as.factor(stimulus),
    # SCALE THE DISPARITY, NOT THE RAW ANGLE
    angle_s  = as.numeric(scale(ang_disparity)), 
    trial_s  = as.numeric(scale(trial_number))
  )

#  Issues: the above code is only only catches rabbit counter-clockwise items (rn040, rn080 etc.), and misses duck and shape items. It is also not necessary to deconstruct item_original as the info. we need is already there. item_original already contains the full angle directly (e.g., an11240 = 240).

# Fionnuala checking
# Check if column exists
"item_original" %in% names(mrot_rt_trial_no_retest)

# If it does, see what's in it
mrot_rt_trial_no_retest$item_original |> unique() |> sort()

# problem
# str_detect(item_original, "^rn") is wrong — it only catches rabbit counter-clockwise, missing dn, an, bn etc.

#diagnose
# Step 1: Check is_ccw detection by stimulus
mrot_rich_data %>%
  distinct(stimulus, item_original, is_ccw, base_angle, angle) %>%
  arrange(stimulus, item_original) %>%
  print(n = 50)

mrot_rich_data %>%
  count(stimulus, angle) %>%
  pivot_wider(names_from = stimulus, values_from = n) %>%
  arrange(angle) %>%
  print(n = 20)
```
-->