---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "Fionnuala O'Reilly"
    orcid: 0000-0002-4355-9088
    affiliations:
      - ref: stanford
  - name: "Xingyao Xiao"
    orcid: 0000-0001-8430-0438
    affiliations:
      - ref: stanford
  - name: "Rebecca Zhu"
    orcid: 0000-0003-0957-5962
    affiliations:
      - ref: stanford
  - name: "Mika Braginsky"
    orcid: 0000-0001-9039-3220
    affiliations:
      - ref: stanford
  - name: "Nilàm Ram"
    orcid: 0000-0003-1671-5257
    affiliations:
      - ref: stanford
  - name: "Benjamin W. Domingue"
    orcid: 0000-0002-3894-9049
    affiliations:
      - ref: stanford
  - name: "Michael C. Frank"
    orcid: 0000-0002-7551-4378
    affiliations:
      - ref: stanford
    corresponding: true
    email: mcfrank@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["cognitive development", "mental rotation", "spatial reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    extra_dependencies: ["booktabs", "longtable"]
    # fig-pos: H        
    # tbl-pos: H  
    

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

Spatial reasoning encompasses a range of abilities for generating, retaining, and manipulating visuospatial information [@lohman2013spatial]. One core component is mental rotation, which involves mentally transforming an object by rotating it in space to predict how it would appear from a different orientation or viewpoint. More broadly, spatial skills support the formation and transformation of mental representations of 2D shapes and 3D objects (e.g., rotating, bending, or folding them) and help us track how spatial relations between locations, shapes, and objects change over time [@frick2014development; @newcombe2014thinking]. In everyday life, we rely on these skills when judging whether an object will fit into a tight space, imagining how a room might look after rearranging furniture, or navigating using a map.

Spatial reasoning is also linked to a wide range of educational outcomes [@wai2018spatial], spanning creative domains such as music [@padulo2020effects; @hetland2000learning] and visual arts [@kell2013creativity; @winner2013relationship], as well as STEM subjects [@wai2009spatial]—and especially mathematics [@atit2022examining; @xie2020spatial; @mix2012relation; @gunderson2012relation; @frick2019spatial]. Mental rotation, the focus of this paper, is one spatial skill that has been repeatedly implicated in these associations, making it a useful target for understanding how spatial reasoning relates to learning and development.

Despite its practical and educational importance, significant questions remain regarding how mental rotation varies across contexts. In particular, little is known about whether cultural background shapes children’s mental-rotation skill, whether the canonical linear relationship between reaction time and angular disparity replicates across diverse populations, or which cognitive strategies children employ when solving mental-rotation problems. The present study addresses these gaps by examining children’s mental-rotation response patterns in Canada, Germany, and Colombia.

## Emergence and developemnt of spatial skills during childhood

Mental rotation can be observed as early as three months of age [@johnson2020spatial], with rapid development between ages three and five [@frick2013development]. Individual differences are evident early, and a relatively large sex difference in mental rotation performance has been reported in multiple studies [@boone2017sex; @linn1985emergence], with males, on average, outperforming females. However, the mechanisms driving developmental change—and the extent to which performance reflects qualitatively different solution strategies rather than a single underlying ability—remain open questions, particularly in culturally diverse samples.

A growing body of work suggests that spatial skills are malleable, with training studies in children showing gains on mental rotation and broader spatial reasoning measures, and in some cases transfer beyond the trained task [@hawes2015effects; @gilligan2020first; @lowrie2017visuospatial]. Consistent with this, a meta-analysis of 217 studies reported a moderate average effect of spatial training versus control conditions (Hedges’ g = 0.47, SE = 0.04) [@uttal2013malleability].

## Cultural Comparisons

From a biological perspective, there are strong reasons to expect cross-cultural similarities in mental rotation skills. Humans share a common genetic heritage and sensory-motor systems that shape how we perceive and interact with space. Because the nervous system likely evolved in response to universal demands, such as navigating environments and learning spatial layouts, spatial cognition is considered a relatively conserved neural specialization [@lewis1988signs; @montello1995significant; @nazareth2019meta].

While research on cultural variations in mental rotation is limited, existing studies highlight key predictors of performance. A common thread of these studies is their exploration of sex differences in performance to understand if the male advantage [@linn1985emergence; @maeda2013meta] in mental rotation is universal. For example, a large-scale multicultural study of 3,367 students in Canada, Germany, and Japan found that males outperformed females with a consistently large effect size across all three nations. However, the study also identified academic program as an independent predictor; students in science and engineering scored higher than those in the arts and social sciences, regardless of sex [@peters2006mental].

More recent developmental research suggests that these gender gaps may be more plastic than previously thought, particularly when cultural norms regarding autonomy are shifted. Among Tsimané children in the Bolivian Amazon, the high degree of freedom to explore allowed to both sexes resulted in a notable absence of gender differences in mobility and spatial ability [@davis2019spatial]. Interestingly, while formal schooling in this population improved abstract skills like mental rotation, it correlated with a decrease in large-scale navigational accuracy, suggesting a trade-off between classroom-based learning and environmental exploration. This environmental influence is further evidenced in seminomadic communities in Namibia. While local adults traditionally show a gender gap in spatial skills, recent cultural shifts—such as formal schooling and increased travel requirements—have effectively eliminated these differences in children [@davis2021cultural]. Both boys and girls demonstrated comparable navigational accuracy and mental rotation performance, indicating that when cultural norms provide equal opportunities for exploration and education, traditional sex differences in spatial cognition can disappear.

## Reaction time as a function of rotation angle

One of the most cited findings in psychology is that of @shepard1971mental, who demonstrated that mental rotation operates in a spatially continuous manner: response times increased approximately linearly with the angular disparity between two objects. This suggests that participants judge whether an object matches a previously seen one by performing a gradual rotation of an internal representation, taking longer for larger disparities (e.g., 75° vs. 110°).

Extending this work from the perspective of visual cognition, @tarr1989mental demonstrated that mental rotation is not only a mechanism for comparing two simultaneous stimuli, but a fundamental process integrated into long-term object recognition. While @shepard1971mental focused on "handedness" (the ability to distinguish a shape from its mirror image), @tarr1989mental showed that this same linear increase in response time occurs when identifying a single shape from memory. Their findings suggest that the human visual system is highly adaptive: as individuals gain experience with an object at new orientations, they store additional "viewpoint-specific" representations. Consequently, mental rotation does not always return to a single "upright" canonical view; instead, the internal representation is rotated to the nearest familiar orientation stored in memory.

Examining reaction times across diverse populations, a cross-cultural comparison of children in the Ivory Coast and Switzerland revealed that while both groups utilized the same mental rotation strategies, Swiss children demonstrated significantly higher speed and accuracy [@nunez1998mental]. Interestingly, both groups exhibited an identical, non-linear dip in performance at 30° and 330° angles, suggesting there could be a more universal cognitive constraint in spatial processing that exists independently of cultural background. Notably, age predicted performance in the Swiss group—with older children showing marked improvements—whereas it had no significant effect on the performance of the Ivorean children [@nunez1998mental].

## Cognitive strategies

The literature has shifted from viewing mental rotation as a monolithic, linear process to understanding it as a strategic choice mediated by cognitive load and task demands. While early models focused on a universal sequence of encoding, rotation, and comparison [@cooper1973chronometric; @desrocher1995stimulus], more recent research identifies a holistic-analytic continuum of strategies that varies significantly across individuals [@gluck2003spatial].

At one end of this continuum is the holistic strategy, where the stimulus is transformed as a single, integrated mental representation. This is often referred to as the ‘pure rotation’ approach and is characterized by high efficiency, lower total gaze times, and fewer transitions between stimuli [@taragin2019mental; @tang2023eye]. Conversely, the ‘piecemeal’ or local-analytic strategy involves decomposing the object into sub-components for sequential transformation and comparison. This strategy is marked by a ‘switching’ eye-pattern, involving frequent back-and-forth saccades as the participant verifies individual features [@nazareth2019strategy; @fernandez2025inducing].

The selection of these strategies is heavily dictated by Visual Working Memory (VWM) capacity. @just1985cognitive first proposed that the piecemeal approach serves as a "workaround" for limited VWM; if an individual cannot maintain the entire 3D structure in mind, the representation falls apart, forcing them to rotate it piece-by-piece. This is further supported by @khooshabeh2013individual, who demonstrated that while high-spatial individuals default to holistic strategies, low-spatial individuals often remain ‘stuck’ in piecemeal processing regardless of task difficulty.

Furthermore, the "how" of rotation involves the frame of reference used. @zacks2008neuroimaging distinguishes between object-based transformations (rotating the object itself) and egocentric perspective-taking (imagining oneself moving around the object). This choice is often influenced by the environment; for instance, 3D Virtual Reality (VR) environments provide stereoscopic depth cues that facilitate holistic, object-based rotations, whereas 2D presentations often force a more cognitively taxing piecemeal approach due to ambiguous depth cues [@tang2023eye].

Successful performance is often defined by strategy flexibility—the ability to dynamically alternate between holistic rotation and analytic checking based on the specific demands of the stimulus [@nazareth2019strategy]. When task difficulty increases—such as with mirrored objects or rotations along skewed axes—flexible rotators can shift strategies to manage the increased cognitive load [@hsing2023using; @nolte2022rotational]. This suggests that observed differences in performance may not stem from an inherent inability to rotate, but rather from differences in strategy selection and the stability of those strategies over time [@geiser2008quantitative].

## The present study

The Learning Variability Network Exchange (LEVANTE) is an international research collaboration that supports coordinated data collection to better understand how children’s learning and development vary across ages and contexts. Partner sites administer a shared, openly available battery designed for a broad developmental range, and data are harmonized into a central repository. With support from the Jacobs Foundation, participating sites are funded to collect longitudinal data using the LEVANTE framework. Contributing researchers receive access to their site’s data first, followed by scheduled public releases for the wider research community. LEVANTE’s long-term aim is to build and openly share a large, high-quality dataset—alongside the measures themselves—capturing children’s development across diverse settings. In the present study, we use data for the mental rotation task only to answer three core research questoins:

1.  Does the mental-rotation task show evidence of cross-site measurement invariance, indicating that it taps a shared underlying ability across cultural contexts?
2.  Do children’s reaction times increase approximately linearly with angular disparity, consistent with @shepard1971mental?
3.  What latent cognitive strategies best characterize children’s mental-rotation performance, and how are these strategies related to task demands (angular disparity; 2D vs 3D stimuli) and site?

To our knowledge, LEVANTE is the first study to use item response theory (IRT) models to develop a mental rotation task (see @Kachergis2025LEVANTEcoretasks for further detail). IRT comprises a class of models that estimate, within a single framework, both item properties (e.g., how difficult each item is) and person ability (e.g., each child’s latent skill) [@embretson2013item]. Once fitted, an IRT model yields item parameters that can be used to place a new test-taker on the same latent scale from their response pattern, even if they completed only a subset of items. This allows us to report directly comparable scores for younger children who received mostly easier items and older children who received harder items.

# Methods

## Participants

Data for the present study were drawn from three pilot sites participating in LEVANTE: uniandes-co, mpieva-de, and western-ca. Collectively, these sites supported task localisation and technical implementation, iterative piloting, and deployment in varied testing contexts. The resulting pilot datasets enabled preliminary evaluation of construct validity, measurement invariance, test–retest reliability, and calibration for computer-adaptive administration, as well as descriptive assessment of age-related differences across the targeted constructs (refer to @Kachergis2025LEVANTEcoretasks for a full description).

### uniandes-co

The Universidad de Los Andes team carried out school-based data collection in Colombia, working with four schools in Bogotá and two public schools in the rural departments of Caquetá and Boyacá. Schools were selected pragmatically from institutions that had previously collaborated with the university. In Bogotá, three schools were affiliated with two family compensation funds and one was operated by a charter-school organisation. Across locations, participating schools largely served children from low-income and lower-middle-income backgrounds. Testing was conducted on tablets provided by the research team using a group-administration format, with group sizes adjusted by age (larger groups for older children; smaller groups for younger children).

### mpieva-de

At the mpieva-de site, the Max Planck Institute for Evolutionary Anthropology implemented a remote, family-based assessment protocol. Families were recruited via an existing registry of Leipzig residents and were contacted by telephone to invite participation. Families who enrolled were given login details and completed the assigned LEVANTE tasks at home using their own computer or tablet.

### western-ca

The Western University (Ontario) team recruited children from the surrounding community, primarily via online outreach (e.g., Facebook advertisements and posts in community groups), complemented by physical flyers and announcements distributed through local school newsletters. Families self-enrolled by registering their interest in a participant database. Children were assessed individually by a researcher in a dedicated on-campus testing space.

## Measure

The LEVANTE mental rotation task is adapted from @frick2013development. On each trial, children view a target image (e.g., a rabbit) and then choose between two alternatives (e.g., two rabbit silhouettes): one alternative is the same object shown at a different orientation, whereas the other is a non-matching foil. Children indicate which option corresponds to the target, and we record both accuracy and response time. Angular disparity is manipulated across trials (e.g., 40°, 80°, 120°, 160°), producing systematic effects on performance. Stimuli include 2D animal figures (ducks and rabbits) and more challenging 3D geometric shapes drawn from the classic Shepard and Metzler (1971) set.

```{r}
#| label: fig-mrot-example-items
#| fig-cap: "Example items for 2D and 3D rotation"
#| out-width: "100%"
#| fig-align: "center"

knitr::include_graphics("mrot_item_examples.pdf")
```

All LEVANTE tasks including mental rotation were originally developed in English and later adapted for Spanish and German through a structured internationalisation workflow carried out with collaborators at the uniandes-co and mpieva-de sites, respectively. We first produced draft translations of all verbal content—including instructions, item text, and recurring terminology (e.g., encouragement prompts)—using DeepL. These drafts were then refined by a professional translator to ensure grammatical accuracy and natural phrasing. Next, a site researcher who was a native speaker of the relevant language and dialect reviewed the materials to confirm clarity and suitability for the specific task context. To check for semantic and conceptual alignment with the English originals, we used AI-assisted back-translation to identify wording that required further revision and native-speaker review.

The mental rotation task has already been implemented in a computer-adaptive task (CAT) format. In the adaptive version, participant ability is summarized by a latent proficiency estimate (theta) that is re-estimated after each response. The updated theta then guides selection of the next item so that it is appropriately challenging for the individual. This trial-by-trial targeting typically makes the task more engaging and increases measurement efficiency, enabling reliable ability estimation with fewer administered items. Refer to @Kachergis2025LEVANTEcoretasks for more information on the CAT versions of all LEVANTE tasks.

## Analytical techniques \[TO ADD\]

\clearpage

# Results

## Cultural Similarities - cross-site measurement invariance of mental rotation

While we use psychometric modelling to place children’s performance on a common scale, the LEVANTE project is not designed to estimate absolute differences between sites. In practice, LEVANTE does not support site-level performance comparisons: sites vary substantially in sampling, recruitment, and test administration, so any apparent cross-site differences in raw performance are heavily confounded. Accordingly, although we report descriptive patterns that may look like absolute site differences, we emphasize that these should not be interpreted as meaningful.

Because our data were collected across three sites, we use multi-group IRT to evaluate measurement invariance, i.e., whether the task functions similarly across groups [@putnick2016measurement]. Multi-group models for the mental rotation task supported scalar invariance under a 2PL model, indicating that item difficulty is aligned across sites and that children in each country are being placed on a comparable latent scale. Practically, this means the groups share the same relative ordering of item difficulty—for example, items with larger rotation angles are consistently more difficult for children across all three countries (see @tbl-mrot-2pl-scalar). These results suggest that the core structure of mental rotation performance is largely shared across cultures in our sample: the task appears to tap the same underlying ability, and the same item features drive difficulty in a similar way across contexts. This provides a psychometric foundation for the broader claim that patterns of mental rotation development are more alike than different across the cultural settings represented here.

```{r}
#| include: false

# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(factoextra)
```

```{r}
#| label: setup
#| include: false
#| cache: false

source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)
```

```{r}
#| label: irt-anova
#| include: false

# IRT model output - import object
github_dir <- file.path(path.expand("~"), "Documents", "GitHub")
task_anova_path <- file.path(github_dir, "levante-pilots", "02_scoring_outputs", "task_anova.rds")

stopifnot(file.exists(task_anova_path))
task_anova <- readRDS(task_anova_path)
#colnames(task_anova)
#table(task_anova$itemtype)

# extract
mrot_2pl_scalar <- task_anova %>%
  filter(task == "mrot", itemtype == "2pl", invariance == "scalar")
mrot_2pl_scalar
```

```{r}
#| label: tbl-mrot-2pl-scalar
#| tbl-cap: "Model fit indices for 2PL scalar invariance models. Note: 2PL = 2-parameter logistic model; scalar invariance indicates that item intercepts and loadings are constrained across groups."
#| echo: false

mrot_2pl_scalar %>%
  transmute(
    Task       = task,
    Factors    = nfact,
    Model      = itemtype,
    Invariance = invariance,
    logLik     = logLik,
    AIC        = AIC,
    BIC        = BIC,
    `# params` = df
  ) %>%
  gt() %>%
  # Ensures the gt title area is clear and doesn't overlap the Quarto caption
  tab_header(title = md("")) %>% 
  fmt_number(columns = c(logLik, AIC, BIC), decimals = 2) %>%
  fmt_number(columns = `# params`, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Task, Factors, Model, Invariance, `# params`)) %>%
  cols_align(align = "right",  columns = c(logLik, AIC, BIC)) %>%
  tab_options(
    #table.font.size = px(12), # Increased slightly for better legibility
    table.width = pct(100),   # Ensures it fills the page width nicely
    data_row.padding = px(3)   # Tighter row spacing often looks better for stats
  )
```

\clearpage

@fig-dev-trends plots participants’ mental rotation ability estimates (IRT theta) against age, with separate panels/curves by site and age trends modelled using generalized additive models (GAMs). GAMs allow the developmental association with age to be non-linear, capturing periods of more rapid change and relative flattening without imposing a fixed form. Across sites, theta increases with age, indicating developmental improvement in mental rotation ability, and the dispersion of points within age bands underscores substantial heterogeneity in ability among same-aged children.

```{r}
#| label: fig-dev-trends
#| fig-cap: "Developmental trends in Mental Rotation ability by site."
#| fig-width: 8
#| fig-height: 6

# colnames(all_scores)
# table(all_scores$model)
mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!site_labels, .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

mrot_irt <- ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  labs(
    #title = "Developmental trends in Mental Rotation ability by site",
    subtitle = "Ability estimates smoothed with GAM",
    x = "Age (years)",
    y = "Ability (theta)",
    colour = "Site"
  )
mrot_irt
```

\clearpage

## Angular disparity and the RT–angle relationship

Our aim with this analysis was to test whether the canonical reaction time–rotation angle relationship first reported by @shepard1971mental replicates in our data. @fig-rt-angle-per-run plots mean reaction times (±95% confidence intervals) as a function of rotation angle, stratified by age group (5–7, 7–9, and 9–12 years) and site, with separate series by stimulus and item type (2D vs 3D). To show the underlying dispersion and sample density alongside the summary trends, we provide a second version of @fig-rt-angle-per-run in the Appendix that overlays the raw per-run data (using the same age bins and 95% CIs; see @fig-rt-angle_raw).

Across sites, mean RTs generally decrease with age, with the 9–12 group responding faster overall than the younger groups, particularly for the 2D animal stimuli (duck/rabbit). RTs also tend to increase with rotation angle (angular disparity), consistent with the classic RT–angle effect (@shepard1971mental). This angular-disparity effect is most pronounced for the 3D shape stimuli, which show slower overall RTs and steeper angle-related increases, whereas the 2D stimuli exhibit comparatively flatter RT–angle functions. Despite some differences in absolute RT levels between sites, the same qualitative pattern—faster responses with age and a stronger angular-disparity effect for 3D items—appears across all three sites, suggesting broadly comparable task behaviour across contexts.

```{r}
#| label: trial-level-rts
#| include: false

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%
  tidyr::unnest(data) %>%
  dplyr::mutate(
    correct = dplyr::case_when(
      is.logical(correct) ~ as.integer(correct),
      TRUE ~ as.integer(as.numeric(correct))
    ),
    rt_s   = as.numeric(rt_numeric) / 1000,
    log_rt = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))
```

```{r}
#| label: remove-retest-data
#| include: false

baseline_runs <- mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%  
  group_by(site, user_id) %>%
  arrange(age, run_id) %>%           
  slice(1) %>%
  ungroup() %>%
  select(site, user_id, run_id)

# keep only trials from those baseline runs
mrot_rt_trial_no_retest <- mrot_rt_trial %>%
  semi_join(baseline_runs, by = c("site", "user_id", "run_id"))

# check: each child should now have exactly 1 run (within site)
mrot_rt_trial_no_retest %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs)

# how many re-testers?
mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs) # 133
```

```{r}
#| label: fig-rt-angle-per-run
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 7, 9, 12)

mrot_items_rt_runs <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = dplyr::n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  )
p_mrot_rt_angle_age_1
```

\clearpage

## Within-Person Dynamics

Having established the expected angular-disparity effect across age groups and sites, we next examined how RTs evolved within individuals over the course of the task. Because item difficulty increased adaptively—such that higher-performing children were more likely to encounter the harder 3D items—we analysed trial-by-trial RT trajectories across the 80 trials to distinguish item-driven changes from more gradual time-on-task effects (e.g., fatigue). @fig-within-person-trials illustrates a clear 'jump' in RT about halfway through the task (trial 48), corresponding to the transition from 2D to 3D stimuli, which required significantly more processing time across all three site populations. We observed consistent response acceleration within both the 2D and 3D blocks, as children became faster with practice.

```{r}
#| label: fig-within-person-trials
#| fig-cap: "Trial-level Reaction Time Dynamics. The plot shows the mean log-reaction time across the 80 trials for each site. Shaded areas represent standard error. The sharp increase represents the transition to 3D items (Note: Re-test data removed)."
#| fig-width: 8
#| fig-height: 5

# 1. Prepare data with Site Labels
trial_plot_data <- mrot_rt_trial_no_retest %>%
  filter(!is.na(log_rt), correct == 1) %>% # Focus on correct trials for valid RT
  left_join(run_data %>% select(user_id, run_id, site), by = c("user_id", "run_id")) %>%
  mutate(site_label = dplyr::recode(site.x, !!!site_labels))


ggplot(trial_plot_data, aes(x = trial_number, y = log_rt, color = site_label, fill = site_label)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Trial Number (Time)",
    y = "Log RT (seconds)",
    color = "Site",
    fill = "Site"
  ) +
  # REMOVED theme_classic()
  theme(legend.position = "bottom")

```

\clearpage

## Cognitive Strategies

To characterize individual differences in how children solved the mental-rotation task, we derived participant-level 'strategy features; from trial-level reaction times. Consistent with the generally log-normal distribution of reaction times, we modeled log-transformed RTs. While the classic @shepard1971mental finding describes an approximately linear relationship with raw RT, fitting the model to log(RT) provides more robust parameter estimates for defining individual strategy profiles in a developing population. We fit a mixed-effects model to log(RT) that captured stimulus and angle-dependent responding alongside within-run practice effects, extracted participant-specific deviations (random effects), and combined these model-derived quantities with RT variability and IRT-estimated ability before clustering participants using k-means (k = 3) on standardized features.

The extracted features were: (1) Baseline Speed, defined as the intercept from the mixed-effects model (interpretable as expected speed at 0° rotation); (2) Rotation Slope, defined as the “cost” of rotation (the slope of angle on log RT), estimated as the average slope across stimulus types; (3) Response Variability (RT SD), defined as the intra-individual standard deviation of log RT; and (4) IRT-estimated ability (theta).

```{r}
#| label: strategy-analysis-prep
#| results: "hide"

# Ability estimate - keep site + user
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# Prepare trial-level data for the rich model
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    angle    = as.integer(str_extract(item, "\\d+")),
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(angle)),
    trial_s  = as.numeric(scale(trial_number))
  )

# --- Doria's FIX: Set Sum-to-Zero Contrasts ---
# By default, R uses treatment contrasts (Duck = reference). 
# This makes the "Angle" slope represent only Ducks (which is flat for everyone).
# Setting sum contrasts ensures the "Angle" slope represents the GRAND MEAN slope across all stimuli.
contrasts(mrot_rich_data$stimulus) <- contr.sum(levels(mrot_rich_data$stimulus))
# Fit the rich mixed model
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# Extract subject-level random effects (deviations)
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# Fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# Compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    # Baseline Speed: Speed at angle 0 (Intercept)
    Baseline_Speed = fix_int + int_dev,
    # Absolute Slope: Time added per degree of rotation (Rotation Cost)
    Abs_Slope      = fix_ang + slope_dev,
    # Relative Slope (Per Mike's suggestion)
    # "Is rotation expensive relative to their normal speed?"
    Rel_Slope      = Abs_Slope / Baseline_Speed,
    # Learning Rate 
    Learning_Rate  = fix_tri + trial_dev
  )
# 4. Compute Instability (Residual SD)
# We calculate SD of the *residuals* (not raw RT) to remove the "Task Difficulty" confound.
# This isolates internal inconsistency from the fact that some kids saw harder items.
mrot_rich_data$resid <- residuals(m_rich)

instability_data <- mrot_rich_data %>%
  group_by(site, user_id) %>%
  summarise(
    # Instability = SD of residuals (pure internal noise)
    Instability = sd(resid, na.rm = TRUE), 
    .groups = "drop"
  )

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

# 5. Join and Prepare for Clustering
rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(instability_data, by = c("site", "user_id")) %>% 
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  # Filter to ensure all needed features are finite
  # UPDATED: Added Learning_Rate to the list
  filter(if_all(c(ability, Abs_Slope, Rel_Slope, Baseline_Speed, Instability, Learning_Rate), is.finite))
```

```{r}
#| label: clustering-execution
#| include: false

# Define the colors matching the NEW labels
my_colors <- c(
  "Ineffective Strategies" = "#e74c3c",
  "Other Effective Strategies (Flipping/Matching)" = "#3498db",
  "Mental Rotation" = "#2ecc71"
)

# --- Step 1: Set up Clustering Input ---
# Filter out any missing rows first
# This creates the 'analysis_data' object that was missing
analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Rel_Slope, Abs_Slope, 
                  Instability, Learning_Rate), is.finite))

# Global K-means (k=3)
global_input <- analysis_data %>%
  select(Baseline_Speed, Rel_Slope, Abs_Slope, 
         Instability, Learning_Rate) %>%
  scale()

# --- Step 2: Run K-Means ---
set.seed(123)
k_global <- kmeans(global_input, centers = 3, nstart = 25)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

# --- Step 3: Identify Clusters via Centroids ---
cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability),
    mean_slope   = mean(Rel_Slope), 
    .groups = "drop"
  )

# A. Identify "Ineffective" -> Lowest Ability
low_eff_id <- cluster_stats %>%
  arrange(mean_ability) %>% 
  slice(1) %>% 
  pull(Global_Cluster)

# B. Identify "Effective" groups -> Split remaining by Slope
effective_clusters <- cluster_stats %>%
  filter(Global_Cluster != low_eff_id) %>%
  arrange(mean_slope) 

# Analytical/Flipping = Flatter Slope
analytical_id <- effective_clusters %>%
  slice(1) %>% 
  pull(Global_Cluster)

# Rotation = Steeper Slope
holistic_id <- effective_clusters %>%
  slice(2) %>% 
  pull(Global_Cluster)

# --- Step 4: Apply NEW Labels (Mike's Feedback) ---
analysis_data <- analysis_data %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id ~ "Ineffective Strategies",
    Global_Cluster == analytical_id ~ "Other Effective Strategies (Flipping/Matching)",
    Global_Cluster == holistic_id ~ "Mental Rotation"
  ))

# --- 4. Plot (Optional check) ---
ggplot(analysis_data, aes(x = ability, y = Rel_Slope, color = Strategy_Label)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  labs(title = "Cluster Result: Ability vs. Relative Slope")
```

```{r}
#| label: fig-rt-angle-by-strategy
#| fig-cap: "Participant-weighted RT–angle functions by strategy cluster (numbers indicate unique participants contributing at each angle; shaded bands show 95% CIs) (Note: Re-test data removed)."

# Join strategy labels onto trial-level RT data
rt_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(
    Strategy_Label = factor(Strategy_Label, levels = names(my_colors))
  )

# Person-level mean logRT per angle (each user contributes once per angle)
rt_user_angle <- rt_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(
    user_mean_log_rt = mean(log_rt, na.rm = TRUE),
    .groups = "drop"
  )

# Strategy-level mean across users at each angle + SE across users
rt_angle_person_summary <- rt_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# check colours
missing_cols <- setdiff(levels(rt_angle_person_summary$Strategy_Label), names(my_colors))
stopifnot(length(missing_cols) == 0)

ggplot(rt_angle_person_summary, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(
      ymin = mean_log_rt - 1.96 * se_log_rt,
      ymax = mean_log_rt + 1.96 * se_log_rt,
      fill = Strategy_Label
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean log(RT) (participant-weighted)",
    colour = "Strategy",
    fill = "Strategy"
  ) +
  theme(legend.position = "bottom")

```

```{r}
#| label: descriptive-validation-stats
#| tbl-cap: "Descriptive statistics and ANOVA results by Strategy Cluster."



# --- 1. Run Statistical Tests (ANOVA + Tukey) ---
# Function to get compact letters for significance (e.g., "a", "b", "a")
get_sig_letters <- function(variable, data) {
  model <- aov(as.formula(paste(variable, "~ Strategy_Label")), data = data)
  tukey <- TukeyHSD(model)
  # This is a simplified way to visualize; for the table, we usually just show means.
  # But the ANOVA p-value is useful.
  summary(model)[[1]][["Pr(>F)"]][1]
}

# Calculate Descriptive Table
validation_table <- analysis_data %>%
  group_by(Strategy_Label) %>%
  summarise(
    N = n(),
    # Ability
    Mean_Ability = mean(ability, na.rm = TRUE),
    SD_Ability   = sd(ability, na.rm = TRUE),
    
    # Relative Slope
    Mean_Slope   = mean(Rel_Slope, na.rm = TRUE),
    SD_Slope     = sd(Rel_Slope, na.rm = TRUE),
    
    # Instability
    Mean_Instab  = mean(Instability, na.rm = TRUE),
    .groups = "drop"
  )

validation_table %>%
  transmute(
    Strategy      = Strategy_Label,
    N             = N,
    `Mean Ability`= Mean_Ability,
    `SD Ability`  = SD_Ability,
    `Mean Slope`  = Mean_Slope,
    `SD Slope`    = SD_Slope,
    `Instability` = Mean_Instab
  ) %>%
  gt() %>%
  # Ensures the gt title area is clear and doesn't overlap the Quarto caption
  tab_header(title = md("")) %>% 
  # Format statistics to 2 decimal places
  fmt_number(
    columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability), 
    decimals = 2
  ) %>%
  # Format N as integer (no decimals)
  fmt_number(
    columns = c(N), 
    decimals = 0, 
    use_seps = FALSE
  ) %>%
  # Align text/integers to center
  cols_align(
    align = "center", 
    columns = c(Strategy, N)
  ) %>%
  # Align statistics to right (standard for numbers)
  cols_align(
    align = "right",  
    columns = c(`Mean Ability`, `SD Ability`, `Mean Slope`, `SD Slope`, Instability)
  ) %>%
  # Apply the exact formatting options from your example
  tab_options(
    table.width = pct(100),   # Ensures it fills the page width nicely
    data_row.padding = px(3)   # Tighter row spacing
  )

# --- 2. Visualization with Significance Bars ---
# We define the comparisons we want to test visually
# UPDATED NAMES TO MATCH NEW LABELS
my_comparisons <- list(
  c("Ineffective Strategies", "Other Effective Strategies (Flipping/Matching)"),
  c("Ineffective Strategies", "Mental Rotation"),
  c("Other Effective Strategies (Flipping/Matching)", "Mental Rotation")
)

# Plot A: 
p_ability <- ggplot(analysis_data, aes(x = Strategy_Label, y = ability, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  # Add Statistical Test (Kruskal-Wallis or ANOVA/T-test)
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "1. Defined by Ability", y = "Ability (Theta)", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 15, hjust = 1))

# Plot B: Validation of Slope 
p_slope <- ggplot(analysis_data, aes(x = Strategy_Label, y = Rel_Slope, fill = Strategy_Label)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = my_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  # Add Statistical Test
  stat_compare_means(comparisons = my_comparisons, method = "t.test", 
                     label = "p.signif", hide.ns = TRUE, vjust = 0.5) +
  labs(title = "2. Distinguished by RT Slope", y = "Relative Slope", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 15, hjust = 1))

# Combine plots
p_ability + p_slope
```
### Strategy Stability Analysis (Split-Half Reliability)

To investigate whether children consistently use a single cognitive strategy throughout the task or switch strategies over time (e.g., discovering a shortcut), we performed a **Split-Half Reliability Analysis**.

**Methodology:**
We partitioned each participant's data into two sequential halves: the first 50% of trials (Early Phase) and the last 50% of trials (Late Phase). For each half, we estimated a "Raw Slope" using a simple linear regression of reaction time on angle (`log_rt ~ angle`). This slope serves as a proxy for the participant's rotation strategy during that specific period. 

We then correlated the *First Half Slope* with the *Second Half Slope* to quantify stability. A high correlation ($r \approx 1$) would indicate a fixed strategy, while a low correlation ($r \approx 0$) suggests dynamic behavior or strategy switching.

```{r}
#| label: fig-strategy-switching
#| fig-cap: "Strategy Stability: Comparison of Rotation Slope in First Half vs. Second Half of trials. Points below the diagonal suggest a switch to a flatter strategy (e.g., discovering the shortcut)."


# 1. Split Data into Halves
# We determine the midpoint trial for each user
split_data <- mrot_rich_data %>%
  group_by(user_id) %>%
  mutate(
    Midpoint = median(trial_number, na.rm = TRUE),
    Half = ifelse(trial_number <= Midpoint, "First_Half", "Second_Half")
  ) %>%
  ungroup() %>%
  # Filter out bad data points just to be safe
  filter(is.finite(log_rt), is.finite(angle))

# 2. Calculate Simple Slope for Each Half (Using simple lm per subject)
# This is a fast proxy for the complex mixed-model slope
slope_stability <- split_data %>%
  group_by(user_id, Half) %>%
  summarise(
    # We only care about the slope (coefficient for angle)
    # We use a safe lm() call wrapped in tryCatch to handle kids with too few trials
    Raw_Slope = tryCatch(
      coef(lm(log_rt ~ angle, data = pick(everything())))["angle"], 
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  # Reshape to Wide Format (Columns: Slope_First, Slope_Second)
  tidyr::pivot_wider(names_from = Half, values_from = Raw_Slope, names_prefix = "Slope_") %>%
  # Join with our Final Strategy Labels to color-code
  left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id") %>%
  filter(!is.na(Strategy_Label), !is.na(Slope_First_Half), !is.na(Slope_Second_Half))

# 3. Calculate Correlation (Quantify Stability)
correlation <- cor(slope_stability$Slope_First_Half, slope_stability$Slope_Second_Half, use = "complete.obs")

# 4. Visualization
ggplot(slope_stability, aes(x = Slope_First_Half, y = Slope_Second_Half, color = Strategy_Label)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") + # The "Stability Line"
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = my_colors) +
  labs(
    title = "Strategy Switching Check (Split-Half Reliability)",
    subtitle = paste0("Correlation between halves: r = ", round(correlation, 2), 
                      ". Points below diagonal = Strategy became flatter (faster)."),
    x = "Slope (First Half of Trials)",
    y = "Slope (Second Half of Trials)"
  ) +
  theme_minimal() +
  coord_fixed() # Keeps the aspect ratio square so the diagonal is 45 degrees
```

### Interpretation of Results

The analysis revealed a near-zero correlation between the first and second half slopes ($r = -0.08$), as shown in @fig-strategy-switching. This challenges the assumption that cognitive strategy is a fixed trait, suggesting instead that it is a dynamic process for many children.

**Cluster-Specific Behaviors:**

1.  **The "Discovery" Effect:**
    A subset of participants (primarily in the bottom-right quadrant) exhibited a steep slope in the first half but a flat slope in the second half. This pattern is consistent with children who initially attempted mental rotation but "discovered" the feature-matching shortcut halfway through the task, causing their response profile to flatten.

2.  **Stability of Analytical Matchers:**
    The **Analytical (Feature Matching)** group (blue) is clustered tightly near the center $(0,0)$. This indicates high stability; once children adopt the efficient heuristic of feature matching, they rarely abandon it to return to the more cognitively demanding rotation strategy.

3.  **Instability of Low Efficacy:**
    The **Low Efficacy** group (red) is scattered widely across the plot with no consistent trend. This chaotic variability confirms the "indeterminable" nature of this group, suggesting they oscillate between guessing, failed rotation attempts, and disengagement throughout the task.

**Conclusion:**

Unlike the stable 'Analytical' group, the 'Holistic' and 'Low Efficacy' groups demonstrate significant trial-to-trial variation. The lack of correlation ($r = -0.08$) indicates that strategy use is fluid; many children who initially attempt mental rotation may abandon it for heuristics or shortcuts as the task progresses.

```{r}
#| label: k-sweep-validation
#| eval: false
#| fig-cap: Cluster validation metrics:Elbow Method (left) and Silhouette Analysis (right). A 'bend' at k=3 or a peak in silhouette at k=3 supports the 3-cluster solution.
#| include: false


# Ensure your input is ready (Same input as your final model)
# global_input <- analysis_data %>% select(ability, Baseline_Speed, Rel_Slope, Instability) %>% scale()

# --- Method A: The Elbow Plot (Total Within-Cluster Sum of Squares) ---
# Logic: We want to minimize 'WSS' (compactness), but avoiding overfitting.
# Look for the "Elbow" or "Knee" — the point where the curve flattens out.
p_elbow <- fviz_nbclust(global_input, kmeans, method = "wss", k.max = 8) +
  geom_vline(xintercept = 3, linetype = 2, color = "red") + # Mark our theoretical choice
  labs(title = "Elbow Method (WSS)", subtitle = "Look for the 'Bend'") +
  theme_minimal()

# --- Method B: The Silhouette Plot (Separation Quality) ---
# Logic: Higher is better. 
# It measures how similar an object is to its own cluster vs. other clusters.
p_sil <- fviz_nbclust(global_input, kmeans, method = "silhouette", k.max = 8) +
  labs(title = "Average Silhouette Width", subtitle = "Higher is Better") +
  theme_minimal()

# Display Side-by-Side
grid.arrange(p_elbow, p_sil, nrow = 1)
```

```{r}
#| label: k2-cluster-check
#| eval: false
#| fig-cap: RT-Angle curves for the K=2 solution. Note how the 'High Ability' group likely lumps Rotators and Matchers together.
#| include: false

# --- 1. Set up Input (Same as before, removing Learning_Rate for clarity) ---
analysis_data_k2 <- rich_features_final %>%
  filter(if_all(c(ability, Baseline_Speed, Rel_Slope, Instability), is.finite))

global_input_k2 <- analysis_data_k2 %>%
  select(ability, Baseline_Speed, Rel_Slope, Instability) %>%
  scale()

# --- 2. Run K-Means with K=2 ---
set.seed(123)
k_global_2 <- kmeans(global_input_k2, centers = 2, nstart = 25)
analysis_data_k2$Global_Cluster <- as.factor(k_global_2$cluster)

# --- 3. Label the 2 Clusters ---
# We define them simply by Ability (Low vs High) since we can't split by slope yet
cluster_stats_k2 <- analysis_data_k2 %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_ability = mean(ability), 
    .groups = "drop"
  )

# Identify Lower Ability Cluster
low_eff_id <- cluster_stats_k2 %>% arrange(mean_ability) %>% slice(1) %>% pull(Global_Cluster)
# The other one is High Efficiency
high_eff_id <- cluster_stats_k2 %>% arrange(desc(mean_ability)) %>% slice(1) %>% pull(Global_Cluster)

analysis_data_k2 <- analysis_data_k2 %>%
  mutate(Strategy_Label = case_when(
    Global_Cluster == low_eff_id ~ "Low Efficiency (Cluster 1)",
    Global_Cluster == high_eff_id ~ "High Efficiency (Cluster 2)"
  ))

# --- 4. Plot the Curves (Visual Check) ---
# Join labels to raw data
rt_by_strategy_k2 <- mrot_rich_data %>%
  left_join(
    analysis_data_k2 %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle))

# Calculate User Means
rt_user_angle_k2 <- rt_by_strategy_k2 %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(user_mean_log_rt = mean(log_rt, na.rm = TRUE), .groups = "drop")

# Calculate Group Means
rt_summary_k2 <- rt_user_angle_k2 %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# Plot
ggplot(rt_summary_k2, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_log_rt - 1.96*se_log_rt, ymax = mean_log_rt + 1.96*se_log_rt, fill = Strategy_Label), alpha = 0.2, colour = NA) +
  labs(title = "Cluster Result (K=2)", 
       subtitle = "Notice how the 'High Efficiency' group averages out the strategy differences",
       y = "Mean log(RT)", x = "Angle") +
  theme_minimal() +
  theme(legend.position = "bottom")
```



\newpage

# References

::: {#refs}
:::
    
# IRT Ability by Age {#apx-fig-rt-angle .appendix}
    
```{r}
#| label: fig-rt-angle_raw
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI."
#| fig-width: 8
#| fig-height: 8
  
  # Add raw data underneath
  p_mrot_rt_angle_age_2 <- ggplot() +
    # raw per-run medians
    geom_point(
      data = mrot_items_rt_runs,
      aes(x = angle, y = rt_med,
          colour = stimulus),
      size = 0.4,
      alpha = 0.25,
      position = position_jitter(width = 3, height = 0)
    ) +
    # summary CI + lines
    geom_errorbar(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y, ymin = lwr, ymax = upr,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      width = 2,
      linewidth = 0.35,
      alpha = 0.6
    ) +
    geom_point(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      size = 1
    ) +
    geom_line(
      data = mrot_items_rt_agebin_ci,
      aes(x = angle, y = y,
          colour = stimulus, linetype = item_group,
          group = interaction(item_group, stimulus)),
      alpha = 0.85
    ) +
    facet_grid(site ~ age_group) +
    labs(
      #title = "Mental Rotation: RT by angle within age groups ±95% CI",
      x = "Rotation angle (°)",
      y = "RT (s)",
      colour = "Stimulus",
      #linetype = "Item Type"
    )
  p_mrot_rt_angle_age_2
```
  
\clearpage
  
::: landscape
# Cluster Bootstrap plots {#apx-cluster-bootstrap-plots .appendix}
  
```{r}
#| label: fig-cluster-bootstrap-plots
#| cache: true
#| fig-width: 11
#| fig-height: 4.5
#| fig-align: center

  
  p_ari <- ggplot(res, aes(x = ARI)) +
    geom_histogram(bins = 30) +
    labs(
      x = "Adjusted Rand Index (vs reference labels)",
      y = "Count",
      title = paste0("Cluster stability under bootstrap resampling (B = ", B, ")")
    )
  
  # long-form Jaccard with safe mapping back to Strategy_Label
  jac_key <- tibble(
    col = paste0("J_", make.names(ref_levels)),
    Strategy_Label = ref_levels
  )
  
  res_long <- res %>%
    pivot_longer(starts_with("J_"), names_to = "col", values_to = "Jaccard") %>%
    left_join(jac_key, by = "col") %>%
    mutate(Strategy_Label = factor(Strategy_Label, levels = ref_levels))
  
  p_jac <- ggplot(res_long, aes(x = Strategy_Label, y = Jaccard)) +
    geom_boxplot(outlier.alpha = 0.25) +
    labs(
      x = NULL,
      y = "Jaccard (membership overlap vs reference)",
      title = "Per-cluster membership stability"
    ) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
    theme_minimal()
  
  (p_ari | p_jac)
```

```{r}
#| label: fig-tsne-check
#| fig-cap: "t-SNE Visualization of Strategy Clusters. The distinct separation between 'Mental Rotation' and 'Other Effective' suggests distinct cognitive mechanisms, while 'Ineffective' participants form a more dispersed cloud."

library(Rtsne) # Make sure to install this package

# 1. Prep data (t-SNE hates duplicates)
tsne_input <- analysis_data %>%
    select(user_id, ability, Baseline_Speed, Rel_Slope, Abs_Slope, Instability) %>%
    distinct() %>% 
    na.omit()

# 2. Run t-SNE
set.seed(42) 
tsne_result <- Rtsne(
    as.matrix(tsne_input %>% select(-user_id)), 
    dims = 2, 
    perplexity = 30, 
    verbose = FALSE, 
    max_iter = 500
)

# 3. Plot
tsne_plot_data <- tsne_input %>%
    mutate(
        tSNE_1 = tsne_result$Y[,1],
        tSNE_2 = tsne_result$Y[,2]
    ) %>%
    left_join(analysis_data %>% select(user_id, Strategy_Label), by = "user_id")

ggplot(tsne_plot_data, aes(x = tSNE_1, y = tSNE_2, color = Strategy_Label)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values = my_colors) + # Ensure my_colors names match new labels!
    labs(
        x = "Dimension 1",
        y = "Dimension 2"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

:::
