---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "Fionnuala O'Reilly"
    orcid: 0000-0002-4355-9088
    affiliations:
      - ref: stanford
  - name: "Xingyao Xiao"
    orcid: 0000-0001-8430-0438
    affiliations:
      - ref: stanford
  - name: "Rebecca Zhu"
    orcid: 0000-0003-0957-5962
    affiliations:
      - ref: stanford
  - name: "Mika Braginsky"
    orcid: 0000-0001-9039-3220
    affiliations:
      - ref: stanford
  - name: "Nilàm Ram"
    orcid: 0000-0003-1671-5257
    affiliations:
      - ref: stanford
  - name: "Benjamin W. Domingue"
    orcid: 0000-0002-3894-9049
    affiliations:
      - ref: stanford
  - name: "Michael C. Frank"
    orcid: 0000-0002-7551-4378
    affiliations:
      - ref: stanford
    corresponding: true
    email: mcfrank@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["cognitive development", "mental rotation", "spatial reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    # fig-pos: H        
    # tbl-pos: H  
    

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

Spatial reasoning encompasses a range of abilities for generating, retaining, and manipulating visuospatial information [@lohman2013spatial]. One core component is mental rotation, which involves mentally transforming an object by rotating it in space to predict how it would appear from a different orientation or viewpoint. More broadly, spatial skills support the formation and transformation of mental representations of 2D shapes and 3D objects (e.g., rotating, bending, or folding them) and help us track how spatial relations between locations, shapes, and objects change over time [@frick2014development; @newcombe2014thinking]. In everyday life, we rely on these skills when judging whether an object will fit into a tight space, imagining how a room might look after rearranging furniture, or navigating using a map.

Spatial reasoning is also linked to a wide range of educational outcomes [@wai2018spatial], spanning creative domains such as music [@padulo2020effects; @hetland2000learning] and visual arts [@kell2013creativity; @winner2013relationship], as well as STEM subjects [@wai2009spatial]—and especially mathematics [@atit2022examining; @xie2020spatial; @mix2012relation; @gunderson2012relation; @frick2019spatial]. Mental rotation, the focus of this paper, is one spatial skill that has been repeatedly implicated in these associations, making it a useful target for understanding how spatial reasoning relates to learning and development.

Despite its practical and educational importance, significant questions remain regarding how mental rotation varies across contexts. In particular, little is known about whether cultural background shapes children’s mental-rotation skill, whether the canonical linear relationship between reaction time and angular disparity replicates across diverse populations, or which cognitive strategies children employ when solving mental-rotation problems. The present study addresses these gaps by examining children’s mental-rotation response patterns in Canada, Germany, and Colombia.

## Emergence and developemnt of spatial skills during childhood

Mental rotation can be observed as early as three months of age [@johnson2020spatial], with rapid development between ages three and five [@frick2013development]. Individual differences are evident early, and a relatively large sex difference in mental rotation performance has been reported in multiple studies [@boone2017sex; @linn1985emergence], with males, on average, outperforming females. However, the mechanisms driving developmental change—and the extent to which performance reflects qualitatively different solution strategies rather than a single underlying ability—remain open questions, particularly in culturally diverse samples.

A growing body of work suggests that spatial skills are malleable, with training studies in children showing gains on mental rotation and broader spatial reasoning measures, and in some cases transfer beyond the trained task [@hawes2015effects; @gilligan2020first; @lowrie2017visuospatial]. Consistent with this, a meta-analysis of 217 studies reported a moderate average effect of spatial training versus control conditions (Hedges’ g = 0.47, SE = 0.04) [@uttal2013malleability].

## Cultural Comparisons

From a biological perspective, there are strong reasons to expect cross-cultural similarities in mental rotation skills. Humans share a common genetic heritage and sensory-motor systems that shape how we perceive and interact with space. Because the nervous system likely evolved in response to universal demands, such as navigating environments and learning spatial layouts, spatial cognition is considered a relatively conserved neural specialization [@lewis1988signs; @montello1995significant; @nazareth2019meta].

While research on cultural variations in mental rotation is limited, existing studies highlight key predictors of performance. A common thread of these studies is their exploration of sex differences in performance to understand if the male advantage [@linn1985emergence; @maeda2013meta] in mental rotation is universal. For example, a large-scale multicultural study of 3,367 students in Canada, Germany, and Japan found that males outperformed females with a consistently large effect size across all three nations. However, the study also identified academic program as an independent predictor; students in science and engineering scored higher than those in the arts and social sciences, regardless of sex [@peters2006mental].

More recent developmental research suggests that these gender gaps may be more plastic than previously thought, particularly when cultural norms regarding autonomy are shifted. Among Tsimané children in the Bolivian Amazon, the high degree of freedom to explore allowed to both sexes resulted in a notable absence of gender differences in mobility and spatial ability [@davis2019spatial]. Interestingly, while formal schooling in this population improved abstract skills like mental rotation, it correlated with a decrease in large-scale navigational accuracy, suggesting a trade-off between classroom-based learning and environmental exploration. This environmental influence is further evidenced in seminomadic communities in Namibia. While local adults traditionally show a gender gap in spatial skills, recent cultural shifts—such as formal schooling and increased travel requirements—have effectively eliminated these differences in children [@davis2021cultural]. Both boys and girls demonstrated comparable navigational accuracy and mental rotation performance, indicating that when cultural norms provide equal opportunities for exploration and education, traditional sex differences in spatial cognition can disappear.

## Reaction time as a function of rotation angle

One of the most cited findings in psychology is that of @shepard1971mental, who demonstrated that mental rotation operates in a spatially continuous manner: response times increased approximately linearly with the angular disparity between two objects. This suggests that participants judge whether an object matches a previously seen one by performing a gradual rotation of an internal representation, taking longer for larger disparities (e.g., 75° vs. 110°).

Extending this work from the perspective of visual cognition, @tarr1989mental demonstrated that mental rotation is not only a mechanism for comparing two simultaneous stimuli, but a fundamental process integrated into long-term object recognition. While @shepard1971mental focused on "handedness" (the ability to distinguish a shape from its mirror image), @tarr1989mental showed that this same linear increase in response time occurs when identifying a single shape from memory. Their findings suggest that the human visual system is highly adaptive: as individuals gain experience with an object at new orientations, they store additional "viewpoint-specific" representations. Consequently, mental rotation does not always return to a single "upright" canonical view; instead, the internal representation is rotated to the nearest familiar orientation stored in memory.

Examining reaction times across diverse populations, a cross-cultural comparison of children in the Ivory Coast and Switzerland revealed that while both groups utilized the same mental rotation strategies, Swiss children demonstrated significantly higher speed and accuracy [@nunez1998mental]. Interestingly, both groups exhibited an identical, non-linear dip in performance at 30° and 330° angles, suggesting there could be a more universal cognitive constraint in spatial processing that exists independently of cultural background. Notably, age predicted performance in the Swiss group—with older children showing marked improvements—whereas it had no significant effect on the performance of the Ivorean children [@nunez1998mental].

## Cognitive strategies

The literature has shifted from viewing mental rotation as a monolithic, linear process to understanding it as a strategic choice mediated by cognitive load and task demands. While early models focused on a universal sequence of encoding, rotation, and comparison [@cooper1973chronometric; @desrocher1995stimulus], more recent research identifies a holistic-analytic continuum of strategies that varies significantly across individuals [@gluck2003spatial].

At one end of this continuum is the holistic strategy, where the stimulus is transformed as a single, integrated mental representation. This is often referred to as the ‘pure rotation’ approach and is characterized by high efficiency, lower total gaze times, and fewer transitions between stimuli [@taragin2019mental; @tang2023eye]. Conversely, the ‘piecemeal’ or local-analytic strategy involves decomposing the object into sub-components for sequential transformation and comparison. This strategy is marked by a ‘switching’ eye-pattern, involving frequent back-and-forth saccades as the participant verifies individual features [@nazareth2019strategy; @fernandez2025inducing].

The selection of these strategies is heavily dictated by Visual Working Memory (VWM) capacity. @just1985cognitive first proposed that the piecemeal approach serves as a "workaround" for limited VWM; if an individual cannot maintain the entire 3D structure in mind, the representation falls apart, forcing them to rotate it piece-by-piece. This is further supported by @khooshabeh2013individual, who demonstrated that while high-spatial individuals default to holistic strategies, low-spatial individuals often remain ‘stuck’ in piecemeal processing regardless of task difficulty.

Furthermore, the "how" of rotation involves the frame of reference used. @zacks2008neuroimaging distinguishes between object-based transformations (rotating the object itself) and egocentric perspective-taking (imagining oneself moving around the object). This choice is often influenced by the environment; for instance, 3D Virtual Reality (VR) environments provide stereoscopic depth cues that facilitate holistic, object-based rotations, whereas 2D presentations often force a more cognitively taxing piecemeal approach due to ambiguous depth cues [@tang2023eye].

Successful performance is often defined by strategy flexibility—the ability to dynamically alternate between holistic rotation and analytic checking based on the specific demands of the stimulus [@nazareth2019strategy]. When task difficulty increases—such as with mirrored objects or rotations along skewed axes—flexible rotators can shift strategies to manage the increased cognitive load [@hsing2023using; @nolte2022rotational]. This suggests that observed differences in performance may not stem from an inherent inability to rotate, but rather from differences in strategy selection and the stability of those strategies over time [@geiser2008quantitative].

## The present study

The Learning Variability Network Exchange (LEVANTE) is an international research collaboration that supports coordinated data collection to better understand how children’s learning and development vary across ages and contexts. Partner sites administer a shared, openly available battery designed for a broad developmental range, and data are harmonized into a central repository. With support from the Jacobs Foundation, participating sites are funded to collect longitudinal data using the LEVANTE framework. Contributing researchers receive access to their site’s data first, followed by scheduled public releases for the wider research community. LEVANTE’s long-term aim is to build and openly share a large, high-quality dataset—alongside the measures themselves—capturing children’s development across diverse settings. In the present study, we use data for the mental rotation task only to answer three core research questoins:

1.  Does the mental-rotation task show evidence of cross-site measurement invariance, indicating that it taps a shared underlying ability across cultural contexts?
2.  Do children’s reaction times increase approximately linearly with angular disparity, consistent with @shepard1971mental?
3.  What latent cognitive strategies best characterize children’s mental-rotation performance, and how are these strategies related to task demands (angular disparity; 2D vs 3D stimuli) and site?

To our knowledge, LEVANTE is the first study to use item response theory (IRT) models to develop a mental rotation task (see @Kachergis2025LEVANTEcoretasks for further detail). IRT comprises a class of models that estimate, within a single framework, both item properties (e.g., how difficult each item is) and person ability (e.g., each child’s latent skill) [@embretson2013item]. Once fitted, an IRT model yields item parameters that can be used to place a new test-taker on the same latent scale from their response pattern, even if they completed only a subset of items. This allows us to report directly comparable scores for younger children who received mostly easier items and older children who received harder items.

# Methods

## Participants

Data for the present study were drawn from three pilot sites participating in LEVANTE: uniandes-co, mpieva-de, and western-ca. Collectively, these sites supported task localisation and technical implementation, iterative piloting, and deployment in varied testing contexts. The resulting pilot datasets enabled preliminary evaluation of construct validity, measurement invariance, test–retest reliability, and calibration for computer-adaptive administration, as well as descriptive assessment of age-related differences across the targeted constructs (refer to @Kachergis2025LEVANTEcoretasks for a full description).

### uniandes-co

The Universidad de Los Andes team carried out school-based data collection in Colombia, working with four schools in Bogotá and two public schools in the rural departments of Caquetá and Boyacá. Schools were selected pragmatically from institutions that had previously collaborated with the university. In Bogotá, three schools were affiliated with two family compensation funds and one was operated by a charter-school organisation. Across locations, participating schools largely served children from low-income and lower-middle-income backgrounds. Testing was conducted on tablets provided by the research team using a group-administration format, with group sizes adjusted by age (larger groups for older children; smaller groups for younger children).

### mpieva-de

At the mpieva-de site, the Max Planck Institute for Evolutionary Anthropology implemented a remote, family-based assessment protocol. Families were recruited via an existing registry of Leipzig residents and were contacted by telephone to invite participation. Families who enrolled were given login details and completed the assigned LEVANTE tasks at home using their own computer or tablet.

### western-ca

The Western University (Ontario) team recruited children from the surrounding community, primarily via online outreach (e.g., Facebook advertisements and posts in community groups), complemented by physical flyers and announcements distributed through local school newsletters. Families self-enrolled by registering their interest in a participant database. Children were assessed individually by a researcher in a dedicated on-campus testing space.

## Measure

The LEVANTE mental rotation task is adapted from @frick2013development. On each trial, children view a target image (e.g., a rabbit) and then choose between two alternatives (e.g., two rabbit silhouettes): one alternative is the same object shown at a different orientation, whereas the other is a non-matching foil. Children indicate which option corresponds to the target, and we record both accuracy and response time. Angular disparity is manipulated across trials (e.g., 40°, 80°, 120°, 160°), producing systematic effects on performance. Stimuli include 2D animal figures (ducks and rabbits) and more challenging 3D geometric shapes drawn from the classic Shepard and Metzler (1971) set.

```{r}
#| label: fig-mrot-example-items
#| fig-cap: "Example items for 2D and 3D rotation"
#| out-width: "100%"
#| fig-align: "center"

knitr::include_graphics("mrot_item_examples.pdf")
```

All LEVANTE tasks including mental rotation were originally developed in English and later adapted for Spanish and German through a structured internationalisation workflow carried out with collaborators at the uniandes-co and mpieva-de sites, respectively. We first produced draft translations of all verbal content—including instructions, item text, and recurring terminology (e.g., encouragement prompts)—using DeepL. These drafts were then refined by a professional translator to ensure grammatical accuracy and natural phrasing. Next, a site researcher who was a native speaker of the relevant language and dialect reviewed the materials to confirm clarity and suitability for the specific task context. To check for semantic and conceptual alignment with the English originals, we used AI-assisted back-translation to identify wording that required further revision and native-speaker review.

The mental rotation task has already been implemented in a computer-adaptive task (CAT) format. In the adaptive version, participant ability is summarized by a latent proficiency estimate (theta) that is re-estimated after each response. The updated theta then guides selection of the next item so that it is appropriately challenging for the individual. This trial-by-trial targeting typically makes the task more engaging and increases measurement efficiency, enabling reliable ability estimation with fewer administered items. Refer to @Kachergis2025LEVANTEcoretasks for more information on the CAT versions of all LEVANTE tasks.

## Analytical techniques \[TO ADD\]

\clearpage

# Results

## Cultural Similarities - cross-site measurement invariance of mental rotation

While we use psychometric modelling to place children’s performance on a common scale, the LEVANTE project is not designed to estimate absolute differences between sites. In practice, LEVANTE does not support site-level performance comparisons: sites vary substantially in sampling, recruitment, and test administration, so any apparent cross-site differences in raw performance are heavily confounded. Accordingly, although we report descriptive patterns that may look like absolute site differences, we emphasize that these should not be interpreted as meaningful.

Because our data were collected across three sites, we use multi-group IRT to evaluate measurement invariance, i.e., whether the task functions similarly across groups [@putnick2016measurement]. Multi-group models for the mental rotation task supported scalar invariance under a 2PL model, indicating that item difficulty is aligned across sites and that children in each country are being placed on a comparable latent scale. Practically, this means the groups share the same relative ordering of item difficulty—for example, items with larger rotation angles are consistently more difficult for children across all three countries (see @tbl-mrot-2pl-scalar). These results suggest that the core structure of mental rotation performance is largely shared across cultures in our sample: the task appears to tap the same underlying ability, and the same item features drive difficulty in a similar way across contexts. This provides a psychometric foundation for the broader claim that patterns of mental rotation development are more alike than different across the cultural settings represented here.

```{r}
#| include: false

# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
```

```{r}
#| label: setup
#| include: false
#| cache: false

source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)
```

```{r}
#| label: irt-anova
#| include: false

# IRT model output - import object
github_dir <- file.path(path.expand("~"), "Documents", "GitHub")
task_anova_path <- file.path(github_dir, "levante-pilots", "02_scoring_outputs", "task_anova.rds")

stopifnot(file.exists(task_anova_path))
task_anova <- readRDS(task_anova_path)
#colnames(task_anova)
#table(task_anova$itemtype)

# extract
mrot_2pl_scalar <- task_anova %>%
  filter(task == "mrot", itemtype == "2pl", invariance == "scalar")
mrot_2pl_scalar
```

```{r}
#| label: tbl-mrot-2pl-scalar
#| tbl-cap: "Model fit indices for 2PL scalar invariance models. Note: 2PL = 2-parameter logistic model; scalar invariance indicates that item intercepts and loadings are constrained across groups."
#| echo: false

library(gt)
mrot_2pl_scalar %>%
  transmute(
    Task       = task,
    Factors    = nfact,
    Model      = itemtype,
    Invariance = invariance,
    logLik     = logLik,
    AIC        = AIC,
    BIC        = BIC,
    `# params` = df
  ) %>%
  gt() %>%
  # Ensures the gt title area is clear and doesn't overlap the Quarto caption
  tab_header(title = md("")) %>% 
  fmt_number(columns = c(logLik, AIC, BIC), decimals = 2) %>%
  fmt_number(columns = `# params`, decimals = 0, use_seps = FALSE) %>%
  cols_align(align = "center", columns = c(Task, Factors, Model, Invariance, `# params`)) %>%
  cols_align(align = "right",  columns = c(logLik, AIC, BIC)) %>%
  tab_options(
    #table.font.size = px(12), # Increased slightly for better legibility
    table.width = pct(100),   # Ensures it fills the page width nicely
    data_row.padding = px(3)   # Tighter row spacing often looks better for stats
  )
```
\clearpage

@fig-dev-trends plots participants’ mental rotation ability estimates (IRT theta) against age, with separate panels/curves by site and age trends modelled using generalized additive models (GAMs). GAMs allow the developmental association with age to be non-linear, capturing periods of more rapid change and relative flattening without imposing a fixed form. Across sites, theta increases with age, indicating developmental improvement in mental rotation ability, and the dispersion of points within age bands underscores substantial heterogeneity in ability among same-aged children.

```{r}
#| label: fig-dev-trends
#| fig-cap: "Developmental trends in Mental Rotation ability by site."
#| fig-width: 8
#| fig-height: 6

# colnames(all_scores)
# table(all_scores$model)
mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!site_labels, .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

mrot_irt <- ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  labs(
    #title = "Developmental trends in Mental Rotation ability by site",
    subtitle = "Ability estimates smoothed with GAM",
    x = "Age (years)",
    y = "Ability (theta)",
    colour = "Site"
  )
mrot_irt
```

\clearpage

## Angular disparity and the RT–angle relationship

Our aim with this analysis was to test whether the canonical reaction time–rotation angle relationship first reported by @shepard1971mental replicates in our data. @fig-rt-angle-per-run plots mean reaction times (±95% confidence intervals) as a function of rotation angle, stratified by age group (5–7, 7–9, and 9–12 years) and site, with separate series by stimulus and item type (2D vs 3D). To show the underlying dispersion and sample density alongside the summary trends, we provide a second version of @fig-rt-angle-per-run in the Appendix that overlays the raw per-run data (using the same age bins and 95% CIs; see @fig-rt-angle_raw).

Across sites, mean RTs generally decrease with age, with the 9–12 group responding faster overall than the younger groups, particularly for the 2D animal stimuli (duck/rabbit). RTs also tend to increase with rotation angle (angular disparity), consistent with the classic RT–angle effect (@shepard1971mental). This angular-disparity effect is most pronounced for the 3D shape stimuli, which show slower overall RTs and steeper angle-related increases, whereas the 2D stimuli exhibit comparatively flatter RT–angle functions. Despite some differences in absolute RT levels between sites, the same qualitative pattern—faster responses with age and a stronger angular-disparity effect for 3D items—appears across all three sites, suggesting broadly comparable task behaviour across contexts.

```{r}
#| label: trial-level-rts
#| include: false

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%
  tidyr::unnest(data) %>%
  dplyr::mutate(
    correct = dplyr::case_when(
      is.logical(correct) ~ as.integer(correct),
      TRUE ~ as.integer(as.numeric(correct))
    ),
    rt_s   = as.numeric(rt_numeric) / 1000,
    log_rt = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))
```

```{r}
#| label: remove-retest-data
#| include: false

baseline_runs <- mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%  
  group_by(site, user_id) %>%
  arrange(age, run_id) %>%           
  slice(1) %>%
  ungroup() %>%
  select(site, user_id, run_id)

# keep only trials from those baseline runs
mrot_rt_trial_no_retest <- mrot_rt_trial %>%
  semi_join(baseline_runs, by = c("site", "user_id", "run_id"))

# check: each child should now have exactly 1 run (within site)
mrot_rt_trial_no_retest %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs)

# how many re-testers?
mrot_rt_trial %>%
  distinct(site, user_id, run_id) %>%
  count(site, user_id, name = "n_runs") %>%
  count(n_runs) # 133
```

```{r}
#| label: fig-rt-angle-per-run
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 7, 9, 12)

mrot_items_rt_runs <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle = as.integer(stringr::str_extract(item, "(?<=_)\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  filter(!is.na(rt_med)) %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med),
    sd  = sd(rt_med),
    n   = dplyr::n(),
    se  = sd / sqrt(n),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age_1 <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  )
p_mrot_rt_angle_age_1
```

\clearpage

## Within-Person Dynamics

Having established the expected angular-disparity effect across age groups and sites, we next examined how RTs evolved within individuals over the course of the task. Because item difficulty increased adaptively—such that higher-performing children were more likely to encounter the harder 3D items—we analysed trial-by-trial RT trajectories across the 80 trials to distinguish item-driven changes from more gradual time-on-task effects (e.g., fatigue). @fig-within-person-trials illustrates a clear 'jump' in RT about halfway through the task (trial 48), corresponding to the transition from 2D to 3D stimuli, which required significantly more processing time across all three site populations. We observed consistent response acceleration within both the 2D and 3D blocks, as children became faster with practice.

```{r}
#| label: fig-within-person-trials
#| fig-cap: "Trial-level Reaction Time Dynamics. The plot shows the mean log-reaction time across the 80 trials for each site. Shaded areas represent standard error. The sharp increase represents the transition to 3D items (Note: Re-test data removed)."
#| fig-width: 8
#| fig-height: 5

# 1. Prepare data with Site Labels
trial_plot_data <- mrot_rt_trial_no_retest %>%
  filter(!is.na(log_rt), correct == 1) %>% # Focus on correct trials for valid RT
  left_join(run_data %>% select(user_id, run_id, site), by = c("user_id", "run_id")) %>%
  mutate(site_label = dplyr::recode(site.x, !!!site_labels))


ggplot(trial_plot_data, aes(x = trial_number, y = log_rt, color = site_label, fill = site_label)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Trial Number (Time)",
    y = "Log RT (seconds)",
    color = "Site",
    fill = "Site"
  ) +
  # REMOVED theme_classic()
  theme(legend.position = "bottom")

```

\clearpage

## Cognitive Strategies

To characterize individual differences in how children solved the mental-rotation task, we derived participant-level 'strategy features; from trial-level reaction times. Consistent with the generally log-normal distribution of reaction times, we modeled log-transformed RTs. While the classic @shepard1971mental finding describes an approximately linear relationship with raw RT, fitting the model to log(RT) provides more robust parameter estimates for defining individual strategy profiles in a developing population. We fit a mixed-effects model to log(RT) that captured stimulus and angle-dependent responding alongside within-run practice effects, extracted participant-specific deviations (random effects), and combined these model-derived quantities with RT variability and IRT-estimated ability before clustering participants using k-means (k = 3) on standardized features.

The extracted features were: (1) Baseline Speed, defined as the intercept from the mixed-effects model (interpretable as expected speed at 0° rotation); (2) Rotation Slope, defined as the “cost” of rotation (the slope of angle on log RT), estimated net of stimulus effects; (3) Response Variability (RT SD), defined as the intra-individual standard deviation of log RT; and (4) IRT-estimated ability (theta).
```{r}
#| label: strategy-analysis-prep
#| results: "hide"

# Ability estimate - keep site + user
ability_user <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>% 
  mutate(
    metric_value = as.numeric(metric_value),
    age = as.numeric(age)
  ) %>%
  group_by(site, user_id) %>%
  summarise(
    ability = mean(metric_value, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# Prepare trial-level data for the rich model
mrot_rich_data <- mrot_rt_trial_no_retest %>%
  mutate(
    stimulus = str_extract(item, "^[a-z]+"),
    angle    = as.integer(str_extract(item, "\\d+")),
    stimulus = as.factor(stimulus),
    angle_s  = as.numeric(scale(angle)),
    trial_s  = as.numeric(scale(trial_number))
  )

# Fit the rich mixed model
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s +
    (1 + angle_s + trial_s | user_id) +
    (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# Extract subject-level random effects (deviations)
ran_effs <- ranef(m_rich)$user_id %>%
  tibble::rownames_to_column("user_id") %>%
  rename(
    int_dev   = `(Intercept)`,
    slope_dev = angle_s,
    trial_dev = trial_s
  )

# Fixed effects (global averages)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]


# Compute cognitive features (on log-RT scale)
rich_features_raw <- ran_effs %>%
  mutate(
    Baseline_Speed = fix_int + int_dev,
    Abs_Slope      = fix_ang + slope_dev,
    Learning_Rate  = fix_tri + trial_dev
  )

# Add RT variability + site mapping (site-aware) + ability
rt_variability <- mrot_rt_trial_no_retest %>%
  group_by(site, user_id) %>%
  summarise(RT_SD = sd(log_rt, na.rm = TRUE), .groups = "drop")

user_site_map <- mrot_rt_trial_no_retest %>%
  distinct(site, user_id)

stopifnot(
  user_site_map %>% count(user_id) %>% summarise(max(n)) %>% pull() == 1
)

rich_features_final <- rich_features_raw %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(rt_variability, by = c("site", "user_id")) %>%
  left_join(ability_user,  by = c("site", "user_id")) %>%
  mutate(
    Site = str_replace_all(site, c("pilot_" = "", "_" = "-"))
  ) %>%
  filter(if_all(c(ability, Abs_Slope, Baseline_Speed, RT_SD), is.finite))
```

```{r}
#| label: clustering-execution
#| include: false

my_colors <- c(
  "Low Efficiency" = "#e74c3c",
  "Analytical (feature matching)" = "#3498db",
  "Holistic (rotation)" = "#2ecc71"
)

# 1. Set up clustering
analysis_data <- rich_features_final %>%
  filter(if_all(c(ability, Abs_Slope, Baseline_Speed, RT_SD), is.finite))

# Global K-means (k=3)
# Scale inputs to ensure equal weighting
global_input <- analysis_data %>%
  select(ability, Abs_Slope, Baseline_Speed, RT_SD) %>%
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

# Assign labels
cluster_stats <- analysis_data %>%
  group_by(Global_Cluster) %>%
  summarise(
    mean_abil  = mean(ability, na.rm = TRUE),
    mean_slope = mean(Abs_Slope, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_abil)

# Identify "Low Efficiency" (Lowest Ability)
low_efficacy_id <- cluster_stats %>% 
  slice(1) %>% 
  pull(Global_Cluster)

# Separate the remaining two ("Effective") clusters
effective_clusters <- cluster_stats %>% 
  filter(Global_Cluster != low_efficacy_id) %>%
  arrange(mean_slope) # Sort by slope (Lower Slope = Pattern Matching)

# Lower Slope = Pattern Matching
pattern_id <- effective_clusters %>% 
  slice(1) %>% 
  pull(Global_Cluster)

# Higher Slope = Rotation
rotation_id <- effective_clusters %>% 
  slice(2) %>% 
  pull(Global_Cluster)

# apply labels + enforce factor levels that match my_colors
analysis_data <- analysis_data %>%
  mutate(
    Strategy_Label = case_when(
      Global_Cluster == low_efficacy_id  ~ "Low Efficiency",
      Global_Cluster == pattern_id  ~ "Analytical (feature matching)",
      Global_Cluster == rotation_id ~ "Holistic (rotation)"
    ),
    Strategy_Label = factor(Strategy_Label, levels = names(my_colors))
  )

stopifnot(setequal(levels(analysis_data$Strategy_Label), names(my_colors)))
```

```{r}
#| label: fig-rt-angle-by-strategy
#| fig-cap: "Participant-weighted RT–angle functions by strategy cluster (numbers indicate unique participants contributing at each angle; shaded bands show 95% CIs) (Note: Re-test data removed)."

# Join strategy labels onto trial-level RT data
rt_by_strategy <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(
    Strategy_Label = factor(Strategy_Label, levels = names(my_colors))
  )

# Person-level mean logRT per angle (each user contributes once per angle)
rt_user_angle <- rt_by_strategy %>%
  group_by(Strategy_Label, user_id, angle) %>%
  summarise(
    user_mean_log_rt = mean(log_rt, na.rm = TRUE),
    .groups = "drop"
  )

# Strategy-level mean across users at each angle + SE across users
rt_angle_person_summary <- rt_user_angle %>%
  group_by(Strategy_Label, angle) %>%
  summarise(
    n_users     = n_distinct(user_id),
    mean_log_rt = mean(user_mean_log_rt, na.rm = TRUE),
    se_log_rt   = sd(user_mean_log_rt, na.rm = TRUE) / sqrt(n_users),
    .groups = "drop"
  )

# check colours
missing_cols <- setdiff(levels(rt_angle_person_summary$Strategy_Label), names(my_colors))
stopifnot(length(missing_cols) == 0)

ggplot(rt_angle_person_summary, aes(x = angle, y = mean_log_rt, colour = Strategy_Label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_ribbon(
    aes(
      ymin = mean_log_rt - 1.96 * se_log_rt,
      ymax = mean_log_rt + 1.96 * se_log_rt,
      fill = Strategy_Label
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_text(aes(label = n_users), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = my_colors, drop = FALSE) +
  scale_fill_manual(values = my_colors, drop = FALSE) +
  labs(
    x = "Rotation angle (degrees)",
    y = "Mean log(RT) (participant-weighted)",
    colour = "Strategy",
    fill = "Strategy"
  ) +
  theme(legend.position = "bottom")
```
### Strategy clusters show distinct RT–angle signatures

Figure @fig-rt-angle-by-strategy shows participant-weighted RT–angle functions by strategy cluster (each participant contributes once per angle; shaded bands show 95% CIs). The Holistic (rotation) cluster exhibited the steepest RT increase with angular disparity, consistent with progressively greater processing demands at larger rotations The Analytical (feature matching) cluster showed a flatter RT–angle relationship, with relatively little change in RT as angular disparity increased. The Low Efficiency cluster showed little to no RT increase with angular disparity, and in some cases a slight decrease, indicating a weak or absent RT–angle relationship.

```{r}
#| label: emmeans-slope-extract
#| include: false

library(emmeans)
# Trial-level data with strategy labels for slope validation (site-aware join)
d_emm <- mrot_rich_data %>%
  left_join(
    analysis_data %>% select(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label), is.finite(log_rt), is.finite(angle)) %>%
  mutate(
    Strategy_Label = factor(Strategy_Label, levels = names(my_colors)),
    angle_s = as.numeric(scale(angle))  # 1sd increase in angle
  )

m_check <- lmer(
  log_rt ~ angle_s * Strategy_Label + (1 | user_id) + (1 | item),
  data = d_emm
)

# slopes by strategy
emm_tr_df <- as.data.frame(emtrends(m_check, ~ Strategy_Label, var = "angle_s"))

# pairwise comparisons of slopes (Tukey-adjusted)
emm_pairs_df <- as.data.frame(pairs(emtrends(m_check, ~ Strategy_Label, var = "angle_s")))

# helpers for reporting
fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_p <- function(p) ifelse(is.na(p), NA,
                            ifelse(p < 0.0001, "< 0.0001", fmt_num(p, 4)))

get_row <- function(df, label) df[df$Strategy_Label == label, , drop = FALSE]
get_contrast <- function(df, contrast_label) df[df$contrast == contrast_label, , drop = FALSE]

# Labels (match my_colors)
lab_low  <- "Low Efficiency"
lab_anal <- "Analytical (feature matching)"
lab_hol  <- "Holistic (rotation)"

# Per-group slopes
r_low  <- get_row(emm_tr_df, lab_low)
r_anal <- get_row(emm_tr_df, lab_anal)
r_hol  <- get_row(emm_tr_df, lab_hol)

s_low  <- r_low$angle_s.trend;  se_low  <- r_low$SE;  lcl_low  <- r_low$asymp.LCL;  ucl_low  <- r_low$asymp.UCL
s_anal <- r_anal$angle_s.trend; se_anal <- r_anal$SE; lcl_anal <- r_anal$asymp.LCL; ucl_anal <- r_anal$asymp.UCL
s_hol  <- r_hol$angle_s.trend;  se_hol  <- r_hol$SE;  lcl_hol  <- r_hol$asymp.LCL;  ucl_hol  <- r_hol$asymp.UCL

# Extract slope differences (re-express for readability)
# emmeans contrasts are printed as "A - B"
c_la <- get_contrast(emm_pairs_df, paste0(lab_low, " - ", lab_anal)) # Low - Anal
c_lh <- get_contrast(emm_pairs_df, paste0(lab_low, " - ", lab_hol))  # Low - Hol
c_ah <- get_contrast(emm_pairs_df, paste0(lab_anal, " - ", lab_hol)) # Anal - Hol

# Re-express as:
# Analytical - Low, Holistic - Low, Holistic - Analytical
d_a_l  <- -c_la$estimate; se_a_l <- c_la$SE; z_a_l <- abs(c_la$z.ratio); p_a_l <- c_la$p.value
d_h_l  <- -c_lh$estimate; se_h_l <- c_lh$SE; z_h_l <- abs(c_lh$z.ratio); p_h_l <- c_lh$p.value
d_h_a  <- -c_ah$estimate; se_h_a <- c_ah$SE; z_h_a <- abs(c_ah$z.ratio); p_h_a <- c_ah$p.value

```
Estimated RT–angle slopes (change in log(RT) per 1 SD increase in rotation angle) differed across strategy clusters. The holistic (rotation) cluster showed the steepest positive angle dependence (*b* = `r fmt_num(s_hol)`, SE = `r fmt_num(se_hol)`, 95% CI [`r fmt_num(lcl_hol)`, `r fmt_num(ucl_hol)`]), the analytical (feature matching) cluster showed a smaller positive slope (*b* = `r fmt_num(s_anal)`, SE = `r fmt_num(se_anal)`, 95% CI [`r fmt_num(lcl_anal)`, `r fmt_num(ucl_anal)`]), and the low efficacy cluster showed a near-zero/negative slope (*b* = `r fmt_num(s_low)`, SE = `r fmt_num(se_low)`, 95% CI [`r fmt_num(lcl_low)`, `r fmt_num(ucl_low)`]). Tukey-adjusted trend contrasts confirmed that the holistic slope exceeded the analytical slope by Δ*b* = `r fmt_num(d_h_a)` (SE = `r fmt_num(se_h_a)`, *z* = `r fmt_num(z_h_a, 2)`, *p* `r fmt_p(p_h_a)`), and that both the holistic (rotation) and analytical (feature matching) clusters exceeded the low efficacy cluster (analytical vs low efficacy: Δ*b* = `r fmt_num(d_a_l)`, SE = `r fmt_num(se_a_l)`, *z* = `r fmt_num(z_a_l, 2)`, *p* `r fmt_p(p_a_l)`; holistic vs low efficacy: Δ*b* = `r fmt_num(d_h_l)`, SE = `r fmt_num(se_h_l)`, *z* = `r fmt_num(z_h_l, 2)`, *p* `r fmt_p(p_h_l)`). 

```{r}
#| label: verifying-low-efficiency
#| include: false

# This checks whether the Low Efficiency cluster is near chance performance,
# rather than simply being a slower/less effective but still systematic strategy.

# define clustered sample
cluster_sample_ids <- analysis_data %>%
  distinct(site, user_id)

# Compute participant-level raw accuracy from baseline-only RT trials
acc_user <- mrot_rt_trial_no_retest %>%
  group_by(site, user_id) %>%
  summarise(
    n_trials = sum(!is.na(correct)),
    acc      = mean(correct, na.rm = TRUE),
    chance   = mean(chance, na.rm = TRUE),
    .groups  = "drop"
  )

# check
acc_user %>%
  summarise(
    n_users     = n(),
    mean_acc    = mean(acc, na.rm = TRUE),
    mean_chance = mean(chance, na.rm = TRUE)
  )

# Restrict accuracy dataset to the clustered sample
acc_user_in_sample <- acc_user %>%
  semi_join(cluster_sample_ids, by = c("site", "user_id"))

# Summarise raw accuracy by strategy cluster
acc_by_strategy <- acc_user_in_sample %>%
  left_join(
    analysis_data %>% distinct(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(!is.na(Strategy_Label)) %>%
  group_by(Strategy_Label) %>%
  summarise(
    n_users     = n(),
    mean_acc    = mean(acc, na.rm = TRUE),
    sd_acc      = sd(acc, na.rm = TRUE),
    mean_chance = mean(chance, na.rm = TRUE),
    .groups     = "drop"
  )

acc_by_strategy

# Low Efficiency: near-chance diagnostics
low_eff_acc <- acc_user_in_sample %>%
  mutate(acc_minus_chance = acc - chance) %>%
  left_join(
    analysis_data %>% distinct(site, user_id, Strategy_Label),
    by = c("site", "user_id")
  ) %>%
  filter(Strategy_Label == "Low Efficiency")

low_eff_acc %>%
  summarise(
    n_users                = n(),
    prop_at_or_below_chance = mean(acc <= chance, na.rm = TRUE),
    prop_within_5pct_chance = mean(acc <= chance + 0.05, na.rm = TRUE),
    median_acc_minus_chance = median(acc_minus_chance, na.rm = TRUE)
  )

# save stats for para
low_eff_stats <- low_eff_acc %>%
  summarise(
    prop_at_or_below_chance = mean(acc <= chance, na.rm = TRUE),
    prop_within_5pct_chance = mean(acc <= chance + 0.05, na.rm = TRUE),
    median_acc_minus_chance = median(acc_minus_chance, na.rm = TRUE)
  )
```
Because the Low Efficiency label was assigned on the basis of lowest estimated ability in the clustering step, we examined raw accuracy relative to chance as an external validity check that is diagnostic of near-random responding--whereas for the analytical (feature-matching) and holistic (rotation) clusters the key distinguishing prediction is their RT–angle signature (slope), not their absolute accuracy. Low-efficiency accuracy was modestly above chance (mean accuracy = `r fmt_num(acc_by_strategy$mean_acc[acc_by_strategy$Strategy_Label == "Low Efficiency"], 3)`; mean accuracy − chance = `r fmt_num(acc_by_strategy$mean_acc[acc_by_strategy$Strategy_Label == "Low Efficiency"] - acc_by_strategy$mean_chance[acc_by_strategy$Strategy_Label == "Low Efficiency"], 3)`). However, performance in this cluster was rarely consistent with chance responding: `r fmt_num(100 * low_eff_stats$prop_at_or_below_chance, 1)`% of participants performed at or below chance and `r fmt_num(100 * low_eff_stats$prop_within_5pct_chance, 1)`% were within 5 percentage points of chance; the median above-chance accuracy was `r fmt_num(low_eff_stats$median_acc_minus_chance, 3)`. These results therefore support interpreting the lowest-performing cluster as low efficiency rather than literal guessing.

```{r}
#| label: cluster-stability-bootstrap
#| include: false

# inputs
vars   <- c("ability", "Abs_Slope", "Baseline_Speed", "RT_SD")
k      <- 3
B      <- 500 # number of bootstrap resamples
nstart <- 25 # kmeans stability within each resample
set.seed(123)

# data: one row per participant
df_full <- analysis_data %>%
  select(user_id, Strategy_Label, all_of(vars)) %>%
  distinct(user_id, .keep_all = TRUE) %>%
  filter(if_all(all_of(vars), is.finite)) %>%
  mutate(
    Strategy_Label = factor(
      Strategy_Label,
      levels = levels(analysis_data$Strategy_Label)
    )
  )

stopifnot(nlevels(df_full$Strategy_Label) == 3)
stopifnot(n_distinct(df_full$user_id) == nrow(df_full))

ref_levels <- levels(df_full$Strategy_Label)
y_ref      <- df_full$Strategy_Label
ids        <- df_full$user_id
n          <- nrow(df_full)

# Scaling: lock to full-sample mean/sd (comparable across runs)
mu  <- df_full %>% summarise(across(all_of(vars), mean, na.rm = TRUE)) %>% as.list()
sdv <- df_full %>% summarise(across(all_of(vars), sd,   na.rm = TRUE)) %>% as.list()

X_full <- df_full %>%
  mutate(across(all_of(vars), ~ (. - mu[[cur_column()]]) / sdv[[cur_column()]])) %>%
  select(all_of(vars)) %>%
  as.matrix()

# helpers
ARI <- function(a, b) {
  a <- as.integer(factor(a))
  b <- as.integer(factor(b))
  tab <- table(a, b)
  n <- sum(tab)
  if (n <= 1) return(NA_real_)

  comb2 <- function(x) x * (x - 1) / 2
  sum_ij <- sum(comb2(tab))
  sum_i  <- sum(comb2(rowSums(tab)))
  sum_j  <- sum(comb2(colSums(tab)))
  expected <- (sum_i * sum_j) / comb2(n)
  max_idx  <- 0.5 * (sum_i + sum_j)
  denom <- (max_idx - expected)
  if (denom == 0) return(NA_real_)
  (sum_ij - expected) / denom
}

jaccard <- function(set_a, set_b) {
  inter <- length(intersect(set_a, set_b))
  uni   <- length(union(set_a, set_b))
  if (uni == 0) NA_real_ else inter / uni
}

# permutations for k=3 label matching
all_perms_3 <- list(
  c(1,2,3), c(1,3,2),
  c(2,1,3), c(2,3,1),
  c(3,1,2), c(3,2,1)
)

# match cluster IDs -> reference labels by maximum membership overlap
best_perm_by_overlap <- function(cl_hat, y_ref, ref_levels, all_perms_3) {
  tab <- table(cl_hat, factor(y_ref, levels = ref_levels))
  scores <- sapply(all_perms_3, function(p) sum(tab[cbind(1:3, p)]))
  all_perms_3[[which.max(scores)]]
}

# reference k-means centres (aligned to reference labels)
set.seed(123)
km_ref  <- kmeans(X_full, centers = k, nstart = nstart)
cl_ref  <- km_ref$cluster
perm_ref <- best_perm_by_overlap(cl_ref, y_ref, ref_levels, all_perms_3)
ref_centers <- km_ref$centers[perm_ref, , drop = FALSE]  # rows correspond to ref_levels

# bootstrap loop
res <- map_dfr(1:B, function(b) {

  set.seed(1000 + b)

  # bootstrap participants (with replacement)
  idx <- sample.int(n, size = n, replace = TRUE)

  # diagnostics: unique coverage + representation of reference clusters
  idx_unique <- unique(idx)
  tab_ref_unique <- table(factor(y_ref[idx_unique], levels = ref_levels))
  min_ref_unique <- min(tab_ref_unique)
  min_ref_prop   <- min_ref_unique / length(idx_unique)
  n_unique       <- length(idx_unique)

  X_b <- X_full[idx, , drop = FALSE]

  # fit kmeans on bootstrap sample
  km <- kmeans(X_b, centers = k, nstart = nstart)

  # assign all participants to nearest bootstrap center
  centers_b <- km$centers
  dists <- sapply(1:k, function(j) {
    rowSums((X_full - matrix(centers_b[j, ],
                             nrow = nrow(X_full),
                             ncol = ncol(X_full),
                             byrow = TRUE))^2)
  })
  cl_hat <- max.col(-dists)  # 1..k

  # resolve label switching with maximal overlap with reference labels
  perm  <- best_perm_by_overlap(cl_hat, y_ref, ref_levels, all_perms_3)
  y_hat <- factor(ref_levels[perm[cl_hat]], levels = ref_levels)

  # overall agreement
  ari <- ARI(y_ref, y_hat)

  # per-cluster Jaccard
  jac <- map_dbl(ref_levels, function(lab) {
    jaccard(ids[y_ref == lab], ids[y_hat == lab])
  })
  names(jac) <- paste0("J_", make.names(ref_levels))

  # centroid displacement vs reference
  centers_b_aligned <- centers_b[perm, , drop = FALSE]  # align rows to ref_levels
  center_shift <- mean(sqrt(rowSums((centers_b_aligned - ref_centers)^2)))

  tibble(
    replicate = b,
    ARI = ari,
    n_unique = n_unique,
    min_ref_unique = min_ref_unique,
    min_ref_prop = min_ref_prop,
    center_shift = center_shift,
    !!!as.list(jac)
  )
})

# diagnostics plots
# p_underrep <- ggplot(res, aes(x = min_ref_prop, y = ARI)) +
#   geom_point(alpha = 0.3) +
#   geom_smooth(se = FALSE) +
#   labs(
#     x = "Min proportion of unique participants from any reference cluster in resample",
#     y = "ARI (vs reference)",
#     title = "ARI vs under-representation of a reference cluster"
#   )
# 
# p_shift <- ggplot(res, aes(x = center_shift, y = ARI)) +
#   geom_point(alpha = 0.3) +
#   geom_smooth(se = FALSE) +
#   labs(
#     x = "Mean centroid displacement (aligned to reference centres)",
#     y = "ARI (vs reference)",
#     title = "Low-ARI runs coincide with larger shifts in estimated cluster centres"
#   )
# 
# p_underrep
# p_shift

# summaries
stability_summary <- res %>%
  summarise(
    B = n(),
    ARI_median = median(ARI, na.rm = TRUE),
    ARI_IQR    = IQR(ARI, na.rm = TRUE),
    center_shift_median = median(center_shift, na.rm = TRUE),
    center_shift_IQR    = IQR(center_shift, na.rm = TRUE),
    across(starts_with("J_"),
           list(median = ~median(.x, na.rm = TRUE),
                IQR    = ~IQR(.x, na.rm = TRUE)))
  )

stability_summary

# tail rates for a sentence in text
ari_tail <- res %>%
  summarise(
    pct_ARI_below_0.70 = mean(ARI < 0.70, na.rm = TRUE) * 100,
    pct_ARI_below_0.60 = mean(ARI < 0.60, na.rm = TRUE) * 100,
    min_ARI = min(ARI, na.rm = TRUE)
  )

ari_tail
```

```{r}
#| label: cluster-stability-reporting
#| include: false
fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_p   <- function(p) ifelse(is.na(p), NA,
                              ifelse(p < 0.0001, "< 0.0001", fmt_num(p, 4)))
# Pull ARI stats
ari_med <- stability_summary$ARI_median
ari_iqr <- stability_summary$ARI_IQR

pct_ari_70 <- ari_tail$pct_ARI_below_0.70
pct_ari_60 <- ari_tail$pct_ARI_below_0.60
min_ari    <- ari_tail$min_ARI

# Pull Jaccard medians + IQRs (columns produced by stability_summary)
j_med <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_median")]]
})
j_iqr <- sapply(ref_levels, function(lab) {
  stability_summary[[paste0("J_", make.names(lab), "_IQR")]]
})
names(j_med) <- ref_levels
names(j_iqr) <- ref_levels

# For inline text
j_low_med <- j_med["Low Efficiency"];                 j_low_iqr <- j_iqr["Low Efficiency"]
j_ana_med <- j_med["Analytical (feature matching)"];  j_ana_iqr <- j_iqr["Analytical (feature matching)"]
j_hol_med <- j_med["Holistic (rotation)"];            j_hol_iqr <- j_iqr["Holistic (rotation)"]

# quantify relation between centroid shift and ARI (spearman is robust)
if ("center_shift" %in% names(res)) {
  rho_shift <- suppressWarnings(cor(res$ARI, res$center_shift, method = "spearman", use = "complete.obs"))
} else {
  rho_shift <- NA_real_
}

```
## Robutness Checks

We conducted two robustness checks to evaluate whether the three-strategy solution was stable and generalizable. First, we assessed cluster stability under participant bootstrap resampling; second, we compared the pooled (global) clustering solution to site-specific (local) clustering to test whether the same strategy structure emerged within each site.

### Cluster robustness
We assessed robustness of the three-strategy solution via participant bootstrap resampling (B = `r stability_summary$B`), refitting k-means to each resample and assigning all participants to the recovered centers (with cluster labels aligned to the reference solution by maximal membership overlap). The recovered partition closely matched the reference clustering (median ARI = `r fmt_num(ari_med, 3)`; IQR = `r fmt_num(ari_iqr, 3)`), indicating that the three-cluster structure is not dependent on a particular sample composition. Cluster membership was also stable at the strategy level (median Jaccard overlap: Low Efficiency = `r fmt_num(j_low_med, 3)` [IQR = `r fmt_num(j_low_iqr, 3)`], Analytical (feature matching) = `r fmt_num(j_ana_med, 3)` [IQR = `r fmt_num(j_ana_iqr, 3)`], Holistic (rotation) = `r fmt_num(j_hol_med, 3)` [IQR = `r fmt_num(j_hol_iqr, 3)`]). Although a minority of resamples yielded lower agreement (`r fmt_num(pct_ari_70, 1)`% with ARI < 0.70; `r fmt_num(pct_ari_60, 1)`% with ARI < 0.60; minimum ARI = `r fmt_num(min_ari, 3)`), this tail was not explained by simple under-representation of any reference cluster among unique resampled participants. Instead, lower-ARI runs tended to coincide with larger shifts in the estimated cluster centers (Spearman ρ = `r fmt_num(rho_shift, 3)`), consistent with occasional bootstrap draws leading k-means to an alternative local solution and/or reassignment of boundary-case participants, while the typical resample recovered the same three-strategy partition. See @fig-cluster-bootstrap-plots in Appendix 2 for bootstrap stability plots (ARI distribution and per-cluster Jaccard overlap).

```{r}
#| label: local-clustering-agreement
#| include: false

vars <- c("ability", "Abs_Slope", "Baseline_Speed", "RT_SD")
stopifnot(all(vars %in% names(analysis_data)))
stopifnot(all(c("Site", "user_id", "Strategy_Label") %in% names(analysis_data)))
stopifnot(nrow(analysis_data) == nrow(dplyr::distinct(analysis_data, Site, user_id)))

# Local k-means (labels same as global model)
get_local_strategy_labels <- function(df_site,
                                     k = 3,
                                     seed = 123,
                                     nstart = 25,
                                     min_n = 30) {

  if (nrow(df_site) < min_n) return(NULL)
  set.seed(seed)

  X <- df_site %>%
    dplyr::select(dplyr::all_of(vars)) %>%
    scale() %>%
    as.matrix()

  km <- kmeans(X, centers = k, nstart = nstart)

  tmp <- df_site %>%
    mutate(Local_Cluster = factor(km$cluster))

  local_stats <- tmp %>%
    group_by(Local_Cluster) %>%
    summarise(
      mean_abil  = mean(ability, na.rm = TRUE),
      mean_slope = mean(Abs_Slope, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(mean_abil)

  # Identify clusters locally using the SAME decision rules as global:
  # Low Efficiency = lowest mean ability
  # Among remaining: lower slope = Analytical; higher slope = Holistic
  low_id <- local_stats %>% slice(1) %>% pull(Local_Cluster)

  other_two <- local_stats %>%
    filter(Local_Cluster != low_id) %>%
    arrange(mean_slope)

  analytical_id <- other_two %>% slice(1) %>% pull(Local_Cluster)
  holistic_id   <- other_two %>% slice(2) %>% pull(Local_Cluster)

  tmp %>%
    mutate(
      Strategy_Local = dplyr::case_when(
        Local_Cluster == low_id        ~ "Low Efficiency",
        Local_Cluster == analytical_id ~ "Analytical (feature matching)",
        Local_Cluster == holistic_id   ~ "Holistic (rotation)"
      ),
      Strategy_Local = factor(Strategy_Local, levels = levels(analysis_data$Strategy_Label))
    ) %>%
    select(Site, user_id, Strategy_Local)
}

# fit local models (k=3) independently within each site
local_labels_all <- analysis_data %>%
  group_by(Site) %>%
  group_split() %>%
  purrr::map_dfr(get_local_strategy_labels)

analysis_local <- analysis_data %>%
  select(-dplyr::any_of("Strategy_Local")) %>%
  left_join(local_labels_all, by = c("Site", "user_id"))

# agreement (3-class)
agreement_3class <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_by(Site) %>%
  summarise(
    n = n(),
    pct_agree = mean(Strategy_Local == Strategy_Label),
    ARI = ARI(Strategy_Label, Strategy_Local),
    .groups = "drop"
  )

agreement_3class

# confusion matrices per site (Local rows x Global cols)
confusion_by_site <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  group_split(Site) %>%
  purrr::set_names(purrr::map_chr(., ~ unique(.x$Site))) %>%
  purrr::map(~ table(.x$Strategy_Local, .x$Strategy_Label))

confusion_by_site

# recall (given global label) + purity (given local label)
recall_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Label, Strategy_Local) %>%
  group_by(Site, Strategy_Label) %>%
  mutate(recall = n / sum(n)) %>%
  ungroup()

purity_tbl <- analysis_local %>%
  filter(!is.na(Strategy_Local)) %>%
  count(Site, Strategy_Local, Strategy_Label) %>%
  group_by(Site, Strategy_Local) %>%
  mutate(purity = n / sum(n)) %>%
  ungroup()

diag_summary <- recall_tbl %>%
  filter(Strategy_Label == Strategy_Local) %>%
  transmute(Site, Strategy = Strategy_Label, recall) %>%
  left_join(
    purity_tbl %>%
      filter(Strategy_Local == Strategy_Label) %>%
      transmute(Site, Strategy = Strategy_Local, purity),
    by = c("Site", "Strategy")
  )

diag_summary

```

```{r}
#| label: local-vs-pooled-reporting
#| cache: true
#| include: false

fmt_num <- function(x, d = 3) formatC(x, format = "f", digits = d)
fmt_pct <- function(x, d = 1) paste0(formatC(100*x, format = "f", digits = d), "%")

get_row <- function(df, site) df[df$Site == site, , drop = FALSE]

# sites
s_mpi <- "mpieva-de"
s_uni <- "uniandes-co"
s_wes <- "western-ca"

# agreement table
a_mpi <- get_row(agreement_3class, s_mpi)
a_uni <- get_row(agreement_3class, s_uni)
a_wes <- get_row(agreement_3class, s_wes)

# Diagonal recall/purity per label
get_diag <- function(site, label) {
out <- diag_summary[diag_summary$Site == site & diag_summary$Strategy == label, , drop = FALSE]
stopifnot(nrow(out) == 1)
out
}

lab_low  <- "Low Efficiency"
lab_ana  <- "Analytical (feature matching)"
lab_hol  <- "Holistic (rotation)"

d_mpi_low <- get_diag(s_mpi, lab_low)
d_mpi_ana <- get_diag(s_mpi, lab_ana)
d_mpi_hol <- get_diag(s_mpi, lab_hol)

d_uni_low <- get_diag(s_uni, lab_low)
d_uni_ana <- get_diag(s_uni, lab_ana)
d_uni_hol <- get_diag(s_uni, lab_hol)

d_wes_low <- get_diag(s_wes, lab_low)
d_wes_ana <- get_diag(s_wes, lab_ana)
d_wes_hol <- get_diag(s_wes, lab_hol)

# Dominant off-diagonal moves (from confusion matrices: rows = local, cols = global)
get_move <- function(tab, local_lab, global_lab) {
stopifnot(local_lab %in% rownames(tab), global_lab %in% colnames(tab))
n <- tab[local_lab, global_lab]
denom <- sum(tab[, global_lab])
list(n = as.numeric(n), denom = as.numeric(denom), prop = as.numeric(n)/as.numeric(denom))
}

tab_mpi <- confusion_by_site[[s_mpi]]
tab_uni <- confusion_by_site[[s_uni]]
tab_wes <- confusion_by_site[[s_wes]]

# Global Holistic -> Local Analytical (the key swap in mpieva + western)
mpi_h2a <- get_move(tab_mpi, lab_ana, lab_hol)
wes_h2a <- get_move(tab_wes, lab_ana, lab_hol)

# Global Analytical -> Local Holistic (key swap in uniandes)
uni_a2h <- get_move(tab_uni, lab_hol, lab_ana)

```

### Cross-site generalizability: pooled vs site-local clustering
To evaluate whether the three-strategy structure depends on pooling across sites, we refit k-means clustering (k = 3) independently within each site using the same four participant-level features (ability, RT–angle slope, baseline speed, and within-person RT variability), and labelled the recovered clusters using the same rule as in the pooled analysis: the lowest-ability cluster was labelled Low Efficiency, and the remaining two clusters were distinguished by RT–angle slope as Analytical (shallower slope) versus Holistic (steeper slope).

Agreement between the site-specific and pooled solutions was moderate overall (mpieva-de: n = `r a_mpi$n`, % agreement = `r fmt_pct(a_mpi$pct_agree, 1)`, ARI = `r fmt_num(a_mpi$ARI, 3)`; uniandes-co: n = `r a_uni$n`, % agreement = `r fmt_pct(a_uni$pct_agree, 1)`, ARI = `r fmt_num(a_uni$ARI, 3)`; western-ca: n = `r a_wes$n`, % agreement = `r fmt_pct(a_wes$pct_agree, 1)`, ARI = `r fmt_num(a_wes$ARI, 3)`).

Across sites, the Low Efficiency cluster showed high recall (`r fmt_num(d_mpi_low$recall, 3)`–`r fmt_num(d_wes_low$recall, 3)`) and moderate-to-high purity (`r fmt_num(d_mpi_low$purity, 3)`–`r fmt_num(d_uni_low$purity, 3)`), indicating that the lowest-ability subgroup is consistently recovered under local optimisation. In contrast, disagreement between local and pooled solutions was concentrated in reassignment between the Analytical and Holistic clusters.

In mpieva-de and western-ca, a subset of globally Holistic participants was reassigned to the local Analytical cluster (mpieva-de: `r mpi_h2a$n`/`r mpi_h2a$denom`, `r fmt_pct(mpi_h2a$prop, 1)`; western-ca: `r wes_h2a$n`/`r wes_h2a$denom`, `r fmt_pct(wes_h2a$prop, 1)`), while in uniandes-co a subset of globally Analytical participants was reassigned to the local Holistic cluster (`r uni_a2h$n`/`r uni_a2h$denom`, `r fmt_pct(uni_a2h$prop, 1)`). Together, these results suggest that the primary separation of the lowest-ability cluster is robust across local models, whereas the boundary between Analytical and Holistic strategies is comparatively more site-sensitive.


# Discussion \[TO ADD\]

\newpage

# References

::: {#refs}
:::

# IRT Ability by Age {#apx-fig-rt-angle .appendix}

```{r}
#| label: fig-rt-angle_raw
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI."
#| fig-width: 8
#| fig-height: 8

# Add raw data underneath
p_mrot_rt_angle_age_2 <- ggplot() +
  # raw per-run medians
  geom_point(
    data = mrot_items_rt_runs,
    aes(x = angle, y = rt_med,
        colour = stimulus),
    size = 0.4,
    alpha = 0.25,
    position = position_jitter(width = 3, height = 0)
  ) +
  # summary CI + lines
  geom_errorbar(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y, ymin = lwr, ymax = upr,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    width = 2,
    linewidth = 0.35,
    alpha = 0.6
  ) +
  geom_point(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    size = 1
  ) +
  geom_line(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    alpha = 0.85
  ) +
  facet_grid(site ~ age_group) +
  labs(
    #title = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",
    colour = "Stimulus",
    #linetype = "Item Type"
  )
p_mrot_rt_angle_age_2
```

\clearpage

::: {.landscape}
# Cluster Bootstrap plots {#apx-cluster-bootstrap-plots .appendix}
```{r}
#| label: fig-cluster-bootstrap-plots
#| cache: true
#| fig-width: 11
#| fig-height: 4.5
#| fig-align: center

library(patchwork)

p_ari <- ggplot(res, aes(x = ARI)) +
  geom_histogram(bins = 30) +
  labs(
    x = "Adjusted Rand Index (vs reference labels)",
    y = "Count",
    title = paste0("Cluster stability under bootstrap resampling (B = ", B, ")")
  )

# long-form Jaccard with safe mapping back to Strategy_Label
jac_key <- tibble(
  col = paste0("J_", make.names(ref_levels)),
  Strategy_Label = ref_levels
)

res_long <- res %>%
  pivot_longer(starts_with("J_"), names_to = "col", values_to = "Jaccard") %>%
  left_join(jac_key, by = "col") %>%
  mutate(Strategy_Label = factor(Strategy_Label, levels = ref_levels))

p_jac <- ggplot(res_long, aes(x = Strategy_Label, y = Jaccard)) +
  geom_boxplot(outlier.alpha = 0.25) +
  labs(
    x = NULL,
    y = "Jaccard (membership overlap vs reference)",
    title = "Per-cluster membership stability"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  theme_minimal()

(p_ari | p_jac)
```
:::
