---
title: "Measuring the development of mental rotation ages 5-12 across three populations"
shorttitle: "LEVANTE Mental Rotation"

author: 
  - name: "Fionnuala O'Reilly"
    orcid: 0000-0002-4355-9088
    affiliations:
      - ref: stanford
  - name: "Xingyao Xiao"
    orcid: 0000-0001-8430-0438
    affiliations:
      - ref: stanford
  - name: "Rebecca Zhu"
    orcid: 0000-0003-0957-5962
    affiliations:
      - ref: stanford
  - name: "Mika Braginsky"
    orcid: 0000-0001-9039-3220
    affiliations:
      - ref: stanford
  - name: "Nilàm Ram"
    orcid: 0000-0003-1671-5257
    affiliations:
      - ref: stanford
  - name: "Benjamin W. Domingue"
    orcid: 0000-0002-3894-9049
    affiliations:
      - ref: stanford
  - name: "Michael C. Frank"
    orcid: 0000-0002-7551-4378
    affiliations:
      - ref: stanford
    corresponding: true
    email: mcfrank@stanford.edu
    # role:
    #   - conceptualization
    #   - writing
    #   - editing
    #   - supervision

affiliations:
  - id: stanford
    name: "Stanford University"

author-note:
  disclosures:
    financial-support: "This work was supported by the Jacobs Foundation."
    # conflict-of-interest: The author has no conflict of interest to declare.

abstract: To add.

keywords: ["cognitive development", "mental rotation", "spatial reasoning"]
word-count: true

bibliography: library_mrot.bib

floatsintext: true
numbered-lines: true
# draft: false
mask: false

# appendix
number-sections: false
appendix-style: default

# figurelist: no
# tablelist: no
# footnotelist: no

format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    include-in-header: preamble.tex
    fig-format: pdf
    # fig-pos: H        
    # tbl-pos: H  

knitr:
  opts_chunk:
    ft.arraystretch: 1.25

execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
---

# Introduction

Spatial reasoning, broadly defined as the ability to generate, retain, and manipulate visuospatial information is a fundamental cognitive skill [@lohman2013spatial]. It allows us to form and transform mental images of 2D shapes and 3D objects (e.g., by rotating, bending, or folding them), and to represent how spatial relations between locations, shapes, and objects change over time [@frick2014development; @newcombe2014thinking].

Spatial ability is important for a wide range of educational outcomes [@wai2018spatial], including creative domains such as music [@padulo2020effects; @hetland2000learning] and visual arts [@kell2013creativity; @winner2013relationship] as well as STEM subjects [@wai2009spatial] and, most notably, math [@atit2022examining; @xie2020spatial; @mix2012relation; @gunderson2012relation]. Given it’s importance for educational outcomes, a number of attempts have been made to train spatial reasoning skills. For example, @hawes2015effects reported that, compared with an active control group, 6–8-year-old children who received mental rotation training showed significant gains on two separate 2D mental rotation measures, as well as more general improvements in spatial reasoning on an untrained spatial task. @uttal2013malleability conducted a meta-analysis of 217 studies with child and adult samples to examine the size and generalizability of spatial skills training effects. Across studies, the mean effect size for training compared with control conditions was Hedges’ g = 0.47 (SE = 0.04). @gilligan2020first showed that training spatial reasoning skills had	near,	intermediate	and	far	transfer	effects in 8-year old children [@gilligan2020first]. Similarly,in a sample of elementary school students, @lowrie2017visuospatial showed that those who participated in a 10-week visuospatial reasoning (VSR) program achieved substantially higher spatial reasoning scores than students in the control group, and also demonstrated improved mathematics performance.  	

to be continued

## Reaction time

## Cross-country comparisons

## Strategies

```{r}
# packages
library(tidyverse)
library(glue)
library(here)
library(forcats)
library(lme4)       
library(brms)       
library(ggeffects)  
library(gt)         
library(patchwork)  
library(brms)
library(ggplot2)
library(dplyr)
```

```{r}
#| echo: false
source(here::here("..", "levante-pilots", "plot_settings.R"))
source(here::here("..", "levante-pilots", "03_explore_tasks", "explore_helper.R"))
```

```{r}
#| label: load-data
#| echo: false
#| message: false
#| cache: true

site_labels <- c("uniandes-co" = "pilot_uniandes_co",
                 "mpieva-de" = "pilot_mpieva_de",
                 "western-ca" = "pilot_western_ca")

site_map <- tibble::tibble(
  site       = names(site_labels),
  site_label = unname(site_labels)
)

# core_tasks   <- c("matrix","mrot","math","hf","mg","sds",
#                   "trog","vocab","tom")

# Run data
run_data <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data/run_data.rds"))
# colnames(run_data)

# RT data
all_sum_scores <- readr::read_rds(here::here("..", "levante-pilots", "01_fetched_data","task_data_nested.rds"))
# colnames(all_sum_scores)
# glimpse(all_sum_scores$data)

# IRT data
all_scores <- readr::read_rds(here::here("..", "levante-pilots", "02_scoring_outputs","scores","scores_combined.rds"))
# colnames(all_scores)
# table(all_scores$item_task)
# table(all_scores$metric_type)
```

```{r}
#| label: dev-trends
#| fig-cap: "Developmental trends in Mental Rotation ability by site."
#| fig-width: 8
#| fig-height: 6

# colnames(all_scores)
# table(all_scores$model)
mrot_df <- all_scores %>%
  filter(item_task == "mrot", metric_type == "ability") %>%
  mutate(
    age          = as.numeric(age),
    metric_value = as.numeric(metric_value),
    site_label   = dplyr::recode(site, !!!site_labels, .default = site)
  ) %>%
  filter(is.finite(age), is.finite(metric_value))

mrot_irt <- ggplot(mrot_df, aes(x = age, y = metric_value, colour = site_label)) +
  geom_point(alpha = 0.35, size = 1.1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  labs(
    #title = "Developmental trends in Mental Rotation ability by site",
    subtitle = "Ability estimates smoothed with GAM",
    x = "Age (years)",
    y = "Ability (theta)",
    colour = "Site"
  )
mrot_irt
```

```{r}
#| label: ability-rt
#| fig-cap: "IRT Ability (Theta) vs log RT."
#| fig-width: 8
#| fig-height: 6

# get shape and rotation data
mrot_rt_sum <- all_sum_scores %>%
  filter(item_task == "mrot") %>%
  select(site, language, data) %>%
  unnest(data) %>%                         
  mutate(
    correct = as.numeric(correct),
    rt_s    = as.numeric(rt_numeric) / 1000
  ) %>%
  filter(is.finite(rt_s), rt_s > 0) %>%     
  mutate(log_rt = log(rt_s))

# collapse to run level RT summaries
mrot_rt_runs <- mrot_rt_sum %>%
  left_join(run_data %>% 
              select(site, user_id, run_id, age), 
            by = c("site", "user_id", "run_id")) %>%
  group_by(site, user_id, run_id) %>%
  summarise(
    # trial-level RT summaries for correct items
    logrt_median = median(log_rt[correct == 1], na.rm = TRUE),
    logrt_mean   = mean(log_rt[correct == 1],  na.rm = TRUE),
    age = first(age),
    n_items = n_distinct(item_uid),
    .groups = "drop"
  )

ability_runs <- all_scores %>% 
  filter(item_task == "mrot", metric_type == "ability") %>%
  select(site, user_id, run_id, ability = metric_value, age) %>%
  group_by(site, user_id, run_id) %>%               
  summarise(
    ability = mean(ability, na.rm = TRUE),
    age     = first(age),
    .groups = "drop"
  )

# join RT and ability
mrot_speed_ability <- ability_runs %>%
  inner_join(
    mrot_rt_runs %>% select(site, user_id, run_id, logrt_mean),
    by = c("site", "user_id", "run_id")
  )

p_mrot_speed_ability <- mrot_speed_ability %>%
  mutate(age_binned = as_factor(floor(age))) %>%
  ggplot(aes(x = logrt_mean, y = ability, colour = age_binned)) +
  geom_point(alpha = 0.55, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ site) +
  labs(
    #title = "IRT Ability (Theta) vs log RT",
    subtitle = "Each point = participant run; site-wise linear fits",
    x = "log(RT, seconds)",
    y = "IRT ability (theta)",
    colour = "Age (years)"
  )
p_mrot_speed_ability
```
\clearpage

```{r}
#| label: trial-level-rts
#| echo: false
#| message: false
#| cache: true

mrot_rt_trial <- all_sum_scores %>%
  dplyr::filter(item_task == "mrot") %>%
  dplyr::select(site, data) %>%        
  tidyr::unnest(data) %>%              
  dplyr::mutate(
    correct = as.numeric(correct),
    rt_s    = as.numeric(rt_numeric) / 1000,
    log_rt  = log(rt_s)
  ) %>%
  dplyr::filter(is.finite(rt_s), rt_s > 0)

# age per run 
run_ages <- run_data %>%
  dplyr::transmute(
    user_id,
    run_id,
    age = as.numeric(age)
  ) %>%
  dplyr::filter(is.finite(age))

```

```{r}
#| label: fig-rt-angle
#| fig-cap: "Reaction Time by angle within age groups ±95% CI."
#| fig-width: 8
#| fig-height: 8

age_breaks <- c(5, 8, 11, 14)

mrot_items_rt_runs <- mrot_rt_trial %>%
  mutate(
    stimulus = stringr::str_extract(item, "^[a-z]+"),
    angle    = as.integer(stringr::str_extract(item, "\\d+"))
  ) %>%
  filter(stimulus %in% c("duck", "rabbit", "shape"), !is.na(angle)) %>%
  left_join(run_ages, by = c("user_id","run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = age_breaks,
                    include.lowest = TRUE, right = TRUE)
  ) %>%
  tidyr::drop_na(age_group) %>%
  group_by(site, user_id, run_id, age, age_group,
           stimulus, item_group, angle) %>%
  summarise(
    rt_med = median(rt_s[correct == 1], na.rm = TRUE),
    .groups = "drop"
  )

mrot_items_rt_agebin_ci <- mrot_items_rt_runs %>%
  group_by(site, stimulus, item_group, angle, age_group) %>%
  summarise(
    y   = mean(rt_med, na.rm = TRUE),
    sd  = sd(rt_med, na.rm = TRUE),
    n   = dplyr::n(),
    se  = sd / sqrt(pmax(n, 1)),
    lwr = y - 1.96 * se,
    upr = y + 1.96 * se,
    .groups = "drop"
  )

p_mrot_rt_angle_age <- ggplot(
  mrot_items_rt_agebin_ci,
  aes(x = angle, y = y,
      group = interaction(item_group, stimulus),
      colour = stimulus, linetype = item_group)
) +
  # raw data layer added below (step 2) ...
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 2,
                linewidth = 0.35, alpha = 0.6) +
  geom_point(size = 1) +
  geom_line(alpha = 0.85) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",                             
    colour = "Stimulus",
    linetype = "Item Type"
  )
p_mrot_rt_angle_age
```
\clearpage

```{r}
#| label: fig-rt-angle-per-run
#| fig-cap: "Reaction Time by angle within age groups with raw per-run data and ±95% CI."
#| fig-width: 8
#| fig-height: 8

# Add raw data underneath
p_mrot_rt_angle_age <- ggplot() +
  # raw per-run medians
  geom_point(
    data = mrot_items_rt_runs,
    aes(x = angle, y = rt_med,
        colour = stimulus),
    size = 0.4,
    alpha = 0.25,
    position = position_jitter(width = 3, height = 0)
  ) +
  # summary CI + lines
  geom_errorbar(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y, ymin = lwr, ymax = upr,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    width = 2,
    linewidth = 0.35,
    alpha = 0.6
  ) +
  geom_point(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    size = 1
  ) +
  geom_line(
    data = mrot_items_rt_agebin_ci,
    aes(x = angle, y = y,
        colour = stimulus, linetype = item_group,
        group = interaction(item_group, stimulus)),
    alpha = 0.85
  ) +
  facet_grid(site ~ age_group) +
  labs(
    #title    = "Mental Rotation: RT by angle within age groups ±95% CI",
    x = "Rotation angle (°)",
    y = "RT (s)",
    colour = "Stimulus",
    #linetype = "Item Type"
  )
p_mrot_rt_angle_age
```
\clearpage

# RT Residualization
To isolate pure processing speed from item difficulty (e.g., the added difficulty of 3D objects), we fitted a Linear Mixed Model (LMM) residualizing log reaction times against rotation angle, trial number, and stimulus type. 

```{r}
#| label: residualization
#| echo: false
#| message: false
#| cache: true
#| results: "hide"

# prepare data for modeling
model_rt_data <- mrot_rt_trial %>%
  mutate(
    # Extract "rabbit" from "rabbit_040" if not already present
    stimulus = stringr::str_extract(item, "^[a-z]+"), 
    angle = as.integer(stringr::str_extract(item, "\\d+")) 
  ) %>%
  filter(!is.na(angle), !is.na(log_rt))

# Controlling for angle, trial, stimulus, and item grouping
m_rt_full <- lmer(
  log_rt ~ angle + trial_number + stimulus + item_group + (1|item), 
  data = model_rt_data
)

# extract residuals (Person Speed): these represent speed after accounting for item difficulty
model_rt_data <- model_rt_data %>%
  mutate(rt_resid_item = residuals(m_rt_full))

rt_residuals_person <- model_rt_data %>%
  group_by(site, user_id, run_id) %>%
  summarise(
    rt_residual_mean = mean(rt_resid_item, na.rm=TRUE),
    rt_residual_sd   = sd(rt_resid_item, na.rm=TRUE),
    n_items = n(),
    .groups = "drop"
  )

# Save for later
#write_csv(rt_residuals_person, here::here("..", "levante-pilots", "02_scoring_outputs", "scores", "rt_residuals_corrected.csv"))
```

```{r}
#| label: fig-rt-resid
#| fig-cap: "Residualized reaction time."
#| fig-width: 8
#| fig-height: 6

rt_resid_person_age <- rt_residuals_person %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%
  filter(!is.finite(age) == FALSE) %>%    
  mutate(
    age_group = cut(
      age,
      breaks = c(5, 8, 11, 14),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  drop_na(age_group)

# Violin plot of residualized RT
ggplot(rt_resid_person_age,
       aes(x = age_group, y = rt_residual_mean, fill = age_group)) +
  #geom_violin(trim = FALSE, alpha = 0.6) +
  geom_jitter(size = 1.5, alpha = 0.4, width = 0.1) + # looking for the z-score to be jittered
  geom_boxplot(width = 0.15, outlier.shape = NA, width = 10) +
  facet_wrap(~ site, nrow = 1) +
  labs(
    x = "Age group (years)",
    y = "Residualized RT (log seconds)"
  ) +
  theme(legend.position = "none")
```
 
\clearpage

```{r}
#| label: resid-rt-diff
#| echo: false
#| message: false
#| cache: true
#| results: "hide"

# quantifying differences in residualized RT with site and age
rt_resid_person_age <- model_rt_data %>%
  group_by(site, user_id, run_id) %>%
  summarise(
    rt_residual_mean = mean(rt_resid_item, na.rm = TRUE),
    rt_residual_sd   = sd(rt_resid_item,   na.rm = TRUE),
    n_items          = n(),
    .groups          = "drop"
  ) %>%
  left_join(
    run_data %>%
      transmute(
        user_id,
        run_id,
        age = as.numeric(age)
      ),
    by = c("user_id", "run_id")
  ) %>%
  filter(is.finite(age),
         age >= 5,
         age <= 13)

m_site_age <- lmer(
  rt_residual_mean ~ site + age + (1 | user_id), # random intercept; each child’s baseline is estimated independently
  data = rt_resid_person_age
)

summary(m_site_age)
anova(m_site_age)
```

```{r}
#| label: fig-site-age-slopes
#| fig-cap: "Cross-Site Trajectories of Residualized Reaction Time from Ages 5–13."
#| fig-width: 8
#| fig-height: 6
#| results: "hide"

# Site-specific age slopes (interaction model)
m_site_age_int <- lmer(
  rt_residual_mean ~ site * age + (1 | user_id),
  data = rt_resid_person_age
)
anova(m_site_age, m_site_age_int)  

ggplot(rt_resid_person_age,
       aes(x = age, y = rt_residual_mean, colour = site)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4), se = TRUE) +
  labs(
    x = "Age (years)",
    y = "Residualized RT (log seconds)"
  )
```
\clearpage

```{r}
#| label: marginal-means
#| echo: false
#| message: false
#| cache: true
#| results: "hide"

# marginal means at 6, 9 and 12 to understand the magnitude of differences
library(emmeans)
emm <- emmeans(m_site_age_int, ~ site | age,
               at = list(age = c(6, 9, 12)))
emm
pairs(emm)  # pairwise site differences at each age with CIs
```

```{r}
#| label: tbl-site-pairs
#| tbl-cap: "Pairwise site differences in residualized RT at ages 6, 9, and 12 years."

pairs_df <- pairs(emm) %>%
  as.data.frame() %>%
  mutate(contrast = gsub("pilot_", "", contrast),
         contrast = gsub("_", " ", contrast)) %>%
  select(contrast, age, estimate, SE, df, t.ratio, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  mutate(p.value = ifelse(p.value < .001, "<.001", sprintf("%.3f", p.value)))

knitr::kable(
  pairs_df,
  booktabs = TRUE,
  col.names = c("Site Contrast", "Age (years)", "Estimate", "SE", "df", "t", "p")
)
```

\clearpage

## Within-Person Dynamics

To understand how performance fluctuates during the task, we examined reaction time trajectories across the 80 trials. The figure below illustrates the trial-by-trial dynamics. A clear "jump" in reaction time is visible halfway through the task (around trial 48), corresponding to the transition from 2D to 3D stimuli, which required significantly more processing time across all three populations.

```{r}
#| label: fig-within-person-trials
#| fig-cap: "Trial-level Reaction Time Dynamics. The plot shows the mean log-reaction time across the 80 trials for each site. Shaded areas represent standard error. The sharp increase represents the transition to 3D items."
#| echo: false
#| message: false
#| warning: false

# 1. Prepare data with Site Labels
trial_plot_data <- mrot_rt_trial %>%
  filter(!is.na(log_rt), correct == 1) %>% # Focus on correct trials for valid RT
  left_join(run_data %>% select(user_id, run_id, site), by = c("user_id", "run_id")) %>%
  mutate(site_label = dplyr::recode(site.x, !!!site_labels))


ggplot(trial_plot_data, aes(x = trial_number, y = log_rt, color = site_label, fill = site_label)) +
  # Mean line
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  # Confidence bands (Standard Error)
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Trial Number (Time)",
    y = "Log Reaction Time (seconds)",
    color = "Site",
    fill = "Site"
  ) +
  theme_classic() +
  theme(legend.position = "top")

```




# Cognitive Strategies Analysis

To determine if children rely on distinct cognitive strategies (e.g., "Effective Rotation" vs. "Impulsive Guessing"), we extracted four cognitive features for each participant using a mixed-effects model:

1.  Baseline Speed: Intercept from the linear model.
2.  Rotation Slope: The "cost" of rotation (purified of stimulus effects).
3.  Response Variability (RT SD): Intra-individual standard deviation.
4.  Accuracy: IRT-estimated ability.

We then applied K-Means clustering ($k=3$) to identify subgroups. To ensure cross-cultural validity, we compared a "Global Model" (pooled data) against "Local Models" (country-specific).


```{r}
#| label: strategy-analysis-prep
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| results: "hide"

library(lme4)
library(broom)

# 1. PREPARE DATA FOR RICH MODEL
#    Ensure stimulus is a factor for the interaction term
mrot_rich_data <- mrot_rt_trial %>%
    mutate(
        stimulus = stringr::str_extract(item, "^[a-z]+"),
        angle    = as.integer(stringr::str_extract(item, "\\d+"))
    )%>%
  mutate(
    angle_s = scale(angle),
    trial_s = scale(trial_number),
    stimulus = as.factor(stimulus)
  )

model_data_corrected <- ability_runs %>%
  inner_join(rt_residuals_person %>% 
               select(user_id, run_id, rt_resid = rt_residual_mean),
             by = c("user_id", "run_id")) %>%
  filter(!is.na(ability), !is.na(rt_resid), !is.na(age)) %>%
  mutate(
    # Recode sites for cleaner output
    site_model = dplyr::recode(site, !!!site_labels)
  )
# 2. FIT THE "RICH MODEL" (Intercept + Angle*Stimulus + Trial)
#    Formula: RT ~ Angle*Stimulus + Trial
#    This separates "Shape Difficulty" (Stimulus) from "Rotation Ability" (Angle)
m_rich <- lmer(
  log_rt ~ angle_s * stimulus + trial_s + (1 + angle_s + trial_s | user_id) + (1 | item),
  data = mrot_rich_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# 3. EXTRACT PARAMETERS
#    Get Individual Deviations (Random Effects)
ran_effs <- ranef(m_rich)$user_id %>%
  rownames_to_column("user_id") %>%
  rename(int_dev = `(Intercept)`, slope_dev = angle_s, trial_dev = trial_s) %>%
  distinct(user_id, .keep_all = TRUE)

#    Get Global Averages (Fixed Effects)
fix_int <- fixef(m_rich)["(Intercept)"]
fix_ang <- fixef(m_rich)["angle_s"]
fix_tri <- fixef(m_rich)["trial_s"]

# 4. CALCULATE COGNITIVE FEATURES
rich_features_raw <- ran_effs %>%
  mutate(
    Baseline_Speed = fix_int + int_dev,
    Abs_Slope = fix_ang + slope_dev,
    Rel_Slope = Abs_Slope / Baseline_Speed,
    Learning_Rate = fix_tri + trial_dev
  )

# 5. ADD VARIABILITY & MERGE
rt_variability_clean <- mrot_rt_trial %>%
  group_by(user_id) %>%
  summarise(RT_SD = sd(log_rt, na.rm = TRUE), .groups = "drop") %>%
  distinct(user_id, .keep_all = TRUE)

user_site_map <- mrot_rt_trial %>% 
  select(user_id, site) %>% 
  distinct(user_id, .keep_all = TRUE)

rich_features_final <- rich_features_raw %>%
  left_join(rt_variability_clean, by = "user_id") %>%
  left_join(user_site_map, by = "user_id") %>%
  left_join(
    model_data_corrected %>% select(user_id, accuracy = ability) %>% distinct(user_id, .keep_all=TRUE), 
    by = "user_id"
  ) %>%
  mutate(
    Country = case_when(
      grepl("mpieva", site) ~ "Germany",
      grepl("uniandes", site) ~ "Colombia",
      grepl("western", site) ~ "Canada",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(across(c(accuracy, Abs_Slope, Baseline_Speed, RT_SD), is.finite))
```

```{r}
#| label: clustering-execution
#| echo: false
#| message: false
#| warning: false

# 1. SETUP CLUSTERING
analysis_data <- rich_features_final %>%
  filter(across(c(accuracy, Abs_Slope, Baseline_Speed, RT_SD), is.finite))

my_colors <- c("Ineffective" = "#c0392b", "Effective" = "grey60")

# 2. RUN GLOBAL K-MEANS
global_input <- analysis_data %>% 
  select(accuracy, Abs_Slope, Baseline_Speed, RT_SD) %>% 
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

# Identify "Ineffective" Global Cluster (Lowest Accuracy)
bad_global_id <- analysis_data %>% 
  group_by(Global_Cluster) %>% 
  summarise(acc = mean(accuracy)) %>% 
  arrange(acc) %>% 
  pull(Global_Cluster) %>% .[1]

analysis_data <- analysis_data %>%
  mutate(Strategy_Type = ifelse(Global_Cluster == bad_global_id, "Ineffective", "Effective"))

# 3. CALCULATE AGREEMENT STATS (For Text)
check_agreement <- function(data, country_name) {
  c_data <- data %>% filter(Country == country_name)
  set.seed(123)
  input_local <- c_data %>% select(accuracy, Abs_Slope, Baseline_Speed, RT_SD) %>% scale()
  k_local <- kmeans(input_local, centers = 3)
  
  bad_local_id <- c_data %>% 
    mutate(cid = k_local$cluster) %>% 
    group_by(cid) %>% 
    summarise(acc = mean(accuracy)) %>% 
    arrange(acc) %>% 
    pull(cid) %>% .[1]
  
  c_data$Local_Label <- ifelse(k_local$cluster == bad_local_id, "Ineffective", "Effective")
  agreement_table <- table(Local = c_data$Local_Label, Global = c_data$Strategy_Type)
  sum(diag(agreement_table)) / sum(agreement_table)
}

res_de <- check_agreement(analysis_data, "Germany")
res_co <- check_agreement(analysis_data, "Colombia")
res_ca <- check_agreement(analysis_data, "Canada")
```

The validation analysis showed high agreement between the Global and Local models for Germany (`r round(res_de*100, 1)`%) and Canada (`r round(res_ca*100, 1)`%). Agreement was slightly lower for Colombia (`r round(res_co*100, 1)`%), suggesting some nuance in strategy classification, but sufficiently high to proceed with the Global definitions.

```{r}
#| label: fig-strategy-faceted
#| fig-cap: "Global Strategy Model. Participants are classified as using an 'Effective' (Grey) or 'Ineffective' (Red) strategy based on clustering of Accuracy, Slope, Speed, and Variability. The 'Ineffective' group is characterized by low accuracy and flat rotation slopes."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5

ggplot(analysis_data, aes(x = Abs_Slope, y = accuracy, color = Strategy_Type)) +
  geom_point(alpha = 0.6, size = 1.8) +
  stat_ellipse(aes(group = Global_Cluster), type = "norm", linetype = "dashed", color = "black", size = 0.3) +
  facet_wrap(~ Country) + 
  theme_bw(base_size = 12) +
  scale_color_manual(values = my_colors) +
  labs(
    x = "Rotation Slope", y = "Accuracy (Theta)",
    color = "Strategy Type"
  ) + 
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```
\clearpage

##Model Selection: Global vs. Local Clustering

To determine whether we can define "ineffective strategies" universally or if we need country-specific definitions, we compared two approaches:

- Local Models: Clustering performed independently within each country (optimizing for local variance).
- Global Model: Clustering performed on the pooled dataset (applying a universal standard).

We then calculated the agreement rate between the two methods to see if the "Global" definition aligns with "Local" realities.

```{r}
#| label: model-comparison-prep
#| echo: false
#| message: false
#| warning: false
#| results: "hide"

library(patchwork)

# 1. PREP DATA (Ensure we have the features)
# (Re-using rich_features_final from previous steps)
analysis_data <- rich_features_final %>%
  filter(across(c(accuracy, Abs_Slope, Baseline_Speed, RT_SD), is.finite))

# 2. FIT LOCAL MODELS (Per Country)
get_local_labels <- function(data, country_name) {
  c_data <- data %>% filter(Country == country_name)
  set.seed(123)
  # Scale locally
  input_local <- c_data %>% select(accuracy, Abs_Slope, Baseline_Speed, RT_SD) %>% scale()
  k_local <- kmeans(input_local, centers = 3)
  
  # Identify "Ineffective" (Lowest Accuracy cluster)
  bad_id <- c_data %>% 
    mutate(cid = k_local$cluster) %>% 
    group_by(cid) %>% 
    summarise(acc = mean(accuracy)) %>% 
    arrange(acc) %>% 
    pull(cid) %>% .[1]
  
  # Return dataframe with ID and Label
  c_data %>% 
    mutate(Local_Label = ifelse(k_local$cluster == bad_id, "Ineffective", "Effective")) %>%
    select(user_id, Local_Label)
}

# Run for all 3
loc_de <- get_local_labels(analysis_data, "Germany")
loc_co <- get_local_labels(analysis_data, "Colombia")
loc_ca <- get_local_labels(analysis_data, "Canada")

# Merge Local Labels back into main data
local_labels_all <- bind_rows(loc_de, loc_co, loc_ca)
analysis_data <- analysis_data %>%
  left_join(local_labels_all, by = "user_id")

# 3. FIT GLOBAL MODEL (Pooled)
global_input <- analysis_data %>% 
  select(accuracy, Abs_Slope, Baseline_Speed, RT_SD) %>% 
  scale()

set.seed(123)
k_global <- kmeans(global_input, centers = 3)
analysis_data$Global_Cluster <- as.factor(k_global$cluster)

# Identify "Ineffective" Global Cluster
bad_global_id <- analysis_data %>% 
  group_by(Global_Cluster) %>% 
  summarise(acc = mean(accuracy)) %>% 
  arrange(acc) %>% 
  pull(Global_Cluster) %>% .[1]

analysis_data <- analysis_data %>%
  mutate(Global_Label = ifelse(Global_Cluster == bad_global_id, "Ineffective", "Effective"))

# 4. CALCULATE AGREEMENT
agreement_stats <- analysis_data %>%
  group_by(Country) %>%
  summarise(
    Agreement = mean(Local_Label == Global_Label),
    .groups = "drop"
  )
```

### 1. Agreement Scorecard

The table below shows the agreement between the universal (Global) definition and the country-specific (Local) definitions. Agreement is very high for Germany and Canada, with moderate divergence in Colombia.

```{r}
#| label: tbl-agreement
#| tbl-cap: "Agreement between Local and Global Cluster Models."
#| echo: false

agreement_stats %>%
  mutate(Agreement = scales::percent(Agreement, accuracy = 0.1)) %>%
  knitr::kable(col.names = c("Country", "Model Agreement %"), align = "lc")

```

### 2. Visualizing the Differences

The plots below compare the Local Models (top row) against the Global Model (bottom row). The Global model enforces a stricter boundary in Colombia, classifying some children as "Effective" who were "Ineffective" relative to their local peers.

```{r}
#| label: fig-model-comparison
#| fig-cap: "Comparison of Local vs. Global Clustering. Red points indicate the 'Ineffective' strategy. The Global model applies a consistent standard across all populations, whereas Local models adapt to the specific variance of each country."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8

my_colors <- c("Ineffective" = "#c0392b", "Effective" = "grey60")

# Helper for plotting
plot_cluster <- function(data, label_col, title_prefix) {
  ggplot(data, aes(x = Abs_Slope, y = accuracy, color = .data[[label_col]])) +
    geom_point(alpha = 0.5, size = 1.5) +
    stat_ellipse(aes(group = .data[[label_col]]), type = "norm", linetype = "dashed", color = "black", size = 0.3) +
    facet_wrap(~ Country) +
    theme_bw(base_size = 11) +
    scale_color_manual(values = my_colors) +
    labs(
      title = title_prefix,
      x = "Rotation Slope", y = "Accuracy (Theta)",
      color = "Strategy"
    ) +
    theme(legend.position = "none", strip.text = element_text(face = "bold"))
}

p_local <- plot_cluster(analysis_data, "Local_Label", "A. Local Models (Country-Specific Definitions)")
p_global <- plot_cluster(analysis_data, "Global_Label", "B. Global Model (Universal Definition)")

# Combine plots vertically
p_local / p_global
```

### 3. Diagnostic: Who are the "Disagreement" Kids?
To understand the divergence in Colombia, we isolated the mismatched cases. The plot below highlights children classified differently by the two models.

- Grey: Models Agree.
- Blue (Global Lenience): Children considered "Ineffective" locally but "Effective" globally (likely due to the Global model having a lower threshold for accuracy).
- Red (Global Strictness): Children considered "Effective" locally but "Ineffective" globally.

```{r}
#| label: fig-mismatch
#| fig-cap: "Analysis of Classification Disagreements in Colombia. Colored points represent children whose strategy classification changes depending on whether a Local or Global model is used."
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5

# Filter for Colombia and Define Status
col_diag <- analysis_data %>% 
  filter(Country == "Colombia") %>%
  mutate(
    Status = case_when(
      Global_Label == Local_Label ~ "Agreement",
      Global_Label == "Ineffective" & Local_Label == "Effective" ~ "Global Strictness",
      Global_Label == "Effective" & Local_Label == "Ineffective" ~ "Global Lenience",
      TRUE ~ "Error"
    )
  ) %>%
  arrange(Status) # Draw Agreement first so mismatches pop on top

# Custom Palette for Diagnostic
diag_colors <- c(
  "Agreement" = "grey90", 
  "Global Strictness" = "#e74c3c",  # Red
  "Global Lenience" = "#3498db"     # Blue
)

ggplot(col_diag, aes(x = Abs_Slope, y = accuracy, color = Status)) +
  geom_point(alpha = 0.8, size = 2.5) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = diag_colors) +
  labs(
    title = "Diagnostic: Colombia Mismatches",
    x = "Rotation Slope", y = "Accuracy",
    color = "Classification Status"
  ) +
  theme(legend.position = "bottom")
```


\clearpage

## Developmental Trajectory of Strategies

We modeled the prevalence of the "Ineffective" strategy as a function of age. Results confirm that this is a developmental phase: the probability of using an impulsive strategy decreases significantly as children get older.

```{r}
#| label: fig-strategy-age
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Developmental Trajectory of Strategy Use. The percentage of children classified as using the 'Ineffective' strategy decreases with age across all populations.

# 1. Prepare Data (Binning Age)
analysis_data <- analysis_data %>%
  mutate(Strategy_Type = ifelse(Global_Cluster == bad_global_id, "Ineffective", "Effective"))

strategy_by_age <- analysis_data %>%
  left_join(model_data_corrected %>% select(user_id, age), by = "user_id") %>%
  mutate(Age_Bin = floor(age)) %>%
  filter(!is.na(Age_Bin), Age_Bin >= 6, Age_Bin <= 12) %>%
  group_by(Country, Age_Bin) %>%
  summarise(
    Total_N = n(),
    Ineffective_N = sum(Strategy_Type == "Ineffective"),
    Ineffective_Pct = Ineffective_N / Total_N,
    SE = sqrt((Ineffective_Pct * (1 - Ineffective_Pct)) / Total_N),
    .groups = "drop"
  )

# 2. Plot
ggplot(strategy_by_age, aes(x = Age_Bin, y = Ineffective_Pct, color = Country)) +
  geom_errorbar(aes(ymin = pmax(0, Ineffective_Pct - SE), ymax = pmin(1, Ineffective_Pct + SE)), 
                width = 0.2, alpha = 0.5) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_x_continuous(breaks = 6:12) + 
  theme_bw(base_size = 12) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Age (Years)",
    y = "% Using Ineffective Strategy"
  )
```

```{r}
#| label: strategy-stats
#| echo: false
#| message: false
#| warning: false

# Statistical Tests
analysis_data <- analysis_data %>%
  mutate(Strategy_Type = ifelse(Global_Cluster == bad_global_id, "Ineffective", "Effective"))

stat_data <- analysis_data %>%
  left_join(model_data_corrected %>% select(user_id, age), by = "user_id") %>%
  filter(!is.na(age), age >= 6, age <= 12) %>%
  mutate(
    Is_Ineffective = ifelse(Strategy_Type == "Ineffective", 1, 0),
    Country = relevel(as.factor(Country), ref = "Colombia")
  )

# Model A: Colombia Specific
m_colombia <- glm(Is_Ineffective ~ age, 
                  data = stat_data %>% filter(Country == "Colombia"), 
                  family = "binomial")

res_col <- tidy(m_colombia, exponentiate = TRUE, conf.int = TRUE) %>% 
  filter(term == "age")

# Model B: Interaction
m_interaction <- glm(Is_Ineffective ~ age * Country, 
                     data = stat_data, 
                     family = "binomial")
summary(m_interaction)
```

Statistical analysis for the Colombian cohort confirms a significant developmental shift (Odds Ratio = `r round(res_col$estimate, 2)`, 95% CI [`r round(res_col$conf.low, 2)`, `r round(res_col$conf.high, 2)`]), indicating that for every year of age, the odds of using an ineffective strategy decrease substantially.

\clearpage
<!--

```{r}
#| label: within-person-rt
#| echo: false
#| message: false
#| cache: true

# 1. Decompose RT for Within-Person Analysis
within_person_data <- model_rt_data %>%
  filter(!is.na(log_rt), !is.na(correct)) %>%
  group_by(user_id) %>%
  mutate(
    rt_person_mean = mean(log_rt, na.rm = TRUE),
    rt_within = log_rt - rt_person_mean # >0 means slower than my usual
  ) %>%
  ungroup()

# 2. Fit GLMM (Logistic Regression)
# Does slowing down predict correctness?
m_within <- glmer(
  correct ~ rt_within + rt_person_mean + angle + item_group + trial_number + (1|user_id) + (1|item),
  data = within_person_data,
  family = binomial,
  nAGQ = 0 # Optimization for speed
)

```

```{r}
#| label: fig-within-rt-trajectories
#| fig-cap: "Within-person trajectories of log-transformed RT across trials."
#| fig-width: 8
#| fig-height: 6

# Add age info + age groups
within_plot_data <- within_person_data %>%
  left_join(run_ages, by = c("user_id", "run_id")) %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(
      age,
      breaks = c(5, 8, 11, 14),
      include.lowest = TRUE,
      right = TRUE
    )
  )

ggplot(within_plot_sample,
       aes(x = trial_number, y = log_rt, group = user_id)) +
  geom_line(alpha = 0.15) +
  geom_smooth(aes(group = 1),
              colour = "black",
              linewidth = 1) +
  facet_grid(site ~ age_group) +
  labs(
    x = "Trial number",
    y = "Log RT (seconds)",
    subtitle = "Thin lines = individual trajectories; thick line = average trend"
  )


```
\clearpage

# Cross-Country Comparisons (Bivariate Model)

## Cross-Population Differences in Ability and Speed

To simultaneously evaluate developmental trends and cross-cultural differences, we fitted a multivariate Bayesian regression model. This model jointly predicts Spatial Ability ($\theta$) and Residualized Processing Speed, accounting for the correlation between the two traits.

```{r}
#| label: bivariate-prep
#| echo: false
#| message: false

# Join existing scores with new residuals
model_data_corrected <- ability_runs %>%
  inner_join(rt_residuals_person %>% 
               select(user_id, run_id, rt_resid = rt_residual_mean),
             by = c("user_id", "run_id")) %>%
  filter(!is.na(ability), !is.na(rt_resid), !is.na(age)) %>%
  mutate(
    # Recode sites for cleaner output
    site_model = dplyr::recode(site, !!!site_labels)
  )
```

```{r}
#| label: fig-development
#| fig-cap: "Developmental Trajectories. Panel A shows the increase in Spatial Ability (Theta) with age. Panel B shows the decrease in Residualized Reaction Time (log seconds) with age, indicating faster processing speed."
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(patchwork)

# 1. Define APA Theme
theme_apa <- theme_classic(base_size = 12) +
  theme(
    text = element_text(family = "serif", color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(face = "bold", size = 12),
    plot.tag = element_text(face = "bold", size = 16), 
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

# 2. Panel A: Spatial Ability by Age
plot_a <- ggplot(model_data_corrected, aes(x = age, y = ability)) +
  geom_jitter(alpha = 0.25, size = 1.2, width = 0.2, color = "grey40") +
  geom_smooth(method = "lm", color = "black", linewidth = 1, se = TRUE, fill = "grey80") +
  labs(
    tag = "A",
    x = "Age (Years)",
    y = expression(paste("Spatial Ability (", theta, ")")) 
  ) +
  theme_apa

# 3. Panel B: Processing Speed by Age
plot_b <- ggplot(model_data_corrected, aes(x = age, y = rt_resid)) +
  geom_jitter(alpha = 0.25, size = 1.2, width = 0.2, color = "grey40") +
  geom_smooth(method = "lm", color = "black", linewidth = 1, se = TRUE, fill = "grey80") +
  labs(
    tag = "B",
    x = "Age (Years)",
    y = "Residualized RT (log seconds)\n(Negative = Faster)"
  ) 
# 4. Combine and Print
plot_a + plot_b
```
\clearpage

```{r}
#| label: run-brms
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| results: hide

# NOTE: This chunk uses cache=TRUE because brms takes time to run.

m_joint_corrected <- brm(
  mvbind(ability, rt_resid) ~ site + age,
  data = model_data_corrected,
  chains = 4, 
  cores = 4, 
  iter = 2000, 
  seed = 123
)
```

```{r}
#| label: tbl-bivariate-results
#| echo: false
#| message: false
#| warning: false

# 1. Extract results for Table
fixef_results <- fixef(m_joint_corrected) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter")

# 2. Define SAFE Helper functions (Returns NA instead of crashing if name is missing)
get_est <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Estimate)
  if(length(val) == 0) return(NA) else return(val)
}

get_low <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Q2.5)
  if(length(val) == 0) return(NA) else return(val)
}

get_up <- function(pattern) {
  val <- fixef_results %>% filter(Parameter == pattern) %>% pull(Q97.5)
  if(length(val) == 0) return(NA) else return(val)
}

# 3. Build Dynamic Dataframe using CONFIRMED parameter names
site_results <- tibble(
  Outcome = c("Spatial Ability (SD)", "Spatial Ability (SD)", 
              "Response Speed (log s)", "Response Speed (log s)"),
  Comparison = c("Colombia vs. Germany", "Canada vs. Germany", 
                 "Colombia vs. Germany", "Canada vs. Germany"),
  
  # THESE NOW MATCH YOUR CONSOLE OUTPUT EXACTLY:
  Param_Name = c("ability_sitepilot_uniandes_co", "ability_sitepilot_western_ca",
                 "rtresid_sitepilot_uniandes_co", "rtresid_sitepilot_western_ca")
) %>%
  rowwise() %>%
  mutate(
    Beta = get_est(Param_Name),
    CI_Lower = get_low(Param_Name),
    CI_Upper = get_up(Param_Name),
    
    # Logic to determine significance automatically
    Sig_Flag = ifelse(sign(CI_Lower) == sign(CI_Upper), "*", "ns"),
    
    # Logic to generate the interpretation text
    Result = case_when(
      is.na(Beta) ~ "Error: Param Not Found",
      Outcome == "Spatial Ability (SD)" & Beta < 0 ~ "Significantly Lower",
      Outcome == "Spatial Ability (SD)" & Beta > 0 ~ "Significantly Higher",
      Outcome == "Response Speed (log s)" & Beta < 0 & Sig_Flag == "*" ~ "Significantly Faster",
      Outcome == "Response Speed (log s)" & Beta > 0 & Sig_Flag == "*" ~ "Significantly Slower",
      Sig_Flag == "ns" ~ "No Difference",
      TRUE ~ "Other"
    )
  ) %>%
  ungroup() %>%
  filter(!is.na(Beta))

# 4. Generate GT Table
site_results %>%
  select(Outcome, Comparison, Beta, CI_Lower, CI_Upper, Result) %>%
  group_by(Outcome) %>%
  gt() %>%
  tab_header(
    title = "Cross-Cultural Differences in Ability and Processing Speed",
    subtitle = "Posterior estimates from Bivariate Bayesian Regression"
  ) %>%
  cols_label(
    Comparison = "Site Comparison",
    Beta = "Estimate (β)",
    CI_Lower = "Lower 95% CI",
    CI_Upper = "Upper 95% CI",
    Result = "Interpretation"
  ) %>%
  fmt_number(columns = c(Beta, CI_Lower, CI_Upper), decimals = 2) %>%
  tab_options(
    table.border.top.color = "black",
    table.border.bottom.color = "black",
    row_group.font.weight = "bold"
  )
```



```{r}
#| label: fig-site-forest
#| fig-cap: "Site-Specific Processing Styles. Forest plot showing posterior estimates for Spatial Ability (Panel A) and Processing Speed (Panel B) relative to the German baseline (dashed line at 0). Error bars represent 95% confidence intervals."
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(dplyr)

# 1. Define APA Theme for Forest Plot
theme_apa_forest <- theme_classic(base_size = 11) +
  theme(
    text = element_text(family = "serif"),
    axis.text.y = element_text(face = "bold", size = 10),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )

# 2. Extract plotting data from our dynamic dataframe
# We clean up the site names for the Y-axis labels
plot_data <- site_results %>%
  mutate(Site = ifelse(grepl("Colombia", Comparison), "Colombia", "Canada"))

# 3. Panel A: Ability Differences
p1 <- ggplot(subset(plot_data, Outcome == "Spatial Ability (SD)"), 
             aes(x = Beta, y = Site)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "#27ae60") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2, color = "#27ae60") +
  labs(
    title = "A. Ability Differences",
    subtitle = "(Reference: Germany = 0)",
    x = "Difference in SD units"
  ) +
  theme_apa_forest

# 4. Panel B: Speed Differences
p2 <- ggplot(subset(plot_data, Outcome == "Response Speed (log s)"), 
             aes(x = Beta, y = Site)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "#2980b9") +
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.2, color = "#2980b9") +
  labs(
    title = "B. Speed Differences",
    subtitle = "(Negative = Faster)",
    x = "Difference in log(seconds)"
  ) +
  theme_apa_forest

# 5. Combine and Print
p1 + p2
```

\clearpage

## Differential Effectiveness of Processing Speed

We further tested whether the accuracy benefits of slowing down varied across cultures by modeling the interaction between site and processing speed. In contrast to the site-specific hypothesis, results indicated a universal benefit to deliberative processing.

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: true
#| results: hide
# 1. Fit the Interaction Model
# NOTE: Using cache=TRUE to avoid re-running this slow model every render
m_strategy <- brm(
  ability ~ age + site_model * rt_resid, 
  data = model_data_corrected,
  chains = 4, 
  cores = 4, 
  iter = 2000, 
  seed = 123
)
```

```{r}
#| label: fig-strategy-interaction
#| fig-cap: "Universal Speed-Accuracy Trade-off. The relationship between residualized reaction time and spatial ability is consistent across sites. In all three populations, slower processing (higher residualized RT) predicts higher spatial ability scores, as indicated by the parallel positive slopes."
#| echo: false
#| message: false
#| warning: false


# 2. Extract trend lines for plotting
effects_data <- conditional_effects(m_strategy, effects = "rt_resid:site_model")[[1]]

# 3. Define APA Theme
theme_apa <- theme_classic(base_size = 12) +
  theme(
    text = element_text(family = "serif", color = "black"),
    axis.line = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black", size = 11),
    axis.title = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Prepare data for plotting
# Note: Ensure site_labels vector exists from previous chunks
effects_data2 <- effects_data %>%
  mutate(site_label = dplyr::recode(as.character(site_model), 
                                    "pilot_uniandes_co" = "Colombia",
                                    "pilot_mpieva_de" = "Germany",
                                    "pilot_western_ca" = "Canada"))

# Position for line-end labels (rightmost x for each site)
label_positions <- effects_data2 %>%
  group_by(site_label) %>%
  slice_max(rt_resid, with_ties = FALSE) %>%
  ungroup()

# 4. Create the Plot
ggplot(effects_data2, aes(x = rt_resid, y = estimate__, linetype = site_label)) +
  # Ribbons (Confidence Intervals)
  geom_ribbon(aes(ymin = lower__, ymax = upper__, fill = site_label), 
              alpha = 0.15, color = NA) +
  # Trend Lines
  geom_line(linewidth = 1.2, color = "black") +
  # Direct labels near lines
  geom_text(
    data = label_positions,
    aes(x = rt_resid + 0.05, y = estimate__, label = site_label),
    inherit.aes = FALSE, hjust = 0, size = 4.5, fontface = "bold"
  ) +
  # Manual Scales
  scale_linetype_manual(values = c("Germany" = "solid", "Colombia" = "longdash", "Canada" = "dotted")) +
  scale_fill_manual(values = c("Germany" = "black", "Colombia" = "black", "Canada" = "black")) +
  # Labels and Layout
  labs(
    x = "Residualized RT (log seconds)\n(Negative = Fast | Positive = Slow)",
    y = "Predicted Spatial Ability (Theta)"
  ) +
  coord_cartesian(clip = "off") + 
  theme_apa +
  theme(legend.position = "none", plot.margin = margin(r = 80))

```
\clearpage

-->

